### 2017-01-12 src/QCAnalysis.Rmd
init

---
title: "Mamary Gland Development Exploratory Data Analysis"
author: "Karsten Bach"
date: '`r Sys.Date()`'
output:
  html_notebook:
    toc: true
    toc_depth: 2
    toc_float: true
    theme: united
    highlight: tango
---
***
# Load Data and Packages
```{r, message=FALSE}
library(scran)
library(dplyr)
library(knitr)
library(ggplot2)
source("functions.R")
dataList <- readRDS("../data/Robjects/ExpressionList.rds")
m <- dataList[["counts"]]
pD <- dataList[["phenoData"]]
fD <- dataList[["featureData"]]

```
# Overview over the data

## Number of Cells
We sequenced less cells than expected, with a total of `r ncol(m)` cells.
```{r}
kable(t(as.matrix(table(pD$SampleID))))
kable(t(as.matrix(table(pD$Condition))))
```

# Quality Meassures
## Sequencing Depth and Genes detected
```{r}
pD$UmiSums<- colSums(m)
pD$GenesDetected <- colSums(m!=0)
genesDetected <- ggplot(pD, aes(x=SampleID,y=log10(GenesDetected),fill=Condition)) +
    geom_violin(draw_quantiles=0.5)+
    theme_bw()
LibrarySize <- ggplot(pD, aes(x=SampleID,y=log10(UmiSums),fill=Condition)) +
    geom_violin(draw_quantiles=0.5)+
    theme_bw()
gdLibSize <- ggplot(pD, aes(y=log10(GenesDetected), x=log10(UmiSums), color=Condition, shape=Replicate)) +
    geom_point() +
    geom_smooth(method=loess,inherit.aes=FALSE,aes(y=log10(GenesDetected),x=log10(UmiSums))) +
    theme_bw()
genesDetected
LibrarySize
gdLibSize
```

## Genewise QC
```{r, fig.height=8}
ntop <- 50 
mRel <- t(t(m)/colSums(m))
rownames(mRel)  <- fD$symbol
topExpressed <- rowMedians(mRel)
names(topExpressed) <- rownames(mRel)
topExpressed <- topExpressed %>% sort(.,decreasing=TRUE) %>% names
plotData <- t(mRel)[,topExpressed[1:ntop]] %>%
    reshape2::melt() %>%
    rename(Cell=Var1,
	   Gene=Var2,
	   RelativeExpression=value)
topGenes <- ggplot(plotData, aes(x=Gene, y=RelativeExpression)) +
    geom_boxplot() +
    coord_flip() +
    theme_bw()
freqOfExp <- m!=0
rownames(freqOfExp) <- fD$symbol
freqOfExp <- sort(rowSums(freqOfExp)/ncol(freqOfExp),decreasing=TRUE)
plotData <- data.frame("Gene"=names(freqOfExp),"Frequency"=freqOfExp)
topFreq <- ggplot(plotData[1:ntop,], aes(x=factor(Gene,levels=Gene),y=Frequency)) +
    geom_bar(stat="identity") +
    coord_flip() +
    xlab("Gene") +
    theme_bw()
topGenes
topFreq
```

## Cell Viability
```{r}
topFreq20 <- filter(fD, symbol %in% plotData[1:20,1]) %>% .$id
pD$prcntOfTop20 <- colSums(m[topFreq20,]!=0)/length(topFreq20)
mMito <- m[fD$Mitochondrial,]
pD$prcntMito <- colSums(mMito)/colSums(m)
cellViability <- ggplot(pD, aes(x=prcntMito, y=GenesDetected, color=Condition, shape=Replicate))+
    geom_point() +
    theme_bw()
cellViability
```
```{r}
smry <- pD %>%
    summarize(medianPrcntMito=median(prcntMito),
	      madPrcntMito=mad(prcntMito),
	      mGenesDetected=median(GenesDetected),
	      madGenesDetected=mad(GenesDetected),
	      mUmiSums=median(UmiSums),
	      madUmiSums=mad(UmiSums))
smryByGroup <- group_by(pD,Condition) %>%
    summarize(medianPrcntMito=median(prcntMito),
	      madPrcntMito=mad(prcntMito),
	      mGenesDetected=median(GenesDetected),
	      madGenesDetected=mad(GenesDetected),
	      mUmiSums=median(UmiSums),
	      madUmiSums=mad(UmiSums))
```

## QC Thresholding
```{r}
#simple thresholds
MitoCutOff <- 0.05
genesDetectedCutOff <- 1000
LibSizeCutOff <- 2000
```
Four different thresholds were applied:

1. Percentage of UMIs in Mitochondrial genes less than `r MitoCutOff`
2. A minimum of `r genesDetectedCutOff` genes detected per cell
3. A minimum of `r LibSizeCutOff` total UMIs per per cell
4. Not more than four standard deviations away from the #Genes~#UMIcount trend away (see below)
```{r}
#remove outliers from the genes detected per UMI trend 
fit <- loess(formula=log10(GenesDetected)~log10(UmiSums),data=pD)
pD$distFromDetectionTrend <- fit$residuals
upperL <- mean(fit$residuals)+4*sd(fit$residuals)
lowerL <- mean(fit$residuals)-4*sd(fit$residuals)
hist(fit$residuals, breaks=50, main="Histogramm of loess residuals \n and thresholds",
     xlab="Residuals")
abline(v=upperL,col="red")
abline(v=lowerL,col="red")

#Thresholding 
pD <- mutate(pD,
		  PassViability = prcntMito < MitoCutOff,
		  PassGenesDet = GenesDetected > genesDetectedCutOff,
		  PassLibSize = UmiSums > LibSizeCutOff,
		  PassDetTrend = distFromDetectionTrend < upperL & 
		      distFromDetectionTrend > lowerL,
		  PassAll= PassViability & PassGenesDet &
		      PassLibSize & PassDetTrend)
#Update QC plots
gdLibSize <- gdLibSize %+% pD 
gdLibSize <- gdLibSize + aes(color=PassDetTrend,shape=NULL) + 
    annotate("rect",xmin=-Inf, xmax=Inf, ymin=-Inf, ymax=log10(genesDetectedCutOff), fill="grey", alpha=0.3) + 
    annotate("rect",ymin=-Inf, ymax=Inf, xmin=-Inf, xmax=log10(LibSizeCutOff), fill="grey", alpha=0.3) 
gdLibSize
cellViability <- cellViability %+% pD
cellViability <- cellViability + aes(color=PassViability) +
    annotate("rect",ymin=-Inf, ymax=Inf, xmax=Inf, xmin=MitoCutOff
	     ,fill="grey", alpha=0.3) 
cellViability

kable(t(as.matrix(table(pD$Condition,pD$PassAll))),caption="Cells passing QC in different Conditions")
#Apply Thresholds
pD <- filter(pD, PassAll)
m <- m[,as.character(pD$barcode)]

#Thresholding on lowly expressed genes
expThreshold <- 0.1
keep <- rowMeans(m) > expThreshold
m <- m[keep,]
fD <- fD[keep,]
#Final matrix
dim(m)
```

## Normalization and first visualization
```{r}
clusters <- quickCluster(m)
pD$sf <- computeSumFactors(m,clusters=clusters,positive=TRUE)
m.norm <- t(t(m)/pD$sf)
# m.norm <- m.norm[,!is.na(colSums(m.norm))]
pD <- filter(pD, barcode %in% colnames(m.norm))
# 
fD <- mutate(fD, meanExpression=rowMeans(m.norm),
	     CV2=apply(m.norm,1,cv2),
	     dm=DM(meanExpression,CV2))

highVar <- dplyr::arrange(fD,desc(dm)) %>% .$id
fD$highVar <- fD$id %in% highVar[1:1000]
VarPlot <- ggplot(fD, aes(x=log10(meanExpression),y=log10(CV2), color=highVar)) +
    geom_point() +
    theme_bw()

fPCA <- m.norm[fD$highVar,]
pc <- prcomp(log2(t(fPCA)+1))
tsn <- Rtsne(log2(t(fPCA)+1),max_iter=5000)

pD$PC1 <- pc$x[,1]
pD$PC2 <- pc$x[,2]
pD$tSNE1 <- tsn$Y[,1]
pD$tSNE2 <- tsn$Y[,2]

Acta2 <- filter(fD, symbol=="Acta2") %>% .$id
pD$Acta2 <- t(m.norm)[,Acta2]
# 
pcaPlot <- ggplot(pD, aes(x=PC1, y=PC2, color=Condition, shape=Replicate)) +
    geom_point(alpha=0.8,size=4) +
    theme_bw() 
pcaPlot

tsnePlot <- ggplot(pD, aes(x=tSNE1, y=tSNE2, color=Condition, shape=Replicate)) +
    geom_point(alpha=0.8,size=4) +
    theme_bw()
tsnePlot
```

### 2017-01-12 src/PrepareData.R

@@ -1,0 +1,24 @@
+library(cellrangerRkit)
+gene_bc_matrix <- load_cellranger_matrix("../data/CellRangerData/MammaryGland/", genome="mm10")
+pDat <- data.frame(pData(gene_bc_matrix))
+fDat <- data.frame(fData(gene_bc_matrix))
+cDat <- as.matrix(exprs(gene_bc_matrix))
+
+#Add more info to phenotype Data
+library(dplyr)
+pDat <- mutate(pDat, SeqID=substr(barcode,18,18)) %>%
+	       mutate(SeqID=mapvalues(SeqID,c("1","2","3","4","5","6","7","8"),
+					 c("A1","B1","C1","D1","E1","F1","G1","H1"))) %>%
+        mutate(SampleID=mapvalues(SeqID,c("A1","B1","C1","D1","E1","F1","G1","H1"),
+				  c("V1","P2","I2","P1","L2","L1","V2","I1"))) %>%
+	mutate(Replicate=as.factor(substr(SampleID,2,2)),
+	       Condition=as.factor(substr(SampleID,1,1)))
+
+#Add more info to the feature Data
+mitoGenes <- read.table("../data/miscData/MitoGenes.txt")
+fDat <- mutate(fDat,
+	       Mitochondrial=id %in% mitoGenes$x)
+
+
+DataList <- list("phenoData"=pDat, "featureData"=fDat, "counts"=cDat)
+saveRDS(DataList,file="../data/Robjects/ExpressionList.rds")

### 2017-01-12 src/functions.R

@@ -1,0 +1,5 @@
+cv2 <- function(x) {
+    # Computes cv2 for a vector
+    out <- (sd(x)/mean(x))^2
+}
+

### 2017-01-12 src/QCAnalysis.Rmd
Updated QC_Analysis, less conservative and included DimRed plots

@@
 library(ggplot2)
+rnd_seed <- 300
+library(Rtsne)
 source("functions.R")
@@
 fD <- dataList[["featureData"]]
-
 ```
@@
 MitoCutOff <- 0.05
-genesDetectedCutOff <- 1000
-LibSizeCutOff <- 2000
+genesDetectedCutOff <- 500
+LibSizeCutOff <- 500
 ```
@@
 #Apply Thresholds
-pD <- filter(pD, PassAll)
-m <- m[,as.character(pD$barcode)]
+pD.filtered <- filter(pD, PassAll)
+m.filtered <- m[,as.character(pD.filtered$barcode)]
 
@@
 expThreshold <- 0.1
-keep <- rowMeans(m) > expThreshold
-m <- m[keep,]
-fD <- fD[keep,]
+keep <- rowMeans(m.filtered) > expThreshold
+m.filtered <- m.filtered[keep,]
+fD.filtered <- fD[keep,]
 #Final matrix
-dim(m)
+dim(m.filtered)
 ```
@@
 ```{r}
-clusters <- quickCluster(m)
-pD$sf <- computeSumFactors(m,clusters=clusters,positive=TRUE)
-m.norm <- t(t(m)/pD$sf)
-# m.norm <- m.norm[,!is.na(colSums(m.norm))]
-pD <- filter(pD, barcode %in% colnames(m.norm))
+clusters <- quickCluster(m.filtered)
+pD.filtered$sf <- computeSumFactors(m.filtered,clusters=clusters,positive=TRUE)
+m.norm <- t(t(m.filtered)/pD.filtered$sf)
+m.norm <- m.norm[,!is.na(colSums(m.norm))]
+pD.filtered <- filter(pD.filtered, barcode %in% colnames(m.norm))
 # 
-fD <- mutate(fD, meanExpression=rowMeans(m.norm),
+fD.filtered <- mutate(fD.filtered, meanExpression=rowMeans(m.norm),
 	     CV2=apply(m.norm,1,cv2),
@@
 
-highVar <- dplyr::arrange(fD,desc(dm)) %>% .$id
-fD$highVar <- fD$id %in% highVar[1:1000]
-VarPlot <- ggplot(fD, aes(x=log10(meanExpression),y=log10(CV2), color=highVar)) +
+highVar <- dplyr::arrange(fD.filtered,desc(dm)) %>% .$id
+topX <- 0.1*length(highVar)
+fD.filtered$highVar <- fD.filtered$id %in% highVar[1:topX]
+VarPlot <- ggplot(fD.filtered, aes(x=log10(meanExpression),y=log10(CV2), color=highVar)) +
     geom_point() +
     theme_bw()
+VarPlot
 
-fPCA <- m.norm[fD$highVar,]
-pc <- prcomp(log2(t(fPCA)+1))
-tsn <- Rtsne(log2(t(fPCA)+1),max_iter=5000)
+fPCA <- log2(t(m.norm[fD.filtered$highVar,])+1)
+fPCA <- scale(fPCA,scale=TRUE,center=TRUE)
+pc <- prcomp(fPCA)
+set.seed(rnd_seed)
+tsn <- Rtsne(fPCA,perplexity=floor(nrow(fPCA)/5))
 
-pD$PC1 <- pc$x[,1]
-pD$PC2 <- pc$x[,2]
-pD$tSNE1 <- tsn$Y[,1]
-pD$tSNE2 <- tsn$Y[,2]
+pD.filtered$PC1 <- pc$x[,1]
+pD.filtered$PC2 <- pc$x[,2]
+pD.filtered$tSNE1 <- tsn$Y[,1]
+pD.filtered$tSNE2 <- tsn$Y[,2]
 
-Acta2 <- filter(fD, symbol=="Acta2") %>% .$id
-pD$Acta2 <- t(m.norm)[,Acta2]
+Acta2 <- filter(fD.filtered, symbol=="Acta2") %>% .$id
+pD.filtered$Acta2 <- t(m.norm)[,Acta2]
 # 
-pcaPlot <- ggplot(pD, aes(x=PC1, y=PC2, color=Condition, shape=Replicate)) +
-    geom_point(alpha=0.8,size=4) +
+pcaPlot <- ggplot(pD.filtered, aes(x=PC1, y=PC2, color=Condition, shape=Replicate)) +
+    geom_point(alpha=0.8,size=2) +
     theme_bw() 
@@
 
-tsnePlot <- ggplot(pD, aes(x=tSNE1, y=tSNE2, color=Condition, shape=Replicate)) +
-    geom_point(alpha=0.8,size=4) +
+pcaPlot <- ggplot(pD.filtered, aes(x=PC1, y=PC2, color=log2(Acta2+1), shape=Replicate)) +
+    geom_point(alpha=0.8,size=2) +
+    theme_bw() 
+pcaPlot
+
+tsnePlot <- ggplot(pD.filtered, aes(x=tSNE1, y=tSNE2, color=Condition, shape=Replicate)) +
+    geom_point(alpha=0.8,size=2) +
+    theme_bw()
+tsnePlot
+
+tsnePlot <- ggplot(pD.filtered, aes(x=tSNE1, y=tSNE2, color=log2(Acta2+1), shape=Replicate)) +
+    geom_point(alpha=0.8,size=2) +
     theme_bw()
@@
 ```
+## Save Data
+```{r}
+out <- list()
+out[[1]] <- m
+out[[2]] <- pD
+out[[3]] <- fD
+saveRDS(out,file="../data/Robjects/ExpressionList_QC.rds")
+```

### 2017-01-16 src/QCAnalysis.Rmd
QC As of now

@@
 ```{r}
-topFreq20 <- filter(fD, symbol %in% plotData[1:20,1]) %>% .$id
-pD$prcntOfTop20 <- colSums(m[topFreq20,]!=0)/length(topFreq20)
 mMito <- m[fD$Mitochondrial,]
@@
 ```{r}
-smry <- pD %>%
-    summarize(medianPrcntMito=median(prcntMito),
-	      madPrcntMito=mad(prcntMito),
-	      mGenesDetected=median(GenesDetected),
-	      madGenesDetected=mad(GenesDetected),
-	      mUmiSums=median(UmiSums),
-	      madUmiSums=mad(UmiSums))
 smryByGroup <- group_by(pD,Condition) %>%
@@
 	      madPrcntMito=mad(prcntMito),
-	      mGenesDetected=median(GenesDetected),
-	      madGenesDetected=mad(GenesDetected),
-	      mUmiSums=median(UmiSums),
-	      madUmiSums=mad(UmiSums))
+	      threshold_PrcntMito=0.05,
+	      mGenesDetected=median(log10(GenesDetected)),
+	      madGenesDetected=mad(log10(GenesDetected)),
+	      threshold_GenesDetected=max(mGenesDetected-3*madGenesDetected,log10(500)),
+	      mUmiSums=median(log10(UmiSums)),
+	      madUmiSums=mad(log10(UmiSums)),
+	      threshold_UmiSums=max(mUmiSums-3*madUmiSums,log10(900))) %>%
+    select(Condition,starts_with("threshold")) 
+kable(smryByGroup)
 ```
@@
 MitoCutOff <- 0.05
-genesDetectedCutOff <- 500
-LibSizeCutOff <- 500
 ```
-Four different thresholds were applied:
+```{r}
+#Thresholding per group
+#inititae Df
+pD <- mutate(pD,
+	     ThresholdViability = 0,
+	     ThresholdGenesDet = 0,
+	     ThresholdLibSize = 0)
 
-1. Percentage of UMIs in Mitochondrial genes less than `r MitoCutOff`
-2. A minimum of `r genesDetectedCutOff` genes detected per cell
-3. A minimum of `r LibSizeCutOff` total UMIs per per cell
-4. Not more than four standard deviations away from the #Genes~#UMIcount trend away (see below)
-```{r}
-#remove outliers from the genes detected per UMI trend 
-fit <- loess(formula=log10(GenesDetected)~log10(UmiSums),data=pD)
-pD$distFromDetectionTrend <- fit$residuals
-upperL <- mean(fit$residuals)+4*sd(fit$residuals)
-lowerL <- mean(fit$residuals)-4*sd(fit$residuals)
-hist(fit$residuals, breaks=50, main="Histogramm of loess residuals \n and thresholds",
-     xlab="Residuals")
-abline(v=upperL,col="red")
-abline(v=lowerL,col="red")
+grps <- as.character(unique(pD$Condition))
+for (grp in grps) {
+    thrs <- filter(smryByGroup, Condition==grp) %>% select(-Condition) %>% t() %>% as.vector()
+    names(thrs) <- filter(smryByGroup, Condition==grp) %>% select(-Condition) %>% t() %>% rownames()
+    pD <- mutate(pD,
+		 ThresholdViability= ifelse(Condition==grp, MitoCutOff, ThresholdViability),
+		 ThresholdGenesDet= ifelse(Condition==grp, 10^thrs["threshold_GenesDetected"],ThresholdGenesDet),
+		 ThresholdLibSize= ifelse(Condition==grp, 10^thrs["threshold_UmiSums"],ThresholdLibSize))
+}
 
-#Thresholding 
 pD <- mutate(pD,
-		  PassViability = prcntMito < MitoCutOff,
-		  PassGenesDet = GenesDetected > genesDetectedCutOff,
-		  PassLibSize = UmiSums > LibSizeCutOff,
-		  PassDetTrend = distFromDetectionTrend < upperL & 
-		      distFromDetectionTrend > lowerL,
-		  PassAll= PassViability & PassGenesDet &
-		      PassLibSize & PassDetTrend)
-#Update QC plots
-gdLibSize <- gdLibSize %+% pD 
-gdLibSize <- gdLibSize + aes(color=PassDetTrend,shape=NULL) + 
-    annotate("rect",xmin=-Inf, xmax=Inf, ymin=-Inf, ymax=log10(genesDetectedCutOff), fill="grey", alpha=0.3) + 
-    annotate("rect",ymin=-Inf, ymax=Inf, xmin=-Inf, xmax=log10(LibSizeCutOff), fill="grey", alpha=0.3) 
-gdLibSize
+	     PassViability=prcntMito < ThresholdViability,
+	     PassGenesDet=GenesDetected > ThresholdGenesDet,
+	     PassLibSize=UmiSums > ThresholdLibSize,
+	     PassAll= PassViability & PassGenesDet & PassLibSize)
+
+gdHist <- ggplot(pD, aes(x=log10(GenesDetected),y=..density..)) +
+    geom_histogram(fill="white",color="black") +
+    geom_vline(data=smryByGroup,aes(xintercept=threshold_GenesDetected),color="red",lty="longdash") +
+    facet_wrap(~Condition) +
+    theme_bw()
+gdHist
+
+libSizeHist <- ggplot(pD, aes(x=log10(UmiSums),y=..density..)) +
+    geom_histogram(fill="white",color="black") +
+    geom_vline(data=smryByGroup,aes(xintercept=threshold_UmiSums),color="red",lty="longdash") +
+    facet_wrap(~Condition) +
+    theme_bw()
+libSizeHist
+
 cellViability <- cellViability %+% pD
@@
 
-kable(t(as.matrix(table(pD$Condition,pD$PassAll))),caption="Cells passing QC in different Conditions")
+kable(t(as.matrix(table(pD$Condition,pD$PassGenesDet))),caption="Cells passing Gene Detection QC")
+kable(t(as.matrix(table(pD$Condition,pD$PassLibSize))),caption="Cells passing Library Size QC")
+kable(t(as.matrix(table(pD$Condition,pD$PassViability))),caption="Cells passing Viability QC")
+kable(t(as.matrix(table(pD$Condition,pD$PassAll))),caption="Cells passing all QC requirements")
 #Apply Thresholds
@@
 #Thresholding on lowly expressed genes
-expThreshold <- 0.1
-keep <- rowMeans(m.filtered) > expThreshold
-m.filtered <- m.filtered[keep,]
-fD.filtered <- fD[keep,]
+isexpThreshold <- 10
+expThreshold <- 100*isexpThreshold
+keep1 <- rowSums(m.filtered!=0) > isexpThreshold
+keep2 <- rowSums(m.filtered) > expThreshold
+fD$keep <- keep1 & keep2
+m.filtered <- m.filtered[fD$keep,]
+fD.filtered <- fD[fD$keep,]
 #Final matrix
 dim(m.filtered)
+
+qcPCA <- select(pD.filtered, GenesDetected, UmiSums, prcntMito) %>%
+    mutate(GenesDetected=log10(GenesDetected),
+	   UmiSums=log10(UmiSums))
+qcpc <- prcomp(qcPCA)
+qcpD <- pD.filtered %>%
+    mutate(PC1=qcpc$x[,1],
+	   PC2=qcpc$x[,2])
+qcpcaPlot <- ggplot(qcpD, aes(x=PC1, y=PC2, color=Condition, shape=Replicate)) +
+    geom_point(alpha=0.8,size=2) +
+    ggtitle("PCA plot on QC stats after QC") +
+    theme_bw() 
+qcpcaPlot
+
 ```
@@
 pD.filtered$sf <- computeSumFactors(m.filtered,clusters=clusters,positive=TRUE)
+plot(log10(colSums(m.filtered))~log10(pD.filtered$sf),main="Library Size versus Size Factors")
 m.norm <- t(t(m.filtered)/pD.filtered$sf)
@@
 highVar <- dplyr::arrange(fD.filtered,desc(dm)) %>% .$id
-topX <- 0.1*length(highVar)
+topX <- floor(0.2*length(highVar))
+
 fD.filtered$highVar <- fD.filtered$id %in% highVar[1:topX]
@@
 pD.filtered$PC2 <- pc$x[,2]
+pD.filtered$PC3 <- pc$x[,3]
 pD.filtered$tSNE1 <- tsn$Y[,1]
@@
 pD.filtered$Acta2 <- t(m.norm)[,Acta2]
+Fcer1g <- filter(fD.filtered, symbol=="Fcer1g") %>% .$id
+pD.filtered$Fcer1g <- t(m.norm)[,Fcer1g]
 # 
@@
 tsnePlot
+
+```
+***
+## Heatmap on top50
+```{r, fig.height=10}
+gns <- highVar[1:50]
+m.ht <- log2(m.filtered[gns,]+1)
+rownames(fD) <- fD$id
+rownames(m.ht) <- fD[rownames(m.ht),"symbol"]
+colanno <- data.frame("Condition"=pD.filtered$Condition)
+rownames(colanno) <- as.character(pD.filtered$barcode)
+library(pheatmap)
+pheatmap(m.ht, show_colnames=FALSE, annotation_col=colanno, show_rownames=TRUE)
+
 ```

### 2017-01-18 src/QCAnalysis.Rmd
happy with QC

@@
 
-# Quality Meassures
+# Quality Overview
 ## Sequencing Depth and Genes detected
@@
 mMito <- m[fD$Mitochondrial,]
+idtop <- fD[fD$symbol %in% names(freqOfExp)[1:ntop],"id"]
+mTop <- m[idtop,]!=0
+pD$prcntTop <- colSums(mTop)/ntop
 pD$prcntMito <- colSums(mMito)/colSums(m)
@@
 cellViability
+prcntTop <- ggplot(pD, aes(x=prcntTop, y=GenesDetected, color=Condition, shape=Replicate))+
+    geom_point() +
+    theme_bw()
+prcntTop
 ```
 ```{r}
+leftmad <- function(x) {
+    m <- median(x)
+    dev <- abs(x-m)
+    leftmad <- median(dev[x<=m])
+    return(leftmad)
+}
 smryByGroup <- group_by(pD,Condition) %>%
@@
 	      mGenesDetected=median(log10(GenesDetected)),
-	      madGenesDetected=mad(log10(GenesDetected)),
-	      threshold_GenesDetected=max(mGenesDetected-3*madGenesDetected,log10(500)),
+	      madGenesDetected=leftmad(log10(GenesDetected)),
+	      threshold_GenesDetected=max(mGenesDetected-4*madGenesDetected,log10(500)),
 	      mUmiSums=median(log10(UmiSums)),
-	      madUmiSums=mad(log10(UmiSums)),
-	      threshold_UmiSums=max(mUmiSums-3*madUmiSums,log10(900))) %>%
+	      madUmiSums=leftmad(log10(UmiSums)),
+	      threshold_UmiSums=max(mUmiSums-4*madUmiSums,log10(1000))) %>%
     select(Condition,starts_with("threshold")) 
@@
 
-## QC Thresholding
+# QC Thresholding
 ```{r}
@@
 MitoCutOff <- 0.05
+prcntTopCutOff <- 0.5
 ```
@@
 	     ThresholdGenesDet = 0,
-	     ThresholdLibSize = 0)
+	     ThresholdLibSize = 0,
+	     ThresholdPrcntTop = prcntTopCutOff)
 
@@
 	     PassLibSize=UmiSums > ThresholdLibSize,
-	     PassAll= PassViability & PassGenesDet & PassLibSize)
+	     PassPrcntTop=prcntTop > ThresholdPrcntTop,
+	     PassAll= PassViability & PassGenesDet & PassLibSize & PassPrcntTop)
 
@@
 cellViability <- cellViability + aes(color=PassViability) +
-    annotate("rect",ymin=-Inf, ymax=Inf, xmax=Inf, xmin=MitoCutOff
-	     ,fill="grey", alpha=0.3) 
+    annotate("rect",ymin=-Inf, ymax=Inf, xmax=Inf, xmin=MitoCutOff,
+	     fill="grey", alpha=0.3) 
 cellViability
+
+prcntTop <- prcntTop %+% pD
+prcntTop <- prcntTop + aes(color=PassPrcntTop) +
+    annotate("rect",ymin=-Inf, ymax=Inf, xmin=-Inf, xmax=prcntTopCutOff,
+	     fill="grey", alpha=0.3) 
+prcntTop
 
@@
 isexpThreshold <- 10
-expThreshold <- 100*isexpThreshold
+expThreshold <- 50*isexpThreshold
 keep1 <- rowSums(m.filtered!=0) > isexpThreshold
@@
 
-qcPCA <- select(pD.filtered, GenesDetected, UmiSums, prcntMito) %>%
-    mutate(GenesDetected=log10(GenesDetected),
-	   UmiSums=log10(UmiSums))
-qcpc <- prcomp(qcPCA)
-qcpD <- pD.filtered %>%
-    mutate(PC1=qcpc$x[,1],
-	   PC2=qcpc$x[,2])
-qcpcaPlot <- ggplot(qcpD, aes(x=PC1, y=PC2, color=Condition, shape=Replicate)) +
-    geom_point(alpha=0.8,size=2) +
-    ggtitle("PCA plot on QC stats after QC") +
-    theme_bw() 
-qcpcaPlot
-
 ```
 
-## Normalization and first visualization
+# A more or less fair representation of the lactation samples
+```{r}
+pD.lac <- filter(pD, Condition=="L")
+cells <- as.character(pD.lac$barcode)
+m.lac <- m[,cells]
+keep <- rowMeans(m.lac) > 0.1
+m.lac <- m.lac[keep,]
+fD.lac <- fD[keep,]
+#Normalization
+clusters.lac <- quickCluster(m.lac)
+pD.lac$sf <- computeSumFactors(m.lac,clusters=clusters.lac)
+m.lac.norm <- t(t(m.lac)/pD.lac$sf)
+
+fD.lac <- mutate(fD.lac, meanExpression=rowMeans(m.lac),
+	     CV2=apply(m.lac,1,cv2),
+	     dm=DM(meanExpression,CV2))
+
+highVar.lac <- dplyr::arrange(fD.lac,desc(dm)) %>% .$id
+topX <- floor(0.2*length(highVar.lac))
+fD.lac$highVar <- fD.lac$id %in% highVar.lac[1:topX]
+
+fPCA.lac <- log2(t(m.lac.norm[fD.lac$highVar,])+1)
+fPCA.lac <- scale(fPCA.lac,scale=TRUE,center=TRUE)
+set.seed(rnd_seed)
+tsn.lac <- Rtsne(fPCA.lac,perplexity=floor(nrow(fPCA.lac)/5))
+
+pD.lac$tSNE1 <- tsn.lac$Y[,1]
+pD.lac$tSNE2 <- tsn.lac$Y[,2]
+
+pD.lac <- arrange(pD.lac, PassAll)
+
+tsnPlot <- ggplot(pD.lac, aes(x=tSNE1, y=tSNE2, color=PassAll, shape=Replicate)) +
+    geom_point(alpha=0.8,size=3) +
+    scale_color_manual(values=c("dodgerblue","red")) +
+    theme_bw() 
+tsnPlot
+
+kable(table(pD.lac$PassAll,pD.lac$Replicate))
+```
+
+# Normalization and first visualization
 ```{r}
 clusters <- quickCluster(m.filtered)
-pD.filtered$sf <- computeSumFactors(m.filtered,clusters=clusters,positive=TRUE)
+pD.filtered$sf <- computeSumFactors(m.filtered,clusters=clusters)
 plot(log10(colSums(m.filtered))~log10(pD.filtered$sf),main="Library Size versus Size Factors")
 m.norm <- t(t(m.filtered)/pD.filtered$sf)
-m.norm <- m.norm[,!is.na(colSums(m.norm))]
-pD.filtered <- filter(pD.filtered, barcode %in% colnames(m.norm))
 # 
@@
 pD.filtered$PC3 <- pc$x[,3]
+pD.filtered$PC4 <- pc$x[,4]
 pD.filtered$tSNE1 <- tsn$Y[,1]
@@
 pD.filtered$Acta2 <- t(m.norm)[,Acta2]
-Fcer1g <- filter(fD.filtered, symbol=="Fcer1g") %>% .$id
-pD.filtered$Fcer1g <- t(m.norm)[,Fcer1g]
+Csn2 <- filter(fD.filtered, symbol=="Csn2") %>% .$id
+pD.filtered$Csn2 <- t(m.norm)[,Csn2]
 # 
 pcaPlot <- ggplot(pD.filtered, aes(x=PC1, y=PC2, color=Condition, shape=Replicate)) +
+    geom_point(alpha=0.8,size=2) +
+    theme_bw() 
+pcaPlot
+
+pcaPlot <- ggplot(pD.filtered, aes(x=PC1, y=PC2, color=log2(Csn2+1), shape=Replicate)) +
     geom_point(alpha=0.8,size=2) +
@@
 
+tsnePlot <- ggplot(pD.filtered, aes(x=tSNE1, y=tSNE2, color=log2(Csn2+1), shape=Replicate)) +
+    geom_point(alpha=0.8,size=2) +
+    theme_bw()
+tsnePlot
+
 ```
 ***
-## Heatmap on top50
-```{r, fig.height=10}
-gns <- highVar[1:50]
-m.ht <- log2(m.filtered[gns,]+1)
-rownames(fD) <- fD$id
-rownames(m.ht) <- fD[rownames(m.ht),"symbol"]
-colanno <- data.frame("Condition"=pD.filtered$Condition)
-rownames(colanno) <- as.character(pD.filtered$barcode)
-library(pheatmap)
-pheatmap(m.ht, show_colnames=FALSE, annotation_col=colanno, show_rownames=TRUE)
-
-```
 ## Save Data

### 2017-01-18 src/functions.R

@@ -3,3 +3,51 @@ cv2 <- function(x) {
     out <- (sd(x)/mean(x))^2
 }
 
+normAndDimR <- function(m, pD, fD, top=0.1, scale_true=TRUE) {
+
+	#Thresholding on lowly expressed genes
+	m <- m[fD$keep,]
+	fD <- fD[fD$keep,]
+
+	clusters <- quickCluster(m)
+	pD$sf <- computeSumFactors(m,clusters=clusters,positive=TRUE)
+	m.norm <- t(t(m)/pD$sf)
+	m.norm <- m.norm[,!is.na(colSums(m.norm))]
+	pD <- filter(pD, as.character(barcode) %in% colnames(m.norm))
+	# 
+
+	fit <- trendVar(log10(m.norm+1), trend="loess")
+	decVar <- decomposeVar(log10(m.norm+1),fit=fit)
+
+	fD <- mutate(fD, meanExpression=rowMeans(m.norm),
+		     CV2=apply(m.norm,1,cv2),
+		     dm=DM(meanExpression,CV2))
+
+	highVar <- dplyr::arrange(fD,desc(dm)) %>% .$id
+	topX <- top*length(highVar)
+	fD$highVar <- fD$id %in% highVar[1:topX]
+	#         fD$highVar <- fD$id %in% rownames(decVar[decVar$FDR <0.05,])
+
+	fPCA <- log2(t(m.norm[fD$highVar,])+1)
+	fPCA <- scale(fPCA,scale=scale_true,center=TRUE)
+	pc <- prcomp(fPCA)
+	set.seed(rnd_seed)
+	tsn <- Rtsne(fPCA,perplexity=floor(ncol(fPCA)/10))
+
+	Acta2 <- filter(fD, symbol=="Acta2") %>% .$id
+	pD$Acta2 <- t(m.norm)[,Acta2]
+	Cd24 <- filter(fD, symbol=="Cd24a") %>% .$id
+	pD$Cd24 <- t(m.norm)[,Cd24]
+	Gata3 <- filter(fD, symbol=="Gata3") %>% .$id
+	pD$Gata3 <- t(m.norm)[,Gata3]
+	Csn2 <- filter(fD, symbol=="Csn2") %>% .$id
+	pD$Csn2 <- t(m.norm)[,Csn2]
+
+	pD$PC1 <- pc$x[,1]
[…]

### 2017-01-26 presentations/LabMeeting27012016.Rmd
First version of presentation

---
title: "Lab Meeting"
author: |
  | Karsten Bach
date: '`r Sys.Date()`'
output:
  beamer_presentation:
    font: "structurebold"
    fig_caption: FALSE
    includes:
     in_header: header.tex
jont-family: 'Helvetica'
---

# Understanding Breast Cancer Heterogeneity
![](img/CellOfOrigin.pdf)

# How to identify the cell of origin
\columnsbegin
\column{0.6\textwidth}
\begin{enumerate}
\item Expression of marker genes
\item Promoter driven oncogenes
\item Lineage tracing
\item Single-cell Genomics
\end{enumerate}
\column{0.5\textwidth}
![](img/CellOfOrigin2.pdf)
\columnsend

# Motivation
\begin{center}
\includegraphics{img/CellOfOrigin3.pdf}
\end{center}

# Experimental Setup
![](img/setup1.pdf)

# Experimental Setup
![](img/setup2.pdf)

# Experimental part
## Conclusions
- 8 samples is the maximum with this protocol 
- Cell number is the main limitation
    - 20 - 100k per sample after ~20min sort
- Accurate cell counting is important

## How to improve
- Optimize digestion 
- Sort in small volume to avoid spinning \& straining
- Count every sample

# Computational Analysis
\vspace{15pt}
\hspace{50pt}
![](img/CompAnalysis.pdf)
```{r echo=FALSE, message=FALSE, cache=TRUE, warning=FALSE, fig.keep="none"} 
library(knitr)
setwd("../src/")
source(purl("QCAnalysis.Rmd",output=tempfile(),quiet=TRUE),
       echo=FALSE, verbose=FALSE, prompt.echo=FALSE)
setwd("../presentations/")
```

# Overall cell numbers
```{r echo=FALSE, message=FALSE}
library(knitr)
library(dplyr)
library(ggplot2)
library(viridis)
kable(t(as.matrix(table(pD$SampleID))), caption="Number of sequenced cells per sample", format="pandoc")
kable(t(as.matrix(table(pD$Condition))), caption="Number of sequenced cells",format="pandoc")
```

# Libraries from lactation sample have low complexity
- [Lactation Sample 2](../data/CellRangerReports/SampleF1_summary.html)
- [Virgin Sample 2](../data/CellRangerReports/SampleG1_summary.html)

# QC stats - Number of Genes detected
```{r echo=FALSE}
genesDetected
```

# QC stats - Number of Genes Threshold
```{r echo=FALSE, warning=FALSE, message=FALSE}
gdHist
```

# QC stats - Library size
```{r echo=FALSE}
LibrarySize
```

# QC stats - Library size Threshold
```{r echo=FALSE, warning=FALSE, message=FALSE}
libSizeHist
```

# QC stats - Most highly expressed genes
```{r echo=FALSE}
topGenes
```

# QC stats - Ratio of Mitochondrial Genes
```{r echo=FALSE}
cellViability
```

# What are we left with
```{r echo=FALSE, warning=FALSE}
kable(table(pD$Condition,pD$PassAll), caption="Cells retained after QC", format="pandoc")
```

# Presumably fair representation of lactation sample
```{r echo=FALSE, warning=FALSE}
tsnPlot
```

# The "data"
```{r echo=FALSE}
knitr::knit_hooks$set(mysize = function(before, options, envir) {
  if (before) 
    return(options$size)
})
```
```{r echo=TRUE, mysize=TRUE, size="\\footnotesize", comment=""}
m.filtered[1:5,1:2]
dim(m.filtered)
```

# Highly variable Genes
```{r echo=FALSE, warning=FALSE}
genes2lbl <- c("Acta2", "Elf5", "Csn2", "Wap","Krt14", "Pgr", "Prlr", "Gata3")
lblDat <- filter(fD.filtered, symbol %in% genes2lbl)
V2 <- ggplot(arrange(fD.filtered,highVar), aes(x=meanExpression,y=CV2, color=highVar)) +
    geom_point(data=lblDat, aes(x=meanExpression, y=CV2), pch=1, size=4) +
    geom_point(alpha=0.8) +
    geom_label(data=lblDat, aes(x=meanExpression, y=CV2, label=symbol), nudge_y=0.1) +
    scale_x_log10() +
    scale_y_log10() +
    theme_bw()
V2
```

# General structure
```{r echo=FALSE, warning=FALSE}
library(viridis)
pcaPlot <- ggplot(pD.filtered, aes(x=PC1, y=PC3, color=Condition, shape=Replicate)) +
    geom_point(alpha=0.8,size=2) +
    theme_bw() 
pcaPlot
```

# General structure - Milk Proteins
```{r echo=FALSE, warning=FALSE}
pcaPlot <- ggplot(pD.filtered, aes(x=PC1, y=PC3, color=log2(Csn2+1), shape=Replicate)) +
    geom_point(alpha=0.8,size=2) +
    scale_color_viridis() +
    theme_bw() 
pcaPlot
```

# General structure - Basal Marker
```{r echo=FALSE, warning=FALSE}
pcaPlot <- ggplot(pD.filtered, aes(x=PC1, y=PC2, color=log2(Acta2+1), shape=Replicate)) +
    geom_point(alpha=0.8,size=2) +
    scale_color_viridis() +
    theme_bw() 
pcaPlot
```

# General structure - tSNE
```{r echo=FALSE, warning=FALSE}
Krt14 <- filter(fD.filtered, symbol=="Krt14") %>% .$id
pD.filtered$Krt14 <- t(m.norm)[,Krt14]
tsnePlot <- ggplot(pD.filtered, aes(x=tSNE1, y=tSNE2, color=Condition, shape=Replicate)) +
    geom_point(alpha=0.8,size=2) +
    theme_bw()
tsnePlot
```

# General structure - Basal cells
```{r echo=FALSE, warning=FALSE}
tsnePlot <- ggplot(pD.filtered, aes(x=tSNE1, y=tSNE2, color=log2(Krt14+1), shape=Replicate)) +
    geom_point(alpha=0.8,size=2) +
    scale_color_viridis() +
    theme_bw()
tsnePlot
```

# General structure - Cd74 cells?
```{r echo=FALSE, warning=FALSE}
Cd74 <- filter(fD.filtered, symbol=="Cd74") %>% .$id
pD.filtered$Cd74 <- t(m.norm)[,Cd74]
tsnePlot <- ggplot(pD.filtered, aes(x=tSNE1, y=tSNE2, color=log2(Cd74+1), shape=Replicate)) +
    geom_point(alpha=0.8,size=2) +
    scale_color_viridis() +
    theme_bw()
tsnePlot
```

# Clustering
![](img/Clustering.pdf)

### 2017-01-27 presentations/LabMeeting27012016.Rmd
final version
Latest update before Labmeeting

@@
 # Experimental Setup
+![](img/setup0.pdf)
+
+# Experimental Setup
 ![](img/setup1.pdf)
@@
 ![](img/setup2.pdf)
+
+# Experimental Setup
+![](img/setup3.pdf)
 
@@
 ![](img/Clustering.pdf)
+
+# Outlook
+## Analysis
+- Identify clusters and relate them to cell types
+- Differentially expressed genes within each cell type
+- Differentiation trajectories
+    - e.g. in the luminal department
+- Testing specific hypotheses
+
+## Future
+- Plan experiment for tumor development
+
+#
+
+```{r echo=FALSE, warning=FALSE}
+Bcl11a <- filter(fD, symbol=="Bcl11a") %>% .$id
+pD.filtered$Bcl11a <- t(m[,as.character(pD.filtered$barcode)])[,Bcl11a]/pD.filtered$sf
+tsnePlot <- ggplot(pD.filtered, aes(x=tSNE1, y=tSNE2, color=log2(Bcl11a+1), shape=Replicate)) +
+    geom_point(alpha=0.8,size=2) +
+    scale_color_viridis() +
+    theme_bw()
+tsnePlot
+```

### 2017-01-27 src/QCAnalysis.Rmd
Latest update before Labmeeting

@@
 pD$GenesDetected <- colSums(m!=0)
-genesDetected <- ggplot(pD, aes(x=SampleID,y=log10(GenesDetected),fill=Condition)) +
+genesDetected <- ggplot(pD, aes(x=SampleID,y=GenesDetected,fill=Condition)) +
     geom_violin(draw_quantiles=0.5)+
+    scale_y_log10() +
     theme_bw()
-LibrarySize <- ggplot(pD, aes(x=SampleID,y=log10(UmiSums),fill=Condition)) +
+LibrarySize <- ggplot(pD, aes(x=SampleID,y=UmiSums,fill=Condition)) +
     geom_violin(draw_quantiles=0.5)+
+    scale_y_log10() +
     theme_bw()
@@
 
-gdHist <- ggplot(pD, aes(x=log10(GenesDetected),y=..density..)) +
+gdHist <- ggplot(pD, aes(x=GenesDetected,y=..density..)) +
     geom_histogram(fill="white",color="black") +
-    geom_vline(data=smryByGroup,aes(xintercept=threshold_GenesDetected),color="red",lty="longdash") +
+    geom_vline(data=smryByGroup,aes(xintercept=10^threshold_GenesDetected),color="red",lty="longdash") +
+    scale_x_log10() +
     facet_wrap(~Condition) +
@@
 
-libSizeHist <- ggplot(pD, aes(x=log10(UmiSums),y=..density..)) +
+libSizeHist <- ggplot(pD, aes(x=UmiSums,y=..density..)) +
     geom_histogram(fill="white",color="black") +
-    geom_vline(data=smryByGroup,aes(xintercept=threshold_UmiSums),color="red",lty="longdash") +
+    geom_vline(data=smryByGroup,aes(xintercept=10^threshold_UmiSums),color="red",lty="longdash") +
+    scale_x_log10() +
     facet_wrap(~Condition) +
@@
 set.seed(rnd_seed)
-tsn.lac <- Rtsne(fPCA.lac,perplexity=floor(nrow(fPCA.lac)/5))
+tsn.lac <- Rtsne(fPCA.lac,perplexity=25)
 
@@
     geom_point(alpha=0.8,size=3) +
-    scale_color_manual(values=c("dodgerblue","red")) +
+    scale_color_manual(values=c("grey50","black")) +
     theme_bw() 
@@
 
-highVar <- dplyr::arrange(fD.filtered,desc(dm)) %>% .$id
-topX <- floor(0.2*length(highVar))
+#HVG definition by Brennecke
+brennecke <- BrenneckeHVG(m.norm,fdr=0.1,minBiolDisp=0.25)
+fD.filtered$highVar <- fD.filtered$id %in% brennecke
 
-fD.filtered$highVar <- fD.filtered$id %in% highVar[1:topX]
-VarPlot <- ggplot(fD.filtered, aes(x=log10(meanExpression),y=log10(CV2), color=highVar)) +
+VarPlot <- ggplot(fD.filtered, aes(x=meanExpression,y=CV2, color=highVar)) +
     geom_point() +
+    scale_x_log10() +
+    scale_y_log10() +
     theme_bw()
@@
 set.seed(rnd_seed)
-tsn <- Rtsne(fPCA,perplexity=floor(nrow(fPCA)/5))
+tsn <- Rtsne(fPCA,perplexity=25)
 
@@
 
-pcaPlot <- ggplot(pD.filtered, aes(x=PC1, y=PC2, color=log2(Csn2+1), shape=Replicate)) +
+pcaPlot <- ggplot(pD.filtered, aes(x=PC1, y=PC3, color=log2(Csn2+1), shape=Replicate)) +
     geom_point(alpha=0.8,size=2) +

### 2017-01-27 src/functions.R

@@ -51,3 +51,138 @@ normAndDimR <- function(m, pD, fD, top=0.1, scale_true=TRUE) {
 	out <- pD
 	return(out)
 }
+
+plotGeneDist <- function(m, pD, fD, genes) {
+    require(reshape)
+    stopifnot(identical(rownames(m),fD$id))
+    rownames(m) <- fD$symbol
+    feat <- filter(fD, id %in% genes) %>% .$symbol
+    if (is.null(feat)) {
+	feat <- genes
+    }
+    exps <- as.data.frame(cbind(t(m)[,feat],"barcode"=colnames(m)))
+    pltDat <- left_join(select(pD, barcode, Condition),exps) %>% melt(.,id=c("barcode","Condition"))
+    plt <- ggplot(pltDat, aes(x=as.factor(variable),y=log2(as.numeric(value)+1))) +
+	geom_violin() +
+	geom_point(position="jitter",alpha=0.3,shape=19,aes(color=Condition)) +
+	theme_bw()
+    return(plt)
+}
+
+compClustering <- function(m,fD,fs="brennecke",dm="spearman",lk="ward.D2",ds=0) {
+
+    #Feature Selection
+    keep <- fD[,fs]
+    trafM <- log2(m[keep,]+1)
+
+    #Dissimilarity Measure
+    if(dm!="euclidean") {
+    dis <- as.dist((1-cor(trafM,method=dm))/2)
+    } else {
+	dis <- dist(t(trafM),method=dm)
+    }
+
+    #Hierarchical Clustering
+    htree <- hclust(dis, method=lk)
+    #Tree Cut
+    cluss <- cutreeDynamic(htree,distM=as.matrix(dis),deepSplit=ds,minClusterSize=10)
+
+    #Compute Validation Statistics
+    clstats <- cluster.stats(d=dis, cluss)
+    asw <- clstats$avg.silwidth
+    rss <- clstats$within.cluster.ss
+    con <- connectivity(distance=dis,clusters=cluss)
+
+    #Compile output
+    out <- data.frame("FeatureSelection"=fs,
+		      "Dissimilarity"=dm,
+		      "Linkage"=lk,
+		      "DeepSplit"=ds,
+		      "Statistic"=c("Average Silhouette Width",
+				    "Within Cluster SS",
+				    "Connectivity"),
[…]

### 2017-01-30 src/QCAnalysis.Rmd
Releved condition and added sf and highVar to output

@@
 fD <- dataList[["featureData"]]
+#relevel pD Condition to get it in logical sequence
+pD$Condition <- factor(pD$Condition,levels=c("V","P","L","I"))
 ```
@@
 
-# Normalization and first visualization
+# Normalization and Highly Variable Gene Selection
 ```{r}
@@
 VarPlot
+```
 
+# First Visualization
+```{r}
 fPCA <- log2(t(m.norm[fD.filtered$highVar,])+1)
@@
 ```{r}
+pD.add <- select(pD.filtered, barcode, sf, tSNE1, tSNE2)
+fD.add <- select(fD.filtered, id, highVar)
+pD <- left_join(pD,pD.add, by="barcode")
+fD <- left_join(fD,fD.add, by="id")
+stopifnot(identical(as.character(pD$barcode),colnames(m)))
+stopifnot(identical(as.character(fD$id),rownames(m)))
 out <- list()

### 2017-02-01 src/functions.R
bugfixes in plotGeneDist

@@ -60,14 +60,15 @@ plotGeneDist <- function(m, pD, fD, genes, colorBy="Condition") {
     if (length(feat)==0) {
 	feat <- genes
     }
-    exps <- as.data.frame(cbind(t(m)[,feat],"barcode"=colnames(m)))
+    exps <- data.frame(t(m)[,feat])
+    exps$barcode <- rownames(exps)
     pltDat <- left_join(select_(pD, "barcode", colorBy),exps) %>% melt(.,id=c("barcode",colorBy))
-    pltDat$value <- log2(as.numeric(pltDat$value)+1)
+    pltDat$value <- log2(pltDat$value+1)
     pltDat[,colorBy] <- as.factor(pltDat[,colorBy])
     plt <- ggplot(pltDat, aes_string(x=colorBy,y="value")) +
-	geom_violin() +
-	geom_point(position="jitter",alpha=0.3,shape=19) +
-	facet_wrap(~variable) +
+	geom_boxplot() +
+	geom_point(position="jitter",alpha=0.2,shape=19) +
+	facet_wrap(~variable, scales="free") +
 	theme_bw()
     return(plt)
 }

### 2017-02-01 src/ClusterComparison.Rmd
Added Clustercomparison markdown

--- 
title: "Interactive Markdown to explore Clustering"
author: "Karsten Bach"
date: '`r Sys.Date()`'
runtime: shiny
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    theme: united
    highlight: tango
---
***

```{r echo=FALSE, message=FALSE}
library(scran)
library(dplyr)
library(gtools)
library(knitr)
library(ggplot2)
library(dynamicTreeCut)
library(Rtsne)
library(pheatmap)
source("functions.R")

rnd_seed <- 300
clustDat <- readRDS("../data/Robjects/ClusterComparison.rds") 
bootDat <- readRDS("../data/Robjects/ClusterBootstrap.rds")
dataList <- readRDS("../data/Robjects/ExpressionList_QC.rds")

#Commence
m <- dataList[[1]]
pD <- dataList[[2]]
fD <- dataList[[3]]
#Gene and Cell filtering
m <- m[fD$keep,pD$PassAll]
pD <- pD[pD$PassAll,]
fD <- fD[fD$keep,]
#Normalization
m <- t(t(m)/pD$sf)
#Identification of HVG 
# fD <- mutate(fD, meanExpression=rowMeans(m),
#              CV2= apply(m,1,cv2))
# 
# genes2lbl <- c("Acta2", "Krt18", "Csn2", "Wap","Krt14")
# lblDat <- filter(fD, symbol %in% genes2lbl)
# V3 <- ggplot(arrange(fD,highVar), aes(x=meanExpression,y=CV2, color=highVar)) +
#     geom_label(data=lblDat, aes(x=meanExpression, y=CV2, label=symbol), nudge_y=0.1) +
#     geom_point(data=lblDat, aes(x=meanExpression, y=CV2), pch=1, size=4) +
#     geom_point(alpha=0.8) +
#     scale_x_log10() +
#     scale_y_log10() +
#     theme_bw()
# V3

#Average Linkage appears to best capture the dendrogram

clustDat <- filter(clustDat, Statistic=="Average Silhouette Width")
compPlot <- ggplot(clustDat, aes(x=as.character(DeepSplit), y=Value, color=Linkage,
			    group=Linkage)) +
geom_point() +
geom_path() +
facet_grid(.~Dissimilarity, scales="free_y")
compPlot

#Which Linkage is best
m.sub <- m[fD$highVar,]
# trafM <- log2(m.sub+1)
# dms <- c("pearson","spearman","euclidean")
# res <- NULL
# for (dm in dms) {
# Dissimilarity Measure
#     if(dm!="euclidean") {
#     dis <- as.dist((1-cor(trafM,method=dm)))
#     } else {
#         dis <- dist(t(trafM),method=dm)
#     }
# 
# Hierarchical Clustering
#     htree.average <- hclust(dis, method="average")
#     htree.single <- hclust(dis, method="single")
#     htree.complete <- hclust(dis, method="complete")
#     htree.ward <- hclust(dis, method="ward.D2")
# 
#     ccc.average <- cor(cophenetic(htree.average),dis)
#     ccc.single <- cor(cophenetic(htree.single),dis)
#     ccc.complete <- cor(cophenetic(htree.complete),dis)
#     ccc.ward <- cor(cophenetic(htree.ward),dis)
# 
#     tmp <- data.frame("Linkage"=c("Average",
#                                          "Single",
#                                          "Complete",
#                                          "Ward.D2"),
#                              "Cophenetic Correlation Coefficient"=
#                                  c(ccc.average,ccc.single,
#                                    ccc.complete,ccc.ward),
#                              "Dissimilarity"=dm)
#     res <- rbind(res,tmp)
# }
# kable(res)

inputPanel(
	   selectInput("dm", label = "Dissimilarity",
		       choices= c("euclidean","pearson","spearman"), selected="pearson"),

	   selectInput("ds", label = "DeepSplit",
		       choices= c(0,1,2,3,4), selected=0),

	   selectInput("lk", label = "Linkage",
		       choices=c("average","complete","single","ward.D2"), selected="average")
	  )

#Visualize Dendrograms
clustering <- reactive({
    dynamicCluster(t(m.sub), dm=input$dm,
				 lk=input$lk,
				 ds=as.numeric(input$ds),
				 output="Cluster")
})

renderTable({
    clust <- clustering()
    x <- table(clust$cluster,pD$Condition)
    out <- data.frame("Cluster"=rownames(x),
	       "Virgin"=unname(x[,"V"]),
	       "Pregnant"=unname(x[,"P"]),
	       "Lactation"=unname(x[,"L"]),
	       "Involution"=unname(x[,"I"]))
})

renderPlot({
    clust <- clustering()
    nclust <- length(unique(clust$cluster))
    palette(rainbow(nclust))
    htreeWithBars(clust,pD)
}, width=2000, height=900)

renderPlot({
    clust <- clustering()
    nclust <- length(unique(clust$cluster))
    pD$cluster <- factor(clust$cluster)
    tsnPlot <- ggplot(pD, aes(x=tSNE1, y=tSNE2, color=cluster, shape=Replicate)) +
	geom_point(size=4) +
	scale_color_discrete(rainbow(nclust))+
	theme_bw()
    tsnPlot
}, width=800, height=800)

renderPlot({
    clust <- clustering()
    nclust <- length(unique(clust$cluster))
    palette(rainbow(nclust))
    slct <- paste(input$dm,input$lk,input$ds,sep="_")
    boot <- bootDat[[slct]]$bootresult
    boxplot(t(boot),ylim=c(0,1),col=c(1:nclust))

}, width=800, height=800)

renderPlot({
    library(cluster)
    clust <- clustering()
    nclust <- length(unique(clust$cluster))
    palette(rainbow(nclust))
    sils  <- silhouette(x=clust$cluster,dist=clust$dis)
    plot(sils,col=c(1:length(unique(clust$cluster))), border=NA)

}, width=1000, height=1800)

renderPlot({
    clust <- clustering()
    pD$cluster <- clust$cluster
    nclust <- length(unique(clust$cluster))
    ord <- arrange(pD, cluster) 
    simMat <- asSim(clust$dis)
    simMat.ord <- as.matrix(simMat)[as.character(ord$barcode),
				    as.character(ord$barcode)]
    library(pheatmap)
    annoCol <- data.frame("cluster"=as.factor(ord$cluster),
			  "Condition"=ord$Condition)
    rownames(annoCol) <- as.character(ord$barcode)
    ann_colors <- list("cluster"=rainbow(nclust))
    names(ann_colors$cluster) <- c(min(ord$cluster):max(ord$cluster))
    pheatmap(simMat.ord,
	     cluster_rows=FALSE,
	     cluster_cols=FALSE,
	     show_rownames=FALSE,
	     show_colnames=FALSE,
	     annotation_col=annoCol,
	     annotation_colors=ann_colors)
}, width=1200, height=1200)

inputPanel(
	   selectInput("gene", label = "GeneOfInterest",
		       choices= mixedsort(fD$symbol), selected=c("Krt18","Acta2","Csn2","Cd74")
,
		       multiple=TRUE, selectize=FALSE)
	  )

renderPlot({
    clust <- clustering()
    pD$cluster <- clust$cluster

    genes <- input$gene
	 plotGeneDist(m, pD, fD, genes, colorBy="cluster")
   }, width=800, height=800)

#Compare Cluster stability 
# boot1 <- bootDat[["euclidean_average_0"]]

#Clustering
# clust1 <- dynamicCluster(m.sub, dm="euclidean",
#                          lk="average",
#                          ds=0,
#                          output="Cluster")
# clust2 <- dynamicCluster(m.sub, dm="pearson",
#                          lk="average",
#                          ds=0,
#                          output="Cluster")
# clust3 <- dynamicCluster(m.sub, dm="pearson",
#                          lk="ward.D2",
#                          ds=0,
#                          output="Cluster")

#Looking a different clusters in depth
# 
# fPCA <- t(trafM)
# fPCA <- scale(fPCA,scale=TRUE,center=TRUE)
# set.seed(rnd_seed)
# tsn <- Rtsne(fPCA,perplexity=10,initial_dims=5)
# pD$tSNE1 <- tsn$Y[,1]
# pD$tSNE2 <- tsn$Y[,2]
# tsnePlot <- ggplot(pD, aes(x=tSNE1, y=tSNE2, shape=Condition, color=as.factor(cluster))) +
#     geom_point(alpha=0.8,size=2) +
#     theme_bw()
# tsnePlot
# 
# tsnePlot <- ggplot(pD.filtered, aes(x=tSNE1, y=tSNE2, color=log2(Acta2+1), shape=Replicate)) +
#     geom_point(alpha=0.8,size=2) +
#     theme_bw()
# tsnePlot
```

### 2017-02-03 src/ClusterComparison.Rmd
as viewed with WTK and SP

@@
 				 ds=as.numeric(input$ds),
-				 output="Cluster")
+				 output="Cluster",
+				 minSize=15)
 })
@@
 
+renderPlot({
+    clust <- clustering()
+    pD$cluster <- clust$cluster
+
+    genes <- c("Acta2","Csn3","Epcam","Krt8")
+	 plotGeneDist(m, pD, fD, genes, colorBy="cluster")
+   }, width=1000, height=1000)
+
 inputPanel(

### 2017-02-28 src/DEAnalysis.R
included PValue ranks in output

@@ -44,7 +44,7 @@ fit <- glmFit(y, de.design)
 #copy cat
 library(doParallel)
 nCores <- 3
-cl <-makeCluster(nCores, type="FORK")
+cl <-makeCluster(nCores, type="FORK"
 registerDoParallel(cl)
 result <- foreach (i=seq_along(levels(cluster))) %dopar% {
     result.logFC <- result.PValue <- list()
@@ -59,13 +59,15 @@ result <- foreach (i=seq_along(levels(cluster))) %dopar% {
 	res <- glmTreat(fit, contrast=contrast, lfc=1)
 	con.name <- paste0('vs.', levels(cluster)[clust])
 	result.logFC[[con.name]] <- res$table$logFC
-	result.PValue[[con.name]] <- res$table$PValue
+	result.PValue[[con.name]] <- rank(res$table$PValue, ties="first")
     }
 
     collected.ranks <- lapply(result.PValue, rank, ties="first")
     min.rank <- do.call(pmin, collected.ranks)
     marker.set <- data.frame(Top=min.rank, Gene=rownames(y),
-			     logFC=do.call(cbind, result.logFC), stringsAsFactors=FALSE)
+			     logFC=do.call(cbind, result.logFC),
+			     top=do.call(cbind, result.PValue),
+			     stringsAsFactors=FALSE)
     marker.set <- marker.set[order(marker.set$Top),]
     return(marker.set)
 }

### 2017-02-28 src/QCAnalysis.Rmd
added tSNE based on cc-annotated genes

@@
 
+#tsnePlot based on cell Cycle
+library(scLVM)
+ens_ids_cc <- getEnsembl('GO:0007049')
+sel <- rownames(m.norm) %in% ens_ids_cc
+fPCA <- log2(t(m.norm[sel,])+1)
+fPCA <- scale(fPCA,scale=TRUE,center=TRUE)
+set.seed(rnd_seed)
+tsn <- Rtsne(fPCA,perplexity=25)
+
+pD.filtered$cc_tSNE1 <- tsn$Y[,1]
+pD.filtered$cc_tSNE2 <- tsn$Y[,2]
+
+tsnePlot <- ggplot(pD.filtered, aes(x=cc_tSNE1, y=cc_tSNE2, color=Condition, shape=Replicate)) +
+    geom_point(alpha=0.8,size=2) +
+    theme_bw()
+tsnePlot
 ```
+
 ***

### 2017-02-28 src/ShinyApp.R
added GO-Analaysis and adjusted it to fit to the new output from DEAnalysis

@@ -1,3 +1,5 @@
+library(topGO)
+library(org.Mm.eg.db)
 library(shiny)
 library(dplyr)
 library(gtools)
@@ -30,11 +32,19 @@ fD <- fD[keep,]
 #Normalization
 m <- t(t(m)/pD$sf)
 
+#Add cell surface annotation
+library(scLVM)
+surface <- getEnsembl("GO:0009986")
+
 deAdd <- select(fD, id,symbol) 
 colnames(deAdd) <- c("Gene","GeneSymbol")
 deList <- lapply(deList, function(x) {
 		      nw <- left_join(x,deAdd)
-		      nw <- nw[,-2]})
+		      nw$Surface <- nw$Gene %in% surface
+		      nw <- nw[,-2]
+		      cols <- c("GeneSymbol",grep("top",colnames(nw),value=TRUE, #reorder cols
+						  ignore.case=TRUE))
+		      nw <- nw[,c(cols,setdiff(colnames(nw), cols))]})
 names(deList) <- c(1,2,3,4,5,6,7,9,10)
 
 ui <- shinyUI(fluidPage(
@@ -49,8 +59,13 @@ ui <- shinyUI(fluidPage(
 			   choices= c("cluster","Condition"), selected="cluster"),
 	       selectInput("cNumber", label = "Cluster",
 			   choices= as.character(c(1,2,3,4,5,6,7,9,10)), selected="1"),
+	       selectInput("ontology", label = "Ontology",
+			   choices= c("BP","CC","MF"), selected="MF"),
 	       radioButtons("order", label = "Order DE-Table",
-			   choices= c("Up against all", "Down against all", "none"), selected="none")
+			   choices= c("Up", "Down", "none"), selected="none"),
+	       selectInput("against", label = "Compare Against",
+			   choices= as.character(c(1,2,3,4,5,6,7,9,10)),
+			   multiple=TRUE)
 			   
 	  ),plotOutput("distPlot1"),
 	    plotOutput("distPlot2")
@@ -62,7 +77,8 @@ ui <- shinyUI(fluidPage(
         tabsetPanel(
 		    tabPanel("Plots",splitLayout(plotOutput("tSNE"), plotOutput("tSNE2")),
 		dataTableOutput("table")),
-		    tabPanel("Table",verbatimTextOutput("table2")))
+		    tabPanel("Table",verbatimTextOutput("table2")),
+		    tabPanel("GO-Analysis",tableOutput("goTable")))
       )
[…]

### 2017-03-08 paper/main.tex
added paper

\documentclass[oneside]{amsart}
\usepackage{geometry}
\usepackage{times}

%% --for bibtex
\bibliographystyle{unsrt}
%\usepackage{cite}
\usepackage{natbib}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsfonts} %\mathbb
\usepackage{bm}
\usepackage{lineno}

\usepackage{subcaption}
\usepackage{graphicx}
\usepackage{hyperref}

\usepackage{siunitx}

\hypersetup{colorlinks=true,linkcolor=black,citecolor=black,urlcolor=black}
\urlstyle{same}

\newcommand{\fixme}[1]{\textit{\textcolor{red}{Fixme: #1}}}
\newcommand{\comment}[1]{\textit{\textcolor{blue}{Comment: #1}}}
%\DeclareSIUnit\Molar{\textsc{m}} % add molar as captial m to siunitx

\title{Single-cell RNA-sequencing reveals cellular dynamics of the developing mammary gland}
\author{Karsten Bach, Sara Pensa, Walid T. Khaled, John C. Marioni}
\begin{document}
\begin{abstract}
    Bla bla bla
\end{abstract}

\maketitle
\tableofcontents
\newpage
\section{Introduction}

\section{Results}
%------------------------------ Section 1 -------------------------------------------------------
\subsection{Unbiased identification of mammary epithelial cells\label{sec:cluster}}
\begin{itemize}
    \item Experimental Setup, 4 stages of the development of an adult gland
    \item Clustering into major cell types in the mammary gland
    \item Luminal cells: Hormone sensing (HS) (progenitor and differentiated), Non-hormone sensing (NHS) progenitor and secretory cells
    \item Basal cells: MaSCs, PROCR\textsuperscript{+} MaSCs and differentiated myoepithelial cells
\end{itemize}

%------------------------------ Figure 1 -------------------------------------------------------
\begin{figure}[h]
\begin{subfigure}{0.45\linewidth}
\centering %\textbf{a}
\fbox{\parbox{0.9\linewidth}{(Experimental overview)}}
%\includegraphics[height=5cm]{../figures/ROC_Staurosporine.pdf}
\caption{}
\end{subfigure}
\begin{subfigure}{0.45\linewidth}
%\centering %\textbf{a}
    \fbox{\parbox{0.9\linewidth}{(tSNE plot colored by Condition)}}
%\includegraphics[height=5cm]{../figures/ROC_Panobinostat.pdf}
\caption{}
\end{subfigure}
\begin{subfigure}{0.45\linewidth}
\centering %\textbf{a}
\fbox{\parbox{1\linewidth}{(Plot to illustrate clusters and marker genes)}}
%\includegraphics[height=5cm]{../figures/ROC_ATP.pdf}
\caption{}
\end{subfigure}
    \caption{Unbiased identification of mammary epithelial cells}
      \label{fig:f1}
  \end{figure}

\newpage
%------------------------------ Section 2 -------------------------------------------------------
\subsection{The luminal compartment is defined by a continuous spectrum of differentiation\label{sec:LumDiff}}
\comment{Here I would start with only looking at virgin and pregnancy arguing that here we would expect all celltypes. 
I would then describe the parity effect in the next section. Maybe there is a more elegant way of disentagling the two.}
\begin{itemize}
    \item Basal cells are separate from the luminal compartment, suggesting either:
	\begin{itemize}
	    \item This is a rare event 
	    \item Intermediate stages are highly unstable and the transdifferentation is rapid
	\end{itemize}
    \item In the virgin gland there is one ``branch'' of differentiation with NHS progenitors on one side and differentiated HS cells on the other end
    \item In pregnancy another branch derives from the NHS progenitors with differentiated secretory cells at the tip
    \item We conclude that there is a luminal hierarchy with NHS progenitors that can give rise to both HS progenitors/differentiated cells and secretory cells
\end{itemize}

%------------------------------ Figure 2 -------------------------------------------------------
\begin{figure}[h]
\begin{subfigure}{0.45\linewidth}
\centering %\textbf{a}
\fbox{\parbox{0.9\linewidth}{(3D scatterplot of DiffusionMap V/P)}}
%\includegraphics[height=5cm]{../figures/ROC_Staurosporine.pdf}
\caption{}
\end{subfigure}
\begin{subfigure}{0.45\linewidth}
%\centering %\textbf{a}
    \fbox{\parbox{0.9\linewidth}{(Only luminal cells in virgin and pregnancy)}}
%\includegraphics[height=5cm]{../figures/ROC_Panobinostat.pdf}
\caption{}
\end{subfigure}
\begin{subfigure}{0.45\linewidth}
\centering %\textbf{a}
\fbox{\parbox{1\linewidth}{(Multiple plots to show gene expression pattern along the two branches)}}
%\includegraphics[height=5cm]{../figures/ROC_ATP.pdf}
\caption{}
\end{subfigure}
\caption{Differentiation hierarchy in the mammary gland}
      \label{fig:f2}
  \end{figure}

\newpage
%------------------------------ Section 3 -------------------------------------------------------
\subsection{Parity exerts cell type specific effects on the transcriptional landscape of luminal cells\label{sec:ParityEffect}}
\begin{itemize}
    \item Effect is stronger on luminal than on basal cells (\% variance explained?) [as has been shown before on epigenitic level]
    \item \fixme{Is it worth mentioning that the lactation sample FACS profile has completely lost its ``basal/luminal'' appearence, yet contains all cell types}
    \item DE genes between e.g. HS differentiated in involution versus virgin
    \item A ``new'' cell type arises after pregnancy that most likely derives from NHS progenitors 
    \item Highlight importance of this cell type as it sits at the top of the hierarchy
\end{itemize}

%------------------------------ Figure 4 -------------------------------------------------------
\begin{figure}[h]
\begin{subfigure}{0.45\linewidth}
\centering %\textbf{a}
\fbox{\parbox{0.9\linewidth}{(Condition factor explains more variance in the luminal compartment)}}
%\includegraphics[height=5cm]{../figures/ROC_Staurosporine.pdf}
\caption{}
\end{subfigure}
\begin{subfigure}{0.45\linewidth}
%\centering %\textbf{a}
    \fbox{\parbox{0.9\linewidth}{(Cluster 4 is closest to 5 and is at the top of the hierarchy after parity)}}
%\includegraphics[height=5cm]{../figures/ROC_Panobinostat.pdf}
\caption{}
\end{subfigure}
\begin{subfigure}{0.45\linewidth}
\centering %\textbf{a}
\fbox{\parbox{1\linewidth}{(Gene expression differences between 4 and 5)}}
%\includegraphics[height=5cm]{../figures/ROC_ATP.pdf}
\caption{}
\end{subfigure}
\caption{Parity exerts cell type specific effects on the transcriptional landscape of luminal cells}
      \label{fig:f3}
  \end{figure}

\section{Discussion}
\section{Methods}

\clearpage

%% --for biblatex
\bibliographystyle{unsrt}
\bibliography{nonparatpp}

%% --for biblatex
%\printbibliography

\end{document}

### 2017-03-08 src/DiffusionMap.R
added DifMap

@@ -1,0 +1,112 @@
+library(plyr)
+library(dplyr)
+library(destiny)
+library(ggplot2)
+source("functions.R")
+library(RColorBrewer)
+library(rgl)
+
+rnd_seed <- 300
+dataList <- readRDS("../data/Robjects/ExpressionList_Clustered.rds")
+m <- dataList[[1]]
+pD <- dataList[[2]]
+fD <- dataList[[3]]
+
+#Pre-Filtering before DE-Analysis
+# Cells
+keepCells <- pD$PassAll & !pD$isImmuneCell & !pD$isOutlier & !(pD$cluster %in% c(6,7,10)) 
+m <- m[,keepCells]
+pD <- pD[keepCells,]
+
+#Gene filtering
+keep <- rowMeans(m)>0.1
+m <- m[keep,]
+fD <- fD[keep,]
+
+# Genes
+#Pre-selected set
+# genes <- c("Csn2","Gata3","Prlr","Procr","Elf5","Esr1","Pgr","Aldh1a3","Wap",
+#            "Tspan8","Krt18","Krt8","Krt14","Krt5","Fgfr1","Areg","Fgfr2",
+#            "Notch1","Notch3","Foxc1","Zeb2")
+# fD$highVar <- fD$symbol %in% genes
+
+#Normalize
+m.norm <- t(t(m)/pD$sf)
+
+brennecke <- BrenneckeHVG(m.norm,suppress.plot=TRUE)
+fD$highVar <- fD$id %in% brennecke
+
+exps <- m.norm[fD$highVar,]
+exps <- t(log(exps+1))
+dm <- DiffusionMap(exps,n_eigs=20,k=50)
+dpt <- DPT(dm, branching=TRUE)
+plot(eigenvalues(dm)[1:20])
+
+root <- which(dpt@tips[,1])[3]
+rootdpt <- paste0("DPT",root)
+plot(dpt,pch=20,dcs=c(1,2),col_by=rootdpt)
+plot(dpt,pch=20,dcs=c(1,2),col_by="branch")
+
+# palet <- brewer.pal(length(levels(pD$Condition)),name="Paired")
+# cols <- mapvalues(pD$Condition,levels(pD$Condition),palet)
+# plot3d(eigenvectors(dm)[,c(1,2,3)],col=cols,radius=0.001, type="s")
+
+
+#Expression of Genes in the differentiation trajectories
+branch <- dpt@branch[,1]
+pseudot<- dpt[[rootdpt]]
[…]

### 2017-03-13 paper/main.tex
started to write

@@
 \newcommand{\comment}[1]{\textit{\textcolor{blue}{Comment: #1}}}
+\newcommand{\fref}[1]{Fig. \ref{#1}}
 %\DeclareSIUnit\Molar{\textsc{m}} % add molar as captial m to siunitx
@@
 \title{Single-cell RNA-sequencing reveals cellular dynamics of the developing mammary gland}
-\author{Karsten Bach, Sara Pensa, Walid T. Khaled, John C. Marioni}
+\author{Karsten Bach, Sara Pensa, David J. Adams, John C. Marioni, Walid T. Khaled}
 \begin{document}
 \begin{abstract}
-    Bla bla bla
+    \dots
 \end{abstract}
@@
 \section{Introduction}
-
+\textbf{Motivation:}
+\begin{itemize}
+    \item An unbiased, systematic overview of the different cell types in the gland 
+    \item Understand the ``stability'' of cell types across development and whether (known, e.g. from bulk seq) developmental effects are cell type specific
+    \item Highlight weaknesses of traditional approaches e.g. single-gene based lineage tracings, marker-based sortings/stainings etc. and controversies caused by this
+\end{itemize}
 \section{Results}
@@
 \subsection{Unbiased identification of mammary epithelial cells\label{sec:cluster}}
+In order to get a systematic overview over the cell types of the mammary gland and their dynamics across development we isolated and sequenced single-cells from four developmental stages of the adult mouse (\fref{fig:f1}a).
+Epithelial cells from the mammary glands were isolated by digesting the tissue for two hour followed by FACS sorting for EpCam\textsuperscript{+}, Cd49f\textsuperscript{+}, Lin\textsuperscript{-}.
+The single-cell RNA-sequencing libraries were prepared using 10x Genomics Chromium system.
+We sequenced a total of 5598 cells, 3471 of which passed the quality control criteria (2044 cells from the lactation samples were removed due to low gene count, see Sup1).
+
 \begin{itemize}
@@
     \item Basal cells: MaSCs, PROCR\textsuperscript{+} MaSCs and differentiated myoepithelial cells
+    \item highlight a few interesting examples, e.g.:
+	\begin{itemize}
+	    \item Prolactin receptor expressing cells are not secretory (\fixme{Check what isoforms we sequence})
+	    \item Although FACS profile completely different in lactation, (almost) all cell types represented
+	    \item some interesting genes expressed in various clusters
+	    \item contribution of conditions to each cluster
+	\end{itemize}
 \end{itemize}
@@
 \begin{figure}[h]
-\begin{subfigure}{0.45\linewidth}
-\centering %\textbf{a}
-\fbox{\parbox{0.9\linewidth}{(Experimental overview)}}
-%\includegraphics[height=5cm]{../figures/ROC_Staurosporine.pdf}
-\caption{}
-\end{subfigure}
-\begin{subfigure}{0.45\linewidth}
-%\centering %\textbf{a}
-    \fbox{\parbox{0.9\linewidth}{(tSNE plot colored by Condition)}}
-%\includegraphics[height=5cm]{../figures/ROC_Panobinostat.pdf}
-\caption{}
-\end{subfigure}
-\begin{subfigure}{0.45\linewidth}
-\centering %\textbf{a}
-\fbox{\parbox{1\linewidth}{(Plot to illustrate clusters and marker genes)}}
-%\includegraphics[height=5cm]{../figures/ROC_ATP.pdf}
-\caption{}
-\end{subfigure}
-    \caption{Unbiased identification of mammary epithelial cells}
+    \includegraphics[width=\linewidth]{figures/F1.pdf}
+    \caption{Unbiased identification of mammary epithelial cells.
+	(\textbf{a}) Experimental setup.
+	(\textbf{b}) tSNE plot of 3471 epithelial cells from all four timepoints.
+	(\textbf{c}) Cluster identified by hierarchical clustering with euclidean distances.
+	(\textbf{d}) Expression of marker genes for cell types of the luminal compartment.
+	(\textbf{e}) Marker gene expression in the basal compartment.
+}
 \label{fig:f1}
@@
 %------------------------------ Section 2 -------------------------------------------------------
-\subsection{The luminal compartment is defined by a continuous spectrum of differentiation\label{sec:LumDiff}}
-\comment{Here I would start with only looking at virgin and pregnancy arguing that here we would expect all celltypes. 
-I would then describe the parity effect in the next section. Maybe there is a more elegant way of disentagling the two.}
+\subsection{The luminal compartment defines a continuous spectrum of differentiation\label{sec:LumDiff}}
+\comment{Here I would start with only looking at virgin and pregnancy. 
+I would then describe the parity effect in the next section. Maybe there is a more elegant way of disentangling the two.}
 \begin{itemize}
@@
     \item We conclude that there is a luminal hierarchy with NHS progenitors that can give rise to both HS progenitors/differentiated cells and secretory cells
+    \item Genes and especially TFs that are differentially expressed along the branches
 \end{itemize}
@@
 \end{subfigure}
-\caption{Differentiation hierarchy in the mammary gland}
+\caption{Differentiation hierarchy of luminal cells in the mammary gland}
       \label{fig:f2}
@@
 \subsection{Parity exerts cell type specific effects on the transcriptional landscape of luminal cells\label{sec:ParityEffect}}
+\comment{Is it fair to call this parity induced effects? Is this distinction between Virgin/Pregnancy versus Lactation/Post-Involution meaningful?}
 \begin{itemize}
     \item Effect is stronger on luminal than on basal cells (\% variance explained?) [as has been shown before on epigenitic level]
-    \item \fixme{Is it worth mentioning that the lactation sample FACS profile has completely lost its ``basal/luminal'' appearence, yet contains all cell types}
     \item DE genes between e.g. HS differentiated in involution versus virgin
@@
 
-%------------------------------ Figure 4 -------------------------------------------------------
+%------------------------------ Figure 3 -------------------------------------------------------
 \begin{figure}[h]

### 2017-04-05 src/QCAnalysis.Rmd
fixes to accomodate changes in upstream scripts

@@
 #relevel pD Condition to get it in logical sequence
-pD$Condition <- factor(pD$Condition,levels=c("V","P","L","I"))
+# pD$Condition <- factor(pD$Condition,levels=c("V","P","L","I"))
 ```
@@
 ```{r}
-pD.add <- select(pD.filtered, barcode, sf, tSNE1, tSNE2)
-fD.add <- select(fD.filtered, id, highVar)
+pD.add <- dplyr::select(pD.filtered, barcode, sf, tSNE1, tSNE2)
+fD.add <- dplyr::select(fD.filtered, id, highVar)
 pD <- left_join(pD,pD.add, by="barcode")

### 2017-04-06 src/ClusterBootsrap.R
Major formatting changes

@@ -24,10 +24,7 @@ pD$sf <- computeSumFactors(m,clusters=clusters)
 m <- t(t(m)/pD$sf)
 # 
 
-#Run various clustering combinations
-
 #Feature Selection
-library(fpc)
 m.sub <- t(m[fD$highVar,])
 dms <- c("euclidean","pearson","spearman")
 lks <- c("average","ward.D2","complete")

### 2017-04-06 src/DiffusionMapVP.R

@@ -1,250 +1,0 @@
-library(plyr)
-library(pheatmap)
-library(reshape2)
-library(splines)
-library(dplyr)
-library(destiny)
-library(ggplot2)
-source("functions.R")
-library(RColorBrewer)
-library(lmtest)
-
-#REad in data
-rnd_seed <- 300
-dataList <- readRDS("../data/Robjects/ExpressionList_Clustered.rds")
-m <- dataList[[1]]
-pD <- dataList[[2]]
-fD <- dataList[[3]]
-
-#load transcription factors
-tfCheck <- read.table("../data/miscData/TFcheckpoint_WithENSID.tsv",
-		header=TRUE, sep="\t")
-
-tfs <- filter(fD, id %in% tfCheck$ensembl_gene_id) %>% .$symbol
-
-# Pre-Filtering  selecting only virgin and pregnancy
-# Cells
-keepCells <- pD$PassAll & !pD$isImmuneCell & !pD$isOutlier & !(pD$cluster %in% c(6,7,10)) &
-    pD$Condition %in% c("V","P")
-m <- m[,keepCells]
-pD <- pD[keepCells,]
-
-#Gene filtering
-keep <- rowMeans(m)>0.1
-m <- m[keep,]
-fD <- fD[keep,]
-
-#Normalize
-m.norm <- t(t(m)/pD$sf)
-
-# Selection of HVGs
-brennecke <- BrenneckeHVG(m.norm,suppress.plot=TRUE)
-fD$highVar <- fD$id %in% brennecke
-
-#Expression matrix for DiffusionMap
-exps <- m.norm[fD$highVar,]
-exps <- t(log(exps+1))
-
-# Compute Diffusion map
-set.seed(rnd_seed)
-dm <- DiffusionMap(exps,n_eigs=20,k=50)
-plot(eigenvalues(dm)[1:20])
-
-# Pseudotime and branching
-set.seed(rnd_seed)
-dcs <- eigenvectors(dm)[,1:2]
-t1 <- which.min(dcs[,2])
-t2 <- which.min(dcs[,1])
-t3 <- which.max(dcs[,1])
-dpt <- DPT(dm, branching=TRUE, tips=c(t1,t2,t3))
-root <- which(dpt@tips[,1])[3]
-rootdpt <- paste0("DPT",root)
-plot(dpt,pch=20,dcs=c(1,2),col_by=rootdpt)
-
-# Rename branches
-branch <- dpt@branch[,1]
-branch[is.na(branch)]  <- "Decision stage"
[…]

### 2017-04-06 src/Figure2.R

@@ -1,3 +1,8 @@
+#########################################
+#
+#Script to reproduce Figure 2
+#
+#########################################
 library(plyr)
 library(splines)
 library(destiny)
@@ -7,72 +12,58 @@ library(ggplot2)
 library(cowplot)
 library(viridis)
 library(RColorBrewer)
+library(scatterplot3d)
+library(gridGraphics)
+library(gridExtra)
 source("functions.R")
 
+#Load Data
 rnd_seed <- 300
 dataList <- readRDS("../data/Robjects/ExpressionList_Clustered.rds")
 m <- dataList[[1]]
 pD <- dataList[[2]]
 fD <- dataList[[3]]
 
-pD$Condition <- mapvalues(pD$Condition, from=c("V","P","L","I"),
-			  to=c("Virgin", "Pregnancy",
-			       "Lactation", "Post-Involution"))
-
-# Cells
+#Remove QC-fails,outlier and immune cells
 keepCells <- pD$PassAll & !pD$isImmuneCell & !pD$isOutlier 
 m <- m[,keepCells]
 pD <- pD[keepCells,]
 
-pD$cluster <- mapvalues(pD$cluster, from=c(1,2,3,4,5,6,7,9,10),
-			  to=c(1,2,3,4,5,6,7,8,9))
-#Normalize
-m.norm <- t(t(m)/pD$sf)
-rownames(m.norm) <- fD$symbol
 
 
-##########################################
-#
-#
-# Diffusion Maps
-#
-#
-##########################################
-
-
-condComb <- c("Virgin","Pregnancy")
-
-# For Basal and Luminal cells
-#
-#
-
+# ---- BasalAndLuminal ----
+condComb <- c("NP","G")
 keepCells <- pD$Condition %in% condComb
 m.vp <- m[,keepCells]
 pD.vp <- pD[keepCells,]
 
-#Gene filtering
+#Remove lowly expressed genes
 keep <- rowMeans(m.vp)>0.1
 m.vp <- m.vp[keep,]
 fD.vp <- fD[keep,]
 
-#
+#Normalize 
 m.norm <- t(t(m.vp)/pD.vp$sf)
 
+#compute HVGs
 brennecke <- BrenneckeHVG(m.norm,suppress.plot=TRUE)
 fD.vp$highVar <- fD.vp$id %in% brennecke
 
[…]

### 2017-04-06 src/Figure3.R

@@ -1,3 +1,8 @@
+#########################################
+#
+#Script to reproduce Figure 3
+#
+#########################################
 library(plyr)
 library(cowplot)
 library(pheatmap)
@@ -8,97 +13,73 @@ library(destiny)
 library(ggplot2)
 source("functions.R")
 library(RColorBrewer)
+library(viridis)
 library(lmtest)
+library(gridExtra)
 
-#REad in data
 rnd_seed <- 300
 dataList <- readRDS("../data/Robjects/ExpressionList_Clustered.rds")
 m <- dataList[[1]]
 pD <- dataList[[2]]
 fD <- dataList[[3]]
 
-#load transcription factors
-tfCheck <- read.table("../data/miscData/TFcheckpoint_WithENSID.tsv",
-		header=TRUE, sep="\t")
-
-tfs <- filter(fD, id %in% tfCheck$ensembl_gene_id) %>% .$symbol
-
-#Rename clusters
-pD$cluster <- mapvalues(pD$cluster, from=c(1,2,3,4,5,6,7,9,10),
-			  to=c(1,2,3,4,5,6,7,8,9))
-
-# Pre-Filtering  selecting only virgin and pregnancy
+# ---- OnlyLuminal ----
+condComb <- c("NP","G")
 # Cells
-keepCells <- pD$PassAll & !pD$isImmuneCell & !pD$isOutlier & !(pD$cluster %in% c(6,7,9)) &
-    pD$Condition %in% c("V","P")
+keepCells <- pD$PassAll & !pD$isImmuneCell & !pD$isOutlier & !(pD$cluster %in% c(6,7,9)) & pD$Condition %in% condComb
 m <- m[,keepCells]
 pD <- pD[keepCells,]
 
-#Gene filtering
+# Genes 
 keep <- rowMeans(m)>0.1
 m <- m[keep,]
 fD <- fD[keep,]
 
-#Normalize
+# Normalize
 m.norm <- t(t(m)/pD$sf)
 
-# Selection of HVGs
+# HVGs
 brennecke <- BrenneckeHVG(m.norm,suppress.plot=TRUE)
 fD$highVar <- fD$id %in% brennecke
 
-#Expression matrix for DiffusionMap
+# Prepare expression matrix by selecting only HVGs and log-transformation
 exps <- m.norm[fD$highVar,]
 exps <- t(log(exps+1))
 
 # Compute Diffusion map
 set.seed(rnd_seed)
[…]

### 2017-04-06 src/MethylatedGenes.R

@@ -1,68 +1,0 @@
-library(dplyr)
-library(gtools)
-library(knitr)
-library(ggplot2)
-source(file.path("functions.R"))
-
-rnd_seed <- 300
-dataList <- readRDS(file.path("../data/Robjects/ExpressionList_Clustered.rds"))
-deList <- readRDS(file.path("../data/Robjects/DEList.rds"))
-deMethGenes <- read.table("../data/methylationData/HypoMethInParous.txt", header=TRUE, stringsAsFactor=FALSE) %>% .$Gene
-
-
-#Commence
-m <- dataList[[1]]
-pD <- dataList[[2]]
-fD <- dataList[[3]]
-
-# Cells
-keepCells <- pD$PassAll & ! (pD$isImmuneCell | pD$isOutlier)
-m <- m[,keepCells]
-pD <- pD[keepCells,]
-
-# Genes
-keep1 <- rowSums(m!=0)>10
-keep2 <- rowSums(m)>20
-keep <- keep1 & keep2
-m <- m[keep,]
-fD <- fD[keep,]
-
-#Normalization
-m <- t(t(m)/pD$sf)
-
-#FCs
-genes <- c("B4galt1",
-	   "Bax",
-	   "Stat5a",
-	   "Vdr",
-	   "Atp7b",
-	   "Med16",
-	   "Socs2",
-	   "Med13",
-	   "Hif1a",
-	   "Med12l",
-	   "Med13l",
-	   "Eif2ak3",
-	   "Agpat6",
-	   "Csn3",
-	   "Med18",
-	   "Cdo1",
-	   "Xdh",
-	   "Vegfa",
-	   "Stat5b",
-	   "Med1",
-	   "Birc2")
-
-genes <- filter(fD, symbol %in% genes) %>% .$id
-
-deResults <- deList[[4]]
-forPlot <- deResults[,c("Gene","logFC.vs.5","top.vs.5")]
-forPlot <- mutate(forPlot, isHypoMeth= Gene %in% deMethGenes) %>% 
-	   rename(foldChange=logFC.vs.5)
-
-x <- filter(forPlot, foldChange >0, top.vs.5 <= 1000)
-y <- filter(forPlot, foldChange <0, top.vs.5 <= 1000)
-
-p <- ggplot(forPlot, aes(y=foldChange, x=isHypoMeth)) +
-    geom_boxplot() +
-    theme_bw()

### 2017-04-06 src/QCAnalysis.R

@@ -1,4 +1,9 @@
-######### QC-Analysis ########
+####################
+#
+#QC-Analysis
+#
+###################
+
 library(scran)
 library(dplyr)
 library(knitr)
@@ -7,25 +12,20 @@ library(Rtsne)
 library(cowplot)
 source("functions.R")
 
-
-#Load Data
+# Load Data
 dataList <- readRDS("../data/Robjects/ExpressionList.rds")
 m <- dataList[["counts"]]
 pD <- dataList[["phenoData"]]
 fD <- dataList[["featureData"]]
 rnd_seed <- 300
 
-## Number of Cells
+# ---- QCOverview ----
+
+# Number of Cells
 table(pD$SampleID)
 table(pD$Condition)
 
-##################################
-#
-# Overview on QC-stats
-#
-#################################
-
-## Sequencing Depth and Genes detected
+# Sequencing Depth and Genes detected
 pD$UmiSums<- colSums(m)
 pD$GenesDetected <- colSums(m!=0)
 genesDetected <- ggplot(pD, aes(x=SampleID,y=GenesDetected,fill=Condition)) +
@@ -39,14 +39,14 @@ LibrarySize <- ggplot(pD, aes(x=SampleID,y=UmiSums,fill=Condition)) +
     ylab("Total number of molecules") +
     theme_bw()
 
-## Cell Viability
+# Cell Viability
 mMito <- m[fD$Mitochondrial,]
 pD$prcntMito <- colSums(mMito)/colSums(m)
 cellViability <- ggplot(pD, aes(x=prcntMito, y=GenesDetected, color=Condition, shape=Replicate))+
     geom_point() +
     theme_bw()
 
-## Genewise QC
+# Genewise QC
 ntop <- 50 
 mRel <- t(t(m)/colSums(m))
 rownames(mRel)  <- fD$symbol
@@ -77,14 +77,10 @@ plot_grid(genesDetected,LibrarySize,cellViability,
 	  topGenes,topFreq)
 
 
-##################################
-#
-# QC-Thresholding
-#
-#################################
+# ---- QCThresholding ----
 
 
-#Left MAD function for thresholding
+# Left MAD function for thresholding
 leftmad <- function(x) {
     m <- median(x)
     dev <- abs(x-m)
@@ -92,7 +88,7 @@ leftmad <- function(x) {
     return(leftmad)
 }
 
-#Define hard thresholds
+# Define hard thresholds
[…]

### 2017-04-06 src/ClusterComparison.Rmd
Major formatting changes

@@
     out <- data.frame("Cluster"=rownames(x),
-	       "Virgin"=unname(x[,"V"]),
-	       "Pregnant"=unname(x[,"P"]),
+	       "Virgin"=unname(x[,"NP"]),
+	       "Pregnant"=unname(x[,"G"]),
 	       "Lactation"=unname(x[,"L"]),
-	       "Involution"=unname(x[,"I"]))
+	       "Involution"=unname(x[,"PI"]))
 })

### 2017-04-06 src/Clustering.R

@@ -1,4 +1,8 @@
+###########################
+#
 #Clustering
+#
+############################
 library(plyr)
 library(scran)
 library(dplyr)
@@ -10,7 +14,7 @@ library(Rtsne)
 library(pheatmap)
 source("functions.R")
 
-#Reading Data
+# Load Data
 rnd_seed <- 300
 clustDat <- readRDS("../data/Robjects/ClusterComparison.rds") 
 bootDat <- readRDS("../data/Robjects/ClusterBootstrap.rds")
@@ -20,22 +24,25 @@ m.full <- dataList[[1]]
 pD.full <- dataList[[2]]
 fD.full <- dataList[[3]]
 
-#Gene and Cell filtering
+# Gene and cell filtering
 m <- m.full[fD.full$keep,pD.full$PassAll]
 pD <- pD.full[pD.full$PassAll,]
 fD <- fD.full[fD.full$keep,]
 
-#Normalization
+# Normalize
 m <- t(t(m)/pD$sf)
-
-#Clustering on highVar genes
+# Subset to HVGs
 m.sub <- m[fD$highVar,]
 
+# ---- Clustering ----
+
+# Set parameters
 dm <- "euclidean"
 lk <- "average"
 ds <- 1
 minSize <- 15
 
+# Clustering
 clust <- dynamicCluster(t(m.sub), dm=dm,
 				 lk=lk,
 				 ds=ds,
@@ -44,26 +51,23 @@ clust <- dynamicCluster(t(m.sub), dm=dm,
 
 pD$cluster <- clust$cluster
 
-#Rename clusters according to nomenclature in paper
+# Rename clusters according to nomenclature in paper
 pD$cluster <- as.factor(mapvalues(pD$cluster,c(0,1,2,3,4,5,6,7,9,10,8),
 			c(0,1,2,3,4,5,6,7,8,9,10)))
 
-#Clusters per condition and sample
+# Clusters versus condition and sample
 table(pD$cluster,pD$Condition)
 table(pD$cluster,pD$SampleID)
 
-# Removal of Immune Cells from the DataSet
+# ---- RemoveImmuneCellsAndOutlier ----
+
 pD.add <- select(pD, barcode, cluster)
 pD.new <- left_join(pD.full,pD.add, by="barcode")
-stopifnot(identical(colnames(m.full),
-		    as.character(pD.full$barcode)))
[…]

### 2017-04-06 src/DiffusionMap.R

@@ -1,112 +1,0 @@
-library(plyr)
-library(dplyr)
-library(destiny)
-library(ggplot2)
-source("functions.R")
-library(RColorBrewer)
-library(rgl)
-
-rnd_seed <- 300
-dataList <- readRDS("../data/Robjects/ExpressionList_Clustered.rds")
-m <- dataList[[1]]
-pD <- dataList[[2]]
-fD <- dataList[[3]]
-
-#Pre-Filtering before DE-Analysis
-# Cells
-keepCells <- pD$PassAll & !pD$isImmuneCell & !pD$isOutlier & !(pD$cluster %in% c(6,7,10)) 
-m <- m[,keepCells]
-pD <- pD[keepCells,]
-
-#Gene filtering
-keep <- rowMeans(m)>0.1
-m <- m[keep,]
-fD <- fD[keep,]
-
-# Genes
-#Pre-selected set
-# genes <- c("Csn2","Gata3","Prlr","Procr","Elf5","Esr1","Pgr","Aldh1a3","Wap",
-#            "Tspan8","Krt18","Krt8","Krt14","Krt5","Fgfr1","Areg","Fgfr2",
-#            "Notch1","Notch3","Foxc1","Zeb2")
-# fD$highVar <- fD$symbol %in% genes
-
-#Normalize
-m.norm <- t(t(m)/pD$sf)
-
-brennecke <- BrenneckeHVG(m.norm,suppress.plot=TRUE)
-fD$highVar <- fD$id %in% brennecke
-
-exps <- m.norm[fD$highVar,]
-exps <- t(log(exps+1))
-dm <- DiffusionMap(exps,n_eigs=20,k=50)
-dpt <- DPT(dm, branching=TRUE)
-plot(eigenvalues(dm)[1:20])
-
-root <- which(dpt@tips[,1])[3]
-rootdpt <- paste0("DPT",root)
-plot(dpt,pch=20,dcs=c(1,2),col_by=rootdpt)
-plot(dpt,pch=20,dcs=c(1,2),col_by="branch")
-
-# palet <- brewer.pal(length(levels(pD$Condition)),name="Paired")
-# cols <- mapvalues(pD$Condition,levels(pD$Condition),palet)
-# plot3d(eigenvectors(dm)[,c(1,2,3)],col=cols,radius=0.001, type="s")
-
-
-#Expression of Genes in the differentiation trajectories
-branch <- dpt@branch[,1]
-pseudot<- dpt[[rootdpt]]
[…]

### 2017-04-06 src/Figure4.R

@@ -9,6 +9,9 @@ library(dynamicTreeCut)
 library(Rtsne)
 library(reshape2)
 library(pheatmap)
+library(RColorBrewer)
+library(edgeR)
+library(cowplot)
 source("functions.R")
 
 rnd_seed <- 300
@@ -31,27 +34,13 @@ fD <- fD[keep,]
 #Normalize
 m.norm <- t(t(m)/pD$sf)
 
-#HVG
-brennecke <- BrenneckeHVG(m.norm,suppress.plot=TRUE)
-
-#Rename condition and cluster
-pD$Condition <- mapvalues(pD$Condition, from=c("V","P","L","I"),
-			  to=c("Virgin", "Pregnancy",
-			       "Lactation", "Post-Involution"))
-
-pD$cluster <- mapvalues(pD$cluster, from=c(1,2,3,4,5,6,7,9,10),
-			  to=c(1,2,3,4,5,6,7,8,9))
-
-
+pD$cluster <- factor(pD$cluster) #drop unused levels
 out <- data.frame("C1"=numeric(nrow(m)))
-
-for (clust in levels(pD$cluster)[-1]) {
+for (clust in levels(pD$cluster)) {
     expr <- rowMeans(log2(m.norm[,pD$cluster==clust]+1))
     colname <- paste0("C",clust)
     out[,colname] <- expr
 }
-library(pheatmap)
-library(RColorBrewer)
 
 meanSim <- asSim(as.matrix(dist(t(out))))
 simMat <- pheatmap(meanSim,
@@ -60,34 +49,19 @@ simMat <- pheatmap(meanSim,
 		   treeheight_row=0)
 dev.off()
 
-######################
-#
-#DE-Analysis
-#
-######################
 
+# ---- DEProgenitorVsRest ----
+#reload data
 m <- dataList[[1]]
 pD <- dataList[[2]]
 fD <- dataList[[3]]
 
-#Pre-Filtering before DE-Analysis
-# Cells
 keepCells <- pD$PassAll & !(pD$isImmuneCell | pD$isOutlier)
 m <- m[,keepCells]
 pD <- pD[keepCells,]
 
-#Rename condition and cluster
-pD$Condition <- mapvalues(pD$Condition, from=c("V","P","L","I"),
-			  to=c("Virgin", "Pregnancy",
-			       "Lactation", "Post-Involution"))
-
-pD$cluster <- mapvalues(pD$cluster, from=c(1,2,3,4,5,6,7,9,10),
[…]

### 2017-04-06 src/S2.R

@@ -1,40 +1,31 @@
 library(scran)
-library(plyr)
 library(dplyr)
-library(knitr)
 library(ggplot2)
 library(cowplot)
 library(Rtsne)
 source("functions.R")
+
+# Load Data
 dataList <- readRDS("../data/Robjects/ExpressionList_Clustered.rds")
 m <- dataList[[1]]
 pD <- dataList[[2]]
 fD <- dataList[[3]]
-#relevel pD Condition to get it in logical sequence
 rnd_seed <- 300
 
-pD$Condition <- mapvalues(pD$Condition, from=c("V","P","L","I"),
-			  to=c("NP", "G",
-			       "L", "PI"))
-pD$Condition <- factor(pD$Condition, levels=c("NP","G","L","PI"))
-pD$SampleID <- mapvalues(pD$SampleID, from=c("V1","V2","P1","P2","L1","L2","I1","I2"),
-			  to=c("NP1","NP2","G1","G2","L1","L2","PI1","PI2"))
-pD$SampleID <- factor(pD$SampleID,levels=c("NP1","NP2","G1","G2","L1","L2","PI1","PI2")) 
-
-
-# A more or less fair representation of the lactation samples
+# Subset to lactation sample
 pD.lac <- filter(pD, Condition=="L")
 cells <- as.character(pD.lac$barcode)
 m.lac.all <- m[,cells]
 keep <- rowMeans(m.lac.all) > 0.1
 m.lac <- m.lac.all[keep,]
 fD.lac <- fD[keep,]
-#Normalization
+
+# Normalize
 clusters.lac <- quickCluster(m.lac)
 pD.lac$sf <- computeSumFactors(m.lac,clusters=clusters.lac,positive=FALSE)
 m.lac.norm <- t(t(m.lac)/pD.lac$sf)
 
-
+# Compute tSNE
 fPCA.lac <- log2(t(m.lac.norm)+1)
 fPCA.lac <- scale(fPCA.lac,scale=TRUE,center=TRUE)
 set.seed(rnd_seed)
@@ -43,6 +34,7 @@ tsn.lac <- Rtsne(fPCA.lac,perplexity=25)
 pD.lac$tSNE1 <- tsn.lac$Y[,1]
 pD.lac$tSNE2 <- tsn.lac$Y[,2]
 
+# Plot
 pD.lac <- arrange(pD.lac, PassAll) %>%
     rename(QCpass=PassAll) %>%
     mutate(QCpass=ifelse(QCpass,"Pass","Fail"))
[…]

### 2017-04-06 src/S4.R

@@ -1,3 +1,9 @@
+######################
+#
+# Figure S4
+#
+######################
+
 library(scran)
 library(plyr) 
 library(dplyr)
@@ -8,34 +14,30 @@ library(cowplot)
 library(dynamicTreeCut)
 library(Rtsne)
 library(pheatmap)
+library(gridGraphics)
+library(gridExtra)
 source("functions.R")
 
+# Load Data
 rnd_seed <- 300
 clustDat <- readRDS("../data/Robjects/ClusterComparison.rds") 
 bootDat <- readRDS("../data/Robjects/ClusterBootstrap.rds")
 dataList <- readRDS("../data/Robjects/ExpressionList_QC.rds")
 
-#Commence
 m.full <- dataList[[1]]
 pD.full <- dataList[[2]]
 fD.full <- dataList[[3]]
-#Gene and Cell filtering
+
+# Gene and Cell filtering
 m <- m.full[fD.full$keep,pD.full$PassAll]
 pD <- pD.full[pD.full$PassAll,]
 fD <- fD.full[fD.full$keep,]
-#relevel pD Condition to get it in logical sequence
-pD$Condition <- mapvalues(pD$Condition, from=c("V","P","L","I"),
-			  to=c("NP", "G",
-			       "L", "PI"))
-pD$Condition <- factor(pD$Condition, levels=c("NP","G","L","PI"))
-pD$SampleID <- mapvalues(pD$SampleID, from=c("V1","V2","P1","P2","L1","L2","I1","I2"),
-			  to=c("NP1","NP2","G1","G2","L1","L2","PI1","PI2"))
-pD$SampleID <- factor(pD$SampleID,levels=c("NP1","NP2","G1","G2","L1","L2","PI1","PI2")) 
 
-#Normalization
+# Normalization
 m <- t(t(m)/pD$sf)
 m.sub <- m[fD$highVar,]
 
+# ---- Clustering ----
 dm <- "euclidean"
 lk <- "average"
 ds <- 1
@@ -46,16 +48,15 @@ clust <- dynamicCluster(t(m.sub), dm=dm,
 				 output="Cluster",
 				 minSize=minSize)
 
-## Extract as grob for plot
-library(gridGraphics)
-library(gridExtra)
+# Funciton to extract plot as grob
 grab_grob <- function(){
   grid.echo()
[…]

### 2017-06-09 src/PrepareData.R
added file for getMito

@@ -30,7 +30,7 @@ tfCheck <- read.table("../data/miscData/TFcheckpoint_WithENSID.tsv",
 		header=TRUE, sep="\t")
 
 fDat <- mutate(fDat,
-	       Mitochondrial=id %in% mitoGenes$x,
+	       Mitochondrial=id %in% mitoGenes$V1,
 	       TranscriptionFactor=id %in% tfCheck$ensembl_gene_id)
 
 

### 2017-06-09 src/getMito.sh

@@ -1,0 +1,11 @@
+#!/bin/bash
+
+# download gtf file from ensembl
+wget -O ../data/miscData/mm10.gtf.gz "ftp://ftp.ensembl.org/pub/release-89/gtf/mus_musculus/Mus_musculus.GRCm38.89.chr.gtf.gz" 
+gunzip ../data/miscData/mm10.gtf.gz
+
+# extract mitochondrial genes
+grep ^MT ../data/miscData/mm10.gtf | grep -o 'ENSMUSG[0-9]*' | uniq > ../data/miscData/MitoGenes.txt
+
+# delete gtf
+rm ../data/miscData/mm10.gtf 

### 2017-06-09 paper/main.tex
added paper

--- 2017-03-08 paper/main.tex
@@
-\documentclass[oneside]{amsart}
+\documentclass[titlepage, 12pt, oneside]{amsart}
+\usepackage[backend=bibtex,
+	    natbib=true,
+	    autocite=superscript,
+	    style=nature,
+	    sorting=none,
+	    isbn=false]{biblatex} 
+\addbibresource{references.bib}
 \usepackage{geometry}
-\usepackage{times}
+\usepackage{amssymb,amsmath,amsaddr}
+\usepackage{siunitx}
+\usepackage{fixltx2e} % provides \textsubscript
+\usepackage{hyperref}
+\PassOptionsToPackage{usenames,dvipsnames}{color} % color is loaded by hyperref
+\hypersetup{unicode=true,
+            pdfborder={0 0 0},
+            breaklinks=true}
+\urlstyle{same}  % don't use monospace font for urls
+\setlength{\parindent}{0pt}
+\setlength{\parskip}{6pt plus 2pt minus 1pt}
+\setcounter{secnumdepth}{0}
 
-%% --for bibtex
-\bibliographystyle{unsrt}
-%\usepackage{cite}
-\usepackage{natbib}
+%---------------- fake figures ------------------------
+\newcommand{\fakefigure}[1]% #1 = label name
+{\refstepcounter{figure}\label{#1}}
+%------------------------------------------------------
 
-\usepackage{amsmath}
-\usepackage{amssymb}
-\usepackage{amsfonts} %\mathbb
-\usepackage{bm}
-\usepackage{lineno}
+%---------------- fake tables ------------------------
+\newcommand{\faketable}[1]% #1 = label name
+{\refstepcounter{table}\label{#1}}
+%------------------------------------------------------
 
-\usepackage{subcaption}
-\usepackage{graphicx}
-\usepackage{hyperref}
+\begin{document}
 
-\usepackage{siunitx}
+\begin{titlepage}
+\large
+\textbf{Differentiation dynamics of mammary epithelial cells revealed by
+single-cell RNA-sequencing}
+\normalsize
 
-\hypersetup{colorlinks=true,linkcolor=black,citecolor=black,urlcolor=black}
-\urlstyle{same}
+\small
+Karsten Bach\textsuperscript{1,2,5}, Sara Pensa\textsuperscript{1,5},
+Marta Grzelak\textsuperscript{2,5}, James Hadfield\textsuperscript{2,5},
+David J. Adams\textsuperscript{3,5}, John C.
+Marioni\textsuperscript{2,3,4}, Walid T. Khaled\textsuperscript{1,5 }
 
-\newcommand{\fixme}[1]{\textit{\textcolor{red}{Fixme: #1}}}
-\newcommand{\comment}[1]{\textit{\textcolor{blue}{Comment: #1}}}
-%\DeclareSIUnit\Molar{\textsc{m}} % add molar as captial m to siunitx
+1. Department of Pharmacology, University of Cambridge, Cambridge, UK\\
+2. Cancer Research UK Cambridge Institute, University of Cambridge,
+Cambridge, UK\\
+3. Wellcome Trust Sanger Institute, Wellcome Genome Campus, Hinxton,
+Cambridge, UK\\
+4. European Bioinformatics Institute, European Molecular Biology
+Laboratory, Hinxton, UK\\
+5. Cancer Research UK Cambridge Cancer Centre, Cambridge, UK
 
-\title{Single-cell RNA-sequencing reveals cellular dynamics of the developing mammary gland}
-\author{Karsten Bach, Sara Pensa, Walid T. Khaled, John C. Marioni}
-\begin{document}
-\begin{abstract}
-    Bla bla bla
-\end{abstract}
+Correspondence to: Walid T. Khaled
+(\href{mailto:wtk22@cam.ac.uk}{\emph{wtk22@cam.ac.uk}}) or John C.
+Marioni (\href{mailto:marioni@ebi.ac.uk}{\emph{marioni@ebi.ac.uk}})
 
-\maketitle
-\tableofcontents
-\newpage
+\normalsize
+\textbf{Keywords: Single cell RNAseq, Development, Mammary gland }
+\end{titlepage}
+
+\textbf{Abstract:}
+\textbf{The mammary gland is a unique organ as it undergoes most of its development during puberty and adulthood.
+Characterising the hierarchy of the various mammary epithelial cells and how they are regulated in response to gestation, lactation and involution is important for understanding how breast cancer develops.
+Recent studies have used various markers to enrich, isolate and characterise the different epithelial cell compartments within the adult mammary gland.
+However, in all of these studies only a handful of markers were used to isolate, define and trace cell populations.
+Therefore, there is a need for an unbiased and comprehensive description of mammary epithelial cells within the gland at different developmental stages.
+To this end we herein use single cell RNA sequencing (scRNAseq) to determine the gene expression profile of individual mammary epithelial cells across four adult developmental stages; nulliparous, mid gestation, lactation and post weaning (full natural involution).
+Our data from 3,471 individual cells identifies 9 distinct mammary epithelial cell populations and allows their hierarchical structure across development to be charted.
+Interestingly, the effect of gestation and lactation appeared to be more pronounced for some cell types.
+For example, our analysis revealed a cluster of luminal progenitor cells in post involution glands, which is distinct from progenitors found in nulliparous glands.
+The data also showed that only few clusters could be fully characterized by a single marker gene.
+We argue instead that the epithelial cells -- especially in the luminal compartment -- should rather be conceptualized as being part of a continuous spectrum of differentiation.
+This view highlights the plasticity of the tissue and might help to explain some of the conflicting results from lineage tracing studies.}
+
 \section{Introduction}
+The purpose of the mammary gland is to provide nourishment and passive immunity to the young until they are capable of feeding themselves.
+From a developmental biology perspective, the mammary gland is a unique organ as it undergoes most of its development during puberty and adulthood\autocite{Watson2008,Hennighausen2005,Hennighausen1998,Inman2015}
+In the pre-pubertal mouse the mammary gland consists of a rudimentary epithelial ductal structure embedded within a mammary fatpad, which is connected to the nipple\autocite{Mikkola2006,Hens2005}
+At the onset of puberty and in response to hormonal changes, the rudimentary ductal structure will proliferate and migrate to fill the entire mammary fatpad leaving a developed network of ductal structures that later serve as channels for milk transport during lactation.
+At the onset of pregnancy a highly proliferative stage is initiated characterised by further ductal side-branching and widespread lobuloalveolar development\autocite{Watson2008}.
+Differentiation of the epithelial cells within alveoli prepares the gland for milk production and secretion.
+Towards the end of pregnancy the gland is extremely dense and primarily occupied by epithelial cells and very little fat.
+This morphology is largely maintained throughout lactation.
+However, in response to cessation of suckling the gland undergoes involution, which is characterised by extensive cell death and tissue remodelling \autocite{Inman2015,Watson2006} .
+Towards the end of involution the gland reaches a morphology resembling that prior to pregnancy and subsequent pregnancies will trigger the same chain of events.
+
+Recent efforts have focused on the identification and characterisation of the various mammary epithelial cell lineages within the gland that contribute to this developmental homeostasis.
+Pioneering fat-pad transplantation studies nearly 70 years ago were the first to demonstrate the regenerative and differentiation capacity of small numbers of cells\autocite{Faulkin1960,Daniel197,Smalley1998}.
+More recently the use of cell surface markers coupled with flow cytometry has been used to enrich for various progenitor and stem cell compartments\autocite{Smalley1998,Stingl2006,Shackleton2006,Asselin2007} and shown that the imbalance of such cell populations results in cellular transformation and subsequently breast cancer\autocite{Lim2009,Molyneux2010}.
+Other studies, inspired by breast cancer transcriptomic profiling, have identified transcriptional regulators of mammary epithelial cell types such as \textit{GATA3} in the luminal cells\autocite{Asselin2007,Kouros2006}.
+More recently, elegant lineage-tracing studies used key markers to address the contribution of each lineage to adult mammary epithelial cell homeostasis\autocite{Inman2015}.
+However, in all of these studies only a handful of markers and genes were used to define the cellular hierarchy of the mammary epithelial cells, with a principal focus on the nulliparous developmental stage.
+Therefore, to properly understand its changing role throughout life, there is a need for an unbiased and comprehensive characterisation of mammary epithelial cell compartments at different developmental stages.
+
+To this end we herein use single cell RNA sequencing (scRNAseq) to map the cellular dynamics of mammary epithelial cells across four adult developmental stages; nulliparous, mid gestation, lactation and post weaning (full natural involution).
+Our data from 3,471 individual cells, identifies 9 distinct cell populations within the gland and allows their hierarchical structure across developmental time points to be charted.
 
 \section{Results}
-%------------------------------ Section 1 -------------------------------------------------------
-\subsection{Unbiased identification of mammary epithelial cells\label{sec:cluster}}
-\begin{itemize}
-    \item Experimental Setup, 4 stages of the development of an adult gland
-    \item Clustering into major cell types in the mammary gland
-    \item Luminal cells: Hormone sensing (HS) (progenitor and differentiated), Non-hormone sensing (NHS) progenitor and secretory cells
-    \item Basal cells: MaSCs, PROCR\textsuperscript{+} MaSCs and differentiated myoepithelial cells
-\end{itemize}
 
-%------------------------------ Figure 1 -------------------------------------------------------
-\begin{figure}[h]
-\begin{subfigure}{0.45\linewidth}
-\centering %\textbf{a}
-\fbox{\parbox{0.9\linewidth}{(Experimental overview)}}
-%\includegraphics[height=5cm]{../figures/ROC_Staurosporine.pdf}
-\caption{}
-\end{subfigure}
-\begin{subfigure}{0.45\linewidth}
-%\centering %\textbf{a}
-    \fbox{\parbox{0.9\linewidth}{(tSNE plot colored by Condition)}}
-%\includegraphics[height=5cm]{../figures/ROC_Panobinostat.pdf}
-\caption{}
-\end{subfigure}
-\begin{subfigure}{0.45\linewidth}
-\centering %\textbf{a}
-\fbox{\parbox{1\linewidth}{(Plot to illustrate clusters and marker genes)}}
-%\includegraphics[height=5cm]{../figures/ROC_ATP.pdf}
-\caption{}
-\end{subfigure}
-    \caption{Unbiased identification of mammary epithelial cells}
-      \label{fig:f1}
-  \end{figure}
+\textbf{Single cell RNA sequencing identifies 9 clusters of mammary epithelial cells}
 
-\newpage
-%------------------------------ Section 2 -------------------------------------------------------
-\subsection{The luminal compartment is defined by a continuous spectrum of differentiation\label{sec:LumDiff}}
-\comment{Here I would start with only looking at virgin and pregnancy arguing that here we would expect all celltypes. 
-I would then describe the parity effect in the next section. Maybe there is a more elegant way of disentagling the two.}
-\begin{itemize}
-    \item Basal cells are separate from the luminal compartment, suggesting either:
-	\begin{itemize}
-	    \item This is a rare event 
-	    \item Intermediate stages are highly unstable and the transdifferentation is rapid
-	\end{itemize}
-    \item In the virgin gland there is one ``branch'' of differentiation with NHS progenitors on one side and differentiated HS cells on the other end
-    \item In pregnancy another branch derives from the NHS progenitors with differentiated secretory cells at the tip
-    \item We conclude that there is a luminal hierarchy with NHS progenitors that can give rise to both HS progenitors/differentiated cells and secretory cells
-\end{itemize}
+First, we isolated mammary epithelial cells from four developmental time points; nulliparous (NP), day 14.5 gestation (G), day 6 lactation (L) and two weeks post natural involution (PI).
+For each time point we sorted mammary epithelial cells based on the Epcam cell surface marker from two independent mice (\autoref{F1}a).
+All samples were then prepared for single-cell RNA-sequencing using the 10x Chromium platform\autocite{Zheng2017}.
+Following quality control (materials and methods), this yielded an average of 11,175 unique transcripts and 3,113 genes detected from 3,471 cells (1,681 from NP, 609 from G, 604 from L and 605 from PI) (Supplementary \autoref{F1}a).
 
-%------------------------------ Figure 2 -------------------------------------------------------
-\begin{figure}[h]
-\begin{subfigure}{0.45\linewidth}
-\centering %\textbf{a}
-\fbox{\parbox{0.9\linewidth}{(3D scatterplot of DiffusionMap V/P)}}
-%\includegraphics[height=5cm]{../figures/ROC_Staurosporine.pdf}
-\caption{}
-\end{subfigure}
-\begin{subfigure}{0.45\linewidth}
-%\centering %\textbf{a}
-    \fbox{\parbox{0.9\linewidth}{(Only luminal cells in virgin and pregnancy)}}
-%\includegraphics[height=5cm]{../figures/ROC_Panobinostat.pdf}
-\caption{}
-\end{subfigure}
-\begin{subfigure}{0.45\linewidth}
-\centering %\textbf{a}
-\fbox{\parbox{1\linewidth}{(Multiple plots to show gene expression pattern along the two branches)}}
-%\includegraphics[height=5cm]{../figures/ROC_ATP.pdf}
-\caption{}
-\end{subfigure}
-\caption{Differentiation hierarchy in the mammary gland}
-      \label{fig:f2}
-  \end{figure}
+Visual inspection of the data using t-distributed stochastic neighbour embedding (t-SNE) suggested that although there is some grouping of cells by time-point, there are also other factors that underlie structure within the dataset (\autoref{F1}b and Supplementary Fig 2a).
+We used unsupervised hierarchical clustering based on the expression of 594 highly variable genes to identify 9 clusters of epithelial cells (C1-C9).
+Importantly, these clusters were consistent with the groups observed in the t-SNE (\autoref{F1}c).
+Based on the expression of \textit{Krt18,} \textit{Krt8, Krt5, Krt14 and Acta2} we grouped the clusters into 6 luminal (C1, C2, C3, C4, C5 \& C8) and 3 basal clusters (C6, C7 \& C9) (\autoref{F1}D, Supplementary Fig 2b and 2c).
+Differential gene expression analysis allowed us to further characterise the clusters and infer putative identities (\autoref{F1}d, \autoref{T1} and Supplementary Table 1).
+For example, C2 expresses high levels of beta-casein, \textit{Csn2,} and is composed exclusively from cells from the gestation and lactation time-points, suggesting that these cells are secretory alveolar cells (\autoref{F1}d).
+C1 and C3 on the other hand represent the clusters with the highest expression of hormone receptors.
+C5 expresses high levels of the luminal progenitor marker \textit{Aldh1a3} suggesting progenitor function, while C7 shows characteristics of myoepithelial cells such as high levels of \textit{Acta2}, \textit{Oxtr} and \textit{Krt15} (\autoref{F1}d).
 
-\newpage
-%------------------------------ Section 3 -------------------------------------------------------
-\subsection{Parity exerts cell type specific effects on the transcriptional landscape of luminal cells\label{sec:ParityEffect}}
-\begin{itemize}
-    \item Effect is stronger on luminal than on basal cells (\% variance explained?) [as has been shown before on epigenitic level]
-    \item \fixme{Is it worth mentioning that the lactation sample FACS profile has completely lost its ``basal/luminal'' appearence, yet contains all cell types}
-    \item DE genes between e.g. HS differentiated in involution versus virgin
-    \item A ``new'' cell type arises after pregnancy that most likely derives from NHS progenitors 
-    \item Highlight importance of this cell type as it sits at the top of the hierarchy
-\end{itemize}
+\textbf{Reconstruction of the luminal differentiation hierarchy}
 
-%------------------------------ Figure 4 -------------------------------------------------------
-\begin{figure}[h]
-\begin{subfigure}{0.45\linewidth}
-\centering %\textbf{a}
-\fbox{\parbox{0.9\linewidth}{(Condition factor explains more variance in the luminal compartment)}}
-%\includegraphics[height=5cm]{../figures/ROC_Staurosporine.pdf}
-\caption{}
-\end{subfigure}
-\begin{subfigure}{0.45\linewidth}
-%\centering %\textbf{a}
-    \fbox{\parbox{0.9\linewidth}{(Cluster 4 is closest to 5 and is at the top of the hierarchy after parity)}}
-%\includegraphics[height=5cm]{../figures/ROC_Panobinostat.pdf}
-\caption{}
-\end{subfigure}
-\begin{subfigure}{0.45\linewidth}
-\centering %\textbf{a}
-\fbox{\parbox{1\linewidth}{(Gene expression differences between 4 and 5)}}
-%\includegraphics[height=5cm]{../figures/ROC_ATP.pdf}
-\caption{}
-\end{subfigure}
-\caption{Parity exerts cell type specific effects on the transcriptional landscape of luminal cells}
-      \label{fig:f3}
-  \end{figure}
+We focused on cells from the NP and G time-points to investigate mammary epithelial differentiation states of the gland.
+These differentiation states and the transitions between them can be computationally reconstructed using diffusion maps.
+Briefly, the method embeds the data in a low-dimensional space, where distances between cells represent the progression through a gradual but stochastic process such as differentiation.
+In diffusion maps constructed from all epithelial cells, we observed a clear segregation between the luminal and basal clusters, with virtually no transition states between the two (\autoref{F2}a).
+This supports the hypothesis that during normal tissue homoeostasis the two lineages are largely self-maintained, which is in agreement with the majority of the lineage tracing studies\autocite{VanKeymeulen2011,VanAmerongen2012,Davis2016}.
+In contrast, the luminal compartment showed a distinct structure, with gradual transitions between different clusters and cells originating from a common origin (\autoref{F2}b).
+We confirmed the robustness of this bifurcation by verifying that it was present when different methods of feature selection, algorithms for trajectory inference and down-sampling were employed (Supplementary Fig.
+3).
+
+The expression of the progenitor marker \textit{Aldh1a3} gradually decreased as cells progressed away from their common origin (\autoref{F2}b right panel), which was largely composed of C5.
+We further noted that the left arm of the differentiation trajectory terminates at C2 and shows increasing expression of \textit{Csn2} (\autoref{F2}b bottom panel), consistent with a secretory phenotype.
+Between C5 and C2 we found cells from C8 (\autoref{F2}b).
+C8 is the second rarest cell population after C9 and is found in all developmental stages in approximately equal numbers (6 cells in NP, 5 cells in G, 8 cells in L and 7 cells in PI).
+C8 cells express \textit{Aldh3a1, Elf5, Cd14 and Kit} similar to C5 but in addition express high levels of prolactin induced protein \textit{(Pip)} and apolipoprotein D \textit{(Apod)} (\autoref{F1}d).
+On the right arm of the differentiation trajectory, cells in C5 transitioned to C3 and C1, during which the expression of \textit{Esr1} and \textit{Pgr} increase suggesting that this branch represents differentiation towards hormone-sensing luminal cells (\autoref{F2}b, bottom panel).
+
+Being confident that the diffusion map recapitulates the luminal differentiation process, we then computationally inferred the two branches and ordered the cells according to their progression through ``pseudotime''\autocite{Haghverdi2016} (see materials and methods for further explanation) (\autoref{F3}a).
+We found 806 genes that showed a pseudotime-dependent expression with the same directionality along both differentiation trajectories (\autoref{F3}b, top panel of heatmap, Supplementary Table 2, Supplementary Fig 4a).
+These included genes associated with known progenitor characteristics such as \textit{Aldh1a3}, \textit{Kit}, \textit{Cd14} and \textit{Tspan8} as well as transcription factors that have not previously been associated with luminal differentiation such as \textit{Creb5}, \textit{Hey1} and \textit{Fosl1} (\autoref{F3}c).
+In addition, we identified 1478 genes with branch-specific expression patterns (\autoref{F3}b, bottom panel of heatmap, Supplementary Table 3, Supplementary Fig 4b), including milk related genes (\textit{Wap}, \textit{Csn1s1}, \textit{Csn3}) as well as genes involved in hormone-receptor signalling (\textit{Pgr, Prlr, S100a6, Cited1, Areg}).
+Amongst the branch-specific transcription factors we found for example \textit{Runx1, Tox2} and \textit{Bhlhe41} (also known as \textit{Sharp1}) to be transcribed during differentiation towards the hormone-sensing lineage (\autoref{F3}d) and \textit{Elf5, Foxs1} and \textit{Ehf} in the secretory lineage (\autoref{F3}e).
+Interestingly, \textit{Runx1} is a known repressor of \textit{Elf5} and its deletion has been shown to be deleterious for ductal morphogenesis\autocite{VanBragt2014}.
+The expression of the known progenitor master regulator \textit{Elf5} is maintained and further increased during secretory differentiation suggesting that its transcriptional level is fine-tuned in luminal progenitors (\autoref{F3}e).
+
+\textbf{Cluster 4 represents a post involution specific progenitor cell}
+
+From the initial clustering analysis we noticed that C4 appeared only after the animals had undergone a full pregnancy and was predominantly made up of cells from PI (0 cells from NP, 0 cells from G, 83 cells from L and 339 cells from PI) (\autoref{F1}b).
+Moreover, multiple progenitor marker genes from C5 were also to be present in C4 (\autoref{F1}c, Supplementary Fig 5).
+The similarity between the two clusters also held true on transcriptome-wide level (\autoref{F4}a, Supplementary Fig 10c).
+We thus hypothesised that C4 represents the luminal progenitor population in the post-parous gland, which is supported by the observation that C5 virtually disappears after parity (Table 1).
+To further support this hypothesis, we identified genes that distinguish C5 from the rest of the luminal compartment to see if they are also characteristic for the proposed post-parity progenitor population C4.
+Indeed, we find genes that are differentially expressed between C5 and the rest of the luminal compartment to show the same trend between C4 and the luminal compartment (\autoref{F4}b).
+In a similar manner we can distinguish C4 from the rest of the PI gland in a principal component analysis (PCA) using the identified progenitor genes (Supplementary Fig 6a and b).
+Given this high degree of similarity we were not surprised that only 120 genes were differentially expressed between C4 and C5 (Supplementary Table 4).
+Interestingly, genes that were up-regulated in C4 were significantly enriched for pathways that are involved in the immune response and lactation (\autoref{F4}c, Supplementary Fig 6c).
+Furthermore, the expression of milk genes was surprisingly maintained at high levels in C4 even after lactation (\autoref{F4}d).
+Of note, the genes of the casein locus (\textit{Csn2}, \textit{Csn1s1}, \textit{Csn1s2a}, \textit{Csn3}) have previously been reported to be up-regulated in the parous gland, most likely due to changes in chromatin accessibility\autocite{Dos2015,Rijnkels2013}.
+However, it has not been shown before that this effect is confined to the progenitor population of the luminal compartment (Supplementary Fig 7).
+Together, the data suggest that luminal progenitor cells maintain memory of having undergone gestation and involution.
+This memory could potentially prime progenitor cells towards the alveolar fate to facilitate alveologenesis in subsequent pregnancies.
 
 \section{Discussion}
-\section{Methods}
 
-\clearpage
+We have reported here the use of single cell RNA sequencing to comprehensively map the transcriptomes of thousands of mammary epithelial cells across four developmental time-points.
+Our analysis identified 9 clusters of epithelial cells, some of which are only present during specific developmental stage (e.g. C2 and C4).
+This study provides a rich dataset that can be mined online (see link in materials and methods) to identify marker genes and lineage specific promoters that can be used to trace populations of cells \textit{in vivo}.
+We note, however, that only some of the clusters can be fully characterized by a single marker gene.
+Instead we argue that the epithelial cells - especially in the luminal compartment - should rather be conceptualized as being part of a continuous spectrum of differentiation as visualised in \autoref{F2}.
+This view highlights the plasticity of the tissue and might help to explain some of the conflicting results from lineage tracing studies\autocite{Inman2015}.
+In this study we could not provide any evidence for contribution of a putative mutlipotent stem cell to the day-to-day homoeostasis of the gland.
+We do note, however, that the cluster C9 is placed between the luminal and basal cells in \autoref{F2}a.
+C9 represents the cluster with the fewest numbers of cells captured and expresses low levels of both luminal and basal markers, high levels of the stem cell marker \textit{Procr}\autocite{Wang2015} and high levels of the luminal progenitor marker \textit{Notch3}\autocite{Lafkas2013} (\autoref{F1}d).
+Yet, we did not capture cells in transition between C9 and any other clusters suggesting that if these cells are indeed, multi-potent progenitors their involvement in tissue homoeostasis is negligible.
+Alternatively, we cannot exclude the possibility that C9 is enriched for doublets.
+Based on the gene expression data presented here, the luminal compartment appears to have one common progenitor population (C5).
+We also identified a previously described progenitor population with expression of hormone-receptor related genes (C3) that appears to represent a meta-stable state on the differentiation path from luminal progenitors towards fully differentiated hormone-sensing cells (C1).
+Furthermore, we characterised gene expression patterns along the differentiation hierarchy, thus enabling to disentangle the cellular hierarchy in the mammary gland.
+By analysing the mammary gland at various stages of development we were also able to describe the molecular effects of parity at cellular resolution.
+We found that the luminal progenitor compartment undergoes lasting changes at the transcriptional level.
+This is especially interesting in the light of the protective effect of pregnancies against breast cancer and the role of luminal progenitors as cell of origin.
+The majority of the changes were related to pathways involved in immunity and lactation, suggesting that, in particular, the luminal progenitors maintain a memory of gestation and involution.
+It is reasonable to assume that C4 overlaps with the previously parity-induced mammary epithelial cells (PI-MECs)\autocite{Wagner2002}.
 
-%% --for biblatex
-\bibliographystyle{unsrt}
-\bibliography{nonparatpp}
+In summary, the study provides a novel view of mammary gland development.
+This unbiased approach helps support some previously formed hypotheses in the mammary gland field and describes differentiation processes at a high cellular resolution.
+The dataset will be a useful resource for future studies that aim to understand the relationship of the different cell types in the gland and how breast cancer develops and progresses.
 
-%% --for biblatex
-%\printbibliography
+\section{Figure legends}
 
+%---------------------------------------------Figure 1--------------------------------------------------------------
+\textbf{Figure 1. Single cell RNA sequencing identifies 9 clusters of
+mammary epithelial cells }
+\fakefigure{F1}
+\textbf{(a)} Schematic diagram highlighting the experimental setup for isolating and sequencing the RNA of single cells using the 10X chromium system.
+\textbf{(b)} t-SNE plot of 3,471 cells visualizes general structure in the data.
+Cells are coloured by the four developmental time points Pink=NP, Dark Green=G, Light Green=L, Purple=PI.
+\textbf{(c)} Same as \textbf{(b)} but colouring cells by clusters.
+\textbf{(d)} Heatmap highlighting some key marker genes.
+Colour scale represents log-transformed and normalized UMI counts scaled to a maximum of 1 per row.
+The upper panel shows genes that were used to distinguish between luminal and basal cells.
+Upper bars represent the cluster assignment and stages for the individual cells.
+For visualization purposes only 100 randomly selected cells were shown for each of the large clusters.
+The list of differentially expressed genes between clusters can be found in Supplementary Table 1.
+%-------------------------------------------------------------------------------------------------------------------
+
+%---------------------------------------------Figure 2--------------------------------------------------------------
+\textbf{Figure 2. Computational reconstruction of differentiation
+processes in the mammary gland}
+\fakefigure{F2}
+\textbf{(a)} Diffusion map of epithelial cells from the NP and G time points, showing the first three diffusion components.
+\textbf{(b)} Differentiation trajectory of the luminal compartment based on the first two diffusion components.
+The right panel depicts the log-transformed normalized expression levels of the luminal progenitor marker \textit{Aldh1a3} in all cells along component 2, the bottom panel shows the levels of \textit{Csn2} and \textit{Pgr} along component 1.
+Expression trends represent natural cubic spline fits with five degrees of freedom.
+%-------------------------------------------------------------------------------------------------------------------
+
+%---------------------------------------------Figure 3--------------------------------------------------------------
+\textbf{Figure 3. Pseudotime ordering identifies genes associated with
+luminal differentiation}
+\fakefigure{F3}
+\textbf{(a)} Definition of the hormone-sensing and secretory differentiation branch.
+Cells are coloured by their progression through pseudotime, where low values represent undifferentiated cells.
+\textbf{(b)} Heatmap of the top 50 pseudotime-dependent genes (rows) with the same overall trend along both branches (top panel) or branch-specific trends (bottom panel, see materials and methods for definition).
+The black vertical line represents the common origin from the progenitor population.
+Arrows highlight differentiation direction.
+Pseudotime and the cluster assignment are annotated above the heatmap.
+The values in the heatmap represent z-scaled, spline-smoothed expression values.
+\textbf{(c-e)} Examples of transcription factors with pseudotime-dependent expression with the same overall trend on both branches \textbf{(c)} or branch specific trends \textbf{(d,e)}.
+%-------------------------------------------------------------------------------------------------------------------
+
+%---------------------------------------------Figure 4--------------------------------------------------------------
+\textbf{Figure 4. The effect of parity on the transcriptomic landscape
+of the luminal progenitor compartment}
+\fakefigure{F4}
+\textbf{(a)} Colour-coded similarity matrix between the log-transformed mean expression values of the 9 epithelial cell clusters.
+Similarities are based on Euclidean distances.
+\textbf{(b)} Comparison of fold changes from C5 versus luminal compartment and fold changes from C4 versus the luminal cells.
+The genes represent the top 500 differentially expressed genes between C5 and luminal cells.
+\textbf{(c)} Volcano plot illustrates differential expression between C5 and C4, coloured dots represent significant genes with known function in lactation and immunity, dashed lines highlight the P value threshold of 0.01 and a log fold change of 1.
+P values are adjusted for multiple testing using Benjamini-Hochberg.
+\textbf{(d)} Visualization of expression difference for some lactation related genes from \textbf{(c)}, where C4 is split according to the developmental time point.
+%-------------------------------------------------------------------------------------------------------------------
+
+\textbf{Table 1. Summary of mammary epithelial cell clusters}
+\fakefigure{T1}
+Overview of the different clusters including number of cells captured for each time-point and key genes that were used to infer their identities.
+
+\textbf{Supplementary Figure 1. Quality control of sequencing data}
+
+\textbf{(a)} Table summarizing quality control criteria per sample.
+Number of unique molecules, genes detected and number of reads represent the median value for each sample.
+\textbf{(b-c)} Histograms for the four conditions showing the distributions of the number of genes detected \textbf{(b)} or total number of molecules \textbf{(c)} metric with indicated threshold (dashed line).
+\textbf{(d)} Scatterplot of number of genes detected versus percentage of mitochondrial RNA molecules, the threshold at 5\% is indicated.
+For choice of thresholds see materials and methods.
+
+\textbf{Supplementary Figure 2. Biological replicates explain little
+structure in the data }
+
+\textbf{(a)} t-SNE plot coloured by the eight different samples.
+Some minor structures (e.g.
+two groups PI1 and PI2 at the top) are mainly replicate dependent, but all of the clusters identified in the analysis are represented by multiple replicates.
+\textbf{(b-c)} t-SNE plots highlighting the expression of luminal \textbf{(b)} and basal \textbf{(c)} genes in all 3,471 cells.
+
+\textbf{Supplementary Figure 3. Testing the robustness of the diffusion
+map}
+
+\textbf{(a-b)} Eigenvalues for the diffusion map used in \autoref{F2}a
+\textbf{(a)} and \autoref{F2}b \textbf{(b)}. 
+\textbf{(c)} The differentiation trajectory as determined by Monocle (see materials and methods) coloured by cluster assignment.
+\textbf{(d)} The diffusion map is robust towards down-sampling of cells (100, 50 or 25\% of all cells were used in the left, middle or right panel, respectively) as well as the method of feature selection (all= all genes with mean expression level above 0.1, HVG= highly variable genes, PCA= first 50 components from PCA, selected= a manually selected choice of genes that are known to be involved in luminal cell differentiation).
+The gene list included: \textit{Csn2, Gata3, Prlr, Elf5, Esr1, Pgr, Aldh1a3, Wap, Tspan8, Krt18, Krt8, Areg, Fgfr1, Fgfr2, Notch1, Notch3, Foxc1} and \textit{Zeb2}.
+
+\textbf{Supplementary Figure 4. Full heatmaps of pseudo-time dependent
+expression}
+
+Heatmaps as in \autoref{F3}, showing all genes that are pseudo-time dependent in their expression along the trajectories with the same overall trend \textbf{(a)} or opposing trend \textbf{(b)}.
+
+\textbf{Supplementary Figure 5. Expression of progenitors markers in C4}
+
+Panel of t-SNE plots (compare to \autoref{F1}) with coloured normalized log-expression values of progenitor markers.
+
+\textbf{Supplementary Figure 6. C4 is a post-parity progenitor
+population }
+
+\textbf{(a)} Principal component analysis (PCA) on all luminal cells from the PI time-point.
+The PCA was computed on the top 500 differentially expressed genes between C5 and the rest of the luminal NP gland.
+PC1 separates the progenitors and the differentiated cells, with the progenitors showing negative PC1 values.
+\textbf{(b)} Genes that are higher expressed in C5 compared to the rest of the NP gland also have negative PC1 loadings.
+\textbf{(c)} Top 20 GO-terms (biological processes) that are significantly enriched in upregulated genes from \autoref{F4}c.
+Dashed line indicates p values threshold at 0.001.
+
+\textbf{Supplementary Figure 7. Overexpression of milk genes after
+parity is restricted to the progenitor compartment}
+
+Panel of t-SNE plots (compare to \autoref{F1}) with coloured normalized log-expression values of milk genes from \autoref{F4}d.
+
+\textbf{Supplementary Figure 8. Quality control of cells from the}
+\textbf{Lactation sample}
+
+t-SNE plot computed only on the lactation sample.
+Cells that were flagged as low quality are highlighted in grey and cells that passed the QC are shown in black.
+Although a majority of the sample was removed, the cells that passed represent at least all major groups from this sample (see materials and methods).
+
+\textbf{Supplementary Figure 9. Outline of the clustering strategy }
+
+\textbf{(a)} Dendrogram of the hierarchical clustering using Euclidean distances and average linkage.
+Colour bars indicate the animal (binary, only comparable within one condition), the condition and cluster assignment.
+\textbf{(b)} Cluster stability as determined by the overlap between the original cluster and the most similar cluster in the bootstrap samples.
+The distributions of jaccard indeces are shown as a boxplot for 100 bootstrap samples.
+Stability is shown for three values of the deepSplit parameter (0,1,2).
+Deep split 1 was chosen as this maintained overall a high cluster stability while identifying relevant cell types (see materials and methods).
+
+\textbf{Supplementary Figure 10. Clustering} \textbf{Quality Control and
+removal of C10}
+
+\textbf{(a)} Total molecule count grouped by different clusters.
+\textbf{(b)} Markers that were used to identify C10 as lymphocyte cluster\autocite{Scheele2017} \textbf{(c)} Separation of C10 from all other clusters as shown by a PCA on the cluster average gene expression.
+
+\textbf{Table S1.} Differentially expressed genes between clusters
+
+\textbf{Table S2.} List of genes with pseudo-time dependent gene
+expression with same overall trend in the two branches
+
+\textbf{Table S3.} List of genes with pseudo-time dependent gene
+expression with different trends in the two branches
+
+\textbf{Table S4.} Genes differentially expressed between C4 and C5
+
+\section{Materials and methods}
+
+\textbf{Animals}
+
+All experimental animal work was performed in accordance to the Animals (Scientific Procedures) Act 1986, UK and approved by the Ethics Committee at the Sanger Institute.
+C57BL/6N mice were housed in individually ventilated cages under a 12:12~h light-dark cycle, with water and food available \textit{ad libitum}.
+The experiment was set up to allow for all of the developmental time points to be collected and tissues to be processed at the same time.
+Mice were euthanized by terminal anaesthesia.
+Females were mated with studs and allowed to litter.
+Tissues were then harvested at gestation day 14.5 (G), lactation day 6 (L) and day 11 post natural weaning of the pups (PI).
+Tissue from NP females was harvested at 8 weeks of age.
+Two individual mice per developmental time point were used in the study.
+
+\textbf{Mammary gland dissociation into single-cell suspension}
+
+Lymph node divested mammary glands (excluding the cervical pair and the proximal region of one of the fourth glands) were dissected from the mice and mechanically dissociated.
+The finely minced tissue was transferred to a digestion mix consisting of DMEM/F12 (Gibco) + 10 mM HEPES (Gibco) + 1 mg ml\textsuperscript{-1} collagenase (Roche) + 100 U ml\textsuperscript{-1} hyaluronidase (Sigma) + 50 mycg ml\textsuperscript{-1} gentamicin (Gibco) for 2.5 hours at \SI{37}{\celsius} and vortexed every 30 minutes.
+After the lysis of red blood cells in NH\textsubscript{4}Cl, cells were briefly digested with warm 0.05\% Trypsin-EDTA (Gibco), 5 mg ml\textsuperscript{-1} dispase (Sigma) and 1 mg ml\textsuperscript{-1} DNase (Sigma), and filtered through a 40 mycm cell strainer (BD Biosciences).
+
+\textbf{Cell labelling, flow cytometry and sorting}
+
+Single cell suspensions were incubated in HF medium (Hank's balanced salt solution (Gibco) + 1\% foetal bovine serum, Gibco) + 10\% normal rat serum (Sigma) for 20 min on ice to pre-block before antibody staining.
+All antibody incubations were performed for 10 min on ice in HF media.
+Mammary cells were stained with the following primary antibodies: 1 mycg ml\textsuperscript{-1} CD31-biotin (eBioscience); 1 mycg ml\textsuperscript{-1} CD45-biotin (eBioscience); 1 mycg ml\textsuperscript{-1} Ter119-biotin (eBioscience) and 0.5 mycg ml\textsuperscript{-1} EpCAM--PE (Biolegend).
+Cells were then stained with 0.4 mycg ml\textsuperscript{-1} streptavidin-PE-CF594 (BD-Biosciences).
+Propidium Iodide (PI, 1 mycg ml\textsuperscript{-1}; Sigma) was used to detect dead cells.
+Cells were filtered through a 30 mycm cell strainer (Partec) before sorting.
+Sorting of cells was done using a SH800 sorter (SONY).
+Single-stained control cells were used to perform compensation manually and unstained cells were used to set gates.
+Chip alignment and sorting calibration was performed with automatic setup beads (SONY) immediately prior to sorting.
+Doublets, dead cells and contaminating haematopoietic, endothelial and stromal cells were gated out and EpCAM positive cells were sorted in FACS tubes (BD Biosciences) containing 1 mL of HF.
+After sorting, cells were transferred to a 1.5 ml tube through a 30 mycm cell strainer, spun down and resuspended in 15-50 mycl of HF according to the number of cells sorted per sample.
+Representative samples were manually counted using an improved Neubauer chamber to estimate sample loss from centrifugation and filtering.
+Equal numbers of cells per sample were processed for scRNA library preparation.
+Samples were processed for scRNA library preparation within 10 hours from tissue isolation.
+
+\textbf{Library preparation and sequencing}
+
+Library preparation was performed according to instruction in the 10X chromium single cell kit.
+
+\textbf{RNA-seq data processing }
+
+Read processing was performed as previously reported\autocite{Zheng2017}.
+Briefly, the Cell Ranger Single-Cell Software Suite was used for demultiplexing, barcode assignment and UMI quantification (\href{http://software.10xgenomics.com/single-cell/overview/welcome}{\textit{http://software.10xgenomics.com/single-cell/overview/welcome}}).
+The reads were aligned to the mm10 reference genome using a pre-built annotation package obtained from the 10X Genomics website.
+All lanes per sample were processed using the ``cellranger count'' function.
+The output from different lanes were then aggregated using ``cellranger aggr'' with --normalize set to ``none''.
+
+\textbf{Quality control and preprocessing}
+
+In total the Cell Ranger software identified 5,598 barcodes that contained enough unique molecules to be considered as cells (1,707 in NP, 623 in G, 2,648 in L and 620 in PI).
+Libraries prepared from the NP, G and P time points all showed high quality that was reproducible between the two biological replicates (Supplementary 1a).
+We used the following metrics to flag poor quality cells: number of genes detected, total number of unique molecules (UMIs) and percentage of molecules mapped to mitochondrial genes.
+For these samples we then identified poor quality cells by setting a threshold on the number of genes and number of UMIs that was defined as four median absolute deviations (MAD) below the median for each sample (Supplementary 1b and c).
+The samples from the lactation sample, however, showed on average nearly one order of magnitude fewer genes detected and total number of UMIs.
+As this was specific to both lactation samples, we inferred that this was driven by the biological phenotype of the sample rather than technical artefacts.
+Nonetheless, we flagged a substantial amount of the lactation sample as poor quality cells by setting a fixed threshold of 500 detected genes and 1000 total molecules detected (Supplementary Fig 1b and c).
+This was necessary in order to allow for normalisation of all samples; otherwise, keeping more cells of lower quality would inflate the range of normalisation factors amplifying technical noise due to cell-specific biases and adversely impact downstream analyses.
+We ensured that we still sampled cells from all major structures present in the lactation sample (Supplementary Fig 8).
+Finally, all cells with 5\% or more of UMIs mapping to mitochondrial genes were defined as non-viable or apoptotic and removed form the analysis (Supplementary Fig 1d).
+This left us with a total number of 3,499 cells (1,681 in NP, 609 in G, 604 in L and 605 in P).
+The cells were then normalised by size factors as previously described\autocite{Lun2016}.
+The log-transformed (log2(counts+1)) counts of highly variable genes (HVGs) were used as features for dimensionality reduction and clustering.
+HVGs were defined using the method from Brennecke et al.\autocite{Brennecke2013}.
+Genes with a Benjamini-Hochberg adjusted P value smaller than 0.1 were considered to be significant.
+The t-SNE embedding in \autoref{F1} was computed using the ``Rtsne'' package with default settings and perplexity set to 25 (\href{https://github.com/jkrijthe/Rtsne}{\textit{https://github.com/jkrijthe/Rtsne}}).
+
+\textbf{Clustering}
+
+As we expected the epithelial compartment to show a hierarchical organisation, we used agglomerative hierarchical clustering to identify cell types.
+We first computed the pairwise Euclidean distances between cells based on the log-transformed counts of HVGs.
+The dissimilarity matrix was used to perform hierarchical clustering (Supplementary Fig 9a) with the 'hclust' function in R using average linkage, as this resulted in the highest cophenetic correlation coefficient between the tree and the dissimilarity matrix (average:0.91 single: 0.73, complete: 0.83, ward.D2: 0.73).
+Clusters were then defined with the 'cutreeDynamic' function from the 'dynamicTreeCut' package (minClusterSize= 15, deepSplit=1, method=''hybrid'')\autocite{Langfelder2008}.
+The deepSplit parameter was set to 1 based on evaluating the robustness of the clustering.
+At a deepSplit of 0, all clusters were highly robust as determined by bootstrapping\autocite{Henning2015}.
+Increasing the parameter to 1, led to splitting C1 from C3 as well as C9 from C6.
+C1, C3 and C6 were highly robust in the bootstrap (see Supplementary Fig 9b).
+C9 showed a below average robustness.
+However, due to the expression of marker genes for a previously reported cell type\autocite{Wang2015} we decided to keep it as a separate cluster.
+Interestingly, the putative cell type of C9 is known to have an expression profile that is in between luminal and basal cells (see main text), which might explain its assignment to other clusters during the bootstrap.
+Further increasing the deepSplit parameter led to unstable clusters (Supplementary Fig 9b).
+With this method, we identified ten clusters. In addition, the dynamic tree cut method allows for a ``noise component'' for cells that do not match to any cluster.
+1 cell from the lactation sample was flagged as noise by the algorithm and removed from further analysis.
+C10 (27 cells) expressed lymphocyte related genes was also removed from all downstream analyses (as previously reported\autocite{Scheele2017}, Supplementary Fig 10).
+
+\textbf{Differential expression analysis }
+
+Differential gene expression analysis was performed using ``edgeR''\autocite{Robinson2010}.
+Genes with a mean expression level below 0.1 counts were removed from the analysis.
+A negative binomial generalized log-linear model was fitted to the remaining genes with the cluster assignments as covariate(s).
+The 'glmTreat' function was used to identify genes that have a significantly higher log fold change than 1 at an FDR of 0.01.
+
+\textbf{Diffusion maps and pseudotime inference}
+
+For inferring the differentiation trajectory we used diffusion maps.
+First, we selected all cells from the NP and G time point (\autoref{F2}a) and detected the HVGs as described above.
+The log-transformed (log2(count+1)) gene counts were then used to compute the first twenty diffusion components using the 'DiffusionMap' function (k=50, otherwise the default parameters as in 'destiny'\autocite{Angerer2016}.
+After inspecting the fraction of explained variation for the first twenty components, we decided to retain the first 3.
+In \autoref{F2}b we then focused on the luminal compartment and recomputed the diffusion map based only on the luminal cells, using the aforementioned procedure.
+Here we retained only the first two components (Supplementary Fig 3a).
+Notably, the structure inferred by the diffusion map algorithm was robust to the choice of features and down-sampling of cells (Supplementary Fig 3d).
+The structure of a common origin and the two branches could also be inferred using Monocle with standard settings\autocite{Trapnell2014} (Supplementary Fig 3c).
+For inferring the branches and pseudotime ordering, we defined the following three tips, the cell with the smallest value for the second eigenvector (which was set as root) and the cells with the largest and smallest values for the first eigenvector (compare \autoref{F2}b).
+
+\textbf{Pseudotime-dependent expression}
+
+To identify genes whose expression was significantly associated with the pseudotime we first fitted a natural cubic spline with three degrees of freedom to the log-transformed (log2(counts+1)) expression data in each branch.
+A likelihood ratio test was then used to assess statistical significance of the fit compared to a null (pseudotime-independent) model.
+Genes with a Benjamini-Hochberg corrected P value below 0.001 were considered to be significantly pseudotime-dependent.
+We then used a heuristic definition of branch-specific expression instead of modelling the branch assignment explicitly.
+This was motivated as follows.
+We were interested in general expression trends of genes, i.e.
+increase or decrease along the differentiation towards one of the two cell types, rather than comparing the exact timing of gene activation/inactivation between the branches.
+Any approach trying to do the latter would have been complicated by the different cell densities along the branches and by the difficulty of verifying any such hypotheses \textit{in vivo}.
+Hence, we defined genes to be branch-specific when they were pseudotime-dependent in their expression in at least one of the two branches and when the gradient differed in signs between two branches.
+The gradient was determined as the coefficient of a linear model fit to the spline-smoothed expression values, which was set to 0 if the coefficient was not significantly different from 0 at alpha=0.01.
+Consequently, the gradient of a gene could either be -1 (decreasing), 0 (flat) or 1 (increasing).
+
+\textbf{Gene set enrichment analysis}
+
+A gene set enrichment analysis based on gene-ontology (GO) terms was conducted to characterize genes that were up-regulated in C4 compared to C5.
+Genes with positive log fold changes were compared to all genes that were tested for differential expression using topGO with default settings\autocite{Alexa2016}.
+
+\textbf{Code availability}
+
+All computational analyses were performed in R (Version 3.3.3) using standard functions unless otherwise indicated.
+Code is available online at \href{https://github.com/MarioniLab/MammaryGland}{\textit{https://github.com/MarioniLab/MammaryGland}}.
+Data can be accessed at this link (\url{https://karstenbach.shinyapps.io/webapp/}).
+
+\textbf{Author contribution}
+
+K.B. performed most of the experiments and all the computational analysis.
+S.P. and K.B. setup and collected the mammary epithelial cells.
+M.G. and J.H performed the 10X library production and sequencing.
+J.C.M, D.A., K.B and W.T.K conceptualised the study and wrote the manuscript.
+
+\textbf{Acknowledgements }
+
+We would like to thank the staff at Sanger Institute, Research Service Facility (RSF) for their assistance.
+We would like to thank Dr. Aaron T. Lun (CRUK CI) for helpful discussions and comments on manuscript.
+K.B. is funded by a Cambridge Cancer Centre studentship.
+D.A. is funded by CRUK and Wellcome Trust.
+S.P. is funded by CRUK.
+J.C.M is funded by CRUK and EMBL.
+W.T.K is funded by a CRUK career establishment award (C47525/A17348), University of Cambridge and Magdalene College, Cambridge.
+
+\printbibliography
 \end{document}

### 2017-06-09 src/QCAnalysis.R
added tsne back in

@@ -186,9 +186,16 @@ m.norm <- t(t(m.filtered)/pD.filtered$sf)
 brennecke <- BrenneckeHVG(m.norm,fdr=0.1)
 fD.filtered$highVar <- fD.filtered$id %in% brennecke
 
+# Compute tSNE 
+fPCA <- log2(t(m.norm[brennecke,])+1)
+fPCA <- scale(fPCA,scale=TRUE,center=TRUE)
+set.seed(300)
+tsn <- Rtsne(fPCA,perplexity=25)
+pD.filtered$tSNE1 <- tsn$Y[,1]
+pD.filtered$tSNE2 <- tsn$Y[,2]
 
 # ---- SaveData ----
-pD.add <- select(pD.filtered, barcode, sf)
+pD.add <- select(pD.filtered, barcode, sf, tSNE1, tSNE2)
 fD.add <- select(fD.filtered, id, highVar)
 pD <- left_join(pD,pD.add, by="barcode")
 fD <- left_join(fD,fD.add, by="id")

### 2017-06-09 src/ClusterComparison.Rmd
added scripts for choosing clustering method back in and dynamicCluster takes distance matrix form asDis now

@@
 	   selectInput("dm", label = "Dissimilarity",
-		       choices= c("euclidean","pearson","spearman"), selected="pearson"),
+		       choices= c("euclidean","pearson","spearman"), selected="euclidean"),
 
@@
 clustering <- reactive({
-    dynamicCluster(t(m.sub), dm=input$dm,
-				 lk=input$lk,
+    dis <- asDis(m.sub,dm=input$dm)
+    dynamicCluster(dis, lk=input$lk,
 			ds=as.numeric(input$ds),

### 2017-06-09 src/ClusterBootsrap.R

@@ -2,11 +2,7 @@
 
 library(scran)
 library(dplyr)
-library(knitr)
-library(ggplot2)
 library(dynamicTreeCut)
-library(Rtsne)
-library(pheatmap)
 require(doParallel)
 require(fpc)
 require(clValid)
@@ -27,11 +23,11 @@ fD <- fD[fD$keep,]
 m <- t(t(m)/pD$sf)
 
 # Feature Selection
-m.sub <- t(m[fD$highVar,])
+m.sub <- m[fD$highVar,]
 
 # Combinations of clustering parameters
-dms <- c("euclidean")
-lks <- c("average")
+dms <- c("euclidean","pearson","spearman")
+lks <- c("average","complete")
 dss <- c(0,1,2)
 nCores <- 3
 cl <-makeCluster(nCores, type="FORK")
@@ -41,8 +37,10 @@ result <- foreach(i=seq_along(dss), .combine=c) %dopar% {
     tmp0 <- list()
 	for (lk in lks) {
 	    for (dm in dms) {
-		res <- clusterboot(data=m.sub, B=100, clustermethod=dynamicCluster,
-				   dm=dm, ds=ds, lk=lk)
+
+		dis <- asDis(m.sub,dm=dm)
+		res <- clusterboot(data=dis, B=100, clustermethod=dynamicCluster,
+				   ds=ds, lk=lk)
 		id <- paste(dm,lk,ds,sep="_")
 		tmp0[[id]] <- res
 	    }}

### 2017-06-09 src/Clustering.R

@@ -38,11 +38,12 @@ ds <- 1
 minSize <- 15
 
 # Clustering
-clust <- dynamicCluster(t(m.sub), dm=dm,
-				 lk=lk,
-				 ds=ds,
-				 output="Cluster",
-				 minSize=minSize)
+dis <- asDis(m.sub, dm=dm)
+clust <- dynamicCluster(dis=dis,dm=dm,
+			  	lk=lk,
+				ds=ds,
+				output="Cluster",
+				minSize=minSize)
 
 pD$cluster <- clust$cluster
 

### 2017-06-09 src/ComputeClusterComparison.R

@@ -1,0 +1,49 @@
+library(scran)
+library(dplyr)
+library(knitr)
+library(ggplot2)
+library(dynamicTreeCut)
+library(Rtsne)
+library(pheatmap)
+source("functions.R")
+
+rnd_seed <- 300
+dataList <- readRDS("../data/Robjects/ExpressionList_QC.rds")
+
+m <- dataList[[1]]
+pD <- dataList[[2]]
+fD <- dataList[[3]]
+#Gene and Cell filtering
+m <- m[fD$keep,pD$PassAll]
+pD <- pD[pD$PassAll,]
+fD <- fD[fD$keep,]
+
+#Normalization
+m <- t(t(m)/pD$sf)
+# 
+
+#Run various clustering combinations
+
+fss <- c("highVar")
+dms <- c("pearson","spearman","euclidean")
+lks <- c("complete","average","ward.D2")
+dss <- c(0,1,2,3,4)
+require(doParallel)
+require(fpc)
+require(clValid)
+nCores <- 3
+cl <-makeCluster(nCores, type="FORK")
+registerDoParallel(cl)
+result <- foreach (i=seq_along(dms), .combine=rbind) %dopar% {
+    dm <- dms[i]
+    tmp0 <- NULL
+    for (fs in fss) {
+	for (lk in lks) {
+	    for (ds in dss) {
+		res <- compClustering(m,fD,fs=fs,dm=dm,lk=lk,ds=ds)
+		tmp0 <- rbind(tmp0,res)
+	    }}}
+    return(tmp0)}
+stopCluster(cl)
+
+saveRDS(result,file="../data/Robjects/ClusterComparison.rds")

### 2017-06-09 src/S9_S10.R

@@ -36,8 +36,8 @@ dm <- "euclidean"
 lk <- "average"
 ds <- 1
 minSize <- 15
-clust <- dynamicCluster(t(m.sub), dm=dm,
-				 lk=lk,
+dis <- asDis(m.sub,dm=dm)
+clust <- dynamicCluster(dis=dis, lk=lk,
 				 ds=ds,
 				 output="Cluster",
 				 minSize=minSize)

### 2017-06-09 src/functions.R

@@ -65,18 +65,9 @@ BrenneckeHVG <- function (m, suppress.plot = FALSE, fdr = 0.1,
 	}
 	return(names(means)[sig])
     }
-dynamicCluster <- function(m, dm="euclidean", lk="average", ds=0, output="ForBootstrap", minSize=15) {
-    print("Clustering on n*p matrix")
-    trafM <- log2(m+1)
-
-    #Dissimilarity Measure
-    if(dm=="euclidean") {
-	dis <- dist(trafM,method=dm)
-    } else {
-	dis <- as.dist((1-cor(t(trafM),method=dm))/2)
-    }
-
-
+dynamicCluster <- function(dis, lk="average", ds=0, output="ForBootstrap", minSize=15) {
+    # ensure dis is dis matrix
+    dis <- as.dist(dis)
     #Hierarchical Clustering
     htree <- hclust(dis, method=lk)
     #Tree Cut
@@ -125,7 +116,56 @@ asSim <- function(distance) {
     simMat <- 1-(distance-min(distance))/(max(distance)-min(distance))
 }
 
+asDis <- function(m, dm="euclidean") {
+		logM <- t(log2(m+1))
+		#Dissimilarity Measure
+		if(dm=="euclidean") {
+		    dis <- dist(logM,method=dm)
+		} else {
+		    dis <- as.dist((1-cor(t(logM),method=dm))/2)
+		}
+		return(dis)
+}
+
+
 grab_grob <- function(){
   grid.echo()
   grid.grab()
 }
+compClustering <- function(m,fD,fs="brennecke",dm="spearman",lk="ward.D2",ds=0,minSize=15) {
+
+    #Feature Selection
+    keep <- fD[,fs]
+    trafM <- log2(m[keep,]+1)
+
+    #Dissimilarity Measure
+    if(dm=="euclidean") {
+	dis <- dist(trafM,method=dm)
+    } else {
+	dis <- as.dist((1-cor(t(trafM),method=dm))/2)
+    }
+
+    #Hierarchical Clustering
+    htree <- hclust(dis, method=lk)
+    #Tree Cut
+    cluss <- cutreeDynamic(htree,distM=as.matrix(dis),deepSplit=ds,minClusterSize=minSize)
+
+    #Compute Validation Statistics
+    clstats <- cluster.stats(d=dis, cluss)
+    asw <- clstats$avg.silwidth
+    rss <- clstats$within.cluster.ss
[…]

### 2017-06-14 src/ClusterComparison.Rmd
bugfix in compClustering and minor changes

@@
 			    group=Linkage)) +
-geom_point() +
 geom_path() +
-facet_grid(.~Dissimilarity, scales="free_y")
+geom_point(aes(size=K)) +
+facet_grid(.~Dissimilarity, scales="free_y") +
+ylab("Average Silhouette Width") +
 compPlot
@@
 
-#Compare Cluster stability 
-# boot1 <- bootDat[["euclidean_average_0"]]
-
-#Clustering
-# clust1 <- dynamicCluster(m.sub, dm="euclidean",
-#                          lk="average",
-#                          ds=0,
-#                          output="Cluster")
-# clust2 <- dynamicCluster(m.sub, dm="pearson",
-#                          lk="average",
-#                          ds=0,
-#                          output="Cluster")
-# clust3 <- dynamicCluster(m.sub, dm="pearson",
-#                          lk="ward.D2",
-#                          ds=0,
-#                          output="Cluster")
-
-#Looking a different clusters in depth
-# 
-# fPCA <- t(trafM)
-# fPCA <- scale(fPCA,scale=TRUE,center=TRUE)
-# set.seed(rnd_seed)
-# tsn <- Rtsne(fPCA,perplexity=10,initial_dims=5)
-# pD$tSNE1 <- tsn$Y[,1]
-# pD$tSNE2 <- tsn$Y[,2]
-# tsnePlot <- ggplot(pD, aes(x=tSNE1, y=tSNE2, shape=Condition, color=as.factor(cluster))) +
-#     geom_point(alpha=0.8,size=2) +
-#     theme_bw()
-# tsnePlot
-# 
-# tsnePlot <- ggplot(pD.filtered, aes(x=tSNE1, y=tSNE2, color=log2(Acta2+1), shape=Replicate)) +
-#     geom_point(alpha=0.8,size=2) +
-#     theme_bw()
-# tsnePlot
 ```

### 2017-06-14 src/ClusterBootsrap.R

@@ -27,7 +27,7 @@ m.sub <- m[fD$highVar,]
 
 # Combinations of clustering parameters
 dms <- c("euclidean","pearson","spearman")
-lks <- c("average","complete")
+lks <- c("average","complete","ward.D2")
 dss <- c(0,1,2)
 nCores <- 3
 cl <-makeCluster(nCores, type="FORK")

### 2017-06-14 src/functions.R

@@ -136,7 +136,7 @@ compClustering <- function(m,fD,fs="brennecke",dm="spearman",lk="ward.D2",ds=0,m
 
     #Feature Selection
     keep <- fD[,fs]
-    trafM <- log2(m[keep,]+1)
+    trafM <- t(log2(m[keep,]+1))
 
     #Dissimilarity Measure
     if(dm=="euclidean") {

### 2017-06-14 submissions/resubmission/response.tex
init submissions folder

\documentclass{article}
\usepackage[a4paper]{geometry}
\usepackage{pdfpages}
\begin{document}
\includepdf{letter.pdf}

%---------------------Index-Swapping--------------------------------------
\section*{Index swapping}
After the submission of our manuscript the life science community has been made aware of substantial technical issues of the HiSeq4000.
This is referred to as index-swapping and leads to the missassignment of reads to samples during de-multiplexing.
The full scope of index-swapping has been extensively document by CITE and a response from Illumina can be found HERE.
At the time of submission we were unaware of the mere possibility of this artifact and had thus not looked into this further.
We will briefly summarize what index swapping is and how it affected our data.
We show that 
\subsection*{What is index-swapping}
\subsection*{Impact on the presented data}
\subsection*{Biological conclusion remain unaffected}
\subsection*{Conclusions}

%---------------------Addressing comments--------------------------------------
\section*{Reviewers' Comments}
\subsection*{Reviewer \#1}
\textbf{1. Can it be excluded that Cluster 9 is a non-epithelial contamination? For instance, microvascular or perivascular cells? Cluster 9 cells, similar to myoepithelial cells, express some smooth muscle markers at low/medium level, however, they hardly express any epithelial marker.}\\
\ldots

\textbf{2. Samples from nulliparous and post-involution mice - NP1, NP2, PI1 and PI2. It is not mentioned at what stage of estrus cycle these mice were sacrificed, yet, it is known that gene expression is differentially regulated at different estrus stages. The data shown in Supplementary Figure 2a reveal significant differences in cell distribution between NP1 and NP2, as well as between PI1 and PI2. In pregnancy and lactation, hormonal levels should not differ significantly between individuals 1 and 2. Consistently, the samples 1 and 2 from pregnant and lactating mice, look similar within each stage - L1 and L2 samples, as well as G1 and G2 samples perfectly overlap in Supplementary Figure 2a. Ideally, cycling (NP and PI) mice should have been synchronized prior to cell sorting. This issue should be mentioned in the discussion.}\\
\ldots

\textbf{3. Why is Acta2 relatively high in some luminal clusters shown in Supplementary Figure 2c, – for instance, in Cluster 2, secretory luminal population? Is it “experimental noise”? Could the authors comment.
Page 7, lines 148-149.”… as well as transcription factors that have not previously been associated with luminal differentiation such as Creb5, Hey1…”. This statement is inexact. On the contrary, Hey1 is a well established marker of luminal progenitors (Bouras et al., 2008. Notch Signaling Regulates Mammary Stem Cell Function and Luminal Cell-Fate Commitment, Cell Stem Cell, 3: 429).
Page 9, lines 203-205. In contrast to authors’ statement, it is not clear how the results of this work “might help to explain some of the conflicting results from lineage tracing studies”. The contradictions in lineage tracing data concern mostly the bipotency of basal stem cells, whilst, as the authors mention, the results of this work rather support the concept of lineage-restricted progenitor/stem cells in both compartments (Figure 2a).}\\
\ldots

\textbf{4. Page 4, lines 87-88 “…from two independent mice.” What does “independent” stand for? Littermates?}\\
\ldots

\textbf{5. Page 9, lines 189-190 “…luminal progenitor cells maintain memory of having undergone gestation and involution.” Maybe, “gestation and lactation”?}\\
\ldots

\subsection*{Reviewer \#2}
\textbf{1. A major consideration with this paper is the use of Epcam to sort epithelial cells. Prior seminal publications (eg doi:10.1038/nm.2000) using human cells demonstrate an important contribution of Epcam negative epithelial cells to stem / progenitor activity in the mammary gland. What is the evidence that Epcam quantitatively recovers mouse mammary epithelia? The consequence of this methodology is the possible exclusion of subsets of mammary epithelial cells from the data capture.}\\
\ldots

\textbf{2. Very few studies have used this system to sample diverse epithelial cell types, so it is not yet clear what sampling bias might be present in this method. The authors should compare the proportion of cell types represented in the single cell data with equivalents identified by flow cytometry at these developmental stages to address sampling bias.}\\
\ldots

\textbf{3. The authors conducted analysis on biological replicates, however this is not described in the results section. This is an important consideration with a new technology and the variability between replicates should be appropriately quantitated and addressed in the results. }\\
\ldots

\textbf{4. Given that the authors see this dataset as a resource, data should be made available for download.}\\
\ldots

\textbf{5. Terminal End Buds (TEBs) are a critical mediator of pubertal development, with unique gene expression and function. TEBs start to disappear by the 8 weeks of age point used for nulliparous animals, so these cell types may not have been sampled. Do the authors observe TEBs at this time point?}\\
\ldots

\textbf{Minor issues: \\
1. The link to the scripts on github is not operational \\ 
2. Was estrous stage controlled, measured or considered for the nulliparous mice? This will impact gene expression. \\
3. Please include a summary table of the number of cells captured from each animal, both total and in each cluster \\
4. The description of Fig 1 could use more cross referencing to known markers of each subset. \\
5. The colours of dots in each figure are very hard to discriminate, especially for those with diminished colour vision. Please try to choose a higher contrast palette or the use of patterns to distinguish groups. \\
6. Please label all axes, Eg. Fig A,D,E \\
7. Please conduct pathway and ontology analysis of the genes changing through pseudotime- described in lines 144-151. \\
8. C7 doesn’t appear in Fig 2A. Please correct. \\
9. The genes referred to in lines 134-136 can’t all be seen in Fig 1d.}\\
\ldots

\subsection*{Reviewer \#3}
\textbf{1. The authors need to do independent experiments to confirm that the lactation stage cells really express so low number of genes. I am not convinced that an individual cell at L1 stage only express 400 genes. The authors should did single cell RNA-seq using SMART-seq2 protocol for all of these four stages of cells, probably five to ten single cells for each stage and check if lactation stage cells really express so low number of genes compared with the other three stages of cells.}\\
\ldots

\textbf{2. For Cluster 9, it seems that only 15 cells are there. And for Cluster 8, it seems that only 26 cells are there. Are they real clusters or just doublets? The authors should analyze further of these two clusters. For example, are C8 cells are really intermediate state of C2 \& C5? Or they are merely doublets of C2+C5 cells? Probably the authors should capture all of the differentially expressed genes between C2 and C5 and check if C8 cells in general express these genes at 1/2 of the level of sum of C2+C5. Or maybe do some immunostaining to see if the cells double positive for C2 and C5 markers are really there.}\\
\ldots

\textbf{3. The authors should give the details of the filtration of the poor quality cells: exactly how many genes detected, how many UMI detected, what percentage of reads mapped to mitochondrial genes. Are they the same for all of the four stages of cells? Or different criteria for different stages of cells?}\\
\ldots

\end{document}

### 2017-06-14 submissions/resubmission/letter.tex
init submissions folder

\documentclass{letter}
\usepackage[a4paper]{geometry}
\begin{document}
\signature{Karsten Bach and Walid Khaled}
\begin{letter}{}
    \opening{Dear Ann,}

Thank you for your rapid and constructive review process. We are grateful for the comments from you and the three reviewers, which highlight the importance of our results for the field of mammary gland biology.

We have revised the manuscript to take into account all of the reviewers' comments as well as recently discovered technical problems of Illumina sequencing.
In particular, we have identified that our presented data was in parts compromised by a sequencing artifact referred to as ``index-swapping''.
Below we have outlined in detail how this has affected our data and that our biological conclusion remain unaffected.
In addition, we have repeated the experiment in order to increase our confidence in the presented results.
From this we have learned that \ldots

We hope that you enjoy reading our revised manuscript and look forward to hearing from you soon.
    \closing{Yours sincerely,}

\end{letter}
\end{document}

### 2017-06-14 submissions/resubmission/response.tex
rephrasing

@@
 \documentclass{article}
+\usepackage[backend=bibtex,
+	    natbib=true,
+	    autocite=superscript,
+	    style=nature,
+	    url=true,
+	    sorting=none,
+	    isbn=false]{biblatex} 
+\addbibresource{../../paper/references.bib}
+\usepackage{hyperref}
+\usepackage{url}
 \usepackage[a4paper]{geometry}
@@
 \section*{Index swapping}
-After the submission of our manuscript the life science community has been made aware of substantial technical issues of the HiSeq4000.
+After the submission of our manuscript \citeauthor{Sinha2017} reported a spreading-of-signal from multiplexed samples on the HiSeq 4000 \autocite{Sinha2017}.
 This is referred to as index-swapping and leads to the missassignment of reads to samples during de-multiplexing.
-The full scope of index-swapping has been extensively document by CITE and a response from Illumina can be found HERE.
-At the time of submission we were unaware of the mere possibility of this artifact and had thus not looked into this further.
-We will briefly summarize what index swapping is and how it affected our data.
-We show that 
+The full scope of index-swapping has been extensively documented in \autocite{Sinha2017} to which Illumina has responed in the form of a white paper \autocite{Illumina}.
+At the time of submission we were unaware of the mere possibility of this artifact and have thus only now analysed our data in this regard.
+
+We found that index swapping although strongly affecting our data has not altered our biological conclusions.
+
+For this we have re-sequenced the library on a HiSeq2500 as well as repeated the entire experiment.
+In both instance our biological conclusions held, which we see as strong support for the validity of our claims.
 \subsection*{What is index-swapping}
@@
 
+\printbibliography
 \end{document}

### 2017-06-16 submissions/resubmission/response.tex
first version of response

@@
 \usepackage{hyperref}
+\usepackage{graphicx, subcaption}
 \usepackage{url}
@@
 After the submission of our manuscript \citeauthor{Sinha2017} reported a spreading-of-signal from multiplexed samples on the HiSeq 4000 \autocite{Sinha2017}.
-This is referred to as index-swapping and leads to the missassignment of reads to samples during de-multiplexing.
-The full scope of index-swapping has been extensively documented in \autocite{Sinha2017} to which Illumina has responed in the form of a white paper \autocite{Illumina}.
+This is referred to as index-swapping and leads to wrongly assigning reads to samples during de-multiplexing.
+The full scope of index-swapping has been extensively documented in \autocite{Sinha2017} to which Illumina has responded in the form of a white paper \autocite{Illumina}.
 At the time of submission we were unaware of the mere possibility of this artifact and have thus only now analysed our data in this regard.
+We found that index swapping although substantially affecting our data has not altered our biological conclusions.
+To show this we have re-sequenced the library on a HiSeq 2500, which is almost not affected by index-swapping.
+In addition, we have performed an independent repeat experiment, meaning that we have prepared fresh libraries from new samples and sequenced again on a HiSeq 2500.
+In both instances our biological conclusions held, which we see as strong support for the validity of our findings.
 
-We found that index swapping although strongly affecting our data has not altered our biological conclusions.
+%-------------------------------------------------------------------------
+\subsection*{Introduction}
+As mentioned above the underlying mechanisms and effects on de-multiplexing are extensively covered in \autocite{Sinha2017}, hence we only briefly introduce the topic to improve readability.
+Index-swapping is defined as the priming and extension of library molecules by free indexing primers.
+Mispriming of free indexing primers and extensions lead to the duplication of reads and incorporation of a new indexing sequence (\autoref{F1}A).
+Essentially, this allows information from one sample to bleed into other samples as exemplified in plate based scRNA-seq data from \autocite{Sinha2017}, where reads bleed into the rows or columns due to the swapping of one of the two indices (\autoref{F1}B).\\
+The library design of 10x libraries differs in the regard that the Illumina indices are used to identify samples (in our case e.g. ``Virgin Replicate 1'') instead of cells (\autoref{F1}C).
+Cells are distinguished by a 14bp cellular barcode which is drawn at random from a pool of $\sim$750,000 possible barcodes \autocite{Zheng2017}.
+In the event of index-swapping we would see the same read appearing in two or more samples, in which case the sequence of the cell barcode would also be identical.
+Hence, all reads that have swapped between samples can only be present in cells that share barcode sequences between samples.
+Now there are two potential scenarios.\\
+In the first scenario, two samples contain two cells that share the same barcode by pure chance, which would mean that reads between the two cells could be swapped between samples.
+The impact of this on 10x data is however expected to be small due to the small chance of observing one out of $\sim$750,000 possible barcodes in two samples.
+The second possibility is that a cell from sample A bleeds enough into sample B to provide sufficient reads to create a miniature copy of itself in sample B.
+The amount of bleedthrough necessary to create copies depends on the library size distribution of the cells present in the ``recipient'' sample due to how CellRanger distinguishes between cells and background.
 
-For this we have re-sequenced the library on a HiSeq2500 as well as repeated the entire experiment.
-In both instance our biological conclusions held, which we see as strong support for the validity of our claims.
-\subsection*{What is index-swapping}
+\begin{figure}
+    \begin{subfigure}[c]{0.5\textwidth}
+	\includegraphics[width=\textwidth]{img/swapping.png}
+    \caption{}
+    \end{subfigure}
+    \begin{subfigure}[c]{0.5\textwidth}
+	\includegraphics[width=\textwidth]{img/swapping2.png}
+    \caption{}
+    \end{subfigure}
+    \begin{subfigure}[c]{0.5\textwidth}
+	\includegraphics[width=\textwidth]{img/library.png}
+    \caption{}
+    \end{subfigure}
+    \caption{Overview of index swapping.
+	(\textbf{A}) illustrates the process of index-swapping during cluster generation with the ExAmp chemistry.
+	(\textbf{B}) shows the spreading-of-signal observed when a cell expresses high levels of a gene that is not expressed in other cells on the plate.
+	Swapping of a single barcode on either end then leads to the cross-hair pattern.
+	(\textbf{C}) illustration of the library generated from the 10x, Illumina barcodes that swap are highlighted. 
+	Figures \textbf{A} and \textbf{B} are adopted from \autocite{Sinha2017}.
+    }
+    \label{F1}
+\end{figure}
+
 \subsection*{Impact on the presented data}
+We first noted that the samples in our data share a high number of barcodes (\autoref{F2}A).
+In a one-by-one comparison between samples all possible comparisons show a higher number of barcodes shared than expected by chance (\autoref{F2}B).
+This suggests the presence of the aboved described second scenario, where index swapping causes cells to create minitaure copies in other samples.
+If this is the case we would also expect cells that share barcodes across samples to be highly similar.
+Indeed, we found that cells sharing the same barcode were almost always part of the same cluster, suggesting a high similarity \autoref{F2}C). \\
+Even if the rate of index swapping can be as high as 10\% \autocite{Sinha2017}, the ``pseudo-cells'' should be substantially smaller than the cell that gave rise to it. 
+Conclusively, in order for CellRanger to identify the ``pseudo-cells'' as cells the real cells in this sample need to have contained few cDNA molecules.
+
+\begin{enumerate}
+    \item More shared barcodes than expected
+    \item Cells sharing barcodes are highly similar
+\end{enumerate}
+\begin{enumerate}
+    \item Lactation sample is only composed of swapped barcodes
+    \item Lowest library size
+    \item Cells from all samples created fake cells
+\end{enumerate}
+\begin{figure}
+    \begin{subfigure}[c]{0.5\textwidth}
+	\includegraphics[width=\textwidth]{img/p1.pdf}
+    \caption{}
+    \end{subfigure}
+    \begin{subfigure}[c]{0.5\textwidth}
+	\includegraphics[width=\textwidth]{img/p2.pdf}
+    \caption{}
+    \end{subfigure}
+    \begin{subfigure}[c]{0.5\textwidth}
+	\includegraphics[width=\textwidth]{img/p3.pdf}
+    \caption{}
+    \end{subfigure}
+    \caption{Overview of index swapping.
+    }
+    \label{F2}
+\end{figure}
 \subsection*{Biological conclusion remain unaffected}

### 2017-06-16 src/IndexSwapping.R

@@ -1,0 +1,204 @@
+library(ggplot2)
+library(cowplot)
+
+# Computation of p-values based on hypergeometric test
+compare <- function(barcodes, samples) {
+    out <- NULL
+    ids <- levels(samples)
+    combs <- combn(ids,m=2, simplify=FALSE)
+    for (i in seq_along(combs)) {
+	comb <- combs[[i]]
+	s1 <- comb[1]
+	s2 <- comb[2]
+	bc1 <- as.character(barcodes[samples==s1])
+	bc2 <- as.character(barcodes[samples==s2])
+	x <- length(intersect(bc1,bc2))
+	m <- length(bc1)
+	n <- 750000
+	k <- length(bc2)
+	p.val <- phyper(q=x-1,m=m,n=n,k=k,lower.tail=FALSE)
+	tmp <- data.frame(s1=s1,
+			  s2=s2,
+			  n1=m,
+			  n2=k,
+			  shared=x,
+			  p.val=p.val)
+	out <- rbind(out,tmp)
+    }
+    return(out)
+}
+
+# Load Data
+dList <- readRDS("../data/Robjects/ExpressionList_Clustered.rds")
+pD <- dList[[2]]
+
+# Extract barcodes by removing sampleID from bc
+pD <- pD[pD$PassAll & !pD$isImmuneCell & !pD$isOutlier,]
+barcodes <- as.character(pD$barcode)
+spt <- strsplit(barcodes, split = "-", fixed = T)
+pD$sample <- sapply(spt, function(x) x[2])
+pD$bcs <- sapply(spt, function(x) x[1])
+pD.add <- data.frame(bcs=names(table(pD$bcs)),
+		     bc.obs=(table(pD$bcs)))
+pD.add <- pD.add[,-2]
+pD <- dplyr::left_join(pD,pD.add)
+
+# ------- More Barcodes are shared than expected -------------
+
+# P1
+p1 <- ggplot(pD, aes(x=SampleID, fill=as.factor(bc.obs.Freq))) +
+    geom_bar() +
+    ggtitle("Samples contain many shared barcodes") +
+    scale_fill_discrete(name="# Barcode observed") +
+    theme(legend.position="bottom",
+	  legend.direction="horizontal")
+
+ggsave("../submissions/resubmission/img/p1.pdf",p1)
+
+# P2
+compDf <- compare(pD$bcs, pD$SampleID)
[…]

### 2017-06-19 submissions/resubmission/response.tex
beginning discussing lact

@@
 \usepackage{pdfpages}
+\renewcommand{\thesubfigure}{\Alph{subfigure}}
+\newcommand{\comment}[1]{\textit{\textcolor{blue}{Comment: #1}}}
 \begin{document}
@@
 At the time of submission we were unaware of the mere possibility of this artifact and have thus only now analysed our data in this regard.
-We found that index swapping although substantially affecting our data has not altered our biological conclusions.
+We found that index-swapping although substantially affecting our data has not altered our biological conclusions.
 To show this we have re-sequenced the library on a HiSeq 2500, which is almost not affected by index-swapping.
@@
 Hence, all reads that have swapped between samples can only be present in cells that share barcode sequences between samples.
-Now there are two potential scenarios.\\
-In the first scenario, two samples contain two cells that share the same barcode by pure chance, which would mean that reads between the two cells could be swapped between samples.
-The impact of this on 10x data is however expected to be small due to the small chance of observing one out of $\sim$750,000 possible barcodes in two samples.
-The second possibility is that a cell from sample A bleeds enough into sample B to provide sufficient reads to create a miniature copy of itself in sample B.
-The amount of bleedthrough necessary to create copies depends on the library size distribution of the cells present in the ``recipient'' sample due to how CellRanger distinguishes between cells and background.
+Conceptually, we can distinguish three potential scenarios.\\
+In the first scenario, two samples contain two cells that share the same barcode by pure chance, which would mean that reads between the two cells could be swapped between samples in either direction.
+The impact of this is however expected to be small due to the small chance of observing one out of $\sim$750,000 possible barcodes in two samples.
+In the second scenario, only one of two samples (sample A) contains a cell with a specific barcode whereas the other does not (sample B).
+In this case, the swapping is mostly uni-directional, where sample A bleeds into B.
+Sample B only contains some background levels of reads for this barcode and the absolute number of reads swapped into the other sample should be insignificant.
+The third possibility arises when reads from sample A bleed enough into sample B to provide sufficient reads to create a miniature copy of itself in sample B.
+The amount of bleedthrough necessary to create copies depends on the library size distribution of the cells present in the ``recipient'' sample due to how CellRanger distinguishes between cells and background.\\
+In summary, the first scenario is a rare event that only affects very few cells.
+The effect of the second case on the data is negligible as this might only increase the general noise in the data to a small extent.
+The third scenario can be more severe depending on the library size distribution of the samples that were run together on the HiSeq 4000.
 
@@
     \end{subfigure}
-    \caption{Overview of index swapping.
+    \caption{Overview of index-swapping.
 	(\textbf{A}) illustrates the process of index-swapping during cluster generation with the ExAmp chemistry.
@@
 \subsection*{Impact on the presented data}
-We first noted that the samples in our data share a high number of barcodes (\autoref{F2}A).
+\paragraph{Index swapping occurred}
+First, we noted that the samples in our data share a high number of barcodes (\autoref{F2}A).
 In a one-by-one comparison between samples all possible comparisons show a higher number of barcodes shared than expected by chance (\autoref{F2}B).
-This suggests the presence of the aboved described second scenario, where index swapping causes cells to create minitaure copies in other samples.
-If this is the case we would also expect cells that share barcodes across samples to be highly similar.
-Indeed, we found that cells sharing the same barcode were almost always part of the same cluster, suggesting a high similarity \autoref{F2}C). \\
-Even if the rate of index swapping can be as high as 10\% \autocite{Sinha2017}, the ``pseudo-cells'' should be substantially smaller than the cell that gave rise to it. 
-Conclusively, in order for CellRanger to identify the ``pseudo-cells'' as cells the real cells in this sample need to have contained few cDNA molecules.
+This suggests the presence of the third scenario, where index-swapping creates ``pseudo-cells''.
+If this is the case we would also expect cells that share barcodes across samples to be highly similar, which again would otherwise be non-expected.
+Indeed, we found that cells sharing the same barcode were almost always part of the same cluster, suggesting a high similarity (\autoref{F2}C).
+Lastly, we also identified a pattern per barcode of one cell with substantially more mRNA molecules than the remaining cells for that barcode (\autoref{F2}D). \\
+All of this suggested to us that index-swapping occurred and has led to the identification of ``pseudo-cells'' that are in fact only copies of other cells from other samples.\\
 
-\begin{enumerate}
-    \item More shared barcodes than expected
-    \item Cells sharing barcodes are highly similar
-\end{enumerate}
-\begin{enumerate}
-    \item Lactation sample is only composed of swapped barcodes
-    \item Lowest library size
-    \item Cells from all samples created fake cells
-\end{enumerate}
 \begin{figure}
@@
     \end{subfigure}
-    \caption{Overview of index swapping.
+    \begin{subfigure}[c]{0.5\textwidth}
+	\includegraphics[width=\textwidth]{img/p4.pdf}
+    \caption{}
+    \end{subfigure}
+    \caption{Index swapping compromised parts of the presented data.
+	(\textbf{A}) Number of barcodes detected in each samples, colored by the number of times each barcode appears in the entire dataset.
+	(\textbf{B}) Volcano-plot based of the number of shared barcodes versus the P value to observe at that many or more barcodes shared between any two samples.
+	P values are based on a hypergeometric test for all possible two-sample comparisons. \comment{add more detail about test?}
+	(\textbf{C}) Histogram of the fraction of cells with the same barcode that were identified in our analysis as being part of one cluster.
+	(\textbf{D}) Kernel density estimates of the distribution of ratios between the largest cell and the mean of the remaining cells for each barcode that was shared between multiple samples.
     }
@@
 \end{figure}
+
+\paragraph{Cells identified in the lactation sample derive from index-swapping}
+We were surprised to find that the lactation samples only contained one single cell with a unique barcode.
+In addition, 313 out of all 314 non-unique barcodes across the entire dataset had at least one barcode appearing in the lactation sample and in 298 case was the cell with the largest library size not coming from the lactation sample (\autoref{F1}A, \autoref{F3}A).
+Further, we can look at the pattern of index-swapping in more detail, assuming that for each barcode the cell with the largest number of UMIs is the cell that gave rise to ``pseudo-cells'' with fewer UMIs.
+Here it becomes evident again that in the majority of cases one or both lactation samples received reads from swapping (\autoref{F3}B).
+This suggested to us that the vast majority of cells in the lactation samples are ``pseudo-cells'' derived from index-swapping from other samples. \\
+\autoref{F3}B also shows that in the few cases were reads were not swapped into the lactation samples, the lactation samples themselves appeared to be the one that were the donor samples.
+
+\begin{figure}
+    \begin{subfigure}[c]{0.5\textwidth}
+	\includegraphics[width=\textwidth]{img/p5.pdf}
+    \caption{}
+    \end{subfigure}
+    \begin{subfigure}[c]{0.5\textwidth}
+	\includegraphics[width=\textwidth]{img/p6.pdf}
+    \caption{}
+    \end{subfigure}
+    \caption{Lactation sample contained no cells.
+    }
+    \label{F3}
+\end{figure}
+
 \subsection*{Biological conclusion remain unaffected}
 \subsection*{Conclusions}
-
+\newpage
 %---------------------Addressing comments--------------------------------------

### 2017-06-19 src/IndexSwapping.R

@@ -48,8 +48,8 @@ pD <- dplyr::left_join(pD,pD.add)
 # P1
 p1 <- ggplot(pD, aes(x=SampleID, fill=as.factor(bc.obs.Freq))) +
     geom_bar() +
-    ggtitle("Samples contain many shared barcodes") +
-    scale_fill_discrete(name="# Barcode observed") +
+    #     ggtitle("Samples contain many shared barcodes") +
+    scale_fill_discrete(name="Times barcode observed") +
     theme(legend.position="bottom",
 	  legend.direction="horizontal")
 
@@ -61,8 +61,8 @@ p2 <- ggplot(compDf, aes(x=shared, y=-log10(p.val))) +
     geom_point() +
     xlab("# shared Barcodes") +
     ylab("-log10(P)") +
-    geom_hline(yintercept=2,lty="dashed",color="red") +
-    ggtitle("Samples share more barcodes than expected by chance") 
+    geom_hline(yintercept=2,lty="dashed",color="red") 
+    #     ggtitle("Samples share more barcodes than expected by chance") 
 
 ggsave("../submissions/resubmission/img/p2.pdf",p2)
 
@@ -70,45 +70,113 @@ bcTab <- as.matrix(table(pD$bcs,pD$cluster))
 dupBcs <- pD[pD$bc.obs.Freq>1,"bcs"]
 bcTab <- bcTab[pD[pD$bcs %in% dupBcs,"bcs"],]
 freq <-  apply(bcTab,1,function(x) ifelse(max(x)>1,max(x),0))/apply(bcTab,1,sum) * 100
+freq <- freq[unique(names(freq))]
 pD.add <- data.frame("bcs"=names(freq),
 		   "freqSameCluster"=freq)
 pD <- dplyr::left_join(pD,pD.add,by="bcs")
 
 p3 <- ggplot(pD, aes(x=freqSameCluster)) +
-    geom_bar(color="black",fill="white") +
+    geom_bar(color="black",fill="grey") +
     xlab("Fraction of cells per shared barcode in the same cluster") +
-    ylab("Count") +
-    ggtitle("Cells that share barcodes cluster together")
+    ylab("Count") 
+    #     ggtitle("Cells that share barcodes cluster together")
 
 ggsave("../submissions/resubmission/img/p3.pdf",p3)
 
[…]

### 2017-06-20 submissions/resubmission/response.tex
updates

@@
 This suggests the presence of the third scenario, where index-swapping creates ``pseudo-cells''.
-If this is the case we would also expect cells that share barcodes across samples to be highly similar, which again would otherwise be non-expected.
+If this is the case we would also expect cells that share barcodes across samples to be highly similar, which again would otherwise not be expected.
 Indeed, we found that cells sharing the same barcode were almost always part of the same cluster, suggesting a high similarity (\autoref{F2}C).
-Lastly, we also identified a pattern per barcode of one cell with substantially more mRNA molecules than the remaining cells for that barcode (\autoref{F2}D). \\
+Lastly, we also identified a pattern per barcode where one cell has substantially more mRNA molecules than the remaining cells for that barcode (\autoref{F2}D). \\
 All of this suggested to us that index-swapping occurred and has led to the identification of ``pseudo-cells'' that are in fact only copies of other cells from other samples.\\
@@
 \paragraph{Cells identified in the lactation sample derive from index-swapping}
-We were surprised to find that the lactation samples only contained one single cell with a unique barcode.
-In addition, 313 out of all 314 non-unique barcodes across the entire dataset had at least one barcode appearing in the lactation sample and in 298 case was the cell with the largest library size not coming from the lactation sample (\autoref{F1}A, \autoref{F3}A).
+\autoref{F1}A suggests that the lactation samples in particular were severely affected with only a single cell having a unique barcode.
+In addition, 313 out of all 314 non-unique barcodes across the entire dataset appeared at least once in the lactation sample and in 298 case was the cell with the largest library size not coming from the lactation sample (\autoref{F1}A, \autoref{F3}A).
+This suggests that these were the samples that were mainly involved in index swapping by soaking up reads from other samples.
 Further, we can look at the pattern of index-swapping in more detail, assuming that for each barcode the cell with the largest number of UMIs is the cell that gave rise to ``pseudo-cells'' with fewer UMIs.
 Here it becomes evident again that in the majority of cases one or both lactation samples received reads from swapping (\autoref{F3}B).
-This suggested to us that the vast majority of cells in the lactation samples are ``pseudo-cells'' derived from index-swapping from other samples. \\
-\autoref{F3}B also shows that in the few cases were reads were not swapped into the lactation samples, the lactation samples themselves appeared to be the one that were the donor samples.
+Together, this suggested to us that the vast majority of cells in the lactation samples are ``pseudo-cells'' derived from index-swapping.
+The most likely explanation for this is that the cells form the lactation sample either did not lyse properly or lysed before they were encapsulated in the beads.
+This explains the low amount of cDNA we observed for these samples (data not shown).
+On a HiSeq 2500 this would have led to the detection of only few cells.
+However, due to the index-swapping on the HiSeq 4000 these samples acted as a pool that soaked up index-swapped reads from other samples.\\
+\autoref{F3}B also shows that in the few cases were reads were not swapped into the lactation samples, the lactation samples themselves bled into the other samples.
+In total we found 17 barcodes that appeared to stem from true cells in the lactation samples, notably all cells of cluster 7 and 8 were part of this (\autoref{F3}C).
+The fact that the cells from lactation were highly likely to bleed into other samples can be explained by the high relative concentration of their cDNA.
+As all cDNA were pooled at equal amounts for sequencing the lactation samples contained much more cDNA per cell than the other samples, increasing the likelihood of index-swapping.
+This also explains why the ratio of total number of unique molecules in \autoref{F1}C is low when the reads came from lactation.\\
+
+In summary, the data we presented was compromised by index swapping, a technical artifact specific to the Illumina HiSeq 4000 that leads to a spreading-of-signal.
+This had a substantial effect on the lactation samples, which - considering the new evidence - are most likely to have been almost completely avoid of any true cells.
+This means that all inference based on these two sample are meaningless. 
+In addition, two of the identified clusters (C7, C8) are now also questionable as they are made up of much fewer cells than previously thought.\\
+In order to gauge how much this has impacted the biological conclusion we drew from our analysis we re-sequenced the same cDNA library on a HiSeq 2500.
+Furthermore, we have repeated the entire experiment taking some of the reviewers' comments into consideration in order to be fully confident of our statements.
+In the following we will summarize the re-analysis of the HiSeq 2500 data from the first experiment.
+The second experiment can then be found in the revised version of the manuscript.
 
@@
     \end{subfigure}
+    \begin{subfigure}[c]{0.5\textwidth}
+	\includegraphics[width=\textwidth]{img/p7.pdf}
+    \caption{}
+    \end{subfigure}
     \caption{Lactation sample contained no cells.
@@
 
-\subsection*{Biological conclusion remain unaffected}
+\subsection*{Biological conclusions remain unaffected}
 \subsection*{Conclusions}

### 2017-06-20 src/IndexSwapping.R

@@ -101,7 +101,6 @@ ggsave("../submissions/resubmission/img/p4.pdf",p4)
 
 # ---------- Lactation sample did not contain real cells --------------
 
-
 p5 <- ggplot(pD, aes(x=tSNE1, y=tSNE2, color=Condition)) +
     theme_void(base_size=12) +
     geom_point(size=1.5) +
@@ -141,13 +140,15 @@ trueL <- group_by(pD, bcs) %>%
 sbst <- filter(pD, bcs %in% trueL) %>%
     mutate(cluster=paste0("C",cluster))
 
-p7 <- ggplot(sbst, aes(x=bcs,y=UmiSums,fill=SampleID)) +
+p7 <- ggplot(sbst, aes(x=bcs,y=UmiSums /1000,fill=SampleID)) +
     geom_bar(stat="identity",position="dodge") +
     facet_grid(~cluster, scales="free") +
     scale_color_brewer("Paired") +
-    ylab("Library Size") +
+    ylab("Library Size [x1000]") +
     xlab("Barcode") +
     coord_flip()
+
+ggsave("../submissions/resubmission/img/p7.pdf",p7)
 
 
 

### 2017-06-20 submissions/resubmission/response.tex
minor updates

@@
     \end{subfigure}
+    \begin{subfigure}[c]{0.5\textwidth}
+	\includegraphics[width=\textwidth]{img/p10.pdf}
+    \caption{}
+    \end{subfigure}
     \caption{Lactation sample contained no cells.
+	(\textbf{A}) tSNE plot coloured by timepoints.
+	Cells that share a common barcode are connected by dashed lines.
+	(\textbf{B}) Barplot of number of times a given pattern of index-swapping occured.
+	Bars are grouped by the combination of samples that reads bled into and colored by the condition that contained the cell that bled into the others.
+	(\textbf{C}) Library sizes of cells grouped together by barcodes for all cells that appeared to have truly been captured in the lactation sample.
+	Bars are colored by sample identifiers.
+	(\textbf{D}) Same as \autoref{F1}B and \autoref{F1}C after removing the lactation sample and ``pseudo-cells'' that stem from the lactation samples.
     }

### 2017-06-20 src/IndexSwapping.R

@@ -152,8 +152,37 @@ ggsave("../submissions/resubmission/img/p7.pdf",p7)
 
 
 
+# Load Data
+sbst <- pD[pD$Condition !="L" & !(pD$bcs %in% trueL), !(colnames(pD) %in% c("sample","bcs","bc.obs.Freq"))]
+barcodes <- as.character(sbst$barcode)
+spt <- strsplit(barcodes, split = "-", fixed = T)
+sbst$sample <- sapply(spt, function(x) x[2])
+sbst$bcs <- sapply(spt, function(x) x[1])
+sbst.add <- data.frame(bcs=names(table(sbst$bcs)),
+		     bc.obs=(table(sbst$bcs)))
+sbst.add <- sbst.add[,-2]
+sbst <- dplyr::left_join(sbst,sbst.add)
 
 
+p8 <- ggplot(sbst, aes(x=SampleID, fill=as.factor(bc.obs.Freq))) +
+    geom_bar() +
+    #     ggtitle("Samples contain many shared barcodes") +
+    scale_fill_discrete(name="Times barcode observed") +
+    theme(legend.position="bottom",
+	  legend.direction="horizontal")
+compDf <- compare(sbst$bcs, sbst$SampleID)
+p9 <- ggplot(compDf, aes(x=shared, y=-log10(p.val))) +
+    geom_point() +
+    xlab("# shared Barcodes") +
+    ylab("-log10(P)") +
+    geom_hline(yintercept=2,lty="dashed",color="red") 
+    #     ggtitle("Samples share more barcodes than expected by chance")
+
+p10 <- plot_grid(p8,p9,nrow=2)
+
+ggsave("../submissions/resubmission/img/p10.pdf",p10)
+
+# end
 
 
 

### 2017-06-21 submissions/resubmission/response.tex
fig rearranged

@@
 We found that index-swapping although substantially affecting our data has not altered our biological conclusions.
-To show this we have re-sequenced the library on a HiSeq 2500, which is almost not affected by index-swapping.
+To show this we have re-sequenced the cDNA library on a HiSeq 2500, which is almost not affected by index-swapping.
 In addition, we have performed an independent repeat experiment, meaning that we have prepared fresh libraries from new samples and sequenced again on a HiSeq 2500.
@@
 \begin{figure}
-    \begin{subfigure}[c]{0.5\textwidth}
+\begin{minipage}[c][13cm][t]{.5\textwidth}
+  \vspace*{\fill}
+  \centering
   \includegraphics[width=\textwidth]{img/swapping.png}
-    \caption{}
-    \end{subfigure}
-    \begin{subfigure}[c]{0.5\textwidth}
+  \subcaption{}
+\end{minipage}%
+\begin{minipage}[c][11cm][t]{.5\textwidth}
+  \vspace*{\fill}
+  \centering
   \includegraphics[width=\textwidth]{img/swapping2.png}
-    \caption{}
-    \end{subfigure}
-    \begin{subfigure}[c]{0.5\textwidth}
+  \subcaption{}
+  \par\vfill
   \includegraphics[width=\textwidth]{img/library.png}
-    \caption{}
-    \end{subfigure}
+  \subcaption{}
+\end{minipage}
 \caption{Overview of index-swapping.
@@
 The fact that the cells from lactation were highly likely to bleed into other samples can be explained by the high relative concentration of their cDNA.
-As all cDNA were pooled at equal amounts for sequencing the lactation samples contained much more cDNA per cell than the other samples, increasing the likelihood of index-swapping.
-This also explains why the ratio of total number of unique molecules in \autoref{F1}C is low when the reads came from lactation.\\
+As all cDNA libraries were pooled at equal amounts for sequencing the lactation samples had much more cDNA per cell than the other samples, increasing the likelihood of index-swapping.
+\comment{explain why ratio in \autoref{F1}C is low for lactation?}
 
@@
 In the following we will summarize the re-analysis of the HiSeq 2500 data from the first experiment.
-The second experiment can then be found in the revised version of the manuscript.
+The repeat experiment can then be found in the revised version of the manuscript.
 

### 2017-06-23 src/PrepareData.R
bugFix!

@@ -9,17 +9,22 @@ library(dplyr)
 gene_bc_matrix <- load_cellranger_matrix("../data/CellRangerData/secondRun_2500/MammaryGland",
 					 genome="mm10")
 pDat <- data.frame(pData(gene_bc_matrix))
-fDat <- data.frame(fData(gene_bc_matrix))
 cDat <- as.matrix(exprs(gene_bc_matrix))
+
+#reduce size of matrix
+keep <- rowSums(cDat) > 1
+cDat <- cDat[keep,]
+fDat <- data.frame(fData(gene_bc_matrix))[rownames(cDat),]
 
 # ---- Formatting ----
 
 # Add more info to phenotype Data
 pDat <- mutate(pDat, SeqID=substr(barcode,18,18)) %>%
-	       mutate(SeqID=mapvalues(SeqID,c("1","2","3","4","5","6","7","8"),
-					 c("A7","B7","C7","D7","E7","F7","G7","H7"))) %>%
+	mutate(barcode=as.character(barcode)) %>%
+        mutate(SeqID=mapvalues(SeqID,c("1","2","3","4","5","6","7","8"),
+				 c("A7","B7","C7","D7","E7","F7","G7","H7"))) %>%
         mutate(SampleID=mapvalues(SeqID,c("A7","B7","C7","D7","E7","F7","G7","H7"),
-				  c("NP1","G2","PI2","G1","L2","L1","NP2","PI1"))) %>%
+				  c("PI1","PI2","L1","L2","G1","G2","NP1","NP2"))) %>%
 	mutate(Replicate=as.factor(substr(SampleID,nchar(SampleID),nchar(SampleID))),
 	       Condition=factor(substr(SampleID,1,nchar(SampleID)-1),levels=c("NP","G","L","PI"))) %>%
 	mutate(SampleID=factor(SampleID,levels=c("NP1","NP2","G1","G2","L1","L2","PI1","PI2"))) 
@@ -29,11 +34,11 @@ mitoGenes <- read.table("../data/miscData/MitoGenes.txt")
 tfCheck <- read.table("../data/miscData/TFcheckpoint_WithENSID.tsv",
 		header=TRUE, sep="\t")
 
-fDat <- mutate(fDat,
-	       Mitochondrial=id %in% mitoGenes$V1,
-	       TranscriptionFactor=id %in% tfCheck$ensembl_gene_id)
+fDat$Mitochondrial <- fDat$id %in% mitoGenes$V1
[…]

### 2017-06-23 src/QCAnalysis.Rmd
report

--- 2017-02-28 src/QCAnalysis.Rmd
@@
 library(ggplot2)
-rnd_seed <- 300
+library(cowplot)
+library(reshape2)
 library(Rtsne)
 source("functions.R")
-dataList <- readRDS("../data/Robjects/ExpressionList.rds")
+dataList <- readRDS("../data/Robjects/secondRun_2500/ExpressionList.rds")
 m <- dataList[["counts"]]
@@
 fD <- dataList[["featureData"]]
-#relevel pD Condition to get it in logical sequence
-pD$Condition <- factor(pD$Condition,levels=c("V","P","L","I"))
+rm(dataList)
 ```
@@
 
+## 
+```{r}
+smry <- read.csv("../data/CellRangerData/secondRun_2500/Summary.csv")
+smry <- melt(smry)
+smry$SampleID <- factor(smry$SampleID,levels=c("NP1","NP2","G1","G2","L1","L2","PI1","PI2"))
+smPlot <- ggplot(smry, aes(x=as.factor(SampleID),y=value,fill=SampleID)) +
+    geom_bar(stat="identity") +
+    facet_wrap(~variable, scale="free") +
+    theme(axis.text.x=element_blank()) +
+    xlab("Sample")
+
+smPlot
+```
+
 ## Number of Cells
-We sequenced less cells than expected, with a total of `r ncol(m)` cells.
+We sequenced more cells than expected with a total of `r ncol(m)` cells.
 ```{r}
@@
     geom_violin(draw_quantiles=0.5)+
-    scale_y_log10() +
-    theme_bw()
+    scale_y_log10() 
 LibrarySize <- ggplot(pD, aes(x=SampleID,y=UmiSums,fill=Condition)) +
     geom_violin(draw_quantiles=0.5)+
-    scale_y_log10() +
-    theme_bw()
+    scale_y_log10() 
 gdLibSize <- ggplot(pD, aes(y=log10(GenesDetected), x=log10(UmiSums), color=Condition, shape=Replicate)) +
     geom_point() +
-    geom_smooth(method=loess,inherit.aes=FALSE,aes(y=log10(GenesDetected),x=log10(UmiSums))) +
-    theme_bw()
+    geom_smooth(method=loess,inherit.aes=FALSE,aes(y=log10(GenesDetected),x=log10(UmiSums))) 
 genesDetected
@@
     geom_boxplot() +
-    coord_flip() +
-    theme_bw()
+    coord_flip() 
 freqOfExp <- m!=0
@@
     coord_flip() +
-    xlab("Gene") +
-    theme_bw()
+    xlab("Gene") 
 topGenes
 topFreq
+rm(mRel)
 ```
@@
 cellViability <- ggplot(pD, aes(x=prcntMito, y=GenesDetected, color=Condition, shape=Replicate))+
-    geom_point() +
-    theme_bw()
+    geom_point() 
 cellViability
 prcntTop <- ggplot(pD, aes(x=prcntTop, y=GenesDetected, color=Condition, shape=Replicate))+
-    geom_point() +
-    theme_bw()
+    geom_point() 
 prcntTop
+rm(mMito)
+rm(mTop)
 ```
+
+## Thresholds
+
 ```{r}
-leftmad <- function(x) {
-    m <- median(x)
-    dev <- abs(x-m)
-    leftmad <- median(dev[x<=m])
-    return(leftmad)
-}
 smryByGroup <- group_by(pD,Condition) %>%
@@
 	      mGenesDetected=median(log10(GenesDetected)),
-	      madGenesDetected=leftmad(log10(GenesDetected)),
-	      threshold_GenesDetected=max(mGenesDetected-4*madGenesDetected,log10(500)),
+	      madGenesDetected=mad(log10(GenesDetected)),
+	      threshold_GenesDetected=mGenesDetected-2*madGenesDetected,
 	      mUmiSums=median(log10(UmiSums)),
-	      madUmiSums=leftmad(log10(UmiSums)),
-	      threshold_UmiSums=max(mUmiSums-4*madUmiSums,log10(1000))) %>%
+	      madUmiSums=mad(log10(UmiSums)),
+	      threshold_UmiSums=mUmiSums-2*madUmiSums) %>%
     select(Condition,starts_with("threshold")) 
 kable(smryByGroup)
+
+thrGenesDetected <- median(log10(pD$GenesDetected))-3*mad(log10(pD$GenesDetected))
+thrUmiSums <- median(log10(pD$UmiSums))-3*mad(log10(pD$UmiSums))
 ```
 
-# QC Thresholding
+# Per Condition
 ```{r}
-#simple thresholds
 MitoCutOff <- 0.05
-prcntTopCutOff <- 0.5
+gdHist <- ggplot(pD, aes(x=GenesDetected,y=..density..)) +
+    geom_histogram(fill="white",color="black") +
+    geom_vline(data=smryByGroup,aes(xintercept=10^threshold_GenesDetected),color="red",lty="longdash") +
+    scale_x_log10() +
+    xlab("Total number of genes detected") +
+    facet_wrap(~Condition)
+
+libSizeHist <- ggplot(pD, aes(x=UmiSums,y=..density..)) +
+    geom_histogram(fill="white",color="black") +
+    geom_vline(data=smryByGroup,aes(xintercept=10^threshold_UmiSums),color="red",lty="longdash") +
+    scale_x_log10() +
+    facet_wrap(~Condition) +
+    xlab("Total number of unique molecules")
+
+cellViability <- cellViability %+% pD
+cellViability <- cellViability + 
+    annotate("rect",ymin=-Inf, ymax=Inf, xmax=Inf, xmin=MitoCutOff,
+	     fill="grey", alpha=0.3)
+
+gdHist
+libSizeHist
+cellViability
 ```
+
+# In total
+
+```{r}
+gdHist <- ggplot(pD, aes(x=GenesDetected,y=..density..)) +
+    geom_histogram(fill="white",color="black") +
+    geom_vline(aes(xintercept=10^thrGenesDetected),color="red",lty="longdash") +
+    scale_x_log10() +
+    xlab("Total number of genes detected")
+
+libSizeHist <- ggplot(pD, aes(x=UmiSums,y=..density..)) +
+    geom_histogram(fill="white",color="black") +
+    geom_vline(aes(xintercept=10^thrUmiSums),color="red",lty="longdash") +
+    scale_x_log10() +
+    xlab("Total number of unique molecules")
+
+gdHist
+libSizeHist
+pD2 <- mutate(pD,
+	     PassViability=prcntMito < MitoCutOff,
+	     PassGenesDet=GenesDetected > thrGenesDetected ,
+	     PassLibSize=UmiSums > thrUmiSums,
+	     PassAll= PassViability & PassGenesDet & PassLibSize)
+table(pD2$PassViability,pD2$SampleID)
+table(pD2$PassGenesDet,pD2$SampleID)
+table(pD2$PassLibSize,pD2$SampleID)
+table(pD2$PassAll,pD2$SampleID)
+```
+
 ```{r}
@@
 	     ThresholdGenesDet = 0,
-	     ThresholdLibSize = 0,
-	     ThresholdPrcntTop = prcntTopCutOff)
+	     ThresholdLibSize = 0)
 
@@
 	     PassLibSize=UmiSums > ThresholdLibSize,
-	     PassPrcntTop=prcntTop > ThresholdPrcntTop,
-	     PassAll= PassViability & PassGenesDet & PassLibSize & PassPrcntTop)
+	     PassAll= PassViability & PassGenesDet & PassLibSize)
+table(pD$PassViability,pD$SampleID)
+table(pD$PassGenesDet,pD$SampleID)
+table(pD$PassLibSize,pD$SampleID)
+table(pD$PassAll,pD$SampleID)
+```
 
-gdHist <- ggplot(pD, aes(x=GenesDetected,y=..density..)) +
-    geom_histogram(fill="white",color="black") +
-    geom_vline(data=smryByGroup,aes(xintercept=10^threshold_GenesDetected),color="red",lty="longdash") +
-    scale_x_log10() +
-    facet_wrap(~Condition) +
-    theme_bw()
-gdHist
+## Index Swapping
 
-libSizeHist <- ggplot(pD, aes(x=UmiSums,y=..density..)) +
-    geom_histogram(fill="white",color="black") +
-    geom_vline(data=smryByGroup,aes(xintercept=10^threshold_UmiSums),color="red",lty="longdash") +
-    scale_x_log10() +
-    facet_wrap(~Condition) +
-    theme_bw()
-libSizeHist
+```{r}
+barcodes <- as.character(pD$barcode)
+spt <- strsplit(barcodes, split = "-", fixed = T)
+pD$sample <- sapply(spt, function(x) x[2])
+pD$bcs <- sapply(spt, function(x) x[1])
+pD.add <- data.frame(bcs=names(table(pD$bcs)),
+		     bc.obs=(table(pD$bcs)))
+pD.add <- pD.add[,-2]
+pD <- dplyr::left_join(pD,pD.add)
 
-cellViability <- cellViability %+% pD
-cellViability <- cellViability + aes(color=PassViability) +
-    annotate("rect",ymin=-Inf, ymax=Inf, xmax=Inf, xmin=MitoCutOff,
-	     fill="grey", alpha=0.3) 
-cellViability
+# P1
+p1 <- ggplot(pD, aes(x=SampleID, fill=as.factor(bc.obs.Freq))) +
+    geom_bar() +
+    #     ggtitle("Samples contain many shared barcodes") +
+    scale_fill_discrete(name="Times barcode observed") +
+    theme(legend.position="bottom",
+	  legend.direction="horizontal")
 
-prcntTop <- prcntTop %+% pD
-prcntTop <- prcntTop + aes(color=PassPrcntTop) +
-    annotate("rect",ymin=-Inf, ymax=Inf, xmin=-Inf, xmax=prcntTopCutOff,
-	     fill="grey", alpha=0.3) 
-prcntTop
+compare <- function(barcodes, samples) {
+    out <- NULL
+    ids <- levels(samples)
+    combs <- combn(ids,m=2, simplify=FALSE)
+    for (i in seq_along(combs)) {
+	comb <- combs[[i]]
+	s1 <- comb[1]
+	s2 <- comb[2]
+	bc1 <- as.character(barcodes[samples==s1])
+	bc2 <- as.character(barcodes[samples==s2])
+	x <- length(intersect(bc1,bc2))
+	m <- length(bc1)
+	n <- 750000
+	k <- length(bc2)
+	p.val <- phyper(q=x-1,m=m,n=n,k=k,lower.tail=FALSE)
+	tmp <- data.frame(s1=s1,
+			  s2=s2,
+			  n1=m,
+			  n2=k,
+			  shared=x,
+			  p.val=p.val)
+	out <- rbind(out,tmp)
+    }
+    return(out)
+}
 
-kable(t(as.matrix(table(pD$Condition,pD$PassGenesDet))),caption="Cells passing Gene Detection QC")
-kable(t(as.matrix(table(pD$Condition,pD$PassLibSize))),caption="Cells passing Library Size QC")
-kable(t(as.matrix(table(pD$Condition,pD$PassViability))),caption="Cells passing Viability QC")
-kable(t(as.matrix(table(pD$Condition,pD$PassAll))),caption="Cells passing all QC requirements")
+# P2
+compDf <- compare(pD$bcs, pD$SampleID)
+p2 <- ggplot(compDf, aes(x=shared, y=-log10(p.val))) +
+    geom_point() +
+    xlab("# shared Barcodes") +
+    ylab("-log10(P)") +
+    geom_hline(yintercept=2,lty="dashed",color="red") 
+p1
+p2
+```
+
 #Apply Thresholds
-pD.filtered <- filter(pD, PassAll)
-m.filtered <- m[,as.character(pD.filtered$barcode)]
+
+```{r}
+pD <- filter(pD, PassAll)
+m <- m[,as.character(pD$barcode)]
 
@@
 expThreshold <- 50*isexpThreshold
-keep1 <- rowSums(m.filtered!=0) > isexpThreshold
-keep2 <- rowSums(m.filtered) > expThreshold
+keep1 <- rowSums(m!=0) > isexpThreshold
+keep2 <- rowSums(m) > expThreshold
 fD$keep <- keep1 & keep2
-m.filtered <- m.filtered[fD$keep,]
-fD.filtered <- fD[fD$keep,]
+m <- m[fD$keep,]
+fD <- fD[fD$keep,]
 #Final matrix
-dim(m.filtered)
-
-```
-
-# A more or less fair representation of the lactation samples
-```{r}
-pD.lac <- filter(pD, Condition=="L")
-cells <- as.character(pD.lac$barcode)
-m.lac <- m[,cells]
-keep <- rowMeans(m.lac) > 0.1
-m.lac <- m.lac[keep,]
-fD.lac <- fD[keep,]
-#Normalization
-clusters.lac <- quickCluster(m.lac)
-pD.lac$sf <- computeSumFactors(m.lac,clusters=clusters.lac)
-m.lac.norm <- t(t(m.lac)/pD.lac$sf)
-
-fD.lac <- mutate(fD.lac, meanExpression=rowMeans(m.lac),
-	     CV2=apply(m.lac,1,cv2),
-	     dm=DM(meanExpression,CV2))
-
-highVar.lac <- dplyr::arrange(fD.lac,desc(dm)) %>% .$id
-topX <- floor(0.2*length(highVar.lac))
-fD.lac$highVar <- fD.lac$id %in% highVar.lac[1:topX]
-
-fPCA.lac <- log2(t(m.lac.norm[fD.lac$highVar,])+1)
-fPCA.lac <- scale(fPCA.lac,scale=TRUE,center=TRUE)
-set.seed(rnd_seed)
-tsn.lac <- Rtsne(fPCA.lac,perplexity=25)
-
-pD.lac$tSNE1 <- tsn.lac$Y[,1]
-pD.lac$tSNE2 <- tsn.lac$Y[,2]
-
-pD.lac <- arrange(pD.lac, PassAll)
-
-tsnPlot <- ggplot(pD.lac, aes(x=tSNE1, y=tSNE2, color=PassAll, shape=Replicate)) +
-    geom_point(alpha=0.8,size=3) +
-    scale_color_manual(values=c("grey50","black")) +
-    theme_bw() 
-tsnPlot
-
-kable(table(pD.lac$PassAll,pD.lac$Replicate))
+dim(m)
 ```
@@
 ```{r}
-clusters <- quickCluster(m.filtered)
-pD.filtered$sf <- computeSumFactors(m.filtered,clusters=clusters)
-plot(log10(colSums(m.filtered))~log10(pD.filtered$sf),main="Library Size versus Size Factors")
-m.norm <- t(t(m.filtered)/pD.filtered$sf)
+clusters <- quickCluster(m,method="igraph")
+pD$sf <- computeSumFactors(m,clusters=clusters)
+plot(log10(colSums(m))~log10(pD$sf),main="Library Size versus Size Factors")
+m <- t(t(m)/pD$sf)
 # 
-fD.filtered <- mutate(fD.filtered, meanExpression=rowMeans(m.norm),
-	     CV2=apply(m.norm,1,cv2),
-	     dm=DM(meanExpression,CV2))
-
 #HVG definition by Brennecke
 brennecke <- BrenneckeHVG(m.norm,fdr=0.1,minBiolDisp=0.25)
-fD.filtered$highVar <- fD.filtered$id %in% brennecke
+fD$highVar <- fD$id %in% brennecke
 
-VarPlot <- ggplot(fD.filtered, aes(x=meanExpression,y=CV2, color=highVar)) +
+VarPlot <- ggplot(fD, aes(x=meanExpression,y=CV2, color=highVar)) +
     geom_point() +
@@
 ```
-
-***
-## Save Data
-```{r}
-pD.add <- select(pD.filtered, barcode, sf, tSNE1, tSNE2)
-fD.add <- select(fD.filtered, id, highVar)
-pD <- left_join(pD,pD.add, by="barcode")
-fD <- left_join(fD,fD.add, by="id")
-stopifnot(identical(as.character(pD$barcode),colnames(m)))
-stopifnot(identical(as.character(fD$id),rownames(m)))
-out <- list()
-out[[1]] <- m
-out[[2]] <- pD
-out[[3]] <- fD
-saveRDS(out,file="../data/Robjects/ExpressionList_QC.rds")
-```

### 2017-06-23 src/QCAnalysis.Rmd
first QC

@@
     xlab("Sample")
-
 smPlot
@@
 
-## Thresholds
-
-```{r}
-smryByGroup <- group_by(pD,Condition) %>%
-    summarize(medianPrcntMito=median(prcntMito),
-	      madPrcntMito=mad(prcntMito),
-	      threshold_PrcntMito=0.05,
-	      mGenesDetected=median(log10(GenesDetected)),
-	      madGenesDetected=mad(log10(GenesDetected)),
-	      threshold_GenesDetected=mGenesDetected-2*madGenesDetected,
-	      mUmiSums=median(log10(UmiSums)),
-	      madUmiSums=mad(log10(UmiSums)),
-	      threshold_UmiSums=mUmiSums-2*madUmiSums) %>%
-    select(Condition,starts_with("threshold")) 
-kable(smryByGroup)
-
-thrGenesDetected <- median(log10(pD$GenesDetected))-3*mad(log10(pD$GenesDetected))
-thrUmiSums <- median(log10(pD$UmiSums))-3*mad(log10(pD$UmiSums))
-```
-
-# Per Condition
-```{r}
-MitoCutOff <- 0.05
-gdHist <- ggplot(pD, aes(x=GenesDetected,y=..density..)) +
-    geom_histogram(fill="white",color="black") +
-    geom_vline(data=smryByGroup,aes(xintercept=10^threshold_GenesDetected),color="red",lty="longdash") +
-    scale_x_log10() +
-    xlab("Total number of genes detected") +
-    facet_wrap(~Condition)
-
-libSizeHist <- ggplot(pD, aes(x=UmiSums,y=..density..)) +
-    geom_histogram(fill="white",color="black") +
-    geom_vline(data=smryByGroup,aes(xintercept=10^threshold_UmiSums),color="red",lty="longdash") +
-    scale_x_log10() +
-    facet_wrap(~Condition) +
-    xlab("Total number of unique molecules")
-
-cellViability <- cellViability %+% pD
-cellViability <- cellViability + 
-    annotate("rect",ymin=-Inf, ymax=Inf, xmax=Inf, xmin=MitoCutOff,
-	     fill="grey", alpha=0.3)
-
-gdHist
-libSizeHist
-cellViability
-```
-
-# In total
-
-```{r}
-gdHist <- ggplot(pD, aes(x=GenesDetected,y=..density..)) +
-    geom_histogram(fill="white",color="black") +
-    geom_vline(aes(xintercept=10^thrGenesDetected),color="red",lty="longdash") +
-    scale_x_log10() +
-    xlab("Total number of genes detected")
-
-libSizeHist <- ggplot(pD, aes(x=UmiSums,y=..density..)) +
-    geom_histogram(fill="white",color="black") +
-    geom_vline(aes(xintercept=10^thrUmiSums),color="red",lty="longdash") +
-    scale_x_log10() +
-    xlab("Total number of unique molecules")
-
-gdHist
-libSizeHist
-pD2 <- mutate(pD,
-	     PassViability=prcntMito < MitoCutOff,
-	     PassGenesDet=GenesDetected > thrGenesDetected ,
-	     PassLibSize=UmiSums > thrUmiSums,
-	     PassAll= PassViability & PassGenesDet & PassLibSize)
-table(pD2$PassViability,pD2$SampleID)
-table(pD2$PassGenesDet,pD2$SampleID)
-table(pD2$PassLibSize,pD2$SampleID)
-table(pD2$PassAll,pD2$SampleID)
-```
-
-```{r}
-#Thresholding per group
-#inititae Df
-pD <- mutate(pD,
-	     ThresholdViability = 0,
-	     ThresholdGenesDet = 0,
-	     ThresholdLibSize = 0)
-
-grps <- as.character(unique(pD$Condition))
-for (grp in grps) {
-    thrs <- filter(smryByGroup, Condition==grp) %>% select(-Condition) %>% t() %>% as.vector()
-    names(thrs) <- filter(smryByGroup, Condition==grp) %>% select(-Condition) %>% t() %>% rownames()
-    pD <- mutate(pD,
-		 ThresholdViability= ifelse(Condition==grp, MitoCutOff, ThresholdViability),
-		 ThresholdGenesDet= ifelse(Condition==grp, 10^thrs["threshold_GenesDetected"],ThresholdGenesDet),
-		 ThresholdLibSize= ifelse(Condition==grp, 10^thrs["threshold_UmiSums"],ThresholdLibSize))
-}
-
-pD <- mutate(pD,
-	     PassViability=prcntMito < ThresholdViability,
-	     PassGenesDet=GenesDetected > ThresholdGenesDet,
-	     PassLibSize=UmiSums > ThresholdLibSize,
-	     PassAll= PassViability & PassGenesDet & PassLibSize)
-table(pD$PassViability,pD$SampleID)
-table(pD$PassGenesDet,pD$SampleID)
-table(pD$PassLibSize,pD$SampleID)
-table(pD$PassAll,pD$SampleID)
-```
-
 ## Index Swapping
@@
 
-#Apply Thresholds
+## Thresholds
 
 ```{r}
-pD <- filter(pD, PassAll)
-m <- m[,as.character(pD$barcode)]
-
-#Thresholding on lowly expressed genes
-isexpThreshold <- 10
-expThreshold <- 50*isexpThreshold
-keep1 <- rowSums(m!=0) > isexpThreshold
-keep2 <- rowSums(m) > expThreshold
-fD$keep <- keep1 & keep2
-m <- m[fD$keep,]
-fD <- fD[fD$keep,]
-#Final matrix
-dim(m)
+smryByGroup <- group_by(pD,Condition) %>%
+    summarize(threshold_PrcntMito=0.05,
+	      mGenesDetected=median(log10(GenesDetected)),
+	      madGenesDetected=mad(log10(GenesDetected)),
+	      threshold_GenesDetected=mGenesDetected-3*madGenesDetected,
+	      mUmiSums=median(log10(UmiSums)),
+	      madUmiSums=mad(log10(UmiSums)),
+	      threshold_UmiSums=mUmiSums-3*madUmiSums) %>%
+    select(Condition,starts_with("threshold")) 
+kable(smryByGroup)
 ```
 
-# Normalization and Highly Variable Gene Selection
+# Per Condition
 ```{r}
-clusters <- quickCluster(m,method="igraph")
-pD$sf <- computeSumFactors(m,clusters=clusters)
-plot(log10(colSums(m))~log10(pD$sf),main="Library Size versus Size Factors")
-m <- t(t(m)/pD$sf)
-# 
-#HVG definition by Brennecke
-brennecke <- BrenneckeHVG(m.norm,fdr=0.1,minBiolDisp=0.25)
-fD$highVar <- fD$id %in% brennecke
+MitoCutOff <- 0.05
+gdHist <- ggplot(pD, aes(x=GenesDetected,y=..density..)) +
+    geom_histogram(fill="white",color="black") +
+    geom_vline(data=smryByGroup,aes(xintercept=10^threshold_GenesDetected),color="red",lty="longdash") +
+    scale_x_log10() +
+    xlab("Total number of genes detected") +
+    facet_wrap(~Condition)
 
-VarPlot <- ggplot(fD, aes(x=meanExpression,y=CV2, color=highVar)) +
-    geom_point() +
+libSizeHist <- ggplot(pD, aes(x=UmiSums,y=..density..)) +
+    geom_histogram(fill="white",color="black") +
+    geom_vline(data=smryByGroup,aes(xintercept=10^threshold_UmiSums),color="red",lty="longdash") +
     scale_x_log10() +
-    scale_y_log10() +
-    theme_bw()
-VarPlot
+    facet_wrap(~Condition) +
+    xlab("Total number of unique molecules")
+
+cellViability <- cellViability %+% pD
+cellViability <- cellViability + 
+    annotate("rect",ymin=-Inf, ymax=Inf, xmax=Inf, xmin=MitoCutOff,
+	     fill="grey", alpha=0.3)
+
+gdHist
+libSizeHist
+cellViability
 ```
 
-# First Visualization
 ```{r}
-fPCA <- log2(t(m.norm[fD.filtered$highVar,])+1)
-fPCA <- scale(fPCA,scale=TRUE,center=TRUE)
-pc <- prcomp(fPCA)
-set.seed(rnd_seed)
-tsn <- Rtsne(fPCA,perplexity=25)
+#Thresholding per group
+#inititae Df
+pD <- mutate(pD,
+	     ThresholdViability = 0,
+	     ThresholdGenesDet = 0,
+	     ThresholdLibSize = 0)
 
-pD.filtered$PC1 <- pc$x[,1]
-pD.filtered$PC2 <- pc$x[,2]
-pD.filtered$PC3 <- pc$x[,3]
-pD.filtered$PC4 <- pc$x[,4]
-pD.filtered$tSNE1 <- tsn$Y[,1]
-pD.filtered$tSNE2 <- tsn$Y[,2]
+grps <- as.character(unique(pD$Condition))
+for (grp in grps) {
+    thrs <- filter(smryByGroup, Condition==grp) %>% select(-Condition) %>% t() %>% as.vector()
+    names(thrs) <- filter(smryByGroup, Condition==grp) %>% select(-Condition) %>% t() %>% rownames()
+    pD <- mutate(pD,
+		 ThresholdViability= ifelse(Condition==grp, MitoCutOff, ThresholdViability),
+		 ThresholdGenesDet= ifelse(Condition==grp, 10^thrs["threshold_GenesDetected"],ThresholdGenesDet),
+		 ThresholdLibSize= ifelse(Condition==grp, 10^thrs["threshold_UmiSums"],ThresholdLibSize))
+}
 
-Acta2 <- filter(fD.filtered, symbol=="Acta2") %>% .$id
-pD.filtered$Acta2 <- t(m.norm)[,Acta2]
-Csn2 <- filter(fD.filtered, symbol=="Csn2") %>% .$id
-pD.filtered$Csn2 <- t(m.norm)[,Csn2]
-# 
-pcaPlot <- ggplot(pD.filtered, aes(x=PC1, y=PC2, color=Condition, shape=Replicate)) +
-    geom_point(alpha=0.8,size=2) +
-    theme_bw() 
-pcaPlot
+pD <- mutate(pD,
+	     PassViability=prcntMito < ThresholdViability,
+	     PassGenesDet=GenesDetected > ThresholdGenesDet,
+	     PassLibSize=UmiSums > ThresholdLibSize,
+	     PassAll= PassViability & PassGenesDet & PassLibSize & bc.obs.Freq == 1)
+table(pD$PassViability,pD$SampleID)
+table(pD$PassGenesDet,pD$SampleID)
+table(pD$PassLibSize,pD$SampleID)
+table(pD$PassAll,pD$SampleID)
+```
 
-pcaPlot <- ggplot(pD.filtered, aes(x=PC1, y=PC3, color=log2(Csn2+1), shape=Replicate)) +
-    geom_point(alpha=0.8,size=2) +
-    theme_bw() 
-pcaPlot
+# Save Data
 
-pcaPlot <- ggplot(pD.filtered, aes(x=PC1, y=PC2, color=log2(Acta2+1), shape=Replicate)) +
-    geom_point(alpha=0.8,size=2) +
-    theme_bw() 
-pcaPlot
-
-tsnePlot <- ggplot(pD.filtered, aes(x=tSNE1, y=tSNE2, color=Condition, shape=Replicate)) +
-    geom_point(alpha=0.8,size=2) +
-    theme_bw()
-tsnePlot
-
-tsnePlot <- ggplot(pD.filtered, aes(x=tSNE1, y=tSNE2, color=log2(Acta2+1), shape=Replicate)) +
-    geom_point(alpha=0.8,size=2) +
-    theme_bw()
-tsnePlot
-
-tsnePlot <- ggplot(pD.filtered, aes(x=tSNE1, y=tSNE2, color=log2(Csn2+1), shape=Replicate)) +
-    geom_point(alpha=0.8,size=2) +
-    theme_bw()
-tsnePlot
-
-#tsnePlot based on cell Cycle
-library(scLVM)
-ens_ids_cc <- getEnsembl('GO:0007049')
-sel <- rownames(m.norm) %in% ens_ids_cc
-fPCA <- log2(t(m.norm[sel,])+1)
-fPCA <- scale(fPCA,scale=TRUE,center=TRUE)
-set.seed(rnd_seed)
-tsn <- Rtsne(fPCA,perplexity=25)
-
-pD.filtered$cc_tSNE1 <- tsn$Y[,1]
-pD.filtered$cc_tSNE2 <- tsn$Y[,2]
-
-tsnePlot <- ggplot(pD.filtered, aes(x=cc_tSNE1, y=cc_tSNE2, color=Condition, shape=Replicate)) +
-    geom_point(alpha=0.8,size=2) +
-    theme_bw()
-tsnePlot
+```{r}
+stopifnot(identical(as.character(pD$barcode),colnames(m)))
+stopifnot(identical(as.character(fD$id),rownames(m)))
+out <- list()
+out[[1]] <- m
+out[[2]] <- pD
+out[[3]] <- fD
+saveRDS(out,file="../data/Robjects/secondRun_2500/ExpressionList_QC.rds")
 ```

### 2017-06-23 src/QCAnalysis.Rmd
added fdkeep

@@
 table(pD$PassAll,pD$SampleID)
+
+### FD
+fD$keep <- rowSums(m) > 0.01
 ```

### 2017-06-26 src/QCAnalysis.Rmd
bugfix
minor changes

--- 2017-06-23 src/QCAnalysis.Rmd
@@
 	     PassLibSize=UmiSums > ThresholdLibSize,
-	     PassAll= PassViability & PassGenesDet & PassLibSize & bc.obs.Freq == 1)
+	     PassAll= PassViability & PassGenesDet & PassLibSize)
 table(pD$PassViability,pD$SampleID)
@@
 table(pD$PassAll,pD$SampleID)
+
+### FD
+fD$keep <- rowMeans(m) > 0.01
 ```

### 2017-06-26 src/HVGandNorm.R

@@ -1,6 +1,8 @@
 # Estimate size factors using scran
 library(dplyr)
 library(scran)
+library(Rtsne)
+source("functions.R")
 
 dataList <- readRDS("../data/Robjects/secondRun_2500/ExpressionList_QC.rds")
 m <- dataList[[1]]
@@ -13,8 +15,11 @@ m <- m[fD$keep,pD$PassAll]
 pD <- pD[pD$PassAll,]
 fD <- fD[fD$keep,]
 
-clusters <- quickCluster(m,method="igraph")
+test <- dist(m)
+clusters <- quickCluster(m,method="hclust")
 pD$sf <- computeSumFactors(m,clusters=clusters)
+
+# pD$sf <- colSums(m)
 
 # Normalize count matrix
 m <- t(t(m)/pD$sf)
@@ -32,8 +37,12 @@ tsn <- Rtsne(fPCA,perplexity=50)
 pD$tSNE1 <- tsn$Y[,1]
 pD$tSNE2 <- tsn$Y[,2]
 
+pcs <- prcomp(fPCA)
+pD$PC1 <- pcs$x[,1]
+pD$PC2 <- pcs$x[,2]
+
 # save
-pD.add <- pD[,c("barcode","sf","tSNE1","tSNE2")]
+pD.add <- pD[,c("barcode","sf","tSNE1","tSNE2","PC1","PC2")]
 fD.add <- fD[,c("id","highVar")]
 
 

### 2017-06-26 src/functions.R

@@ -169,3 +169,15 @@ compClustering <- function(m,fD,fs="brennecke",dm="spearman",lk="ward.D2",ds=0,m
 		      "m"=length(cluss))
 }
 
+subSample <- function(dataList, cell.filter=NULL, cell.number=5000, group=NULL) {
+    require(dplyr)
+    pD <- dataList[["phenoData"]]
+    if(!is.null(cell.filter)) pD <- pD[cell.filter,]
+    if(!is.null(group)) pD <- group_by_(pD,group)
+    cells <- sample_n(pD,size=cell.number, replace=FALSE) %>% .$barcode
+    out <- list()
+    out[["phenoData"]] <- pD[pD$barcode %in% cells,]
+    out[["featureData"]] <- dataList[["featureData"]]
+    out[["counts"]] <- dataList[["counts"]][,pD$barcode]
+}
+

### 2017-06-26 src/QCAnalysis.Rmd
bugfix

@@
 out <- list()
-out[[1]] <- m
-out[[2]] <- pD
-out[[3]] <- fD
+out[["counts"]] <- m
+out[["phenoData"]] <- pD
+out[["featureData"]] <- fD
 saveRDS(out,file="../data/Robjects/secondRun_2500/ExpressionList_QC.rds")

### 2017-06-27 src/QCAnalysis.Rmd
minor updates

@@
 	     PassLibSize=UmiSums > ThresholdLibSize,
-	     PassAll= PassViability & PassGenesDet & PassLibSize)
+	     PassAll= PassViability & PassGenesDet & PassLibSize & bc.obs.Freq==1)
 table(pD$PassViability,pD$SampleID)

### 2017-06-27 src/ComputeClusterComparison.R

@@ -1,18 +1,16 @@
 library(scran)
 library(dplyr)
-library(knitr)
-library(ggplot2)
 library(dynamicTreeCut)
-library(Rtsne)
-library(pheatmap)
 source("functions.R")
 
 rnd_seed <- 300
-dataList <- readRDS("../data/Robjects/secondRun_2500/ExpressionList_QC.rds")
+dataList <- readRDS("../data/Robjects/secondRun_2500/ExpressionList_QC_norm.rds")
 
-m <- dataList[[1]]
-pD <- dataList[[2]]
-fD <- dataList[[3]]
+m <- dataList[["counts"]]
+pD <- dataList[["phenoData"]]
+fD <- dataList[["featureData"]]
+rm(dataList)
+
 #Gene and Cell filtering
 m <- m[fD$keep,pD$PassAll]
 pD <- pD[pD$PassAll,]
@@ -22,8 +20,11 @@ fD <- fD[fD$keep,]
 m <- t(t(m)/pD$sf)
 # 
 
-#Run various clustering combinations
+#trafM
+m <- t(log2(m[fD$highVar,]+1))
 
+#Run various clustering combinations
+gc()
 fss <- c("highVar")
 dms <- c("pearson","spearman","euclidean")
 lks <- c("complete","average","ward.D2")
@@ -31,7 +32,7 @@ dss <- c(0,1,2,3,4)
 require(doParallel)
 require(fpc)
 require(clValid)
-nCores <- 3
+nCores <- 6
 cl <-makeCluster(nCores, type="FORK")
 registerDoParallel(cl)
 result <- foreach (i=seq_along(dms), .combine=rbind) %dopar% {
@@ -40,7 +41,7 @@ result <- foreach (i=seq_along(dms), .combine=rbind) %dopar% {
     for (fs in fss) {
 	for (lk in lks) {
 	    for (ds in dss) {
-		res <- compClustering(m,fD,fs=fs,dm=dm,lk=lk,ds=ds)
+		res <- compClustering(trafM=m,fs=fs,dm=dm,lk=lk,ds=ds)
 		tmp0 <- rbind(tmp0,res)
 	    }}}
     return(tmp0)}

### 2017-06-27 src/DiffusionMap.R

@@ -5,19 +5,20 @@ source("functions.R")
 
 # Load Data
 rnd_seed <- 300
-dataList <- readRDS("../data/Robjects/secondRun_2500/ExpressionList_Clustered.rds")
+dataList <- readRDS("../data/Robjects/secondRun_2500/ExpressionList_QC_norm.rds")
 m <- dataList[[1]]
 pD <- dataList[[2]]
+rm(dataList)
 
 # Remove QC-fails,outlier and immune cells
-keepCells <- pD$PassAll & !pD$isImmuneCell & !pD$isOutlier 
+keepCells <- pD$PassAll # & !pD$isImmuneCell & !pD$isOutlier 
 m <- m[,keepCells]
 pD <- pD[keepCells,]
 
 # ---- BasalAndLuminalNPandG ----
 
 condComb <- c("NP","G")
-cells <- list("all","luminal")
+cells <- list("all")
 
 for (cell in cells) {
     if(cell=="luminal") {
@@ -31,7 +32,7 @@ for (cell in cells) {
     pD.vp <- pD[keepCells,]
 
     # Remove genes below 0.1 mean
-    keep <- rowMeans(m.vp)>0.1
+    keep <- rowMeans(m.vp)>0.01
     m.vp <- m.vp[keep,]
 
     # Normalize 

### 2017-06-27 src/HVGandNorm.R

@@ -30,7 +30,7 @@ m <- t(t(m)/pD$sf)
 
 
 # Highly variable genes 
-varDf <- technicalCV2(t(t(m)/(1/pD$sf)), is.spike=NA, sf.cell=pD$sf,sf.spike=pD$sf)
+varDf <- technicalCV2(t(t(m)/(1/pD$sf)), is.spike=NA, sf.cell=pD$sf, sf.spike=pD$sf)
 plot(varDf$mean, varDf$cv2, log="xy",pch=19)
 points(varDf$mean, varDf$trend, col="red", pch=16, cex=0.5)
 points(varDf[varDf$FDR < 0.1,"mean"], varDf[varDf$FDR < 0.1, "cv2"], col="blue", pch=16, cex=0.5)
@@ -56,4 +56,4 @@ fD.add <- fD[,c("id","highVar")]
 dataList <- readRDS("../data/Robjects/secondRun_2500/ExpressionList_QC.rds")
 dataList[["phenoData"]] <- left_join(dataList[["phenoData"]],pD.add)
 dataList[["featureData"]] <- left_join(dataList[["featureData"]],fD.add)
-saveRDS(dataList,file="../data/Robjects/secondRun_2500/test.rds")
+saveRDS(dataList,file="../data/Robjects/secondRun_2500/ExpressionList_QC_norm.rds")

### 2017-06-27 src/functions.R

@@ -132,11 +132,7 @@ grab_grob <- function(){
   grid.echo()
   grid.grab()
 }
-compClustering <- function(m,fD,fs="brennecke",dm="spearman",lk="ward.D2",ds=0,minSize=15) {
-
-    #Feature Selection
-    keep <- fD[,fs]
-    trafM <- t(log2(m[keep,]+1))
+compClustering <- function(trafM,fs="brennecke",dm="spearman",lk="ward.D2",ds=0,minSize=15) {
 
     #Dissimilarity Measure
     if(dm=="euclidean") {

### 2017-06-28 src/ClusterComparison.Rmd
updates on clust comparison

@@
 library(ggplot2)
-library(dynamicTreeCut)
-library(Rtsne)
+library(reshape2)
 library(pheatmap)
@@
 
-rnd_seed <- 300
-clustDat <- readRDS("../data/Robjects/ClusterComparison.rds") 
-bootDat <- readRDS("../data/Robjects/ClusterBootstrap.rds")
-dataList <- readRDS("../data/Robjects/ExpressionList_QC.rds")
+clustDat <- readRDS("../data/Robjects/secondRun_2500/ClusterComparison.rds") 
+# bootDat <- readRDS("../data/Robjects/ClusterBootstrap.rds")
+dataList <- readRDS("../data/Robjects/secondRun_2500/ExpressionList_QC_norm.rds")
+set.seed(300)
+dataList <- subSample(dataList,cell.number=1000)
 
 #Commence
-m <- dataList[[1]]
-pD <- dataList[[2]]
-fD <- dataList[[3]]
+m <- dataList[["counts"]]
+pD <- dataList[["phenoData"]]
+fD <- dataList[["featureData"]]
+rm(dataList)
+
 #Gene and Cell filtering
@@
 m <- t(t(m)/pD$sf)
-#Identification of HVG 
-# fD <- mutate(fD, meanExpression=rowMeans(m),
-#              CV2= apply(m,1,cv2))
-# 
-# genes2lbl <- c("Acta2", "Krt18", "Csn2", "Wap","Krt14")
-# lblDat <- filter(fD, symbol %in% genes2lbl)
-# V3 <- ggplot(arrange(fD,highVar), aes(x=meanExpression,y=CV2, color=highVar)) +
-#     geom_label(data=lblDat, aes(x=meanExpression, y=CV2, label=symbol), nudge_y=0.1) +
-#     geom_point(data=lblDat, aes(x=meanExpression, y=CV2), pch=1, size=4) +
-#     geom_point(alpha=0.8) +
-#     scale_x_log10() +
-#     scale_y_log10() +
-#     theme_bw()
-# V3
 
-#Average Linkage appears to best capture the dendrogram
+clustStats <- unlist(clustDat, recursive=FALSE)
+clustStats <- clustStats[grepl("Statistics",names(clustStats))]
+clustStats <- do.call(rbind,clustStats)
 
-clustDat <- filter(clustDat, Statistic=="Average Silhouette Width")
-compPlot <- ggplot(clustDat, aes(x=as.character(DeepSplit), y=Value, color=Linkage,
-			    group=Linkage)) +
+clustStats <- filter(clustStats, Statistic=="Average Silhouette Width")
+compPlot <- ggplot(clustStats, aes(x=as.character(DeepSplit), y=Value, color=Linkage, group=Linkage)) +
     geom_path() +
@@
     facet_grid(.~Dissimilarity, scales="free_y") +
-ylab("Average Silhouette Width") +
+    ylab("Average Silhouette Width")
+
 compPlot
-
-#Which Linkage is best
-m.sub <- m[fD$highVar,]
-# trafM <- log2(m.sub+1)
-# dms <- c("pearson","spearman","euclidean")
-# res <- NULL
-# for (dm in dms) {
-# Dissimilarity Measure
-#     if(dm!="euclidean") {
-#     dis <- as.dist((1-cor(trafM,method=dm)))
-#     } else {
-#         dis <- dist(t(trafM),method=dm)
-#     }
-# 
-# Hierarchical Clustering
-#     htree.average <- hclust(dis, method="average")
-#     htree.single <- hclust(dis, method="single")
-#     htree.complete <- hclust(dis, method="complete")
-#     htree.ward <- hclust(dis, method="ward.D2")
-# 
-#     ccc.average <- cor(cophenetic(htree.average),dis)
-#     ccc.single <- cor(cophenetic(htree.single),dis)
-#     ccc.complete <- cor(cophenetic(htree.complete),dis)
-#     ccc.ward <- cor(cophenetic(htree.ward),dis)
-# 
-#     tmp <- data.frame("Linkage"=c("Average",
-#                                          "Single",
-#                                          "Complete",
-#                                          "Ward.D2"),
-#                              "Cophenetic Correlation Coefficient"=
-#                                  c(ccc.average,ccc.single,
-#                                    ccc.complete,ccc.ward),
-#                              "Dissimilarity"=dm)
-#     res <- rbind(res,tmp)
-# }
-# kable(res)
 
@@
 	   selectInput("lk", label = "Linkage",
-		       choices=c("average","complete","single","ward.D2"), selected="average")
+		       choices=c("average"), selected="average")
 	  )
@@
 clustering <- reactive({
-    dis <- asDis(m.sub,dm=input$dm)
-    dynamicCluster(dis, lk=input$lk,
-			ds=as.numeric(input$ds),
-			output="Cluster",
-			minSize=15)
+    lbl <- paste(input$dm,input$lk,input$ds,sep="_")
+    cluss <- clustDat[[lbl]][["Labels"]]
 })
@@
     clust <- clustering()
-    x <- table(clust$cluster,pD$Condition)
+    x <- table(clust,pD$Condition)
     out <- data.frame("Cluster"=rownames(x),
@@
     clust <- clustering()
-    nclust <- length(unique(clust$cluster))
-    palette(rainbow(nclust))
-    htreeWithBars(clust,pD)
-}, width=2000, height=900)
-
-renderPlot({
-    clust <- clustering()
-    nclust <- length(unique(clust$cluster))
-    pD$cluster <- factor(clust$cluster)
+    nclust <- length(unique(clust))
+    pD$cluster <- factor(clust)
     tsnPlot <- ggplot(pD, aes(x=tSNE1, y=tSNE2, color=cluster, shape=Replicate)) +
@@
 
-renderPlot({
-    clust <- clustering()
-    nclust <- length(unique(clust$cluster))
-    palette(rainbow(nclust))
-    slct <- paste(input$dm,input$lk,input$ds,sep="_")
-    boot <- bootDat[[slct]]$bootresult
-    boxplot(t(boot),ylim=c(0,1),col=c(1:nclust))
+# renderPlot({
+    #     clust <- clustering()
+    #     nclust <- length(unique(clust))
+    #     palette(rainbow(nclust))
+    #     slct <- paste(input$dm,input$lk,input$ds,sep="_")
+    #     boot <- bootDat[[slct]]$bootresult
+    #     boxplot(t(boot),ylim=c(0,1),col=c(1:nclust))
+    # 
+# }, width=800, height=800)
 
-}, width=800, height=800)
+# renderPlot({
+#     library(cluster)
+#     clust <- clustering()
+#     nclust <- length(unique(clust$cluster))
+#     palette(rainbow(nclust))
+#     sils  <- silhouette(x=clust$cluster,dist=clust$dis)
+#     plot(sils,col=c(1:length(unique(clust$cluster))), border=NA)
+# 
+# }, width=1000, height=1800)
+
+# renderPlot({
+#     clust <- clustering()
+#     pD$cluster <- clust
+#     nclust <- length(unique(clust))
+#     ord <- arrange(pD, cluster) 
+#     simMat <- asSim(clust$dis)
+#     simMat.ord <- as.matrix(simMat)[as.character(ord$barcode),
+#                                     as.character(ord$barcode)]
+#     library(pheatmap)
+#     annoCol <- data.frame("cluster"=as.factor(ord$cluster),
+#                           "Condition"=ord$Condition)
+#     rownames(annoCol) <- as.character(ord$barcode)
+#     ann_colors <- list("cluster"=rainbow(nclust))
+#     names(ann_colors$cluster) <- c(min(ord$cluster):max(ord$cluster))
+#     pheatmap(simMat.ord,
+#              cluster_rows=FALSE,
+#              cluster_cols=FALSE,
+#              show_rownames=FALSE,
+#              show_colnames=FALSE,
+#              annotation_col=annoCol,
+#              annotation_colors=ann_colors)
+# }, width=1200, height=1200)
 
 renderPlot({
-    library(cluster)
-    clust <- clustering()
-    nclust <- length(unique(clust$cluster))
-    palette(rainbow(nclust))
-    sils  <- silhouette(x=clust$cluster,dist=clust$dis)
-    plot(sils,col=c(1:length(unique(clust$cluster))), border=NA)
-
-}, width=1000, height=1800)
-
-renderPlot({
-    clust <- clustering()
-    pD$cluster <- clust$cluster
-    nclust <- length(unique(clust$cluster))
-    ord <- arrange(pD, cluster) 
-    simMat <- asSim(clust$dis)
-    simMat.ord <- as.matrix(simMat)[as.character(ord$barcode),
-				    as.character(ord$barcode)]
-    library(pheatmap)
-    annoCol <- data.frame("cluster"=as.factor(ord$cluster),
-			  "Condition"=ord$Condition)
-    rownames(annoCol) <- as.character(ord$barcode)
-    ann_colors <- list("cluster"=rainbow(nclust))
-    names(ann_colors$cluster) <- c(min(ord$cluster):max(ord$cluster))
-    pheatmap(simMat.ord,
-	     cluster_rows=FALSE,
-	     cluster_cols=FALSE,
-	     show_rownames=FALSE,
-	     show_colnames=FALSE,
-	     annotation_col=annoCol,
-	     annotation_colors=ann_colors)
-}, width=1200, height=1200)
-
-renderPlot({
-    clust <- clustering()
-    pD$cluster <- clust$cluster
-
+    pD$cluster <- clustering()
     genes <- c("Acta2","Csn3","Epcam","Krt8")
@@
 renderPlot({
-    clust <- clustering()
-    pD$cluster <- clust$cluster
+    pD$cluster <- clustering()
 

### 2017-06-28 src/ComputeClusterComparison.R

@@ -3,8 +3,9 @@ library(dplyr)
 library(dynamicTreeCut)
 source("functions.R")
 
-rnd_seed <- 300
 dataList <- readRDS("../data/Robjects/secondRun_2500/ExpressionList_QC_norm.rds")
+set.seed(300)
+dataList <- subSample(dataList,cell.number=10000)
 
 m <- dataList[["counts"]]
 pD <- dataList[["phenoData"]]
@@ -25,26 +26,25 @@ m <- t(log2(m[fD$highVar,]+1))
 
 #Run various clustering combinations
 gc()
-fss <- c("highVar")
 dms <- c("pearson","spearman","euclidean")
-lks <- c("complete","average","ward.D2")
+lks <- c("average")
 dss <- c(0,1,2,3,4)
 require(doParallel)
 require(fpc)
 require(clValid)
-nCores <- 6
+nCores <- detectCores()
 cl <-makeCluster(nCores, type="FORK")
 registerDoParallel(cl)
-result <- foreach (i=seq_along(dms), .combine=rbind) %dopar% {
+result <- foreach (i=seq_along(dms)) %dopar% {
     dm <- dms[i]
-    tmp0 <- NULL
-    for (fs in fss) {
+    tmp0 <- list() 
 	for (lk in lks) {
 	    for (ds in dss) {
-		res <- compClustering(trafM=m,fs=fs,dm=dm,lk=lk,ds=ds)
-		tmp0 <- rbind(tmp0,res)
-	    }}}
+		res <- compClustering(trafM=m, dm=dm, lk=lk, ds=ds)
+		tmp0[[paste(dm,lk,ds,sep="_")]] <- res
+	    }}
     return(tmp0)}
 stopCluster(cl)
 
+result <- unlist(result,recursive=FALSE)
 saveRDS(result,file="../data/Robjects/secondRun_2500/ClusterComparison.rds")

### 2017-06-28 src/functions.R

@@ -132,7 +132,7 @@ grab_grob <- function(){
   grid.echo()
   grid.grab()
 }
-compClustering <- function(trafM,fs="brennecke",dm="spearman",lk="ward.D2",ds=0,minSize=15) {
+compClustering <- function(trafM, dm="spearman", lk="ward.D2", ds=0, minSize=15) {
 
     #Dissimilarity Measure
     if(dm=="euclidean") {
@@ -153,8 +153,7 @@ compClustering <- function(trafM,fs="brennecke",dm="spearman",lk="ward.D2",ds=0,
     con <- connectivity(distance=dis,clusters=cluss)
 
     #Compile output
-    out <- data.frame("FeatureSelection"=fs,
-		      "Dissimilarity"=dm,
+    out.df <- data.frame("Dissimilarity"=dm,
 		      "Linkage"=lk,
 		      "DeepSplit"=ds,
 		      "Statistic"=c("Average Silhouette Width",
@@ -163,6 +162,9 @@ compClustering <- function(trafM,fs="brennecke",dm="spearman",lk="ward.D2",ds=0,
 		      "Value"=c(asw,rss,con),
 		      "K"=max(cluss),
 		      "m"=length(cluss))
+    out.cls <- cluss
+    out <- list("Statistics"=out.df,
+		"Labels"=out.cls)
 }
 
 subSample <- function(dataList, cell.filter=NULL, cell.number=5000, group=NULL) {

### 2017-07-04 src/DiffusionMap.R
first update

@@ -1,17 +1,18 @@
 # Diffusion map computed on all cells from NP and G
 
 library(destiny)
+library(scran)
 source("functions.R")
 
 # Load Data
 rnd_seed <- 300
-dataList <- readRDS("../data/Robjects/secondRun_2500/ExpressionList_QC_norm.rds")
+dataList <- readRDS("../data/Robjects/secondRun_2500/ExpressionList_QC_norm_clustered4.rds")
 m <- dataList[[1]]
 pD <- dataList[[2]]
 rm(dataList)
 
 # Remove QC-fails,outlier and immune cells
-keepCells <- pD$PassAll # & !pD$isImmuneCell & !pD$isOutlier 
+keepCells <- pD$PassAll & !pD$IsNonEpithelial
 m <- m[,keepCells]
 pD <- pD[keepCells,]
 
@@ -24,30 +25,34 @@ for (cell in cells) {
     if(cell=="luminal") {
 	excludeClustComb <- c(6,7,9)
     } else{
-	excludeClustComb <- NULL
+excludeClustComb <- NULL
     }
 
-    keepCells <- pD$Condition %in% condComb & !(pD$cluster %in% excludeClustComb)
+    keepCells <- pD$Condition %in% condComb & !(pD$SubCluster %in% excludeClustComb)
     m.vp <- m[,keepCells]
     pD.vp <- pD[keepCells,]
+    pD.vp$SubCluster <- factor(pD.vp$SubCluster)
 
     # Remove genes below 0.1 mean
     keep <- rowMeans(m.vp)>0.01
     m.vp <- m.vp[keep,]
 
     # Normalize 
-    m.norm <- t(t(m.vp)/pD.vp$sf)
+    m.vp <- t(t(m.vp)/pD.vp$sf)
 
-    # Compute HVGs
-    brennecke <- BrenneckeHVG(m.norm,suppress.plot=TRUE)
+    # Highly variable genes 
+    var.des <- trendVar(log2(m.vp+1),trend="semiloess")
+    var.out <- decomposeVar(log2(m.vp+1),var.des)
+    o <- order(var.out$mean)
+    hvg.out <- var.out[which(var.out$FDR <= 0.05 & var.out$bio >=0.5),]
 
     # Prepare expression matrix
-    exps <- m.norm[brennecke,]
-    exps <- t(log(exps+1))
+    m.vp <- m.vp[rownames(hvg.out),]
[…]

### 2017-07-04 src/ClusterComparison.Rmd
updates

@@
 
+rnd_seed <- 300
 clustDat <- readRDS("../data/Robjects/secondRun_2500/ClusterComparison.rds") 
@@
 dataList <- readRDS("../data/Robjects/secondRun_2500/ExpressionList_QC_norm.rds")
-set.seed(300)
-dataList <- subSample(dataList,cell.number=1000)
+# set.seed(rnd_seed)
+# dataList <- subSample(dataList,cell.number=10000)
 
@@
 
+# fD$keep <- rowMeans(m) > 0.01 & fD$keep
+
 #Gene and Cell filtering
@@
 fD <- fD[fD$keep,]
+
+# only keep highVar
+m <- m[fD$highVar,]
+fD <- fD[fD$highVar,]
 #Normalization
@@
 
+tsnPlot1 <- ggplot(pD, aes(x=tSNE1, y=tSNE2, color=Condition, shape=Replicate)) +
+    geom_point(size=1) +
+    theme_bw()
+tsnPlot1
+
 inputPanel(
@@
     lbl <- paste(input$dm,input$lk,input$ds,sep="_")
-    cluss <- clustDat[[lbl]][["Labels"]]
+    print(lbl)
+    cluss <- unname(clustDat[[lbl]][["Labels"]])
 })
@@
 
-# renderPlot({
-#     clust <- clustering()
-#     pD$cluster <- clust
-#     nclust <- length(unique(clust))
-#     ord <- arrange(pD, cluster) 
-#     simMat <- asSim(clust$dis)
-#     simMat.ord <- as.matrix(simMat)[as.character(ord$barcode),
-#                                     as.character(ord$barcode)]
-#     library(pheatmap)
-#     annoCol <- data.frame("cluster"=as.factor(ord$cluster),
-#                           "Condition"=ord$Condition)
-#     rownames(annoCol) <- as.character(ord$barcode)
-#     ann_colors <- list("cluster"=rainbow(nclust))
-#     names(ann_colors$cluster) <- c(min(ord$cluster):max(ord$cluster))
-#     pheatmap(simMat.ord,
-#              cluster_rows=FALSE,
-#              cluster_cols=FALSE,
-#              show_rownames=FALSE,
-#              show_colnames=FALSE,
-#              annotation_col=annoCol,
-#              annotation_colors=ann_colors)
-# }, width=1200, height=1200)
+renderPlot({
+    pD$cluster <- clustering()
+    set.seed(rnd_seed)
+    subsP <- group_by(pD, cluster) %>%
+	do(sample_n(.,min(200,nrow(.)))) %>%
+	arrange(cluster)
+
+    nclust <- length(unique(pD$clust))
+
+    # Compute smaller dis matrix
+    trafM <- t(log2(m[,as.character(subsP$barcode)]+1))
+    if(input$dm=="euclidean") {
+	dis <- dist(trafM,method=input$dm)
+    } else {
+	dis <- as.dist((1-cor(t(trafM),method=input$dm))/2)
+    }
+
+    #traf to simmat
+    simMat <- asSim(as.matrix(dis))
+    simMat <- simMat[as.character(subsP$barcode), as.character(subsP$barcode)]
+
+    #prep heatmap
+    annoCol <- data.frame("cluster"=as.factor(subsP$cluster),
+			  "Condition"=subsP$Condition)
+    rownames(annoCol) <- as.character(subsP$barcode)
+    ann_colors <- list("cluster"=rainbow(nclust))
+    names(ann_colors$cluster) <- c(min(subsP$cluster):max(subsP$cluster))
+    pheatmap(simMat,
+	     cluster_rows=FALSE,
+	     cluster_cols=FALSE,
+	     show_rownames=FALSE,
+	     show_colnames=FALSE,
+	     annotation_col=annoCol,
+	     annotation_colors=ann_colors)
+}, width=1200, height=1200)
+#
+
+# ---- PCAonAvgExpression ----
+
+renderPlot({
+# Compute mean expression values
+pD$cluster <- clustering()
+nclust <- length(unique(pD$cluster))
+out <- data.frame("C0"=numeric(nrow(m)))
+for (clust in unique(pD$cluster)) {
+    expr <- rowMeans(log2(as.matrix(m[,pD$cluster==clust])+1))
+    colname <- paste0("C",clust)
+    out[,colname] <- expr
+}
+
+# Compute euclidean distances
+meandist <- dist(t(out))
+
+# PCA
+pc <- prcomp(t(out))
+
+# Plot
+palette(rainbow(nclust))
+plot(pc$x[,1],pc$x[,2],col=c(1:10),pch=20,cex=2,xlab="PC1",ylab="PC2")
+legend("topleft", legend = c(1:nclust), col = c(1:nclust), pch = 16)
+})
 
@@
     pD$cluster <- clustering()
-    genes <- c("Acta2","Csn3","Epcam","Krt8")
-	 plotGeneDist(m, pD, fD, genes, colorBy="cluster")
+    genes <- c("Des","Sox17","Mcm3","Col3a1")
+	 plotGeneDist(m, pD, fD, genes, colorBy="Condition")
    }, width=1000, height=1000)
@@
 	   selectInput("gene", label = "GeneOfInterest",
-		       choices= mixedsort(fD$symbol), selected=c("Krt18","Acta2","Csn2","Cd74")
+		       choices= mixedsort(fD$symbol), selected=c(fD$symbol[1])
 ,
@@
     pD$cluster <- clustering()
-
     genes <- input$gene

### 2017-08-01 src/S6.R
first version of Supp for F4

@@ -3,7 +3,32 @@ library(ggplot2)
 library(cowplot)
 library("topGO")
 library(org.Mm.eg.db)
+library(dplyr)
+library(edgeR)
+library(reshape2)
 source("functions.R")
+
+# ---- ImpactStrongestOnProgenitor ----
+
+out <- data.frame()
+for (clust in c("C1","C3","C5")) {
+    filen <- sprintf("../data/Robjects/secondRun_2500/%s_NPvsPI.csv",clust)
+    tmp <- read.csv(file=filen)
+    tmp$Cluster <- clust
+    out <- rbind(out,tmp)
+}
+out <- group_by(out, Cluster) %>%
+    summarize(Up=sum(FDR < 0.01 & logFC > 0),
+	      Down=sum(FDR < 0.01 & logFC < 0)) %>%
+    melt()
+
+out$Cluster <- factor(out$Cluster,levels=c("C5","C3","C1"))
+p1 <- ggplot(out, aes(x=Cluster,y=value,fill=variable)) +
+    geom_bar(stat="identity",position="dodge") +
+    coord_flip() +
+    ylab("Number of Genes DE") +
+    labs(fill="")
+
 
 # ---- DEAnalysis4vs5 ----
 output <- data.frame()
@@ -24,40 +49,22 @@ for (clust in c("C1","C3","C5")) { # C5 has to be last for rest of script
     # prepare Data for topGO
     GO.data <- new("topGOdata", description="Lib GO",ontology="BP", allGenes=alG, 
 		   annot=annFUN.org, mapping="org.Mm.eg.db",
-		   nodeSize=10, ID="symbol")
+		   nodeSize=20, ID="symbol")
     result.classic <- runTest(GO.data, statistic="fisher")
-    tmp <- GenTable(GO.data, Fisher.classic=result.classic, orderBy="topgoFisher", topNodes=50, numChar=10000)
+    tmp <- GenTable(GO.data, Fisher.classic=result.classic, orderBy="topgoFisher", topNodes=50, numChar=30)
 
-    tmp$Term <- factor(tmp$Term, levels=rev(tmp$Term))
+    tmp$Term <- factor(tmp$Term, levels=unique(rev(tmp$Term)))
     tmp$Cluster <- clust
     output <- rbind(tmp,output)
 }
 
[…]

### 2017-08-01 submissions/resubmission/response.tex
minor updates

@@
 The authors conducted analysis on biological replicates, however this is not described in the results section.
-This is an important consideration with a new technology and the variability between replicates should be appropriately quantitated and addressed in the results.
-}\\
+This is an important consideration with a new technology and the variability between replicates should be appropriately quantitated and addressed in the results.}\\
+\comment{Add scatter plot of avereage expression values in supps}
 \ldots

### 2017-08-01 src/Figure4.R

@@ -13,27 +13,6 @@ m <- dataList[[1]]
 pD <- dataList[[2]]
 fD <- dataList[[3]]
 
-# ---- ImpactStrongestOnProgenitor ----
-
-out <- data.frame()
-for (clust in c("C1","C3","C5")) {
-    filen <- sprintf("../data/Robjects/secondRun_2500/%s_NPvsPI.csv",clust)
-    tmp <- read.csv(file=filen)
-    tmp$Cluster <- clust
-    out <- rbind(out,tmp)
-}
-out <- group_by(out, Cluster) %>%
-    summarize(Up=sum(FDR < 0.01 & logFC > 0),
-	      Down=sum(FDR < 0.01 & logFC < 0)) %>%
-    melt()
-
-out$Cluster <- factor(out$Cluster,levels=c("C5","C3","C1"))
-p1 <- ggplot(out, aes(x=Cluster,y=value,fill=variable)) +
-    geom_bar(stat="identity",position="dodge") +
-    coord_flip() +
-    ylab("Number of Genes DE") +
-    labs(fill="")
-
 # ---- C5DE ----
 
 topTab <- read.csv(file="../data/Robjects/secondRun_2500/C5_NPvsPI.csv")

### 2017-08-01 src/ProgenitorDE.R

@@ -4,27 +4,25 @@ library(dplyr)
 library(edgeR)
 source("functions.R")
 
-dataList <- readRDS("../data/Robjects/secondRun_2500/ExpressionList_Clustered.rds")
+dataList <- readRDS("../data/Robjects/secondRun_2500/ExpressionList_QC_norm_clustered_clean.rds")
 m <- dataList[[1]]
 pD <- dataList[[2]]
 fD <- dataList[[3]]
 
 # Cells
-keepCells <- pD$PassAll & !(pD$isImmuneCell | pD$isOutlier)
+m <- m[,pD$keep]
+pD <- pD[pD$keep,]
 
-m <- m[,keepCells]
-pD <- pD[keepCells,]
-
-comps <- c(5,4)
+comps <- c("C5-NP","C5-PI")
 out <- list()
 
 # DE for each progenitor cluster versus all other luminal cells
 for (choice in comps) {
     # subset data
-    rmClust <- setdiff(c(5,4),choice)
-    pD.sub <- filter(pD, !(cluster %in% c(6,7,9,rmClust)))
+    rmClust <- setdiff(c("C5-NP","C5-PI"),choice)
+    pD.sub <- filter(pD, !(SubCluster %in% c("C6-G1","C6","C7","C9",rmClust)))
     m.sub <- m[,as.character(pD.sub$barcode)]
-    keep <- rowMeans(m.sub) > 0.1
+    keep <- rowMeans(m.sub) > 0.01
     m.sub <- m.sub[keep,]
     fD.sub <- fD[keep,]
     rownames(m.sub) <- fD.sub$symbol
@@ -39,17 +37,17 @@ for (choice in comps) {
 
 
     # DE
-    cluster <- factor(as.numeric(pD.sub$cluster==choice))
+    cluster <- factor(as.numeric(pD.sub$SubCluster==choice))
     de.design <- model.matrix(~cluster)
     y <- estimateDisp(y, de.design, prior.df=0,trend="none")
     fit <- glmFit(y, de.design)
     res <- glmTreat(fit,lfc=1)
     resTab <- topTags(res,n=Inf,sort.by="PValue")
     topTab <- resTab$table
-    out[[paste0("C",choice)]] <- topTab
+    out[[choice]] <- topTab
 }
-tabNulPar <- out[["C5"]][1:500,]
[…]

### 2017-08-06 paper/main.tex
first version of F4 results, QC methods and Clustering

@@
 
-\textbf{Cluster 4 represents a post involution specific progenitor cell}
-
-From the initial clustering analysis we noticed that C4 appeared only after the animals had undergone a full pregnancy and was predominantly made up of cells from PI (0 cells from NP, 0 cells from G, 83 cells from L and 339 cells from PI) (\autoref{F1}b).
-Moreover, multiple progenitor marker genes from C5 were also to be present in C4 (\autoref{F1}c, Supplementary Fig 5).
-The similarity between the two clusters also held true on transcriptome-wide level (\autoref{F4}a, Supplementary Fig 10c).
-We thus hypothesised that C4 represents the luminal progenitor population in the post-parous gland, which is supported by the observation that C5 virtually disappears after parity (Table 1).
-To further support this hypothesis, we identified genes that distinguish C5 from the rest of the luminal compartment to see if they are also characteristic for the proposed post-parity progenitor population C4.
-Indeed, we find genes that are differentially expressed between C5 and the rest of the luminal compartment to show the same trend between C4 and the luminal compartment (\autoref{F4}b).
-In a similar manner we can distinguish C4 from the rest of the PI gland in a principal component analysis (PCA) using the identified progenitor genes (Supplementary Fig 6a and b).
-Given this high degree of similarity we were not surprised that only 120 genes were differentially expressed between C4 and C5 (Supplementary Table 4).
-Interestingly, genes that were up-regulated in C4 were significantly enriched for pathways that are involved in the immune response and lactation (\autoref{F4}c, Supplementary Fig 6c).
+\textbf{Parity biases luminal progenitors towards the alveolar fate}
+Our data illustrates how the cellular composition of the gland changes during the pregnancy cycle.
+The luminal compartment shifts from giving rise to mainly hormone sensing cells to producing alveolar, milk-producing cells during pregnancy and lactation. 
+The basal compartment on the other hand differentiates to produce oxytocin-sensing, myoepithelial cells that enable duct contraction and milk secretion during lactation.
+At the end of lactation the gland then reverts back to a stage that resembles the pre-pregnancy situation.
+However, we found that in particular the luminal compartment differs from its nulliparous counterpart (\autoref{F1}b and \autoref{F1}d).
+Interestingly, the effect of parity is most pronounced for the less differentiated and more progenitor-like celltypes (C5-PI, \autoref{F1}c, Supplementary Fig 6a).
+To ensure that C5-PI still represents the progenitor population in the post-parous gland, we identified genes that distinguish C5-NP from the rest of the luminal compartment to see if they are also characteristic for the proposed post-parity progenitor population C5-PI.
+Indeed, we find genes that are differentially expressed between C5-NP and the rest of the luminal compartment to show the same trend between C5-PI and the luminal compartment (\autoref{F4}a).
+In a similar manner we can distinguish C5-PI from the rest of the PI gland in a principal component analysis (PCA) using the identified progenitor genes (Supplementary Fig 6c and d).
+From this we conclude, that C5-PI indeed represents the post-parity luminal progenitor population.
+So what's the difference?
+Interestingly, genes that were up-regulated in C5-PI compared to C5-NP were significantly enriched for pathways that are involved in the immune response and lactation (\autoref{F4}b, Supplementary Fig 6c).
+These included genes that play roles in various processes during lactation, e.g.
+milk-proteins (Csn2, Lalba), lipases (Lipa), proteins involved in milk secretion (Xdh, Cd36) and transcriptional regulators of lactation (Cidea).
 Furthermore, the expression of milk genes was surprisingly maintained at high levels in C4 even after lactation (\autoref{F4}d).
@@
 
-In total the Cell Ranger software identified 5,598 barcodes that contained enough unique molecules to be considered as cells (1,707 in NP, 623 in G, 2,648 in L and 620 in PI).
-Libraries prepared from the NP, G and P time points all showed high quality that was reproducible between the two biological replicates (Supplementary 1a).
+In total the Cell Ranger software identified 25,806 barcodes that contained enough unique molecules to be considered as cells (4,376 in NP, 6,021 in G, 9,603 in L and 5,806 in PI).
+Libraries prepared from all time points showed high quality that was reproducible between the two biological replicates (Supplementary 1a).
 We used the following metrics to flag poor quality cells: number of genes detected, total number of unique molecules (UMIs) and percentage of molecules mapped to mitochondrial genes.
-For these samples we then identified poor quality cells by setting a threshold on the number of genes and number of UMIs that was defined as four median absolute deviations (MAD) below the median for each sample (Supplementary 1b and c).
-The samples from the lactation sample, however, showed on average nearly one order of magnitude fewer genes detected and total number of UMIs.
-As this was specific to both lactation samples, we inferred that this was driven by the biological phenotype of the sample rather than technical artefacts.
-Nonetheless, we flagged a substantial amount of the lactation sample as poor quality cells by setting a fixed threshold of 500 detected genes and 1000 total molecules detected (Supplementary Fig 1b and c).
-This was necessary in order to allow for normalisation of all samples; otherwise, keeping more cells of lower quality would inflate the range of normalisation factors amplifying technical noise due to cell-specific biases and adversely impact downstream analyses.
-We ensured that we still sampled cells from all major structures present in the lactation sample (Supplementary Fig 8).
-Finally, all cells with 5\% or more of UMIs mapping to mitochondrial genes were defined as non-viable or apoptotic and removed form the analysis (Supplementary Fig 1d).
-This left us with a total number of 3,499 cells (1,681 in NP, 609 in G, 604 in L and 605 in P).
-The cells were then normalised by size factors as previously described\autocite{Lun2016}.
+Poor quality cells were then identified by setting a threshold on the number of genes and number of UMIs that was defined as three median absolute deviations (MAD) below the median for each sample, while requiring a minimum value of 1,000 total molecules and 500 genes detected.
+This resulted in the following thresholds for total number of genes detected: 1,042 for NP, 836 for G, 500 for L and 759 for PI; and the following for total number of molecules : 2,012 in NP, 1,479 in G, 1,000 in L and 1,379 in G.
+In addition, all cells with 5\% or more of UMIs mapping to mitochondrial genes were defined as non-viable or apoptotic and removed form the analysis (Supplementary Fig 1d).
+We finally also ensured that none of the reads in our dataset are derived from index swapping.
+For this we excluded cells from the analysis with barcodes that appeared in more than one sample (non-unique barcodes).
+The reasoning being that any index swapped read between samples will also share the same cellular barcode, exclusion of these cells hence offers protection against index swapping.
+This left us with a total number of 25,010 cells (4,223 in NP, 5,826 in G, 9,319 in L and 5,642 in P).
+Gene expression values were then normalised by size factors as previously described\autocite{Lun2016}.
 The log-transformed (log2(counts+1)) counts of highly variable genes (HVGs) were used as features for dimensionality reduction and clustering.
-HVGs were defined using the method from Brennecke et al.\autocite{Brennecke2013}.
-Genes with a Benjamini-Hochberg adjusted P value smaller than 0.1 were considered to be significant.
-The t-SNE embedding in \autoref{F1} was computed using the ``Rtsne'' package with default settings and perplexity set to 25 (\href{https://github.com/jkrijthe/Rtsne}{\textit{https://github.com/jkrijthe/Rtsne}}).
+HVGs were identified by first fitting a mean-dependent trend to the gene-specific variances to all genes assuming that this trend is dominated by technical variance.
+This trend was then used to estimate the technical component of the variance and all genes with a biological component (the residual variance) of at least 0.5 and an Benjamini-Hochberg adjusted P value smaller than or equal to 0.05.
+The t-SNE embedding in \autoref{F1} was computed using the ``Rtsne'' package with default settings and perplexity set to 50 (\href{https://github.com/jkrijthe/Rtsne}{\textit{https://github.com/jkrijthe/Rtsne}}).
 
 \textbf{Clustering}
+Due to the large number of cells we used a two-step approach to identify clusters of cells.
+In a first step we clustered cells using a shared-nearest-neighbor graph (SNN-graph) based approach.
+The advantage of this approach is that it does not require the computation of a distance matrix which is computationally inhibitive for large number of cells.
+The SNN-graph was constructed using the 'buildSNNGraph' function from scran with the number of nearest-neighbours set to 20.
+Community-based clustering was then performed on the SNN-graph by multi-level modularity optimization using the 'cluster_louvain' function in igraph (CITATION).
 

### 2017-08-08 paper/main.tex
changed all cluster labels

@@
 This resulted in the following thresholds for total number of genes detected: 1,042 for NP, 836 for G, 500 for L and 759 for PI; and the following for total number of molecules : 2,012 in NP, 1,479 in G, 1,000 in L and 1,379 in G.
-In addition, all cells with 5\% or more of UMIs mapping to mitochondrial genes were defined as non-viable or apoptotic and removed form the analysis (Supplementary Fig 1d).
-We finally also ensured that none of the reads in our dataset are derived from index swapping.
+In addition, all cells with 5\% or more of UMIs mapping to mitochondrial genes were defined as non-viable or apoptotic and removed from the analysis (Supplementary Fig 1d).
+We finally also ensured that none of the reads in our dataset derived from index swapping (CITATION).
 For this we excluded cells from the analysis with barcodes that appeared in more than one sample (non-unique barcodes).
-The reasoning being that any index swapped read between samples will also share the same cellular barcode, exclusion of these cells hence offers protection against index swapping.
+The reasoning being that any index swapped read between samples will also share the same cellular barcode.
+Exclusion of these cells hence offers protection against index swapping.
 This left us with a total number of 25,010 cells (4,223 in NP, 5,826 in G, 9,319 in L and 5,642 in P).
-Gene expression values were then normalised by size factors as previously described\autocite{Lun2016}.
+Gene expression values were then normalised by size factors, that were estimated as previously described\autocite{Lun2016}.
 The log-transformed (log2(counts+1)) counts of highly variable genes (HVGs) were used as features for dimensionality reduction and clustering.
 HVGs were identified by first fitting a mean-dependent trend to the gene-specific variances to all genes assuming that this trend is dominated by technical variance.
-This trend was then used to estimate the technical component of the variance and all genes with a biological component (the residual variance) of at least 0.5 and an Benjamini-Hochberg adjusted P value smaller than or equal to 0.05.
-The t-SNE embedding in \autoref{F1} was computed using the ``Rtsne'' package with default settings and perplexity set to 50 (\href{https://github.com/jkrijthe/Rtsne}{\textit{https://github.com/jkrijthe/Rtsne}}).
+This trend was then used to estimate the technical component of the variance and all genes with a biological component (the residual variance) of at least 0.5 and a Benjamini-Hochberg adjusted P value smaller than or equal to 0.05 were defined as HVG.
+The t-SNE embeddings were computed using the ``Rtsne'' package with default settings and perplexity set to 50 (\href{https://github.com/jkrijthe/Rtsne}{\textit{https://github.com/jkrijthe/Rtsne}}).
 
@@
 In a first step we clustered cells using a shared-nearest-neighbor graph (SNN-graph) based approach.
-The advantage of this approach is that it does not require the computation of a distance matrix which is computationally inhibitive for large number of cells.
-The SNN-graph was constructed using the 'buildSNNGraph' function from scran with the number of nearest-neighbours set to 20.
-Community-based clustering was then performed on the SNN-graph by multi-level modularity optimization using the 'cluster_louvain' function in igraph (CITATION).
-
-As we expected the epithelial compartment to show a hierarchical organisation, we used agglomerative hierarchical clustering to identify cell types.
-We first computed the pairwise Euclidean distances between cells based on the log-transformed counts of HVGs.
-The dissimilarity matrix was used to perform hierarchical clustering (Supplementary Fig 9a) with the 'hclust' function in R using average linkage, as this resulted in the highest cophenetic correlation coefficient between the tree and the dissimilarity matrix (average:0.91 single: 0.73, complete: 0.83, ward.D2: 0.73).
-Clusters were then defined with the 'cutreeDynamic' function from the 'dynamicTreeCut' package (minClusterSize= 15, deepSplit=1, method=''hybrid'')\autocite{Langfelder2008}.
-The deepSplit parameter was set to 1 based on evaluating the robustness of the clustering.
-At a deepSplit of 0, all clusters were highly robust as determined by bootstrapping\autocite{Henning2015}.
-Increasing the parameter to 1, led to splitting C1 from C3 as well as C9 from C6.
-C1, C3 and C6 were highly robust in the bootstrap (see Supplementary Fig 9b).
-C9 showed a below average robustness.
-However, due to the expression of marker genes for a previously reported cell type\autocite{Wang2015} we decided to keep it as a separate cluster.
-Interestingly, the putative cell type of C9 is known to have an expression profile that is in between luminal and basal cells (see main text), which might explain its assignment to other clusters during the bootstrap.
-Further increasing the deepSplit parameter led to unstable clusters (Supplementary Fig 9b).
-With this method, we identified ten clusters. In addition, the dynamic tree cut method allows for a ``noise component'' for cells that do not match to any cluster.
-1 cell from the lactation sample was flagged as noise by the algorithm and removed from further analysis.
-C10 (27 cells) expressed lymphocyte related genes was also removed from all downstream analyses (as previously reported\autocite{Scheele2017}, Supplementary Fig 10).
+This has the advantage of not requiring the computation of a distance matrix which is computationally inhibitive for a large number of cells.
+The SNN-graph was constructed using the \texttt{buildSNNGraph} function from scran with the number of nearest-neighbours set to 20.
+Community-based clustering was then performed on the SNN-graph by multi-level modularity optimization using the \texttt{cluster\_louvain} function in igraph (CITATION).
+This identified a total of 13 clusters.
+Yet, the majority of clusters still contained substructe that was evident from inspecting the t-SNE (SubXa).
+In order to resolve theses structures further we then applied agglomerative hierarchical clustering on each of the 13 clusters (SupXb).
+For this, we first computed the pairwise distances between cells based on the spearman correlation of log-transformed HVG counts.
+The dissimilarity matrix was then used to perform hierarchical clustering with the 'hclust' function in R using average linkage.
+Clusters were defined using the \texttt{cutree} function and the optimal $k$ was determined by maximizing the gap statistic using the \texttt{clusGap} function from the cluster package (CITATION).
+With this we identified a total number of 21 clusters.
+Next, we flagged clusters that expressed clear markers of non-epithelial cells as contaminating cells and removed them from the downstream analysis.
+Cluster 6.1 and 6.2 were tagged as immune cells based on the expression of \textit{Cd74}, \textit{Cd72} and \textit{Cd54}\autocite{Scheele2017} and cluster 10.3 as endothelial cells based on the expression of \textit{Eng}, \textit{S1pr1} and \textit{Emcn} (CITATION,SupXc and SupXd).
+We retained cluster 10.2 as it expressed markers of a previously described celltype in the gland (CITATION), yet it was also positive for 2 out of 4 (\textit{Des} and \textit{Cspg4}) pericyte markers (CITATION) (SupXd).
+The high number of cells that was loaded on to the 10x per sample increased the chance of doublet formation in the droplet encapsulation (CITATION).
+Especially doublets of two cell types can confound downstream analysis.
+We thus identified clusters that appeared at low frequencies in the samples were it was detected (less than the maximum doublet rate of 7\%, SupXa) and was highly similar to at least two other clusters present in the same sample ($p > 0.9$ of mean $log2(counts+1)$).
+The only cluster that fulfilled this criterion was cluster 11.1, which was strongly correlated with XX and XX (SupXb).
+In addition, the cluster was almost perfectly correlated with the average expression values of XX and XX, supporting the hypothesis that it is indeed a mixture of XX and XX (SupXb).
+This was further supported by the fact that XX and XX are the most prevalent clusters in these samples.
+Finally, 11.1 also belonged to the clusters with a high number of genes detected and a high number of unique molecules, which is also indicative for doublet clusters (SupXc).
+We hence excluded 11.1 from any downstream analysis.
+As the last step we grouped clusters of cells together in super structures based on the expression of marker genes for known celltypes.
 

### 2017-08-08 src/DiffusionMap.R

@@ -22,7 +22,7 @@ cells <- list("all","luminal")
 
 for (cell in cells) {
     if(cell=="luminal") {
-	excludeClustComb <- c("C6-G1","C6","C7","C9")
+	excludeClustComb <- c("Bsl-G1","Bsl","Myo","Prc")
     } else{
 	excludeClustComb <- NULL
     }

### 2017-08-08 src/DiffusionMap_PI.R

@@ -13,7 +13,7 @@ pD <- dataList[[2]]
 rm(dataList)
 
 # Remove QC-fails,outlier and immune cells
-excludeClustComb <- c("C6-G1","C6","C7","C9")
+excludeClustComb <- c("Bsl-G1","Bsl","Myo","Prc")
 keep <- pD$keep & pD$Condition=="PI" & !(pD$SubCluster %in% excludeClustComb)
 m.vp <- m[,keep]
 pD.vp <- pD[keep,]

### 2017-08-08 src/ExploreDoublets.R

@@ -78,16 +78,16 @@ for (smpl in c("G1","G2")) {
 	colname <- clust
 	tmp[,colname] <- expr
     }
-    x1 <- data.frame("DoubletCluster"=tmp[,"C6-G2"],
-		     "Singleton"=tmp[,"C2d"],
-		     "Cluster"="C2d")
-    x2 <- data.frame("DoubletCluster"=tmp[,"C6-G2"],
-		     "Singleton"=tmp[,"C6-G1"],
-		     "Cluster"="C6-G1")
+    x1 <- data.frame("DoubletCluster"=tmp[,"Bsl-G2"],
+		     "Singleton"=tmp[,"Avd"],
+		     "Cluster"="Avd")
+    x2 <- data.frame("DoubletCluster"=tmp[,"Bsl-G2"],
+		     "Singleton"=tmp[,"Bsl-G1"],
+		     "Cluster"="Bsl-G1")
     x2 <- rbind(x1,x2)
-    x3 <- data.frame("DoubletCluster"=tmp[,"C6-G2"],
-		     "Singleton"=rowMeans(tmp[,c("C6-G1","C2d")]),
-		     "Cluster"="Mean(C2d,C6-G1)")
+    x3 <- data.frame("DoubletCluster"=tmp[,"Bsl-G2"],
+		     "Singleton"=rowMeans(tmp[,c("Bsl-G1","Avd")]),
+		     "Cluster"="Mean(Avd,Bsl-G1)")
     x4 <- rbind(x2,x3)
     x4$Sample <- smpl
     xout <- rbind(xout,x4)
@@ -100,6 +100,11 @@ p2 <- ggplot(xout, aes(Singleton, DoubletCluster)) +
     xlab("Log-Expression") +
     ylab("Log-Expression") +
     theme_bw()
+
+pD <- group_by(pD, SampleID) %>%
+    dplyr::mutate(GenesDetected=GenesDetected/median(GenesDetected)) %>%
+    dplyr::mutate(UmiSums=UmiSums/median(UmiSums)) %>%
+    ungroup()
 
 ordr <- group_by(pD, SubCluster) %>%
     summarize(tot=median(GenesDetected))  %>%
@@ -132,4 +137,7 @@ p4 <- ggplot(fp4, aes(y=UmiSums, x=SubCluster)) +
 
 library(cowplot)
 
-plot_grid(p1,p2,p3,p4)
+
[…]

### 2017-08-08 src/Figure1.R

@@ -44,7 +44,7 @@ p0 <- ggplot(pD, aes(x=tSNE1, y=tSNE2, color=Condition)) +
 # t-SNE colored by cluster
 p1 <- ggplot(pD, aes(x=tSNE1, y=tSNE2, color=SuperCluster)) +
     geom_point(size=1.5) +
-    scale_color_manual(values=levels(pD$SuperColor))+
+    scale_color_manual(values=levels(pD$SuperColors))+
     #     ggtitle("Cluster") +
     theme_void(base_size=12) +
     guides(colour = guide_legend(override.aes = list(size=3))) +
@@ -72,8 +72,8 @@ dendr <- dis %>% hclust(.,method="ward.D2") %>% as.dendrogram %>%
     set("leaves_pch",19) %>%
     set("leaves_cex",2) %>%
     set("branches_lwd",3)
-
-plot(dendr,yaxt="n",horiz=TRUE)
+par(cex=.9)
+plot(dendr,horiz=TRUE,yaxt="n")
 p.dendr <- grab_grob()
 p.dendr <- grid.arrange(p.dendr)
 dev.off()
@@ -89,24 +89,20 @@ c3 <- c("Foxa1","Ly6a","Aldh1a3","Kit","Cd14")
 c5 <- c("Lypd3")
 c6 <- c("1500015O10Rik","Col7a1","Moxd1","Mia","Emid1","Pdpn","Col9a2","Fbln2","Igfbp3","Fst","Il17b")
 c7 <- c("Oxtr","Krt15","Igfbp6","Igfbp2","Tns1")
-c8 <- c("Pip","Apod")
+# c8 <- c("Pip","Apod")
 c9 <- c("Gng11","Procr","Igfbp7","Nrip2","Notch3","Zeb2")
 
 # Combine in order for heatmap
-genes <- c(general,c1,c3,c5,c2,c8,c6,c7,c9)
+genes <- c(general,c1,c3,c5,c2,c6,c7,c9)
 
 # Subsample cells from large clusters
 set.seed(rnd_seed)
-subsP <- filter(pD, !(SuperCluster %in% c("C8d"))) %>%
-    group_by(SuperCluster) %>%
-    do(sample_n(.,100))
+ord <- group_by(pD, SuperCluster) %>%
+    do(sample_n(.,200)) %>%
+    ungroup() %>%
[…]

### 2017-08-08 src/Figure3.R

@@ -81,7 +81,7 @@ for (i in c(1,2)) {
     rownames(annoCol) <- rownames(pD.ord)
 
     #Set colorscheme for heatmap
-    clustCol <- levels(factor(pD.ord$SuperColor))
+    clustCol <- levels(factor(pD.ord$SuperColors))
     names(clustCol) <- levels(factor(pD.ord$SuperCluster))
     dptcols <- viridis(n=nrow(annoCol),,option="magma",begin=1,end=0)
     names(dptcols) <- c(1:length(dptcols))
@@ -174,7 +174,7 @@ fplot2 <- join(fplot2,raw2,by="barcode") %>%
 
 
 #set colorscale
-clustCol <- levels(factor(pD$SuperColor))[1,3,2,4,5]
+clustCol <- levels(factor(pD$SuperColors))
 
 #Plot dpt-dependent expression for features
 pList <- list()
@@ -232,6 +232,7 @@ expPlot <- plot_grid(branches,expPlot,ncol=1,rel_heights=c(0.3,1))
 htmps <- plot_grid(res[[1]][[1]][[4]],
 		   res[[2]][[1]][[4]],
 		   nrow=2,
+		   scale=0.8,
 		   labels=c("e","f"))
 fullP <- plot_grid(expPlot,htmps,ncol=2)
 

### 2017-08-08 src/Figure4.R

@@ -13,9 +13,9 @@ m <- dataList[[1]]
 pD <- dataList[[2]]
 fD <- dataList[[3]]
 
-# ---- C5DE ----
+# ---- LpDE ----
 
-topTab <- read.csv(file="../data/Robjects/secondRun_2500/C5_NPvsPI.csv")
+topTab <- read.csv(file="../data/Robjects/secondRun_2500/Lp_NPvsPI.csv")
 
 # Highlight genes with lactation/immune annotation
 lac <- read.csv("../data/Robjects/secondRun_2500/LactationGenes.csv", stringsAsFactors=FALSE)$x
@@ -75,12 +75,12 @@ for (i in c(1,2)) {
     add <- select(pD, barcode, SubCluster, SuperCluster, Condition) %>%
 	mutate(barcode=as.character(barcode))
     forPl <- left_join(forPl,add,by="barcode") %>%
-	filter(SubCluster %in% c("C1-NP","C1-PI","C3-NP","C3-PI","C5-NP","C5-PI"))
+	filter(SubCluster %in% c("Hsd-NP","Hsd-PI","Hsp-NP","Hsp-PI","Lp-NP","Lp-PI"))
 
     forPl <- melt(forPl,id=c("barcode","SuperCluster","SubCluster","Condition"))
 
     #Plot
-    forPl$SubCluster <- factor(forPl$SubCluster,levels=c("C5-PI","C5-NP","C3-PI","C3-NP","C1-PI","C1-NP"))
+    forPl$SubCluster <- factor(forPl$SubCluster,levels=c("Lp-PI","Lp-NP","Hsp-PI","Hsp-NP","Hsd-PI","Hsd-NP"))
     out[[i]] <- ggplot(forPl, aes(y=value,x=SubCluster,color=SubCluster)) +
 	geom_jitter(size=0.9) +
 	stat_summary(fun.y = mean, fun.ymin = mean, fun.ymax = mean,
@@ -131,13 +131,13 @@ p2 <- ggplot(progenitorDE, aes(x=NullParFC,y=ParousFC)) +
     geom_point(color="grey50",size=2) +
     geom_point(data=interest, aes(x=NullParFC, y=ParousFC), color="black") +
     geom_label_repel(data=interest, aes(x=NullParFC,y=ParousFC,label=Gene)) +
-    xlab("LFC of C5-NP vs. luminal cells") +
-    ylab("LFC of C5-PI vs. luminal cells") +
[…]

### 2017-08-08 src/IndexSwapping.R

@@ -284,7 +284,7 @@ This leads to cellular barcodes in the lactation sample containing reads from so
 For the CellRanger pipeline it is then essentially impossible to distinguish these barcodes from genuine cells, especially as this sample did (almost) not contain any. 
 The other samples instead contained enough genuine cells to allow the distinction from barcodes that only contained background + index swapping reads.
 Hence, if the lactation samples are removed the number of non-unique barcodes per sample drops almost to zero.
-Nonetheless, some cells are most likely still an artifact, especially cells in C7 and C8.
+Nonetheless, some cells are most likely still an artifact, especially cells in Myo and C8.
 
 ```{r echo=FALSE, message=FALSE, warning=FALSE}
 pD <- pD[pD$PassAll & !pD$isImmuneCell & !pD$isOutlier & pD$Condition!="L",-c(25,26)]

### 2017-08-08 src/Monocle.R

@@ -17,7 +17,7 @@ fD <- dataList[[3]]
 # ---- Prepocessing ----
 
 # Cells
-keepCells <- pD$keep & pD$Condition %in% c("NP","G") & !(pD$SuperCluster %in% c("C6","C6-G1","C7","C9"))
+keepCells <- pD$keep & pD$Condition %in% c("NP","G") & !(pD$SuperCluster %in% c("Bsl","Bsl-G1","Myo","Prc"))
 m <- m[,keepCells]
 pD <- pD[keepCells,]
 

### 2017-08-08 src/NPPIDE.R

@@ -10,12 +10,12 @@ pD <- dataList[[2]]
 fD <- dataList[[3]]
 
 # Loop over all luminal clusters
-comps <- list(C1=c("C1-PI","C1-NP"),
-	      C3=c("C3-PI","C3-NP"),
-	      C5=c("C5-PI","C5-NP"),
-	      C6=c("C6-G1","C6"))
+comps <- list(Hsd=c("Hsd-PI","Hsd-NP"),
+	      Hsp=c("Hsp-PI","Hsp-NP"),
+	      Lp=c("Lp-PI","Lp-NP"),
+	      Bsl=c("Bsl-G1","Bsl"))
 
-for (cname in c("C1","C3","C5","C6")) {
+for (cname in c("Hsd","Hsp","Lp","Bsl")) {
     comp <- comps[[cname]]
     pD.sub <- pD[pD$SubCluster %in% comp,]
     m.sub <- m[,as.character(pD.sub$barcode)]
@@ -45,6 +45,6 @@ for (cname in c("C1","C3","C5","C6")) {
     # Write DE table for supps
     topTab <- select(topTab, id, symbol, logFC, unshrunk.logFC, logCPM, PValue, FDR)
     fname <- sprintf("../data/Robjects/secondRun_2500/%s_NPvsPI.csv",cname)
-    if (cname=="C6") {fname <- sprintf("../data/Robjects/secondRun_2500/%s_NPvsG.csv",cname)}
+    if (cname=="Bsl") {fname <- sprintf("../data/Robjects/secondRun_2500/%s_NPvsG.csv",cname)}
     write.csv(topTab,file=fname, row.names=FALSE)
 }

### 2017-08-08 src/PrepareData.R

@@ -22,8 +22,8 @@ fDat <- data.frame(fData(gene_bc_matrix))[rownames(cDat),]
 pDat <- mutate(pDat, SeqID=substr(barcode,18,18)) %>%
 	mutate(barcode=as.character(barcode)) %>%
         mutate(SeqID=mapvalues(SeqID,c("1","2","3","4","5","6","7","8"),
-				 c("A7","B7","C7","D7","E7","F7","G7","H7"))) %>%
-        mutate(SampleID=mapvalues(SeqID,c("A7","B7","C7","D7","E7","F7","G7","H7"),
+				 c("A7","B7","Myo","D7","E7","F7","G7","H7"))) %>%
+        mutate(SampleID=mapvalues(SeqID,c("A7","B7","Myo","D7","E7","F7","G7","H7"),
 				  c("PI1","PI2","L1","L2","G1","G2","NP1","NP2"))) %>%
 	mutate(Replicate=as.factor(substr(SampleID,nchar(SampleID),nchar(SampleID))),
 	       Condition=factor(substr(SampleID,1,nchar(SampleID)-1),levels=c("NP","G","L","PI"))) %>%

### 2017-08-08 src/ProgenitorDE.R

@@ -13,14 +13,14 @@ fD <- dataList[[3]]
 m <- m[,pD$keep]
 pD <- pD[pD$keep,]
 
-comps <- c("C5-NP","C5-PI")
+comps <- c("Lp-NP","Lp-PI")
 out <- list()
 
 # DE for each progenitor cluster versus all other luminal cells
 for (choice in comps) {
     # subset data
-    rmClust <- setdiff(c("C5-NP","C5-PI"),choice)
-    pD.sub <- filter(pD, !(SubCluster %in% c("C6-G1","C6","C7","C9",rmClust)))
+    rmClust <- setdiff(c("Lp-NP","Lp-PI"),choice)
+    pD.sub <- filter(pD, !(SubCluster %in% c("Bsl-G1","Bsl","Myo","Prc",rmClust)))
     m.sub <- m[,as.character(pD.sub$barcode)]
     keep <- rowMeans(m.sub) > 0.01
     m.sub <- m.sub[keep,]
@@ -46,8 +46,8 @@ for (choice in comps) {
     topTab <- resTab$table
     out[[choice]] <- topTab
 }
-tabNulPar <- out[["C5-NP"]]
-tabPar <- filter(out[["C5-PI"]],symbol %in% tabNulPar$symbol)
+tabNulPar <- out[["Lp-NP"]]
+tabPar <- filter(out[["Lp-PI"]],symbol %in% tabNulPar$symbol)
 rownames(tabNulPar) <- tabNulPar$symbol
 tabNulPar <- tabNulPar[tabPar$symbol,]
 
@@ -58,7 +58,7 @@ progenitorDE <- data.frame("NullParFC"=tabNulPar$logFC,
 write.csv(progenitorDE,file="../data/Robjects/secondRun_2500/ProgenitorDE.csv",
 	  row.names=FALSE)
 
-# ---- DEC4vsC5 ----
+# ---- DEC4vsLp ----
 
 # reload data
 m <- dataList[[1]]
@@ -68,7 +68,7 @@ fD <- dataList[[3]]
 m <- m[,keepCells]
 pD <- pD[keepCells,]
 
-# DE C4 vs C5
+# DE C4 vs Lp
 
 # subset data
 pD.sub <- filter(pD, (cluster %in% c(4,5)))
@@ -99,4 +99,4 @@ topTab <- resTab$table
 
 # Write DE table for supps
[…]

### 2017-08-08 src/S3.R

@@ -21,7 +21,7 @@ fD <- dataList[[3]]
 # ---- ScreePlots ----
 
 condComb <- c("NP","G")
-sets <- list(NULL,c("C6-G1","C6","C7","C9")) #order important for Robustness section
+sets <- list(NULL,c("Bsl-G1","Bsl","Myo","Prc")) #order important for Robustness section
 out <- list()
 for (i in seq_along(sets)) {
     set <- sets[[i]]
@@ -120,15 +120,13 @@ for (feats in features) {
 }
 
 # Set color scheme
-clusts <- as.numeric(as.character(sort(unique(pD.vp$cluster))))
-cols <- brewer.pal(n=9,name="Paired")[clusts]
-names(cols) <- clusts
+cols <- levels(factor(pD.vp$SuperColor))
 
 # Plot
 out$SubSample <- factor(out$SubSample, levels=c("100%","50%","25%"))
 p <- ggplot(out, aes(DC1,DC2, color=SuperCluster)) +
     geom_point(size=0.8) +
-    #     scale_color_manual(values=cols) +
+    scale_color_manual(values=cols) +
     facet_grid(features~SubSample) +
     theme(axis.text=element_blank(),
 	  axis.ticks=element_blank(),
@@ -138,16 +136,13 @@ p <- ggplot(out, aes(DC1,DC2, color=SuperCluster)) +
 	  legend.direction="horizontal",
 	  ) 
 
-# Read in Monocle plot
-pal <- brewer.pal(n=9,name="Paired")[c(1,2,3,5,8,9)]
-
 # Run script if not done yet
 if (!file.exists("../data/Robjects/secondRun_2500/ExpressionList_Monocle.rds")){
     source("Monocle.R")
 }
 
 monoc <- readRDS("../data/Robjects/secondRun_2500/ExpressionList_Monocle.rds")
-monoc.plt <- monoc[["plot"]] %+% guides(color=FALSE)
+monoc.plt <- monoc[["plot"]] %+% guides(color=FALSE) %+% scale_color_manual(values=cols)
 
 # Combine plots
 subp0 <- plot_grid(g1,g2,monoc.plt,nrow=1,labels="auto")

### 2017-08-08 src/S6.R

@@ -11,7 +11,7 @@ source("functions.R")
 # ---- ImpactStrongestOnProgenitor ----
 
 out <- data.frame()
-for (clust in c("C1","C3","C5")) {
+for (clust in c("Hsd","Hsp","Lp")) {
     filen <- sprintf("../data/Robjects/secondRun_2500/%s_NPvsPI.csv",clust)
     tmp <- read.csv(file=filen)
     tmp$Cluster <- clust
@@ -22,7 +22,7 @@ out <- group_by(out, Cluster) %>%
 	      Down=sum(FDR < 0.01 & logFC < 0)) %>%
     melt()
 
-out$Cluster <- factor(out$Cluster,levels=c("C5","C3","C1"))
+out$Cluster <- factor(out$Cluster,levels=c("Lp","Hsp","Hsd"))
 p1 <- ggplot(out, aes(x=Cluster,y=value,fill=variable)) +
     geom_bar(stat="identity",position="dodge") +
     coord_flip() +
@@ -32,7 +32,7 @@ p1 <- ggplot(out, aes(x=Cluster,y=value,fill=variable)) +
 
 # ---- DEAnalysis4vs5 ----
 output <- data.frame()
-for (clust in c("C1","C3","C5")) { # C5 has to be last for rest of script
+for (clust in c("Hsd","Hsp","Lp")) { # Lp has to be last for rest of script
     filen <- sprintf("../data/Robjects/secondRun_2500/%s_NPvsPI.csv",clust)
     topTab <- read.csv(file=filen)
 
@@ -68,7 +68,7 @@ output$Term <- factor(output$Term,levels=unique(ordr))
 
 ## Extract gene lists for volcano plot
 # better hardcode go-terms at one point to define gene lists
-sub.output <- output[output$Cluster=="C5",]
+sub.output <- output[output$Cluster=="Lp",]
 imterm <- sub.output[as.character(c(5,11,12,21,23,29,37,41,42,44,49,50)),1]
 lterm <- sub.output[as.character(c(35,25,6,20)),1]
 im.genes <- intersect(unlist(genesInTerm(GO.data,imterm)),deGenes)
@@ -96,7 +96,7 @@ m <- m[,pD$keep]
 pD <- pD[pD$keep,]
 
 # DE Test on NP cells progenitor versus rest
[…]

### 2017-08-08 src/S9_S10.R

@@ -108,7 +108,7 @@ p3 <- p3 %+% facet_grid(variable~.) %+% xlab("Cluster") %+% ylab("Log-Expression
 # ---- PCAonAvgExpression ----
 
 # Compute mean expression values
-out <- data.frame("C1"=numeric(nrow(m)))
+out <- data.frame("Hsd"=numeric(nrow(m)))
 for (clust in levels(pD$cluster)[-1]) {
     expr <- rowMeans(log2(m[,pD$cluster==clust]+1))
     colname <- paste0("C",clust)

### 2017-08-08 paper/main.tex
changed cluster labels

@@
 \usepackage{hyperref}
+\usepackage{color}
 \PassOptionsToPackage{usenames,dvipsnames}{color} % color is loaded by hyperref
@@
 {\refstepcounter{table}\label{#1}}
+%------------------------------------------------------
+
+%---------------- editing------------------------
+\newcommand{\fixme}[1]{\textit{\textcolor{red}{Fixme: #1}}}
+\newcommand{\comment}[1]{\textit{\textcolor{blue}{Comment: #1}}}
 %------------------------------------------------------
@@
 Visual inspection of the data using t-distributed stochastic neighbour embedding (t-SNE) suggested that although there is some grouping of cells by time-point, there are also other factors that underlie structure within the dataset (\autoref{F1}b and Supplementary Fig 2a).
-We used unsupervised hierarchical clustering based on the expression of 594 highly variable genes to identify 9 clusters of epithelial cells (C1-C9).
+We used unsupervised hierarchical clustering based on the expression of 594 highly variable genes to identify 9 clusters of epithelial cells (Hsd-Prc).
 Importantly, these clusters were consistent with the groups observed in the t-SNE (\autoref{F1}c).
-Based on the expression of \textit{Krt18,} \textit{Krt8, Krt5, Krt14 and Acta2} we grouped the clusters into 6 luminal (C1, C2, C3, C4, C5 \& C8) and 3 basal clusters (C6, C7 \& C9) (\autoref{F1}D, Supplementary Fig 2b and 2c).
+Based on the expression of \textit{Krt18,} \textit{Krt8, Krt5, Krt14 and Acta2} we grouped the clusters into 6 luminal (Hsd, Avd, Hsp, Lp-PI, Lp \& Avp) and 3 basal clusters (Bsl, Myo \& Prc) (\autoref{F1}D, Supplementary Fig 2b and 2c).
 Differential gene expression analysis allowed us to further characterise the clusters and infer putative identities (\autoref{F1}d, \autoref{T1} and Supplementary Table 1).
-For example, C2 expresses high levels of beta-casein, \textit{Csn2,} and is composed exclusively from cells from the gestation and lactation time-points, suggesting that these cells are secretory alveolar cells (\autoref{F1}d).
-C1 and C3 on the other hand represent the clusters with the highest expression of hormone receptors.
-C5 expresses high levels of the luminal progenitor marker \textit{Aldh1a3} suggesting progenitor function, while C7 shows characteristics of myoepithelial cells such as high levels of \textit{Acta2}, \textit{Oxtr} and \textit{Krt15} (\autoref{F1}d).
+For example, Avd expresses high levels of beta-casein, \textit{Csn2,} and is composed exclusively from cells from the gestation and lactation time-points, suggesting that these cells are secretory alveolar cells (\autoref{F1}d).
+Hsd and Hsp on the other hand represent the clusters with the highest expression of hormone receptors.
+Lp expresses high levels of the luminal progenitor marker \textit{Aldh1a3} suggesting progenitor function, while Myo shows characteristics of myoepithelial cells such as high levels of \textit{Acta2}, \textit{Oxtr} and \textit{Krt15} (\autoref{F1}d).
 
@@
 
-The expression of the progenitor marker \textit{Aldh1a3} gradually decreased as cells progressed away from their common origin (\autoref{F2}b right panel), which was largely composed of C5.
-We further noted that the left arm of the differentiation trajectory terminates at C2 and shows increasing expression of \textit{Csn2} (\autoref{F2}b bottom panel), consistent with a secretory phenotype.
-Between C5 and C2 we found cells from C8 (\autoref{F2}b).
-C8 is the second rarest cell population after C9 and is found in all developmental stages in approximately equal numbers (6 cells in NP, 5 cells in G, 8 cells in L and 7 cells in PI).
-C8 cells express \textit{Aldh3a1, Elf5, Cd14 and Kit} similar to C5 but in addition express high levels of prolactin induced protein \textit{(Pip)} and apolipoprotein D \textit{(Apod)} (\autoref{F1}d).
-On the right arm of the differentiation trajectory, cells in C5 transitioned to C3 and C1, during which the expression of \textit{Esr1} and \textit{Pgr} increase suggesting that this branch represents differentiation towards hormone-sensing luminal cells (\autoref{F2}b, bottom panel).
+The expression of the progenitor marker \textit{Aldh1a3} gradually decreased as cells progressed away from their common origin (\autoref{F2}b right panel), which was largely composed of Lp.
+We further noted that the left arm of the differentiation trajectory terminates at Avd and shows increasing expression of \textit{Csn2} (\autoref{F2}b bottom panel), consistent with a secretory phenotype.
+Between Lp and Avd we found cells from Avp (\autoref{F2}b).
+\fixme{C8 is the second rarest cell population after Prc and is found in all developmental stages in approximately equal numbers (6 cells in NP, 5 cells in G, 8 cells in L and 7 cells in PI).
+C8 cells express \textit{Aldh3a1, Elf5, Cd14 and Kit} similar to Lp but in addition express high levels of prolactin induced protein \textit{(Pip)} and apolipoprotein D \textit{(Apod)} (\autoref{F1}d).}
+On the right arm of the differentiation trajectory, cells in Lp transitioned to Hsp and Hsd, during which the expression of \textit{Esr1} and \textit{Pgr} increase suggesting that this branch represents differentiation towards hormone-sensing luminal cells (\autoref{F2}b, bottom panel).
 
@@
 However, we found that in particular the luminal compartment differs from its nulliparous counterpart (\autoref{F1}b and \autoref{F1}d).
-Interestingly, the effect of parity is most pronounced for the less differentiated and more progenitor-like celltypes (C5-PI, \autoref{F1}c, Supplementary Fig 6a).
-To ensure that C5-PI still represents the progenitor population in the post-parous gland, we identified genes that distinguish C5-NP from the rest of the luminal compartment to see if they are also characteristic for the proposed post-parity progenitor population C5-PI.
-Indeed, we find genes that are differentially expressed between C5-NP and the rest of the luminal compartment to show the same trend between C5-PI and the luminal compartment (\autoref{F4}a).
-In a similar manner we can distinguish C5-PI from the rest of the PI gland in a principal component analysis (PCA) using the identified progenitor genes (Supplementary Fig 6c and d).
-From this we conclude, that C5-PI indeed represents the post-parity luminal progenitor population.
+Interestingly, the effect of parity is most pronounced for the less differentiated and more progenitor-like celltypes (Lp-PI, \autoref{F1}c, Supplementary Fig 6a).
+To ensure that Lp-PI still represents the progenitor population in the post-parous gland, we identified genes that distinguish Lp-NP from the rest of the luminal compartment to see if they are also characteristic for the proposed post-parity progenitor population Lp-PI.
+Indeed, we find genes that are differentially expressed between Lp-NP and the rest of the luminal compartment to show the same trend between Lp-PI and the luminal compartment (\autoref{F4}a).
+In a similar manner we can distinguish Lp-PI from the rest of the PI gland in a principal component analysis (PCA) using the identified progenitor genes (Supplementary Fig 6c and d).
+From this we conclude, that Lp-PI indeed represents the post-parity luminal progenitor population.
 So what's the difference?
-Interestingly, genes that were up-regulated in C5-PI compared to C5-NP were significantly enriched for pathways that are involved in the immune response and lactation (\autoref{F4}b, Supplementary Fig 6c).
+Interestingly, genes that were up-regulated in Lp-PI compared to Lp-NP were significantly enriched for pathways that are involved in the immune response and lactation (\autoref{F4}b, Supplementary Fig 6c).
 These included genes that play roles in various processes during lactation, e.g.
 milk-proteins (Csn2, Lalba), lipases (Lipa), proteins involved in milk secretion (Xdh, Cd36) and transcriptional regulators of lactation (Cidea).
-Furthermore, the expression of milk genes was surprisingly maintained at high levels in C4 even after lactation (\autoref{F4}d).
+Furthermore, the expression of milk genes was surprisingly maintained at high levels in Lp-PI even after lactation (\autoref{F4}d).
 Of note, the genes of the casein locus (\textit{Csn2}, \textit{Csn1s1}, \textit{Csn1s2a}, \textit{Csn3}) have previously been reported to be up-regulated in the parous gland, most likely due to changes in chromatin accessibility\autocite{Dos2015,Rijnkels2013}.
@@
 We have reported here the use of single cell RNA sequencing to comprehensively map the transcriptomes of thousands of mammary epithelial cells across four developmental time-points.
-Our analysis identified 9 clusters of epithelial cells, some of which are only present during specific developmental stage (e.g. C2 and C4).
+Our analysis identified 9 clusters of epithelial cells, some of which are only present during specific developmental stage (e.g. Avd and Lp-PI).
 This study provides a rich dataset that can be mined online (see link in materials and methods) to identify marker genes and lineage specific promoters that can be used to trace populations of cells \textit{in vivo}.
@@
 In this study we could not provide any evidence for contribution of a putative mutlipotent stem cell to the day-to-day homoeostasis of the gland.
-We do note, however, that the cluster C9 is placed between the luminal and basal cells in \autoref{F2}a.
-C9 represents the cluster with the fewest numbers of cells captured and expresses low levels of both luminal and basal markers, high levels of the stem cell marker \textit{Procr}\autocite{Wang2015} and high levels of the luminal progenitor marker \textit{Notch3}\autocite{Lafkas2013} (\autoref{F1}d).
-Yet, we did not capture cells in transition between C9 and any other clusters suggesting that if these cells are indeed, multi-potent progenitors their involvement in tissue homoeostasis is negligible.
-Alternatively, we cannot exclude the possibility that C9 is enriched for doublets.
-Based on the gene expression data presented here, the luminal compartment appears to have one common progenitor population (C5).
-We also identified a previously described progenitor population with expression of hormone-receptor related genes (C3) that appears to represent a meta-stable state on the differentiation path from luminal progenitors towards fully differentiated hormone-sensing cells (C1).
+\fixme{We do note, however, that the cluster Prc is placed between the luminal and basal cells in \autoref{F2}a.
+Prc represents the cluster with the fewest numbers of cells captured and expresses low levels of both luminal and basal markers, high levels of the stem cell marker \textit{Procr}\autocite{Wang2015} and high levels of the luminal progenitor marker \textit{Notch3}\autocite{Lafkas2013} (\autoref{F1}d).
+Yet, we did not capture cells in transition between Prc and any other clusters suggesting that if these cells are indeed, multi-potent progenitors their involvement in tissue homoeostasis is negligible.
+Alternatively, we cannot exclude the possibility that Prc is enriched for doublets.}
+Based on the gene expression data presented here, the luminal compartment appears to have one common progenitor population (Lp).
+We also identified a previously described progenitor population with expression of hormone-receptor related genes (Hsp) that appears to represent a meta-stable state on the differentiation path from luminal progenitors towards fully differentiated hormone-sensing cells (Hsd).
 Furthermore, we characterised gene expression patterns along the differentiation hierarchy, thus enabling to disentangle the cellular hierarchy in the mammary gland.
@@
 The majority of the changes were related to pathways involved in immunity and lactation, suggesting that, in particular, the luminal progenitors maintain a memory of gestation and involution.
-It is reasonable to assume that C4 overlaps with the previously parity-induced mammary epithelial cells (PI-MECs)\autocite{Wagner2002}.
+It is reasonable to assume that Lp-PI overlaps with the previously parity-induced mammary epithelial cells (PI-MECs)\autocite{Wagner2002}.
 
@@
 Similarities are based on Euclidean distances.
-\textbf{(b)} Comparison of fold changes from C5 versus luminal compartment and fold changes from C4 versus the luminal cells.
-The genes represent the top 500 differentially expressed genes between C5 and luminal cells.
-\textbf{(c)} Volcano plot illustrates differential expression between C5 and C4, coloured dots represent significant genes with known function in lactation and immunity, dashed lines highlight the P value threshold of 0.01 and a log fold change of 1.
+\textbf{(b)} Comparison of fold changes from Lp versus luminal compartment and fold changes from Lp-PI versus the luminal cells.
+The genes represent the top 500 differentially expressed genes between Lp and luminal cells.
+\textbf{(c)} Volcano plot illustrates differential expression between Lp and Lp-PI, coloured dots represent significant genes with known function in lactation and immunity, dashed lines highlight the P value threshold of 0.01 and a log fold change of 1.
 P values are adjusted for multiple testing using Benjamini-Hochberg.
-\textbf{(d)} Visualization of expression difference for some lactation related genes from \textbf{(c)}, where C4 is split according to the developmental time point.
+\textbf{(d)} Visualization of expression difference for some lactation related genes from \textbf{(c)}, where Lp-PI is split according to the developmental time point.
 %-------------------------------------------------------------------------------------------------------------------
@@
 
-\textbf{Supplementary Figure 5. Expression of progenitors markers in C4}
+\textbf{Supplementary Figure 5. Expression of progenitors markers in Lp-PI}
 
@@
 
-\textbf{Supplementary Figure 6. C4 is a post-parity progenitor
+\textbf{Supplementary Figure 6. Lp-PI is a post-parity progenitor
 population }
@@
 \textbf{(a)} Principal component analysis (PCA) on all luminal cells from the PI time-point.
-The PCA was computed on the top 500 differentially expressed genes between C5 and the rest of the luminal NP gland.
+The PCA was computed on the top 500 differentially expressed genes between Lp and the rest of the luminal NP gland.
 PC1 separates the progenitors and the differentiated cells, with the progenitors showing negative PC1 values.
-\textbf{(b)} Genes that are higher expressed in C5 compared to the rest of the NP gland also have negative PC1 loadings.
+\textbf{(b)} Genes that are higher expressed in Lp compared to the rest of the NP gland also have negative PC1 loadings.
 \textbf{(c)} Top 20 GO-terms (biological processes) that are significantly enriched in upregulated genes from \autoref{F4}c.
@@
 
-\textbf{Table S4.} Genes differentially expressed between C4 and C5
+\textbf{Table S4.} Genes differentially expressed between Lp-PI and Lp
 
@@
 First, we selected all cells from the NP and G time point (\autoref{F2}a) and detected the HVGs as described above.
-The log-transformed (log2(count+1)) gene counts were then used to compute the first twenty diffusion components using the 'DiffusionMap' function (k=50, otherwise the default parameters as in 'destiny'\autocite{Angerer2016}.
+The log-transformed (log2(count+1)) gene counts were then used to compute the first twenty diffusion components using the \texttt{DiffusionMap} function (default parameters as in 'destiny'\autocite{Angerer2016}.
 After inspecting the fraction of explained variation for the first twenty components, we decided to retain the first 3.
@@
 A likelihood ratio test was then used to assess statistical significance of the fit compared to a null (pseudotime-independent) model.
-Genes with a Benjamini-Hochberg corrected P value below 0.001 were considered to be significantly pseudotime-dependent.
+Genes with a Benjamini-Hochberg corrected P value below 0.01 and a minimum change in log2(expression) along pseudotime of 0.5 were considered to be significantly pseudotime-dependent.
 We then used a heuristic definition of branch-specific expression instead of modelling the branch assignment explicitly.
@@
 increase or decrease along the differentiation towards one of the two cell types, rather than comparing the exact timing of gene activation/inactivation between the branches.
-Any approach trying to do the latter would have been complicated by the different cell densities along the branches and by the difficulty of verifying any such hypotheses \textit{in vivo}.
+Any approach trying to do the latter would have been complicated by the different cell densities along the branches, differences in branch length and by the difficulty of verifying any such hypotheses \textit{in vivo}.
 Hence, we defined genes to be branch-specific when they were pseudotime-dependent in their expression in at least one of the two branches and when the gradient differed in signs between two branches.
@@
 
-A gene set enrichment analysis based on gene-ontology (GO) terms was conducted to characterize genes that were up-regulated in C4 compared to C5.
+A gene set enrichment analysis based on gene-ontology (GO) terms was conducted to characterize genes that were up-regulated in Lp-PI compared to Lp.
 Genes with positive log fold changes were compared to all genes that were tested for differential expression using topGO with default settings\autocite{Alexa2016}.
@@
 
-K.B. performed most of the experiments and all the computational analysis.
+K.B. performed the experiments and all the computational analysis.
 S.P. and K.B. setup and collected the mammary epithelial cells.

### 2017-08-08 paper/main.tex
first version of R1

@@
 \usepackage{color}
+\usepackage{setspace}
 \PassOptionsToPackage{usenames,dvipsnames}{color} % color is loaded by hyperref
@@
 \end{titlepage}
-
+\onehalfspacing
 \textbf{Abstract:}
@@
 To this end we herein use single cell RNA sequencing (scRNAseq) to determine the gene expression profile of individual mammary epithelial cells across four adult developmental stages; nulliparous, mid gestation, lactation and post weaning (full natural involution).
-Our data from 3,471 individual cells identifies 9 distinct mammary epithelial cell populations and allows their hierarchical structure across development to be charted.
+Our data from 25,010 individual cells identifies 8 distinct mammary epithelial cell populations and allows their hierarchical structure across development to be charted.
 Interestingly, the effect of gestation and lactation appeared to be more pronounced for some cell types.
@@
 This view highlights the plasticity of the tissue and might help to explain some of the conflicting results from lineage tracing studies.}
+\newpage
 
@@
 To this end we herein use single cell RNA sequencing (scRNAseq) to map the cellular dynamics of mammary epithelial cells across four adult developmental stages; nulliparous, mid gestation, lactation and post weaning (full natural involution).
-Our data from 3,471 individual cells, identifies 9 distinct cell populations within the gland and allows their hierarchical structure across developmental time points to be charted.
+Our data from 25,010 individual cells, identifies 8 distinct cell populations within the gland and allows their hierarchical structure across developmental time points to be charted.
 
@@
 
-\textbf{Single cell RNA sequencing identifies 9 clusters of mammary epithelial cells}
+\textbf{Single cell RNA sequencing identifies 8 clusters of mammary epithelial cells}
 
@@
 All samples were then prepared for single-cell RNA-sequencing using the 10x Chromium platform\autocite{Zheng2017}.
-Following quality control (materials and methods), this yielded an average of 11,175 unique transcripts and 3,113 genes detected from 3,471 cells (1,681 from NP, 609 from G, 604 from L and 605 from PI) (Supplementary \autoref{F1}a).
-
+Following quality control (materials and methods), this yielded an average of 6,175 unique transcripts and 2,118 genes detected from 25,010 cells (4,223 in NP, 5,826 in G, 9,319 in L and 5,642 in P) (Supplementary \autoref{F1}a).
 Visual inspection of the data using t-distributed stochastic neighbour embedding (t-SNE) suggested that although there is some grouping of cells by time-point, there are also other factors that underlie structure within the dataset (\autoref{F1}b and Supplementary Fig 2a).
-We used unsupervised hierarchical clustering based on the expression of 594 highly variable genes to identify 9 clusters of epithelial cells (Hsd-Prc).
-Importantly, these clusters were consistent with the groups observed in the t-SNE (\autoref{F1}c).
-Based on the expression of \textit{Krt18,} \textit{Krt8, Krt5, Krt14 and Acta2} we grouped the clusters into 6 luminal (Hsd, Avd, Hsp, Lp-PI, Lp \& Avp) and 3 basal clusters (Bsl, Myo \& Prc) (\autoref{F1}D, Supplementary Fig 2b and 2c).
+First, we dissected the global structure in the dataset by unsupervised clustering using a shared-nearest-neighbour community clustering approach.
+This identified 13 groups, which still contained evident substructure as visualized by t-SNE (SupX).
+Hence, we used hierarchical clustering on each of the identify groups to further resolve the cellular heterogeneity.
+This identified a total of 15 clusters of epithelial cells (\autoref{F1}c, SupX).
 Differential gene expression analysis allowed us to further characterise the clusters and infer putative identities (\autoref{F1}d, \autoref{T1} and Supplementary Table 1).
-For example, Avd expresses high levels of beta-casein, \textit{Csn2,} and is composed exclusively from cells from the gestation and lactation time-points, suggesting that these cells are secretory alveolar cells (\autoref{F1}d).
-Hsd and Hsp on the other hand represent the clusters with the highest expression of hormone receptors.
-Lp expresses high levels of the luminal progenitor marker \textit{Aldh1a3} suggesting progenitor function, while Myo shows characteristics of myoepithelial cells such as high levels of \textit{Acta2}, \textit{Oxtr} and \textit{Krt15} (\autoref{F1}d).
+Notably, some clusters represent the same putative cell type at different developmental time points (indicated by the NP, G, L or PI suffix) and were thus grouped together (\autoref{F1}c, \autoref{F1}d).
+Based on the expression of \textit{Krt18,} \textit{Krt8, Krt5, Krt14} and \textit{Acta2} we noted that 5 clusters show a luminal profile (Lp, Hsp, Hsd, Avp \& Avd) and 3 a basal profile (Bsl, Myo \& Prc) (\autoref{F1}D, Supplementary Fig 2b and 2c).
+Avd and Avp express high levels of beta-casein, \textit{Csn2,} and are exclusively composed of cells from the gestation and lactation time-points, suggesting that these cells are secretory alveolar cells (\autoref{F1}d).
+Hsd and Hsp on the other hand represent the clusters with the highest expression of hormone receptors, suggesting hormone sensing function.
+Lp expresses high levels of the luminal progenitor marker \textit{Aldh1a3} suggesting progenitor function, while Hsp and Avp also show expression of progenitor markers but to a lesser extent.
+Myo shows characteristics of myoepithelial cells such as high levels of \textit{Acta2}, \textit{Oxtr} and \textit{Krt15} (\autoref{F1}d).
+\comment{Describe Bsl and Prc as well?}
 

### 2017-08-11 paper/main.tex
version1 as sent to WTK

@@
 Following quality control (materials and methods), this yielded an average of 6,175 unique transcripts and 2,118 genes detected from 25,010 cells (4,223 in NP, 5,826 in G, 9,319 in L and 5,642 in P) (Supplementary \autoref{F1}a).
-Visual inspection of the data using t-distributed stochastic neighbour embedding (t-SNE) suggested that although there is some grouping of cells by time-point, there are also other factors that underlie structure within the dataset (\autoref{F1}b and Supplementary Fig 2a).
-First, we dissected the global structure in the dataset by unsupervised clustering using a shared-nearest-neighbour community clustering approach.
-This identified 13 groups, which still contained evident substructure as visualized by t-SNE (SupX).
-Hence, we used hierarchical clustering on each of the identify groups to further resolve the cellular heterogeneity.
-This identified a total of 15 clusters of epithelial cells (\autoref{F1}c, SupX).
+Visual inspection of the data using t-distributed stochastic neighbour embedding (t-SNE) suggested that although there is grouping of cells by time-point, there are also other factors that underlie structure within the dataset (\autoref{F1}b and Supplementary Figure 2a).
+First, we dissected the global structure in the dataset by unsupervised clustering using a shared-nearest-neighbour clustering approach.
+This resulted in a coarse clustering into 13 groups (SupX).
+In a second step we used hierarchical clustering on each of the identified groups to further resolve the cellular heterogeneity (\autoref{F1}c, SupX).
 Differential gene expression analysis allowed us to further characterise the clusters and infer putative identities (\autoref{F1}d, \autoref{T1} and Supplementary Table 1).
@@
 Based on the expression of \textit{Krt18,} \textit{Krt8, Krt5, Krt14} and \textit{Acta2} we noted that 5 clusters show a luminal profile (Lp, Hsp, Hsd, Avp \& Avd) and 3 a basal profile (Bsl, Myo \& Prc) (\autoref{F1}D, Supplementary Fig 2b and 2c).
-Avd and Avp express high levels of beta-casein, \textit{Csn2,} and are exclusively composed of cells from the gestation and lactation time-points, suggesting that these cells are secretory alveolar cells (\autoref{F1}d).
+Avd and Avp express high levels of beta-casein, \textit{Csn2}, and are exclusively composed of cells from the gestation and lactation time-points, suggesting that these cells are secretory alveolar cells (\autoref{F1}d).
 Hsd and Hsp on the other hand represent the clusters with the highest expression of hormone receptors, suggesting hormone sensing function.
-Lp expresses high levels of the luminal progenitor marker \textit{Aldh1a3} suggesting progenitor function, while Hsp and Avp also show expression of progenitor markers but to a lesser extent.
+Lp expresses high levels of the luminal progenitor marker \textit{Aldh1a3} suggesting progenitor function, while Hsp and Avp also express progenitor markers but to a lesser extent.
 Myo shows characteristics of myoepithelial cells such as high levels of \textit{Acta2}, \textit{Oxtr} and \textit{Krt15} (\autoref{F1}d).
@@
 Between Lp and Avd we found cells from Avp (\autoref{F2}b).
-\fixme{C8 is the second rarest cell population after Prc and is found in all developmental stages in approximately equal numbers (6 cells in NP, 5 cells in G, 8 cells in L and 7 cells in PI).
-C8 cells express \textit{Aldh3a1, Elf5, Cd14 and Kit} similar to Lp but in addition express high levels of prolactin induced protein \textit{(Pip)} and apolipoprotein D \textit{(Apod)} (\autoref{F1}d).}
 On the right arm of the differentiation trajectory, cells in Lp transitioned to Hsp and Hsd, during which the expression of \textit{Esr1} and \textit{Pgr} increase suggesting that this branch represents differentiation towards hormone-sensing luminal cells (\autoref{F2}b, bottom panel).
@@
 Being confident that the diffusion map recapitulates the luminal differentiation process, we then computationally inferred the two branches and ordered the cells according to their progression through ``pseudotime''\autocite{Haghverdi2016} (see materials and methods for further explanation) (\autoref{F3}a).
-We found 806 genes that showed a pseudotime-dependent expression with the same directionality along both differentiation trajectories (\autoref{F3}b, top panel of heatmap, Supplementary Table 2, Supplementary Fig 4a).
-These included genes associated with known progenitor characteristics such as \textit{Aldh1a3}, \textit{Kit}, \textit{Cd14} and \textit{Tspan8} as well as transcription factors that have not previously been associated with luminal differentiation such as \textit{Creb5}, \textit{Hey1} and \textit{Fosl1} (\autoref{F3}c).
-In addition, we identified 1478 genes with branch-specific expression patterns (\autoref{F3}b, bottom panel of heatmap, Supplementary Table 3, Supplementary Fig 4b), including milk related genes (\textit{Wap}, \textit{Csn1s1}, \textit{Csn3}) as well as genes involved in hormone-receptor signalling (\textit{Pgr, Prlr, S100a6, Cited1, Areg}).
+We found 456 genes that showed a pseudotime-dependent expression with the same directionality along both differentiation trajectories (Supplementary Table 2).
+These included genes associated with known progenitor characteristics such as \textit{Aldh1a3} and \textit{Tspan8} as well as transcription factors that have not previously been associated with luminal differentiation such as \textit{Creb5}, \textit{Hmga1} and \textit{Fosl1} (\autoref{F3}b).
+Interestingly, \textit{Hmga1} is a known chromatin remodeller with reprogramming activity that has been implicated as a driver in basal-like breast cancer (CHECK AND CITE).
+In addition, we identified 1005 genes with branch-specific expression patterns (\autoref{F3}d, \autoref{F3}e Supplementary Table 3).
+We further clustered the gene expression trend for these on each of the two branches and identified sets of genes that change over pseudotime.
+On the alveolar branch we found two clusters of genes that increase in expression during differentiation, one that is switched on early and another that is activated later in the final stages of differentiation.
+The early cluster is enriched for genes involved in fatty-acid oxidation and lipid biosynthetic processes (e.g. \textit{Lipa}, \textit{Dbi} and \textit{Acox1}) .
+The second, smaller group of genes that is switched on during the late stages of differentiation was enriched for genes involved in fatty-acid transport and lipid homeostasis (including e.g.
+\textit{Olah}, \textit{Cd36}, \textit{Wap} and \textit{Lalba}).
+Genes involved in cell division and translation show a transient phase of up-regulation during the alveolar differentiation process (e.g.
+\textit{Aurka}, \textit{Cenpa}, \textit{Cdk6}), whereas genes regulating cell shape and morphogenesis of a branching epithelium are repressed early on (e.g.
+\textit{Vdr}, \textit{Areg} and \textit{Myc}).
+During the process of differentiation towards hormone-sensing luminal cells, we identified two broad clusters of genes that gradually increase or decrease their expression.
+Genes with increasing expression are involved in hormone metabolic processes and regulation of morphogenesis of an epithelium (e.g. \textit{Vdr}, \textit{Esr1}, \textit{Pgr} and \textit{Msx2}).
 Amongst the branch-specific transcription factors we found for example \textit{Runx1, Tox2} and \textit{Bhlhe41} (also known as \textit{Sharp1}) to be transcribed during differentiation towards the hormone-sensing lineage (\autoref{F3}d) and \textit{Elf5, Foxs1} and \textit{Ehf} in the secretory lineage (\autoref{F3}e).
@@
 
-\textbf{Parity biases luminal progenitors towards the alveolar fate}
+\textbf{Parity primes luminal progenitors towards the alveolar fate}
+
 Our data illustrates how the cellular composition of the gland changes during the pregnancy cycle.
@@
 The basal compartment on the other hand differentiates to produce oxytocin-sensing, myoepithelial cells that enable duct contraction and milk secretion during lactation.
-At the end of lactation the gland then reverts back to a stage that resembles the pre-pregnancy situation.
+At the end of lactation the gland then reverts back to a stage that resembles the pre-pregnancy state.
 However, we found that in particular the luminal compartment differs from its nulliparous counterpart (\autoref{F1}b and \autoref{F1}d).
-Interestingly, the effect of parity is most pronounced for the less differentiated and more progenitor-like celltypes (Lp-PI, \autoref{F1}c, Supplementary Fig 6a).
+Interestingly, the effect of parity is most pronounced for the less differentiated and more progenitor-like cell types (Lp-PI, \autoref{F1}c, Supplementary Fig 6a).
 To ensure that Lp-PI still represents the progenitor population in the post-parous gland, we identified genes that distinguish Lp-NP from the rest of the luminal compartment to see if they are also characteristic for the proposed post-parity progenitor population Lp-PI.
@@
 From this we conclude, that Lp-PI indeed represents the post-parity luminal progenitor population.
-So what's the difference?
 Interestingly, genes that were up-regulated in Lp-PI compared to Lp-NP were significantly enriched for pathways that are involved in the immune response and lactation (\autoref{F4}b, Supplementary Fig 6c).
 These included genes that play roles in various processes during lactation, e.g.
-milk-proteins (Csn2, Lalba), lipases (Lipa), proteins involved in milk secretion (Xdh, Cd36) and transcriptional regulators of lactation (Cidea).
-Furthermore, the expression of milk genes was surprisingly maintained at high levels in Lp-PI even after lactation (\autoref{F4}d).
-Of note, the genes of the casein locus (\textit{Csn2}, \textit{Csn1s1}, \textit{Csn1s2a}, \textit{Csn3}) have previously been reported to be up-regulated in the parous gland, most likely due to changes in chromatin accessibility\autocite{Dos2015,Rijnkels2013}.
-However, it has not been shown before that this effect is confined to the progenitor population of the luminal compartment (Supplementary Fig 7).
+milk-proteins (\textit{Csn2}, \textit{Lalba}), lipases (\textit{Lipa}), proteins involved in milk secretion (\textit{Xdh}, \textit{Cd36}) and transcriptional regulators of lactation (\textit{Cidea}).
+Notably, some of the genes that are involved in the immune response are known regulators of the involution process such as \textit{Ctsc}, \textit{Tgfb3} or \textit{Mfge8}.
+The up-regulation of these genesets was also present in the Hsp and in some cases even the differentiated Hsd cluster, yet the effect remained strongest in the progenitor cells (\autoref{F4}c).
+Finally, we compared the differentiation processes of the luminal compartment of the parous gland to the nulliparous gland by mapping the luminal cells from PI to the trajectory of NP and G cells.
+Interestingly, the Hsp and Hsd cluster generally maintain their position of their nulliparous counterpart in the differentiation hierarchy whereas the Lp-PI cluster is stretched out from the origin down towards the alveolar branch, suggesting that these are biased towards the alveolar fate (\autoref{F4}b).
+\comment{Doesn't fit: Of note, the genes of the casein locus (\textit{Csn2}, \textit{Csn1s1}, \textit{Csn1s2a}, \textit{Csn3}) have previously been reported to be up-regulated in the parous gland, most likely due to changes in chromatin accessibility\autocite{Dos2015,Rijnkels2013}.
+However, it has not been shown before that this effect is confined to the progenitor population of the luminal compartment (Supplementary Fig 7).}
 Together, the data suggest that luminal progenitor cells maintain memory of having undergone gestation and involution.
@@
 We have reported here the use of single cell RNA sequencing to comprehensively map the transcriptomes of thousands of mammary epithelial cells across four developmental time-points.
-Our analysis identified 9 clusters of epithelial cells, some of which are only present during specific developmental stage (e.g. Avd and Lp-PI).
+Our analysis identified 8 clusters of epithelial cells, some of which are only present during specific developmental stage (e.g. Avd and Lp-PI).
 This study provides a rich dataset that can be mined online (see link in materials and methods) to identify marker genes and lineage specific promoters that can be used to trace populations of cells \textit{in vivo}.
@@
 In this study we could not provide any evidence for contribution of a putative mutlipotent stem cell to the day-to-day homoeostasis of the gland.
-\fixme{We do note, however, that the cluster Prc is placed between the luminal and basal cells in \autoref{F2}a.
-Prc represents the cluster with the fewest numbers of cells captured and expresses low levels of both luminal and basal markers, high levels of the stem cell marker \textit{Procr}\autocite{Wang2015} and high levels of the luminal progenitor marker \textit{Notch3}\autocite{Lafkas2013} (\autoref{F1}d).
-Yet, we did not capture cells in transition between Prc and any other clusters suggesting that if these cells are indeed, multi-potent progenitors their involvement in tissue homoeostasis is negligible.
-Alternatively, we cannot exclude the possibility that Prc is enriched for doublets.}
+We do note, however, that the cluster Prc epresents the cluster with the fewest numbers of cells captured and expresses low levels of both luminal and basal markers, high levels of the stem cell marker \textit{Procr}\autocite{Wang2015} and high levels of the luminal progenitor marker \textit{Notch3}\autocite{Lafkas2013} (\autoref{F1}d).
+Yet, our data does not support a central role of this cluster in the day-to-day homeostasis of the gland. 
+In addition, cells from the Prc cluster also express some but not all markers of pericytes and we cannot thus not exclude that these are contaminating non-epithelial cells. 
 Based on the gene expression data presented here, the luminal compartment appears to have one common progenitor population (Lp).
@@
 \textbf{Table 1. Summary of mammary epithelial cell clusters}
-\fakefigure{T1}
+\faketable{T1}
 Overview of the different clusters including number of cells captured for each time-point and key genes that were used to infer their identities.

### 2017-08-11 src/Figure2.R

@@ -96,7 +96,7 @@ m <- m[,pD$barcode]
 m.norm <- t(t(m)/pD$sf)
 
 # Genes to plot for trends
-genes <- c("Aldh1a3","Csn2","Glycam1","Pgr","Esr1")
+genes <- c("Aldh1a3","Csn2","Glycam1","Pgr","Cited1")
 rownames(m.norm) <- fD$symbol
 exps <- log2(m.norm[genes,]+1)
 exps <- t(exps/apply(exps,1,max))

### 2017-08-11 src/Figure3.R

@@ -230,10 +230,13 @@ branches <- plot_grid(NULL,pb1,pb2,NULL,rel_widths=c(0.3,1,1,0.3),nrow=1,labels=
 expPlot <- plot_grid(branches,expPlot,ncol=1,rel_heights=c(0.3,1))
 
 htmps <- plot_grid(res[[1]][[1]][[4]],
+		   NULL,
 		   res[[2]][[1]][[4]],
+		   NULL,
 		   nrow=2,
 		   scale=0.8,
-		   labels=c("e","f"))
+		   labels=c("e","","f",""),
+		   rel_widths=c(1,0.75,1,0.75))
 fullP <- plot_grid(expPlot,htmps,ncol=2)
 
 

### 2017-08-11 src/Figure4.R

@@ -44,9 +44,25 @@ volcano <- ggplot(topTab,aes(x=logFC,y=-log10(FDR))) +
     geom_point(data=interest, aes(x=logFC, y=-log10(FDR)), size=3, color="black", pch=20) +
     geom_point(data=lacinterest, aes(x=logFC, y=-log10(FDR)), size=3, color="dodgerblue", pch=20) +
     geom_point(data=imminterest, aes(x=logFC, y=-log10(FDR)), size=3, color="coral", pch=20) +
-    geom_label_repel(data=interest, aes(x=logFC, y=-log10(FDR), label=symbol)) +
+    geom_text_repel(data=interest, aes(x=logFC, y=-log10(FDR), label=symbol)) +
     xlab("Log2 Fold Change") +
     ylab("-Log10(P value)") 
+
+# ---- ProgenitorvsLuminal ----
+progenitorDE <- read.csv("../data/Robjects/secondRun_2500/ProgenitorDE.csv")
+# genes to highlight
+interest <- filter(progenitorDE, Gene %in% c("Aldh1a3", "Lypd3", "Prlr", "Esr1", "Pgr"))
+
+# FC plot
+p2 <- ggplot(progenitorDE, aes(x=NullParFC,y=ParousFC)) +
+    geom_point(color="grey50",size=2) +
+    geom_point(data=interest, aes(x=NullParFC, y=ParousFC), color="black") +
+    geom_text_repel(data=interest, aes(x=NullParFC,y=ParousFC,label=Gene)) +
+    xlab("LFC of Lp-NP vs. luminal cells") +
+    ylab("LFC of Lp-PI vs. luminal cells") +
+    geom_hline(yintercept=0, lty="dashed") +
+    geom_vline(xintercept=0, lty="dashed") +
+    coord_equal(xlim=c(-5,5),ylim=c(-5,5))
 
 # ---- DEstrongerInProgenitor ----
 
@@ -85,19 +101,20 @@ for (i in c(1,2)) {
 	geom_jitter(size=0.9) +
 	stat_summary(fun.y = mean, fun.ymin = mean, fun.ymax = mean,
 		     geom = "crossbar", width = 1,color="black") +
-	facet_grid(~variable) +
+	facet_wrap(~variable,scales="free_y",nrow=1) +
 	scale_color_brewer(type="qual",palette="Paired") +
-	ylab("Expression") +
+	ylab("") +
 	xlab("") +
[…]

### 2017-08-11 src/S8.R

@@ -1,51 +1,0 @@
-# Figure S8
-
-library(scran)
-library(dplyr)
-library(ggplot2)
-library(cowplot)
-library(Rtsne)
-source("functions.R")
-
-# Load Data
-dataList <- readRDS("../data/Robjects/secondRun_2500/ExpressionList_Clustered.rds")
-m <- dataList[[1]]
-pD <- dataList[[2]]
-fD <- dataList[[3]]
-rnd_seed <- 300
-
-# Subset to lactation sample
-pD.lac <- filter(pD, Condition=="L")
-cells <- as.character(pD.lac$barcode)
-m.lac.all <- m[,cells]
-keep <- rowMeans(m.lac.all) > 0.1
-m.lac <- m.lac.all[keep,]
-fD.lac <- fD[keep,]
-
-# Normalize
-clusters.lac <- quickCluster(m.lac)
-pD.lac$sf <- computeSumFactors(m.lac,clusters=clusters.lac,positive=FALSE)
-m.lac.norm <- t(t(m.lac)/pD.lac$sf)
-
-# Compute tSNE
-fPCA.lac <- log2(t(m.lac.norm)+1)
-fPCA.lac <- scale(fPCA.lac,scale=TRUE,center=TRUE)
-set.seed(rnd_seed)
-tsn.lac <- Rtsne(fPCA.lac,perplexity=25)
-
-pD.lac$tSNE1 <- tsn.lac$Y[,1]
-pD.lac$tSNE2 <- tsn.lac$Y[,2]
-
-# Plot
-pD.lac <- arrange(pD.lac, PassAll) %>%
-    rename(QCpass=PassAll) %>%
-    mutate(QCpass=ifelse(QCpass,"Pass","Fail"))
-tsnPlot <- ggplot(pD.lac, aes(x=tSNE1, y=tSNE2,color=QCpass)) +
-    geom_point(alpha=0.8,size=3) +
-    scale_color_manual(values=c("grey","black")) +
-    theme(legend.title=element_blank())
-
-# cairo_pdf("../paper/figures/S8.pdf")
-tsnPlot
-# dev.off()
-

### 2017-08-12 paper/main.tex
Updates on main text and figures for going back to SubCluster

@@
 
-\textbf{Single cell RNA sequencing identifies 8 clusters of mammary epithelial cells}
+\textbf{Single cell RNA sequencing identifies 15 clusters of mammary epithelial cells}
 
@@
 In a second step we used hierarchical clustering on each of the identified groups to further resolve the cellular heterogeneity (\autoref{F1}c, SupX).
+After removal of enodthelial and immune cell clusters this resulted in a total of 15 clusters of mammary epithelial cells.
 Differential gene expression analysis allowed us to further characterise the clusters and infer putative identities (\autoref{F1}d, \autoref{T1} and Supplementary Table 1).
-Notably, some clusters represent the same putative cell type at different developmental time points (indicated by the NP, G, L or PI suffix) and were thus grouped together (\autoref{F1}c, \autoref{F1}d).
-Based on the expression of \textit{Krt18,} \textit{Krt8, Krt5, Krt14} and \textit{Acta2} we noted that 5 clusters show a luminal profile (Lp, Hsp, Hsd, Avp \& Avd) and 3 a basal profile (Bsl, Myo \& Prc) (\autoref{F1}D, Supplementary Fig 2b and 2c).
-Avd and Avp express high levels of beta-casein, \textit{Csn2}, and are exclusively composed of cells from the gestation and lactation time-points, suggesting that these cells are secretory alveolar cells (\autoref{F1}d).
+Notably, some clusters mainly represent the same putative cell type at different developmental time points (indicated by the NP, G, L or PI suffix, in the following the suffix might be omitted to refer to a group of clusters).
+Based on the expression of \textit{Krt18,} \textit{Krt8, Krt5, Krt14} and \textit{Acta2} we noted that 11 clusters show a luminal profile (Lp, Hsp, Hsd, Avp \& Avd clusters) and 4 a basal profile (Bsl, Myo \& Prc clusters) (\autoref{F1}D, Supplementary Fig 2b and 2c).
+The Avd and Avp clusters express high levels of beta-casein, \textit{Csn2}, and are exclusively composed of cells from the gestation and lactation time-points, suggesting that these cells are secretory alveolar cells (\autoref{F1}d).
 Hsd and Hsp on the other hand represent the clusters with the highest expression of hormone receptors, suggesting hormone sensing function.
@@
 Being confident that the diffusion map recapitulates the luminal differentiation process, we then computationally inferred the two branches and ordered the cells according to their progression through ``pseudotime''\autocite{Haghverdi2016} (see materials and methods for further explanation) (\autoref{F3}a).
+This allows to identify genes whose expression changes during the process of differentiation.
 We found 456 genes that showed a pseudotime-dependent expression with the same directionality along both differentiation trajectories (Supplementary Table 2).
@@
 The early cluster is enriched for genes involved in fatty-acid oxidation and lipid biosynthetic processes (e.g. \textit{Lipa}, \textit{Dbi} and \textit{Acox1}) .
-The second, smaller group of genes that is switched on during the late stages of differentiation was enriched for genes involved in fatty-acid transport and lipid homeostasis (including e.g.
-\textit{Olah}, \textit{Cd36}, \textit{Wap} and \textit{Lalba}).
+The second, smaller group of genes that is switched on during the late stages of differentiation was enriched for genes involved in fatty-acid transport and lipid homeostasis (e.g.
+\textit{Olah}, \textit{Cd36})
 Genes involved in cell division and translation show a transient phase of up-regulation during the alveolar differentiation process (e.g.
@@
 During the process of differentiation towards hormone-sensing luminal cells, we identified two broad clusters of genes that gradually increase or decrease their expression.
-Genes with increasing expression are involved in hormone metabolic processes and regulation of morphogenesis of an epithelium (e.g. \textit{Vdr}, \textit{Esr1}, \textit{Pgr} and \textit{Msx2}).
+Genes with increasing expression are involved in hormone metabolic processes and the regulation of morphogenesis of an epithelium (e.g. \textit{Vdr}, \textit{Esr1}, \textit{Pgr} and \textit{Msx2}).
 Amongst the branch-specific transcription factors we found for example \textit{Runx1, Tox2} and \textit{Bhlhe41} (also known as \textit{Sharp1}) to be transcribed during differentiation towards the hormone-sensing lineage (\autoref{F3}d) and \textit{Elf5, Foxs1} and \textit{Ehf} in the secretory lineage (\autoref{F3}e).
@@
 However, we found that in particular the luminal compartment differs from its nulliparous counterpart (\autoref{F1}b and \autoref{F1}d).
-Interestingly, the effect of parity is most pronounced for the less differentiated and more progenitor-like cell types (Lp-PI, \autoref{F1}c, Supplementary Fig 6a).
+Interestingly, the effect of parity is most pronounced for the less differentiated and more progenitor-like cell types (\autoref{F1}c, Supplementary Fig 6a).
 To ensure that Lp-PI still represents the progenitor population in the post-parous gland, we identified genes that distinguish Lp-NP from the rest of the luminal compartment to see if they are also characteristic for the proposed post-parity progenitor population Lp-PI.
@@
 milk-proteins (\textit{Csn2}, \textit{Lalba}), lipases (\textit{Lipa}), proteins involved in milk secretion (\textit{Xdh}, \textit{Cd36}) and transcriptional regulators of lactation (\textit{Cidea}).
-Notably, some of the genes that are involved in the immune response are known regulators of the involution process such as \textit{Ctsc}, \textit{Tgfb3} or \textit{Mfge8}.
+Of note, the genes of the casein locus (\textit{Csn2}, \textit{Csn1s1}, \textit{Csn1s2a}, \textit{Csn3}) have previously been reported to be up-regulated in the parous gland, most likely due to changes in chromatin accessibility\autocite{Dos2015,Rijnkels2013}.
+However, it has not been shown before that this effect is confined to the progenitor population of the luminal compartment.
+Some of the genes that are involved in the immune response are known regulators of the involution process such as \textit{Ctsc}, \textit{Tgfb3} or \textit{Mfge8}.
 The up-regulation of these genesets was also present in the Hsp and in some cases even the differentiated Hsd cluster, yet the effect remained strongest in the progenitor cells (\autoref{F4}c).
@@
 Interestingly, the Hsp and Hsd cluster generally maintain their position of their nulliparous counterpart in the differentiation hierarchy whereas the Lp-PI cluster is stretched out from the origin down towards the alveolar branch, suggesting that these are biased towards the alveolar fate (\autoref{F4}b).
-\comment{Doesn't fit: Of note, the genes of the casein locus (\textit{Csn2}, \textit{Csn1s1}, \textit{Csn1s2a}, \textit{Csn3}) have previously been reported to be up-regulated in the parous gland, most likely due to changes in chromatin accessibility\autocite{Dos2015,Rijnkels2013}.
-However, it has not been shown before that this effect is confined to the progenitor population of the luminal compartment (Supplementary Fig 7).}
 Together, the data suggest that luminal progenitor cells maintain memory of having undergone gestation and involution.
@@
 %---------------------------------------------Figure 1--------------------------------------------------------------
-\textbf{Figure 1. Single cell RNA sequencing identifies 9 clusters of
-mammary epithelial cells }
+\textbf{Figure 1. Single cell RNA sequencing identifies 15 clusters of mammary epithelial cells }
 \fakefigure{F1}
 \textbf{(a)} Schematic diagram highlighting the experimental setup for isolating and sequencing the RNA of single cells using the 10X chromium system.
-\textbf{(b)} t-SNE plot of 3,471 cells visualizes general structure in the data.
+\textbf{(b)} t-SNE plot of 25,010 cells visualizes general structure in the data.
 Cells are coloured by the four developmental time points Pink=NP, Dark Green=G, Light Green=L, Purple=PI.
-\textbf{(c)} Same as \textbf{(b)} but colouring cells by clusters.
-\textbf{(d)} Heatmap highlighting some key marker genes.
+\textbf{(c)} Dendrogram of clusters based on the mean expression values of the 15 clusters. The tree was computed based on spearman correlation with ward linkage.
+\textbf{(d)} Same as \textbf{(b)} but colouring cells by clusters.
+\textbf{(e)} Heatmap highlighting some key marker genes.
 Colour scale represents log-transformed and normalized UMI counts scaled to a maximum of 1 per row.
@@
 Upper bars represent the cluster assignment and stages for the individual cells.
-For visualization purposes only 100 randomly selected cells were shown for each of the large clusters.
-The list of differentially expressed genes between clusters can be found in Supplementary Table 1.
+For visualization purposes only 100 randomly selected cells were shown for large clusters.
 %-------------------------------------------------------------------------------------------------------------------
@@
 %---------------------------------------------Figure 2--------------------------------------------------------------
-\textbf{Figure 2. Computational reconstruction of differentiation
-processes in the mammary gland}
+\textbf{Figure 2. Computational reconstruction of differentiation processes in the mammary gland}
 \fakefigure{F2}
@@
 \textbf{(b)} Differentiation trajectory of the luminal compartment based on the first two diffusion components.
-The right panel depicts the log-transformed normalized expression levels of the luminal progenitor marker \textit{Aldh1a3} in all cells along component 2, the bottom panel shows the levels of \textit{Csn2} and \textit{Pgr} along component 1.
-Expression trends represent natural cubic spline fits with five degrees of freedom.
+\textbf{(c)} Same plot as in \textbf{(b)}, colored by the normalized and scaled expression values of various genes.
 %-------------------------------------------------------------------------------------------------------------------
@@
 %---------------------------------------------Figure 3--------------------------------------------------------------
-\textbf{Figure 3. Pseudotime ordering identifies genes associated with
-luminal differentiation}
+\textbf{Figure 3. Pseudotime ordering identifies genes associated with luminal differentiation}
 \fakefigure{F3}
@@
 Cells are coloured by their progression through pseudotime, where low values represent undifferentiated cells.
-\textbf{(b)} Heatmap of the top 50 pseudotime-dependent genes (rows) with the same overall trend along both branches (top panel) or branch-specific trends (bottom panel, see materials and methods for definition).
-The black vertical line represents the common origin from the progenitor population.
-Arrows highlight differentiation direction.
+\textbf{(b-d)} Examples of transcription factors with pseudotime-dependent expression with the same overall trend on both branches \textbf{(c)} or branch specific trends \textbf{(d,e)}.
+\textbf{(e-f)} Heatmap of all genes with branch specific, pseudotime-dependent expression for the hormone-sensing lineage \textbf{(e)}a or the secretory lineage \textbf{(f)}.
 Pseudotime and the cluster assignment are annotated above the heatmap.
 The values in the heatmap represent z-scaled, spline-smoothed expression values.
-\textbf{(c-e)} Examples of transcription factors with pseudotime-dependent expression with the same overall trend on both branches \textbf{(c)} or branch specific trends \textbf{(d,e)}.
+Genes in the heatmaps were clustered using hierarchical clustering with a dynamic tree cut.
 %-------------------------------------------------------------------------------------------------------------------
@@
 %---------------------------------------------Figure 4--------------------------------------------------------------
-\textbf{Figure 4. The effect of parity on the transcriptomic landscape
-of the luminal progenitor compartment}
+\textbf{Figure 4. The effect of parity on the transcriptomic landscape of the luminal progenitor compartment}
 \fakefigure{F4}
-\textbf{(a)} Colour-coded similarity matrix between the log-transformed mean expression values of the 9 epithelial cell clusters.
-Similarities are based on Euclidean distances.
-\textbf{(b)} Comparison of fold changes from Lp versus luminal compartment and fold changes from Lp-PI versus the luminal cells.
-The genes represent the top 500 differentially expressed genes between Lp and luminal cells.
-\textbf{(c)} Volcano plot illustrates differential expression between Lp and Lp-PI, coloured dots represent significant genes with known function in lactation and immunity, dashed lines highlight the P value threshold of 0.01 and a log fold change of 1.
+\textbf{(a)} Comparison of fold changes from Lp-NP versus luminal compartment and fold changes from Lp-PI versus the luminal cells.
+The genes represent the top 500 differentially expressed genes between Lp-NP and luminal cells.
+\textbf{(c)} Volcano plot illustrates differential expression between Lp-NP and Lp-PI, coloured dots represent significant genes with known function in lactation and immunity, dashed lines highlight the P value threshold of 0.01 and a log fold change of 1.
 P values are adjusted for multiple testing using Benjamini-Hochberg.
-\textbf{(d)} Visualization of expression difference for some lactation related genes from \textbf{(c)}, where Lp-PI is split according to the developmental time point.
+\textbf{(c-d)} Visualization of expression difference for genes related to lactation \textbf{(c)} or immunity \textbf{(d)} for the six luminal clusters in NP and PI.
+Expression values correspond to normalized UMI counts.
 %-------------------------------------------------------------------------------------------------------------------

### 2017-08-12 src/DiffusionMap.R

@@ -22,7 +22,7 @@ cells <- list("all","luminal")
 
 for (cell in cells) {
     if(cell=="luminal") {
-	excludeClustComb <- c("Bsl-G1","Bsl","Myo","Prc")
+	excludeClustComb <- c("Bsl-G","Bsl","Myo","Prc")
     } else{
 	excludeClustComb <- NULL
     }

### 2017-08-12 src/Figure1.R

@@ -42,17 +42,15 @@ p0 <- ggplot(pD, aes(x=tSNE1, y=tSNE2, color=Condition)) +
 	  legend.title=element_blank()) 
 
 # t-SNE colored by cluster
-p1 <- ggplot(pD, aes(x=tSNE1, y=tSNE2, color=SuperCluster)) +
+p1 <- ggplot(pD, aes(x=tSNE1, y=tSNE2, color=SubCluster)) +
     geom_point(size=1.5) +
-    scale_color_manual(values=levels(pD$SuperColors))+
+    scale_color_manual(values=levels(pD$Colors))+
     #     ggtitle("Cluster") +
     theme_void(base_size=12) +
     guides(colour = guide_legend(override.aes = list(size=3))) +
     theme(legend.position="bottom",legend.direction="horizontal",
 	  legend.title=element_blank())  
 #     facet_wrap(~Condition)
-
-plot_grid(p0,NULL,p1,align="h",nrow=1,rel_widths=c(1,0.2,1))
 
 # ---- Dendrogram ----
 pD$SubCluster <- factor(pD$SubCluster) #drop unused levels
@@ -72,7 +70,7 @@ dendr <- dis %>% hclust(.,method="ward.D2") %>% as.dendrogram %>%
     set("leaves_pch",19) %>%
     set("leaves_cex",2) %>%
     set("branches_lwd",3)
-par(cex=.9)
+par(cex=.8)
 plot(dendr,horiz=TRUE,yaxt="n")
 p.dendr <- grab_grob()
 p.dendr <- grid.arrange(p.dendr)
@@ -97,12 +95,16 @@ genes <- c(general,c1,c3,c5,c2,c6,c7,c9)
 
 # Subsample cells from large clusters
 set.seed(rnd_seed)
-ord <- group_by(pD, SuperCluster) %>%
-    do(sample_n(.,200)) %>%
-    ungroup() %>%
-    mutate(SuperCluster=factor(SuperCluster,levels=c("Hsd","Hsp","Lp","Avp","Avd","Bsl","Myo","Prc"))) %>%
-    arrange(SuperCluster,Condition) 
+subsP <- filter(pD, !SubCluster %in% c("Hsd-G","Avd-L")) %>%
+    group_by(SubCluster) %>%
+    do(sample_n(.,100))
 
+# Combine with remaining clusters and relevel factor according to order in plot
+ord <- filter(pD, SubCluster %in% c("Hsd-G","Avd-L")) %>%
[…]

### 2017-08-12 src/Figure2.R

@@ -21,7 +21,7 @@ pD <- right_join(pD,dms,by="barcode")
 # Plot for luminal and basal cells
 set.seed(300)
 pD <- pD[sample(1:nrow(pD),nrow(pD)),]
-cols <- pD$SuperColor
+cols <- pD$Colors
 
 scatterplot3d(x=pD[,"DC1"],
 	      y=-pD[,"DC2"],
@@ -42,28 +42,28 @@ g <- grid.arrange(g)
 dev.off()
 
 # Create the alternative view inlet
-inlet <- ggplot(pD, aes(DC1,DC2,color=SuperCluster)) +
+inlet <- ggplot(pD, aes(DC1,DC2,color=SubCluster)) +
     geom_point(size=2,pch=20) +
     scale_color_manual(values=levels(cols))+
     xlab("Component 1") +
     ylab("-Component 2") 
 
-library(ggExtra)
-ggMarginal(inlet, type="histogram",margins="x",fill="white",bins=30)
+# library(ggExtra)
+# ggMarginal(inlet, type="histogram",margins="x",fill="white",bins=30)
 # cairo_pdf("../paper/figures/F2inlet.pdf",width=17.54,height=17.54)
-inlet
+# inlet
 # dev.off()
 
 # Create legend for luminal only and luminal/basal plot
 pal <- levels(cols)
 # forcLeg <- filter(pD, !cluster %in% c(4))
-clustLeg <- ggplot(pD, aes(x=tSNE1,y=tSNE2,color=SuperCluster)) +
+clustLeg <- ggplot(pD, aes(x=tSNE1,y=tSNE2,color=SubCluster)) +
     geom_point() +
     scale_color_manual(values=pal) +
     theme(legend.direction="horizontal",
 	  legend.title=element_blank(),
 	  legend.position="bottom")+
-    guides(colour=guide_legend(nrow=1,
+    guides(colour=guide_legend(nrow=2,
 			       override.aes=list(size=3)))
 clustLeg <- get_legend(clustLeg)
 
@@ -72,13 +72,13 @@ dms <- read.csv("../data/Robjects/secondRun_2500/dm_luminal.csv")
 pD <- dataList[[2]]
 pD <- right_join(pD,dms,by="barcode")
 
-cols <- levels(pD$SuperColors)
+cols <- levels(pD$Colors)
 
[…]

### 2017-08-12 src/Figure3.R

@@ -75,14 +75,14 @@ for (i in c(1,2)) {
     pD.ord <- pD
     rownames(pD.ord) <- pD.ord$barcode
     pD.ord <- pD.ord[colnames(m.heat),]
-    annoCol <- data.frame("Cluster"=factor(pD.ord$SuperCluster),
+    annoCol <- data.frame("Cluster"=factor(pD.ord$SubCluster),
 			  "Pseudotime"=rank(pD.ord$dpt,ties.method="first")
 			  )
     rownames(annoCol) <- rownames(pD.ord)
 
     #Set colorscheme for heatmap
-    clustCol <- levels(factor(pD.ord$SuperColors))
-    names(clustCol) <- levels(factor(pD.ord$SuperCluster))
+    clustCol <- levels(factor(pD.ord$Colors))
+    names(clustCol) <- levels(factor(pD.ord$SubCluster))
     dptcols <- viridis(n=nrow(annoCol),,option="magma",begin=1,end=0)
     names(dptcols) <- c(1:length(dptcols))
     annoColors <- list("Cluster"=clustCol,
@@ -174,16 +174,16 @@ fplot2 <- join(fplot2,raw2,by="barcode") %>%
 
 
 #set colorscale
-clustCol <- levels(factor(pD$SuperColors))
+clustCol <- levels(fplot2$Colors)[levels(fplot2$SubCluster)[c(-12,-17,-18,-19,-20)] %in% unique(c(as.character(fplot2$SubCluster),as.character(fplot1$SubCluster)))]
 
+fplot3 <- rbind(fplot1,fplot2)
 #Plot dpt-dependent expression for features
 pList <- list()
 for (feature in features) {
     pnts <- ggname(paste0("raw",feature))
     lns <- ggname(feature)
     p <- ggplot() +
-	geom_point(size=0.8,data=fplot1,aes_string(x="dptNorm",y=pnts, color="SuperCluster")) +
-	geom_point(size=0.8,data=fplot2,aes_string(x="dptNorm",y=pnts, color="SuperCluster")) +
+	geom_point(size=0.8,data=fplot3,aes_string(x="dptNorm",y=pnts, color="SubCluster")) +
 	geom_line(data=fplot1,aes_string(x="dptNorm",y=lns),lty="dashed") +
 	geom_line(data=fplot2,aes_string(x="dptNorm",y=lns)) +
 	ggtitle(feature) +
[…]

### 2017-08-12 src/Figure4.R

@@ -44,7 +44,7 @@ volcano <- ggplot(topTab,aes(x=logFC,y=-log10(FDR))) +
     geom_point(data=interest, aes(x=logFC, y=-log10(FDR)), size=3, color="black", pch=20) +
     geom_point(data=lacinterest, aes(x=logFC, y=-log10(FDR)), size=3, color="dodgerblue", pch=20) +
     geom_point(data=imminterest, aes(x=logFC, y=-log10(FDR)), size=3, color="coral", pch=20) +
-    geom_text_repel(data=interest, aes(x=logFC, y=-log10(FDR), label=symbol)) +
+    geom_text_repel(data=interest, aes(x=logFC, y=-log10(FDR), label=symbol, fontface="italic")) +
     xlab("Log2 Fold Change") +
     ylab("-Log10(P value)") 
 
@@ -57,7 +57,7 @@ interest <- filter(progenitorDE, Gene %in% c("Aldh1a3", "Lypd3", "Prlr", "Esr1",
 p2 <- ggplot(progenitorDE, aes(x=NullParFC,y=ParousFC)) +
     geom_point(color="grey50",size=2) +
     geom_point(data=interest, aes(x=NullParFC, y=ParousFC), color="black") +
-    geom_text_repel(data=interest, aes(x=NullParFC,y=ParousFC,label=Gene)) +
+    geom_text_repel(data=interest, aes(x=NullParFC,y=ParousFC,label=Gene,fontface="italic")) +
     xlab("LFC of Lp-NP vs. luminal cells") +
     ylab("LFC of Lp-PI vs. luminal cells") +
     geom_hline(yintercept=0, lty="dashed") +
@@ -88,12 +88,12 @@ for (i in c(1,2)) {
     #Create DF
     forPl <- data.frame(t(m)[,c(genes)]+1,
 	    barcode=colnames(m))
-    add <- select(pD, barcode, SubCluster, SuperCluster, Condition) %>%
+    add <- select(pD, barcode, SubCluster, SuperCluster, Condition, Colors) %>%
 	mutate(barcode=as.character(barcode))
     forPl <- left_join(forPl,add,by="barcode") %>%
 	filter(SubCluster %in% c("Hsd-NP","Hsd-PI","Hsp-NP","Hsp-PI","Lp-NP","Lp-PI"))
 
[…]

### 2017-08-12 paper/main.tex
all supps done

@@
 To this end we herein use single cell RNA sequencing (scRNAseq) to map the cellular dynamics of mammary epithelial cells across four adult developmental stages; nulliparous, mid gestation, lactation and post weaning (full natural involution).
-Our data from 25,010 individual cells, identifies 8 distinct cell populations within the gland and allows their hierarchical structure across developmental time points to be charted.
+Our data from 25,010 individual cells, identifies 15 distinct cell populations within the gland and allows their hierarchical structure across developmental time points to be charted.
 
@@
 
-First, we isolated mammary epithelial cells from four developmental time points; nulliparous (NP), day 14.5 gestation (G), day 6 lactation (L) and two weeks post natural involution (PI).
+First, we isolated mammary epithelial cells from four developmental time points; nulliparous (NP), day 14.5 gestation (G), day 6 lactation (L) and two weeks post natural involution (PI) (Supplementary Figure 2).
 For each time point we sorted mammary epithelial cells based on the Epcam cell surface marker from two independent mice (\autoref{F1}a).
 All samples were then prepared for single-cell RNA-sequencing using the 10x Chromium platform\autocite{Zheng2017}.
-Following quality control (materials and methods), this yielded an average of 6,175 unique transcripts and 2,118 genes detected from 25,010 cells (4,223 in NP, 5,826 in G, 9,319 in L and 5,642 in P) (Supplementary \autoref{F1}a).
-Visual inspection of the data using t-distributed stochastic neighbour embedding (t-SNE) suggested that although there is grouping of cells by time-point, there are also other factors that underlie structure within the dataset (\autoref{F1}b and Supplementary Figure 2a).
-First, we dissected the global structure in the dataset by unsupervised clustering using a shared-nearest-neighbour clustering approach.
-This resulted in a coarse clustering into 13 groups (SupX).
-In a second step we used hierarchical clustering on each of the identified groups to further resolve the cellular heterogeneity (\autoref{F1}c, SupX).
+Following quality control (materials and methods), this yielded an average of 6,175 unique transcripts and 2,118 genes detected from 25,010 cells (4,223 in NP, 5,826 in G, 9,319 in L and 5,642 in P) (Supplementary 3).
+Visual inspection of the data using t-distributed stochastic neighbor embedding (t-SNE) suggested that although there is grouping of cells by time-point, there are also other factors that underlie structure within the dataset (\autoref{F1}b and Supplementary Figure 4).
+First, we dissected the global structure in the dataset by unsupervised clustering using a shared-nearest-neighbor clustering approach.
+This resulted in a coarse clustering into 13 groups (Supplementary Figure 5a).
+In a second step we used hierarchical clustering on each of the identified groups to further resolve the cellular heterogeneity (\autoref{F1}c, Supplementary Figure 5b).
 After removal of enodthelial and immune cell clusters this resulted in a total of 15 clusters of mammary epithelial cells.
@@
 In contrast, the luminal compartment showed a distinct structure, with gradual transitions between different clusters and cells originating from a common origin (\autoref{F2}b).
-We confirmed the robustness of this bifurcation by verifying that it was present when different methods of feature selection, algorithms for trajectory inference and down-sampling were employed (Supplementary Fig.
-3).
+We confirmed the robustness of this bifurcation by verifying that it was present when different methods of feature selection, algorithms for trajectory inference and down-sampling were employed (Supplementary Figure 6).
 
@@
 In addition, we identified 1005 genes with branch-specific expression patterns (\autoref{F3}d, \autoref{F3}e Supplementary Table 3).
-We further clustered the gene expression trend for these on each of the two branches and identified sets of genes that change over pseudotime.
+We further clustered the gene expression trend for these on each of the two branches and identified sets of genes that change over pseudotime (Supplementary Figure 7).
 On the alveolar branch we found two clusters of genes that increase in expression during differentiation, one that is switched on early and another that is activated later in the final stages of differentiation.
@@
 However, we found that in particular the luminal compartment differs from its nulliparous counterpart (\autoref{F1}b and \autoref{F1}d).
-Interestingly, the effect of parity is most pronounced for the less differentiated and more progenitor-like cell types (\autoref{F1}c, Supplementary Fig 6a).
+Interestingly, the effect of parity is most pronounced for the less differentiated and more progenitor-like cell types (\autoref{F1}c, Supplementary Figure 8).
 To ensure that Lp-PI still represents the progenitor population in the post-parous gland, we identified genes that distinguish Lp-NP from the rest of the luminal compartment to see if they are also characteristic for the proposed post-parity progenitor population Lp-PI.
 Indeed, we find genes that are differentially expressed between Lp-NP and the rest of the luminal compartment to show the same trend between Lp-PI and the luminal compartment (\autoref{F4}a).
-In a similar manner we can distinguish Lp-PI from the rest of the PI gland in a principal component analysis (PCA) using the identified progenitor genes (Supplementary Fig 6c and d).
+In a similar manner we can distinguish Lp-PI from the rest of the PI gland in a principal component analysis (PCA) using the identified progenitor genes (\textbf{Supplementary Fig 8c and d}).
 From this we conclude, that Lp-PI indeed represents the post-parity luminal progenitor population.
-Interestingly, genes that were up-regulated in Lp-PI compared to Lp-NP were significantly enriched for pathways that are involved in the immune response and lactation (\autoref{F4}b, Supplementary Fig 6c).
+Interestingly, genes that were up-regulated in Lp-PI compared to Lp-NP were significantly enriched for pathways that are involved in the immune response and lactation (\autoref{F4}b, \textbf{Supplementary Fig 8c}).
 These included genes that play roles in various processes during lactation, e.g.
@@
 We have reported here the use of single cell RNA sequencing to comprehensively map the transcriptomes of thousands of mammary epithelial cells across four developmental time-points.
-Our analysis identified 8 clusters of epithelial cells, some of which are only present during specific developmental stage (e.g. Avd and Lp-PI).
+Our analysis identified 15 clusters of epithelial cells, some of which are only present during specific developmental stage (e.g. Avd and Lp-PI).
 This study provides a rich dataset that can be mined online (see link in materials and methods) to identify marker genes and lineage specific promoters that can be used to trace populations of cells \textit{in vivo}.
@@
 In this study we could not provide any evidence for contribution of a putative mutlipotent stem cell to the day-to-day homoeostasis of the gland.
-We do note, however, that the cluster Prc epresents the cluster with the fewest numbers of cells captured and expresses low levels of both luminal and basal markers, high levels of the stem cell marker \textit{Procr}\autocite{Wang2015} and high levels of the luminal progenitor marker \textit{Notch3}\autocite{Lafkas2013} (\autoref{F1}d).
+We do note, however, that the cluster Prc represents the cluster with the fewest numbers of cells captured and expresses low levels of both luminal and basal markers, high levels of the stem cell marker \textit{Procr}\autocite{Wang2015} and high levels of the luminal progenitor marker \textit{Notch3}\autocite{Lafkas2013} (\autoref{F1}d).
 Yet, our data does not support a central role of this cluster in the day-to-day homeostasis of the gland. 
-In addition, cells from the Prc cluster also express some but not all markers of pericytes and we cannot thus not exclude that these are contaminating non-epithelial cells. 
+In addition, cells from the Prc cluster also express some but not all markers of pericytes and we thus cannot exclude that these are contaminating non-epithelial cells. 
 Based on the gene expression data presented here, the luminal compartment appears to have one common progenitor population (Lp).
-We also identified a previously described progenitor population with expression of hormone-receptor related genes (Hsp) that appears to represent a meta-stable state on the differentiation path from luminal progenitors towards fully differentiated hormone-sensing cells (Hsd).
-Furthermore, we characterised gene expression patterns along the differentiation hierarchy, thus enabling to disentangle the cellular hierarchy in the mammary gland.
-By analysing the mammary gland at various stages of development we were also able to describe the molecular effects of parity at cellular resolution.
+Lp gives rise to intermediate states of either hormone-sensing (Hsp) or secretory (Avp) cells that express differentiation markers as well as progenitor markers and appear to represent meta-stable states on the differentiation path from luminal progenitors towards fully differentiated cells (Hsd).
+Furthermore, we characterised gene expression patterns along the differentiation hierarchy.
+Here we identified genes that are lineage-specific, thus enabling to disentangle the transcriptional events that regulate differentiation of the luminal compartment.
+By analysing the mammary gland at various stages of development we were also able to describe the effects of parity at cellular resolution.
 We found that the luminal progenitor compartment undergoes lasting changes at the transcriptional level.
@@
 The majority of the changes were related to pathways involved in immunity and lactation, suggesting that, in particular, the luminal progenitors maintain a memory of gestation and involution.
-It is reasonable to assume that Lp-PI overlaps with the previously parity-induced mammary epithelial cells (PI-MECs)\autocite{Wagner2002}.
+It is reasonable to assume that Lp-PI overlaps with the previously described parity-induced mammary epithelial cells (PI-MECs)\autocite{Wagner2002}.
 
@@
 Cells are coloured by the four developmental time points Pink=NP, Dark Green=G, Light Green=L, Purple=PI.
-\textbf{(c)} Dendrogram of clusters based on the mean expression values of the 15 clusters. The tree was computed based on spearman correlation with ward linkage.
+\textbf{(c)} Dendrogram of clusters based on the mean expression values of the 15 clusters. The tree was computed based on Spearman's rank correlation with ward linkage.
 \textbf{(d)} Same as \textbf{(b)} but colouring cells by clusters.
@@
 Overview of the different clusters including number of cells captured for each time-point and key genes that were used to infer their identities.
+\textbf{Missing!}
 
-\textbf{Supplementary Figure 1. Quality control of sequencing data}
+\textbf{Supplementary Figure 1. Estrus cycle}
+Vaginal smears of the mice from the NP and PI timepoints. The animals were classified as (CHECK). The scale bar represents 200$\mu m$.
 
+\textbf{Supplementary Figure 2. Gating strategy to isolate MECs}
+\textbf{(a)} Gating strategy used to select live, lineage\textsuperscript{-} singlets.
+\textbf{(b)} Sorting of epithelial cells based on Epcam. Representative plots for each of the four timepoints are shown.
+
+\textbf{Supplementary Figure 3. Quality control of sequencing data}
 \textbf{(a)} Table summarizing quality control criteria per sample.
 Number of unique molecules, genes detected and number of reads represent the median value for each sample.
-\textbf{(b-c)} Histograms for the four conditions showing the distributions of the number of genes detected \textbf{(b)} or total number of molecules \textbf{(c)} metric with indicated threshold (dashed line).
-\textbf{(d)} Scatterplot of number of genes detected versus percentage of mitochondrial RNA molecules, the threshold at 5\% is indicated.
+\textbf{(b-c)} Histograms for the four conditions showing the distributions of the number of genes detected \textbf{(b)} or total number of molecules \textbf{(c)} metric.
+\textbf{(d)} Scatterplot of number of genes detected versus percentage of mitochondrial RNA molecules.
 For choice of thresholds see materials and methods.
 
-\textbf{Supplementary Figure 2. Biological replicates explain little
-structure in the data }
+\textbf{Supplementary Figure 4. Biological replicates account for little structure in the data.}
+\textbf{(a)} t-SNE plot coloured by the eight different samples. All biological replicates are well mixed and none of the identified clusters is composed of cells from only one replicate.
+\textbf{(b-c)} t-SNE plots highlighting the expression of luminal \textbf{(b)} and basal \textbf{(c)} genes in all 25,010 cells.
 
-\textbf{(a)} t-SNE plot coloured by the eight different samples.
-Some minor structures (e.g.
-two groups PI1 and PI2 at the top) are mainly replicate dependent, but all of the clusters identified in the analysis are represented by multiple replicates.
-\textbf{(b-c)} t-SNE plots highlighting the expression of luminal \textbf{(b)} and basal \textbf{(c)} genes in all 3,471 cells.
+\textbf{Supplementary Figure 5. Clustering strategy}
+\textbf{(a)} First round of clustering using an SNN-Graph and modularity maximization for cluster identification.
+\textbf{(b)} Second round of clustering using hierarchical clustering based on Spearman's rank correlation with average linkage.
+The gap statistic was used as optimization criterion to choose $k$.
+\textbf{(c-e)} Marker genes that were used to remove immune \textbf{(c)} and endothelial \textbf{(d)} cells from the analysis.
+Cluster 10.2 (Prc) showed some degree of expression for some pericyte markers \textbf{(e)}.
 
-\textbf{Supplementary Figure 3. Testing the robustness of the diffusion
-map}
-
-\textbf{(a-b)} Eigenvalues for the diffusion map used in \autoref{F2}a
-\textbf{(a)} and \autoref{F2}b \textbf{(b)}. 
-\textbf{(c)} The differentiation trajectory as determined by Monocle (see materials and methods) coloured by cluster assignment.
-\textbf{(d)} The diffusion map is robust towards down-sampling of cells (100, 50 or 25\% of all cells were used in the left, middle or right panel, respectively) as well as the method of feature selection (all= all genes with mean expression level above 0.1, HVG= highly variable genes, PCA= first 50 components from PCA, selected= a manually selected choice of genes that are known to be involved in luminal cell differentiation).
+\textbf{Supplementary Figure 6. Robustness of bifurcation event in diffusion map}
+\textbf{(a)} The differentiation trajectory as determined by Monocle (see materials and methods) coloured by cluster assignment.
+\textbf{(b)} The diffusion map is robust towards down-sampling of cells (100, 50 or 25\% of all cells were used in the left, middle or right panel, respectively) as well as the method of feature selection (all= all genes with mean expression level above 0.1, HVG= highly variable genes, PCA= first 50 components of PCA, selected= a manually selected choice of genes that are known to be involved in luminal cell differentiation).
 The gene list included: \textit{Csn2, Gata3, Prlr, Elf5, Esr1, Pgr, Aldh1a3, Wap, Tspan8, Krt18, Krt8, Areg, Fgfr1, Fgfr2, Notch1, Notch3, Foxc1} and \textit{Zeb2}.
 
-\textbf{Supplementary Figure 4. Full heatmaps of pseudo-time dependent
-expression}
+\textbf{Supplementary Figure 7. Gene set enrichment analysis of pseudo-time dependent genes}
+\textbf{(a)} Significantly enriched GO-Terms for genes in the two clusters of the hormone-sensing branch.
+\textbf{(b)} Significantly enriched GO-Terms for genes in the four clusters of the secretory branch.
 
-Heatmaps as in \autoref{F3}, showing all genes that are pseudo-time dependent in their expression along the trajectories with the same overall trend \textbf{(a)} or opposing trend \textbf{(b)}.
-
-\textbf{Supplementary Figure 5. Expression of progenitors markers in Lp-PI}
-
-Panel of t-SNE plots (compare to \autoref{F1}) with coloured normalized log-expression values of progenitor markers.
-
-\textbf{Supplementary Figure 6. Lp-PI is a post-parity progenitor
-population }
-
-\textbf{(a)} Principal component analysis (PCA) on all luminal cells from the PI time-point.
+\textbf{Supplementary Figure 8. Lp-PI is a post-parity luminal progenitor cluster}
+\textbf{(a)} Number of differentially expressed genes between the NP and PI clusters of Hsd, Hsp and Lp.
+\textbf{(b)} Principal component analysis (PCA) on all luminal cells from the PI time-point.
 The PCA was computed on the top 500 differentially expressed genes between Lp and the rest of the luminal NP gland.
 PC1 separates the progenitors and the differentiated cells, with the progenitors showing negative PC1 values.
-\textbf{(b)} Genes that are higher expressed in Lp compared to the rest of the NP gland also have negative PC1 loadings.
-\textbf{(c)} Top 20 GO-terms (biological processes) that are significantly enriched in upregulated genes from \autoref{F4}c.
-Dashed line indicates p values threshold at 0.001.
+\textbf{(c)} Genes that are higher expressed in Lp compared to the rest of the NP gland also have negative PC1 loadings.
+\textbf{(d)} Top GO-terms (biological processes) that are significantly enriched in upregulated genes from \textbf{(a)}.
+Dashed line indicates P value threshold at 0.001.
 
-\textbf{Supplementary Figure 7. Overexpression of milk genes after
-parity is restricted to the progenitor compartment}
-
-Panel of t-SNE plots (compare to \autoref{F1}) with coloured normalized log-expression values of milk genes from \autoref{F4}d.
-
-\textbf{Supplementary Figure 8. Quality control of cells from the}
-\textbf{Lactation sample}
-
-t-SNE plot computed only on the lactation sample.
-Cells that were flagged as low quality are highlighted in grey and cells that passed the QC are shown in black.
-Although a majority of the sample was removed, the cells that passed represent at least all major groups from this sample (see materials and methods).
-
-\textbf{Supplementary Figure 9. Outline of the clustering strategy }
-
-\textbf{(a)} Dendrogram of the hierarchical clustering using Euclidean distances and average linkage.
-Colour bars indicate the animal (binary, only comparable within one condition), the condition and cluster assignment.
-\textbf{(b)} Cluster stability as determined by the overlap between the original cluster and the most similar cluster in the bootstrap samples.
-The distributions of jaccard indeces are shown as a boxplot for 100 bootstrap samples.
-Stability is shown for three values of the deepSplit parameter (0,1,2).
-Deep split 1 was chosen as this maintained overall a high cluster stability while identifying relevant cell types (see materials and methods).
-
-\textbf{Supplementary Figure 10. Clustering} \textbf{Quality Control and
-removal of C10}
-
-\textbf{(a)} Total molecule count grouped by different clusters.
-\textbf{(b)} Markers that were used to identify C10 as lymphocyte cluster\autocite{Scheele2017} \textbf{(c)} Separation of C10 from all other clusters as shown by a PCA on the cluster average gene expression.
+\textbf{Supplementary Figure 9. Removal of doublet clusters}
+\textbf{(a)} Frequency of each cluster in all 8 samples, the threshold of 7\% is indicated by the red dashed line.
+\textbf{(b)} Correlation of Bsl-G2 with Avd-G (left panels), Bsl-G (center panel) and the mean of Avd-G and Bsl-G (right panel) for both samples G1 and G2.
+\textbf{(c-d)} Number of genes detected \textbf{(c)} and total number of molecules \textbf{(d)}.
+The values for each cell were normalized to the median value of the sample in which the cell was captured.
 
@@
 
-\textbf{Table S4.} Genes differentially expressed between Lp-PI and Lp
+\textbf{Table S4.} Genes differentially expressed between Lp-PI and Lp-NP
 
@@
 This has the advantage of not requiring the computation of a distance matrix which is computationally inhibitive for a large number of cells.
-The SNN-graph was constructed using the \texttt{buildSNNGraph} function from scran with the number of nearest-neighbours set to 20.
+The SNN-graph was constructed using the \texttt{buildSNNGraph} function from scran with the number of nearest-neighbors set to 20.
 Community-based clustering was then performed on the SNN-graph by multi-level modularity optimization using the \texttt{cluster\_louvain} function in igraph (CITATION).
 This identified a total of 13 clusters.
-Yet, the majority of clusters still contained substructe that was evident from inspecting the t-SNE (SubXa).
-In order to resolve theses structures further we then applied agglomerative hierarchical clustering on each of the 13 clusters (SupXb).
-For this, we first computed the pairwise distances between cells based on the spearman correlation of log-transformed HVG counts.
+Yet, the majority of clusters still contained substructure that was evident from inspecting the t-SNE (Supplementary Figure 5a).
+In order to resolve theses structures further we then applied agglomerative hierarchical clustering on each of the 13 clusters (Supplementary Figure 5b).
+For this, we first computed the pairwise distances between cells based on the Spearman's rank correlation of log-transformed HVG counts.
 The dissimilarity matrix was then used to perform hierarchical clustering with the 'hclust' function in R using average linkage.
@@
 Next, we flagged clusters that expressed clear markers of non-epithelial cells as contaminating cells and removed them from the downstream analysis.
-Cluster 6.1 and 6.2 were tagged as immune cells based on the expression of \textit{Cd74}, \textit{Cd72} and \textit{Cd54}\autocite{Scheele2017} and cluster 10.3 as endothelial cells based on the expression of \textit{Eng}, \textit{S1pr1} and \textit{Emcn} (CITATION,SupXc and SupXd).
-We retained cluster 10.2 as it expressed markers of a previously described celltype in the gland (CITATION), yet it was also positive for 2 out of 4 (\textit{Des} and \textit{Cspg4}) pericyte markers (CITATION) (SupXd).
+Cluster 6.1 and 6.2 were tagged as immune cells based on the expression of \textit{Cd74}, \textit{Cd72} and \textit{Cd54}\autocite{Scheele2017} and cluster 10.3 as endothelial cells based on the expression of \textit{Eng}, \textit{S1pr1} and \textit{Emcn} (CITATION, Supplementary Figure 5c, Supplementary Figure 5d).
+We retained cluster 10.2 as it expressed markers of a previously described celltype in the gland (CITATION), yet it was also positive for 2 out of 4 (\textit{Des} and \textit{Cspg4}) pericyte markers (CITATION) (Supplementary Figure 5e).
 The high number of cells that was loaded on to the 10x per sample increased the chance of doublet formation in the droplet encapsulation (CITATION).
 Especially doublets of two cell types can confound downstream analysis.
-We thus identified clusters that appeared at low frequencies in the samples were it was detected (less than the maximum doublet rate of 7\%, SupXa) and was highly similar to at least two other clusters present in the same sample ($p > 0.9$ of mean $log2(counts+1)$).
-The only cluster that fulfilled this criterion was cluster 11.1, which was strongly correlated with XX and XX (SupXb).
-In addition, the cluster was almost perfectly correlated with the average expression values of XX and XX, supporting the hypothesis that it is indeed a mixture of XX and XX (SupXb).
-This was further supported by the fact that XX and XX are the most prevalent clusters in these samples.
-Finally, 11.1 also belonged to the clusters with a high number of genes detected and a high number of unique molecules, which is also indicative for doublet clusters (SupXc).
-We hence excluded 11.1 from any downstream analysis.
+We thus identified clusters that appeared at low frequencies in the samples were it was detected (less than the maximum doublet rate of 7\%, Supplementary Figure 9b) and was highly similar to at least two other clusters present in the same sample ($p > 0.9$ of mean log2(counts+1)).
+The only cluster that fulfilled this criterion was cluster Bsl-G2, which was strongly correlated with Bsl-G and Avd-G (Supplementary Figure 9b)
+In addition, the cluster was almost perfectly correlated with the average expression values of Bsl-G and Avd-G, supporting the hypothesis that it is indeed a mixture of the two clusters (Supplementary Figure 9b).
+This was further supported by the fact that Bsl-G and Avd-G are the most prevalent clusters in these samples (Table 1).
+Finally, Bsl-G2 also belonged to the clusters with a high number of genes detected and a high number of unique molecules, which is also indicative for doublet clusters (Supplementary Figure 9c).
+We hence excluded Bsl-G2 from any downstream analysis.
 As the last step we grouped clusters of cells together in super structures based on the expression of marker genes for known celltypes.

### 2017-08-12 src/Figure1.R

@@ -70,7 +70,7 @@ dendr <- dis %>% hclust(.,method="ward.D2") %>% as.dendrogram %>%
     set("leaves_pch",19) %>%
     set("leaves_cex",2) %>%
     set("branches_lwd",3)
-par(cex=.8)
+par(cex=.7)
 plot(dendr,horiz=TRUE,yaxt="n")
 p.dendr <- grab_grob()
 p.dendr <- grid.arrange(p.dendr)

### 2017-08-12 src/Monocle.R

@@ -63,7 +63,7 @@ cds <- reduceDimension(cds, max_components=2,norm_method="log")
 cds <- orderCells(cds,reverse=TRUE)
 
 # Plot trajectory colored by clusters
-p0 <- plot_cell_trajectory(cds,x=1,y=2, color_by="SuperCluster", cell_size=1,show_branch_points=FALSE) 
+p0 <- plot_cell_trajectory(cds,x=1,y=2, color_by="SubCluster", cell_size=1,show_branch_points=FALSE) 
 
 # Save
 pD.monoc <- pData(cds)

### 2017-08-12 src/NPPIDE.R

@@ -13,7 +13,7 @@ fD <- dataList[[3]]
 comps <- list(Hsd=c("Hsd-PI","Hsd-NP"),
 	      Hsp=c("Hsp-PI","Hsp-NP"),
 	      Lp=c("Lp-PI","Lp-NP"),
-	      Bsl=c("Bsl-G1","Bsl"))
+	      Bsl=c("Bsl-G","Bsl"))
 
 for (cname in c("Hsd","Hsp","Lp","Bsl")) {
     comp <- comps[[cname]]

### 2017-08-12 src/S10.R

@@ -1,143 +1,0 @@
-# Estimate size factors using scran
-library(dplyr)
-library(scran)
-library(Rtsne)
-library(knitr)
-library(gridExtra)
-source("functions.R")
-
-dataList <- readRDS("../data/Robjects/secondRun_2500/ExpressionList_QC_norm_clustered_clean.rds")
-# set.seed(300)
-# dataList <- subSample(dataList, cell.number=2000)
-m <- dataList[["counts"]]
-pD <- dataList[["phenoData"]]
-fD <- dataList[["featureData"]]
-rm(dataList)
-
-m <- m[fD$keep,pD$PassAll]
-pD <- pD[pD$PassAll,]
-m <- t(t(m)/pD$sf)
-
-# Doublet definition: below 7% in all samples where it was present
-tab <- ftable(pD$SubCluster, pD$SampleID)
-tab <- as.matrix(t(t(tab)/colSums(tab)))
-pot.dob <- rownames(tab)[which(rowMax(tab)<0.07)]
-
-freqDf <- data.frame(as.table(tab))
-ordr <- group_by(freqDf, Var1) %>%
-    summarize(tot=sum(Freq)) %>%
-    dplyr::arrange(desc(tot)) %>%
-    .$Var1 %>%
-    as.character()
-
-freqDf$Var1 <- factor(freqDf$Var1,levels=ordr)
-colnames(freqDf) <- c("Cluster","Sample","Frequency")
-
-p1 <- ggplot(freqDf, aes(x=Cluster,y=Frequency, fill=Sample)) +
-    geom_bar(stat="identity",position="dodge") +
-    geom_hline(yintercept=0.07,color="red",lty="dashed") +
-    theme(axis.text.x=element_text(angle=45, vjust=0.5, size=16))
-
-# for Table
-
-out <- data.frame()
-smpls <- levels(pD$SampleID)
-for (smpl in smpls) {
-    pD.sub <- pD[pD$SampleID==smpl,]
-    keepC <- names(table(pD.sub$SubCluster)[which(table(pD.sub$SubCluster)>1)])
-    pD.sub <- pD.sub[pD.sub$SubCluster %in% keepC,]
-    m.sub <- m[,pD.sub$barcode]
-    cluster <- factor(pD.sub$SubCluster)
-    tmp <- data.frame(numeric(nrow(m)),check.names=FALSE)
[…]

### 2017-08-12 src/S4.R

@@ -1,72 +1,70 @@
-# GO-Analysis
+# S4 and S7
+
+library(plyr)
+library(dplyr)
 library(ggplot2)
 library(cowplot)
-library("topGO")
-library(org.Mm.eg.db)
+library(viridis)
+library(RColorBrewer)
+library(reshape2)
+library(pheatmap)
+source("functions.R")
 
-# ---- DEAnalysis4vs5 ----
-output <- data.frame()
-input <- read.csv("../data/Robjects/secondRun_2500/BranchDECluster.R",
-		  stringsAsFactor=FALSE)
-input$BranchCluster <- paste(input$Branch,input$Cluster,sep="-")
+# Load Data
+rnd_seed <- 300
+dataList <- readRDS("../data/Robjects/secondRun_2500/ExpressionList_QC_norm_clustered_clean.rds")
+m <- dataList[[1]]
+pD <- dataList[[2]]
+fD <- dataList[[3]]
 
-for (clust in unique(input$BranchCluster)) { 
 
-    # DeGenes
-    deGenes <- input[input$BranchCluster==clust,"Gene"]
+# ---- S4 ----
 
-    # Gene universe
-    if (substr(clust,1,3)=="Hrm") i <- 1 else i <- 2
-    m.smooth <- readRDS("../data/Robjects/secondRun_2500/BranchDEList.rds")[[i]][["mSmooth"]]
-    univrs <- rownames(m.smooth[rowMeans(m.smooth)>0.1,])
+# Rename Condition for plot
+pD$Condition <- mapvalues(pD$Condition, from=c("NP","G","L","PI"),
+			  to=c("Nulliparous", "14.5d Gestation",
+			       "6d Lactation", "11d Post Natural Involution"))
 
-    # ---- Data ----
-    alG <- factor(as.numeric(univrs %in% deGenes))
-    names(alG) <- univrs
+# Remove previously identified outlier and immune cells
+m <- m[,pD$keep]
+pD <- pD[pD$keep,]
 
-    # ---- GOanalysis ----
+# put in rnd order for plotting
+set.seed(rnd_seed)
+fp1 <- pD[sample(c(1:nrow(pD)),nrow(pD)),]
+#t-SNE colored by SampleID
+p1 <- ggplot(fp1, aes(x=tSNE1, y=tSNE2, color=SampleID)) +
+    geom_point(size=1) +
+    scale_color_brewer(palette="Paired")+
+    #     ggtitle("Cluster") +
[…]

### 2017-08-12 src/S5.R

@@ -1,0 +1,65 @@
+# S5
+
+library(plyr)
+library(dplyr)
+library(ggplot2)
+library(cowplot)
+library(viridis)
+library(RColorBrewer)
+library(pheatmap)
+library(gridExtra)
+library(gridGraphics)
+source("functions.R")
+
+# Load Data
+rnd_seed <- 300
+dataList <- readRDS("../data/Robjects/secondRun_2500/ExpressionList_QC_norm_clustered.rds")
+m <- dataList[[1]]
+pD <- dataList[[2]]
+fD <- dataList[[3]]
+
+# Remove outlier and immune cells
+m <- m[,pD$PassAll]
+pD <- pD[pD$PassAll,]
+
+m <- t(t(m)/pD$sf)
+
+
+# First step
+p1 <- ggplot(pD,aes(tSNE1,tSNE2,color=Cluster)) +
+    geom_point()
+
+# Second round
+fp2 <- pD
+fp2$SubCluster <- as.character(fp2$SubCluster)
+fp2$SubCluster <- substr(fp2$SubCluster,nchar(fp2$SubCluster),nchar(fp2$SubCluster))
+p2 <- ggplot(fp2,aes(tSNE1,tSNE2,color=SubCluster)) +
+    geom_point(size=.8) +
+    facet_wrap(~Cluster)
+
+subP1 <- plot_grid(p1,p2, labels=c("a","b"))
+
+# ---- PlotImmuneCellMarkers ----
+
+genes <- c("Cd52","Cd74","Cd72")
+p3 <- plotGeneDist(m, pD, fD, genes, colorBy="SubCluster")
+p3 <- p3 %+% facet_grid(variable~.) %+% xlab("SubCluster") %+% ylab("Log-Expression")
+
+# ---- PlotEnodthelialCellMarkers ----
+
+genes <- c("Eng","S1pr1","Emcn")
+p4 <- plotGeneDist(m, pD, fD, genes, colorBy="SubCluster")
+p4 <- p4 %+% facet_grid(variable~.) %+% xlab("Cluster") %+% ylab("Log-Expression")
+
+# ---- PlotPericyteMarkers ----
+
+genes <- c("Pdgfrb","Cspg4","Anpep","Des")
+p5 <- plotGeneDist(m, pD, fD, genes, colorBy="SubCluster")
[…]

### 2017-08-12 src/S5_7.R

@@ -1,129 +1,0 @@
-# S5 and S7
-
-library(plyr)
-library(dplyr)
-library(ggplot2)
-library(cowplot)
-library(viridis)
-library(RColorBrewer)
-library(reshape2)
-library(pheatmap)
-source("functions.R")
-
-# Load Data
-rnd_seed <- 300
-dataList <- readRDS("../data/Robjects/secondRun_2500/ExpressionList_QC_norm_clustered_clean.rds")
-m <- dataList[[1]]
-pD <- dataList[[2]]
-fD <- dataList[[3]]
-
-
-# ---- S4 ----
-
-# Rename Condition for plot
-pD$Condition <- mapvalues(pD$Condition, from=c("NP","G","L","PI"),
-			  to=c("Nulliparous", "14.5d Gestation",
-			       "6d Lactation", "11d Post Natural Involution"))
-
-# Remove previously identified outlier and immune cells
-m <- m[,pD$keep]
-pD <- pD[pD$keep,]
-
-# put in rnd order for plotting
-set.seed(rnd_seed)
-fp1 <- pD[sample(c(1:nrow(pD)),nrow(pD)),]
-#t-SNE colored by SampleID
-p1 <- ggplot(fp1, aes(x=tSNE1, y=tSNE2, color=SampleID)) +
-    geom_point(size=1) +
-    scale_color_brewer(palette="Paired")+
-    #     ggtitle("Cluster") +
-    theme_void(base_size=12) +
-    guides(colour = guide_legend(override.aes = list(size=3))) +
-    theme(legend.position="bottom",legend.direction="horizontal",
-	  legend.title=element_blank()) 
-
-# Normalize
-m.norm <- t(t(m)/pD$sf)
-rownames(m.norm) <- fD$symbol
-
-titles <- genes <- c("Krt18","Krt8","Krt5","Krt14","Acta2")
-add <- data.frame(log2(t(m.norm)[,genes]+1),
-		  barcode=colnames(m))
-colnames(add) <- gsub("X","",colnames(add))
-forPlot <- left_join(add,pD[,c("barcode","tSNE1","tSNE2")])
-forPlot <- melt(forPlot,id=c("barcode","tSNE1","tSNE2")) %>%
-    dplyr::rename(Expression=value)
-plots <- list()
-for(i in seq_along(genes)) {
-    gene <- genes[i]
-    fP <- filter(forPlot,variable==gene) %>% arrange(Expression)
[…]

### 2017-08-12 src/S6.R

@@ -17,52 +17,22 @@ m <- dataList[[1]]
 pD <- dataList[[2]]
 fD <- dataList[[3]]
 
-
-# ---- ScreePlots ----
-
-condComb <- c("NP","G")
-sets <- list(NULL,c("Bsl-G1","Bsl","Myo","Prc")) #order important for Robustness section
-out <- list()
-for (i in seq_along(sets)) {
-    set <- sets[[i]]
-    keepCells <- pD$keep & !(pD$SuperCluster %in% set) & pD$Condition %in% condComb
-
-    # Cell filtering
-    m.vp <- m[,keepCells]
-    pD.vp <- pD[keepCells,]
-
-    # Gene filtering
-    keep <- rowMeans(m.vp)>0.01
-    m.vp <- m.vp[keep,]
-    fD.vp <- fD[keep,]
-
-    # Normalize
-    m.norm <- t(t(m.vp)/pD.vp$sf)
-
-    # Highly variable genes 
-    var.des <- trendVar(log2(m.norm+1),trend="semiloess")
-    var.out <- decomposeVar(log2(m.norm+1),var.des)
-    o <- order(var.out$mean)
-    hvg.out <- var.out[which(var.out$FDR <= 0.05 & var.out$bio >=0.5),]
-
-    # Prepare expression matrix
-    m.norm <- m.norm[rownames(hvg.out),]
-    exps <- t(log(m.norm+1))
-
-    # Compute diffusion map
-    set.seed(rnd_seed)
-    dm <- DiffusionMap(exps, n_eigs=20, rotate=TRUE)
-    plot(eigenvalues(dm),pch=20,xlab="Component",ylab="Eigenvalue")
-    lines(eigenvalues(dm),pch=20,xlab="Component",ylab="Eigenvalue")
-    g <- grab_grob()
-    dev.off()
-    out[[paste0("g",i)]] <- grid.arrange(g)
-}
-
-g1 <- out[["g1"]]
-g2 <- out[["g2"]]
-
 # ---- Robustness -----
+condComb <- c("NP","G")
+set <- c("Bsl-G1","Bsl","Myo","Prc")
+keepCells <- pD$keep & !(pD$SuperCluster %in% set) & pD$Condition %in% condComb
+
+# Cell filtering
+m.vp <- m[,keepCells]
+pD.vp <- pD[keepCells,]
+
+# Gene filtering
+keep <- rowMeans(m.vp)>0.01
[…]

### 2017-08-12 src/S7.R

@@ -1,0 +1,72 @@
+# GO-Analysis
+library(ggplot2)
+library(cowplot)
+library("topGO")
+library(org.Mm.eg.db)
+
+# ---- DEAnalysis4vs5 ----
+output <- data.frame()
+input <- read.csv("../data/Robjects/secondRun_2500/BranchDECluster.R",
+		  stringsAsFactor=FALSE)
+input$BranchCluster <- paste(input$Branch,input$Cluster,sep="-")
+
+for (clust in unique(input$BranchCluster)) { 
+
+    # DeGenes
+    deGenes <- input[input$BranchCluster==clust,"Gene"]
+
+    # Gene universe
+    if (substr(clust,1,3)=="Hrm") i <- 1 else i <- 2
+    m.smooth <- readRDS("../data/Robjects/secondRun_2500/BranchDEList.rds")[[i]][["mSmooth"]]
+    univrs <- rownames(m.smooth[rowMeans(m.smooth)>0.1,])
+
+    # ---- Data ----
+    alG <- factor(as.numeric(univrs %in% deGenes))
+    names(alG) <- univrs
+
+    # ---- GOanalysis ----
+
+    # prepare Data for topGO
+    GO.data <- new("topGOdata", description="Lib GO",ontology="BP", allGenes=alG, 
+		   annot=annFUN.org, mapping="org.Mm.eg.db",
+		   nodeSize=20, ID="symbol")
+    result.classic <- runTest(GO.data, statistic="fisher")
+    tmp <- GenTable(GO.data, Fisher.classic=result.classic, orderBy="topgoFisher", topNodes=50, numChar=300)
+
+    tmp$Term <- factor(tmp$Term, levels=unique(rev(tmp$Term)))
+    tmp$Cluster <- clust
+    output <- rbind(tmp,output)
+}
+
+output <- output[output$Fisher.classic < 0.01,]
+
+library(dplyr)
+# ordr <- group_by(output, Cluster) %>%
+#     arrange(-log10(as.numeric(Fisher.classic))) %>%
+#     .$Term
+
+# output$Term <- factor(output$Term,levels=unique(ordr))
+output1 <- output[grepl("Alv",output$Cluster),]
+output2 <- output[grepl("Hrm",output$Cluster),]
+
+p1 <- ggplot(output1, aes(x=Term, y=-log10(as.numeric(Fisher.classic)))) +
+    geom_bar(stat="identity",color="black",fill="white") +
[…]

### 2017-08-12 src/S8.R

@@ -13,7 +13,7 @@ source("functions.R")
 out <- data.frame()
 for (clust in c("Hsd","Hsp","Lp")) {
     filen <- sprintf("../data/Robjects/secondRun_2500/%s_NPvsPI.csv",clust)
-    tmp <- read.csv(file=filen)
+    tmp <- read.csv(file=filen,stringsAsFactors=FALSE)
     tmp$Cluster <- clust
     out <- rbind(out,tmp)
 }
@@ -34,7 +34,7 @@ p1 <- ggplot(out, aes(x=Cluster,y=value,fill=variable)) +
 output <- data.frame()
 for (clust in c("Hsd","Hsp","Lp")) { # Lp has to be last for rest of script
     filen <- sprintf("../data/Robjects/secondRun_2500/%s_NPvsPI.csv",clust)
-    topTab <- read.csv(file=filen)
+    topTab <- read.csv(file=filen,stringsAsFactors=FALSE)
 
     deGenes <- topTab[topTab$FDR < 0.01 & topTab$logFC > 0,"symbol"] 
     allGenes <- topTab$symbol
@@ -51,7 +51,7 @@ for (clust in c("Hsd","Hsp","Lp")) { # Lp has to be last for rest of script
 		   annot=annFUN.org, mapping="org.Mm.eg.db",
 		   nodeSize=20, ID="symbol")
     result.classic <- runTest(GO.data, statistic="fisher")
-    tmp <- GenTable(GO.data, Fisher.classic=result.classic, orderBy="topgoFisher", topNodes=50, numChar=30)
+    tmp <- GenTable(GO.data, Fisher.classic=result.classic, orderBy="topgoFisher", topNodes=50, numChar=10000)
 
     tmp$Term <- factor(tmp$Term, levels=unique(rev(tmp$Term)))
     tmp$Cluster <- clust
@@ -69,8 +69,8 @@ output$Term <- factor(output$Term,levels=unique(ordr))
 ## Extract gene lists for volcano plot
 # better hardcode go-terms at one point to define gene lists
 sub.output <- output[output$Cluster=="Lp",]
-imterm <- sub.output[as.character(c(5,11,12,21,23,29,37,41,42,44,49,50)),1]
-lterm <- sub.output[as.character(c(35,25,6,20)),1]
[…]

### 2017-08-12 src/S9.R

@@ -1,65 +1,145 @@
-# S9
-
-library(plyr)
+# Estimate size factors using scran
 library(dplyr)
-library(ggplot2)
-library(cowplot)
-library(viridis)
-library(RColorBrewer)
-library(pheatmap)
+library(scran)
+library(Rtsne)
+library(knitr)
 library(gridExtra)
-library(gridGraphics)
 source("functions.R")
 
-# Load Data
-rnd_seed <- 300
-dataList <- readRDS("../data/Robjects/secondRun_2500/ExpressionList_QC_norm_clustered.rds")
-m <- dataList[[1]]
-pD <- dataList[[2]]
-fD <- dataList[[3]]
+dataList <- readRDS("../data/Robjects/secondRun_2500/ExpressionList_QC_norm_clustered_clean.rds")
+# set.seed(300)
+# dataList <- subSample(dataList, cell.number=2000)
+m <- dataList[["counts"]]
+pD <- dataList[["phenoData"]]
+fD <- dataList[["featureData"]]
+rm(dataList)
 
-# Remove outlier and immune cells
-m <- m[,pD$PassAll]
+m <- m[fD$keep,pD$PassAll]
 pD <- pD[pD$PassAll,]
-
 m <- t(t(m)/pD$sf)
 
+# Doublet definition: below 7% in all samples where it was present
+tab <- ftable(pD$SubCluster, pD$SampleID)
+tab <- as.matrix(t(t(tab)/colSums(tab)))
+pot.dob <- rownames(tab)[which(rowMax(tab)<0.07)]
 
-# First step
-p1 <- ggplot(pD,aes(tSNE1,tSNE2,color=Cluster)) +
-    geom_point()
+freqDf <- data.frame(as.table(tab))
+ordr <- group_by(freqDf, Var1) %>%
+    summarize(tot=sum(Freq)) %>%
+    dplyr::arrange(desc(tot)) %>%
+    .$Var1 %>%
+    as.character()
 
-# Second round
-fp2 <- pD
-fp2$SubCluster <- as.character(fp2$SubCluster)
-fp2$SubCluster <- substr(fp2$SubCluster,nchar(fp2$SubCluster),nchar(fp2$SubCluster))
-p2 <- ggplot(fp2,aes(tSNE1,tSNE2,color=SubCluster)) +
-    geom_point(size=.8) +
-    facet_wrap(~Cluster)
[…]

### 2017-08-12 paper/main.tex
:version to be send out for proofread

@@
 All samples were then prepared for single-cell RNA-sequencing using the 10x Chromium platform\autocite{Zheng2017}.
-Following quality control (materials and methods), this yielded an average of 6,175 unique transcripts and 2,118 genes detected from 25,010 cells (4,223 in NP, 5,826 in G, 9,319 in L and 5,642 in P) (Supplementary 3).
+Following quality control (materials and methods), this yielded an average of 6,175 unique transcripts and 2,118 genes detected from 25,010 cells (4,223 in NP, 5,826 in G, 9,319 in L and 5,642 in P) (Supplementary Figure 3).
 Visual inspection of the data using t-distributed stochastic neighbor embedding (t-SNE) suggested that although there is grouping of cells by time-point, there are also other factors that underlie structure within the dataset (\autoref{F1}b and Supplementary Figure 4).
-First, we dissected the global structure in the dataset by unsupervised clustering using a shared-nearest-neighbor clustering approach.
+First, we dissected the global structure by unsupervised clustering using a shared-nearest-neighbor clustering approach.
 This resulted in a coarse clustering into 13 groups (Supplementary Figure 5a).
@@
 Myo shows characteristics of myoepithelial cells such as high levels of \textit{Acta2}, \textit{Oxtr} and \textit{Krt15} (\autoref{F1}d).
-\comment{Describe Bsl and Prc as well?}
 
@@
 
-The expression of the progenitor marker \textit{Aldh1a3} gradually decreased as cells progressed away from their common origin (\autoref{F2}b right panel), which was largely composed of Lp.
-We further noted that the left arm of the differentiation trajectory terminates at Avd and shows increasing expression of \textit{Csn2} (\autoref{F2}b bottom panel), consistent with a secretory phenotype.
+The expression of the progenitor marker \textit{Aldh1a3} gradually decreased as cells progressed away from their common origin (\autoref{F2}c), which was largely composed of Lp.
+We further noted that the left arm of the differentiation trajectory terminates at Avd and shows increasing expression of \textit{Csn2} and \textit{Glycam1} (\autoref{F2}c), consistent with a secretory phenotype.
 Between Lp and Avd we found cells from Avp (\autoref{F2}b).
-On the right arm of the differentiation trajectory, cells in Lp transitioned to Hsp and Hsd, during which the expression of \textit{Esr1} and \textit{Pgr} increase suggesting that this branch represents differentiation towards hormone-sensing luminal cells (\autoref{F2}b, bottom panel).
+On the right arm of the differentiation trajectory, cells in Lp transitioned to Hsp and Hsd, during which the expression of \textit{Cited1} and \textit{Pgr} increase suggesting that this branch represents differentiation towards hormone-sensing luminal cells (\autoref{F2}c).
 
 Being confident that the diffusion map recapitulates the luminal differentiation process, we then computationally inferred the two branches and ordered the cells according to their progression through ``pseudotime''\autocite{Haghverdi2016} (see materials and methods for further explanation) (\autoref{F3}a).
-This allows to identify genes whose expression changes during the process of differentiation.
+This allows us to identify genes whose expression changes during the process of differentiation.
 We found 456 genes that showed a pseudotime-dependent expression with the same directionality along both differentiation trajectories (Supplementary Table 2).
 These included genes associated with known progenitor characteristics such as \textit{Aldh1a3} and \textit{Tspan8} as well as transcription factors that have not previously been associated with luminal differentiation such as \textit{Creb5}, \textit{Hmga1} and \textit{Fosl1} (\autoref{F3}b).
-Interestingly, \textit{Hmga1} is a known chromatin remodeller with reprogramming activity that has been implicated as a driver in basal-like breast cancer (CHECK AND CITE).
+Interestingly, \textit{Hmga1} is a known chromatin remodeller with reprogramming activity that has been implicated in basal-like breast cancer\autocite{Shah2012,Shah2013}.
 In addition, we identified 1005 genes with branch-specific expression patterns (\autoref{F3}d, \autoref{F3}e Supplementary Table 3).
-We further clustered the gene expression trend for these on each of the two branches and identified sets of genes that change over pseudotime (Supplementary Figure 7).
+We further clustered the gene expression trend on each of the two branches and identified sets of genes that change over pseudotime (Supplementary Figure 7).
 On the alveolar branch we found two clusters of genes that increase in expression during differentiation, one that is switched on early and another that is activated later in the final stages of differentiation.
@@
 The second, smaller group of genes that is switched on during the late stages of differentiation was enriched for genes involved in fatty-acid transport and lipid homeostasis (e.g.
-\textit{Olah}, \textit{Cd36})
+\textit{Olah}, \textit{Cd36}).
 Genes involved in cell division and translation show a transient phase of up-regulation during the alveolar differentiation process (e.g.
 \textit{Aurka}, \textit{Cenpa}, \textit{Cdk6}), whereas genes regulating cell shape and morphogenesis of a branching epithelium are repressed early on (e.g.
-\textit{Vdr}, \textit{Areg} and \textit{Myc}).
+\textit{Vdr} and \textit{Areg}).
 During the process of differentiation towards hormone-sensing luminal cells, we identified two broad clusters of genes that gradually increase or decrease their expression.
@@
 Amongst the branch-specific transcription factors we found for example \textit{Runx1, Tox2} and \textit{Bhlhe41} (also known as \textit{Sharp1}) to be transcribed during differentiation towards the hormone-sensing lineage (\autoref{F3}d) and \textit{Elf5, Foxs1} and \textit{Ehf} in the secretory lineage (\autoref{F3}e).
-Interestingly, \textit{Runx1} is a known repressor of \textit{Elf5} and its deletion has been shown to be deleterious for ductal morphogenesis\autocite{VanBragt2014}.
+\textit{Runx1} is a known repressor of \textit{Elf5} and its deletion has been shown to be deleterious for ductal morphogenesis\autocite{VanBragt2014}.
 The expression of the known progenitor master regulator \textit{Elf5} is maintained and further increased during secretory differentiation suggesting that its transcriptional level is fine-tuned in luminal progenitors (\autoref{F3}e).
@@
 Indeed, we find genes that are differentially expressed between Lp-NP and the rest of the luminal compartment to show the same trend between Lp-PI and the luminal compartment (\autoref{F4}a).
-In a similar manner we can distinguish Lp-PI from the rest of the PI gland in a principal component analysis (PCA) using the identified progenitor genes (\textbf{Supplementary Fig 8c and d}).
+In a similar manner we can distinguish Lp-PI from the rest of the PI gland in a principal component analysis (PCA) using the identified progenitor genes (Supplementary Fig 8b and c).
 From this we conclude, that Lp-PI indeed represents the post-parity luminal progenitor population.
-Interestingly, genes that were up-regulated in Lp-PI compared to Lp-NP were significantly enriched for pathways that are involved in the immune response and lactation (\autoref{F4}b, \textbf{Supplementary Fig 8c}).
+Interestingly, genes that were up-regulated in Lp-PI compared to Lp-NP were significantly enriched for pathways that are involved in the immune response and lactation (\autoref{F4}b, Supplementary Fig 8d).
 These included genes that play roles in various processes during lactation, e.g.
@@
 Finally, we compared the differentiation processes of the luminal compartment of the parous gland to the nulliparous gland by mapping the luminal cells from PI to the trajectory of NP and G cells.
-Interestingly, the Hsp and Hsd cluster generally maintain their position of their nulliparous counterpart in the differentiation hierarchy whereas the Lp-PI cluster is stretched out from the origin down towards the alveolar branch, suggesting that these are biased towards the alveolar fate (\autoref{F4}b).
+Interestingly, the Hsp and Hsd cluster generally maintain their position of their nulliparous counterpart in the differentiation hierarchy whereas the Lp-PI cluster is stretched out from the origin down towards the alveolar branch, suggesting that these are biased towards the alveolar fate (\autoref{F4}d).
 Together, the data suggest that luminal progenitor cells maintain memory of having undergone gestation and involution.
@@
 In this study we could not provide any evidence for contribution of a putative mutlipotent stem cell to the day-to-day homoeostasis of the gland.
-We do note, however, that the cluster Prc represents the cluster with the fewest numbers of cells captured and expresses low levels of both luminal and basal markers, high levels of the stem cell marker \textit{Procr}\autocite{Wang2015} and high levels of the luminal progenitor marker \textit{Notch3}\autocite{Lafkas2013} (\autoref{F1}d).
+We do note, however, that the Prc cluster expresses low levels of both luminal and basal markers, high levels of the stem cell marker \textit{Procr}\autocite{Wang2015} and high levels of the luminal progenitor marker \textit{Notch3}\autocite{Lafkas2013} (\autoref{F1}d).
 Yet, our data does not support a central role of this cluster in the day-to-day homeostasis of the gland. 
-In addition, cells from the Prc cluster also express some but not all markers of pericytes and we thus cannot exclude that these are contaminating non-epithelial cells. 
+In addition, cells from the Prc cluster also express some but not all markers of pericytes and we thus cannot exclude that these are non-epithelial cells. 
 Based on the gene expression data presented here, the luminal compartment appears to have one common progenitor population (Lp).
@@
 \textbf{(d)} Same as \textbf{(b)} but colouring cells by clusters.
-\textbf{(e)} Heatmap highlighting some key marker genes.
+\textbf{(e)} Heatmap highlighting some key marker genes to distinguish the main classes of clusters.
 Colour scale represents log-transformed and normalized UMI counts scaled to a maximum of 1 per row.
@@
 Cells are coloured by their progression through pseudotime, where low values represent undifferentiated cells.
-\textbf{(b-d)} Examples of transcription factors with pseudotime-dependent expression with the same overall trend on both branches \textbf{(c)} or branch specific trends \textbf{(d,e)}.
-\textbf{(e-f)} Heatmap of all genes with branch specific, pseudotime-dependent expression for the hormone-sensing lineage \textbf{(e)}a or the secretory lineage \textbf{(f)}.
+\textbf{(b-d)} Examples of transcription factors with pseudotime-dependent expression with the same overall trend on both branches \textbf{(b)} or branch specific trends \textbf{(c,d)}.
+\textbf{(e-f)} Heatmap of all genes with branch specific, pseudotime-dependent expression for the hormone-sensing lineage \textbf{(e)} or the secretory lineage \textbf{(f)}.
 Pseudotime and the cluster assignment are annotated above the heatmap.
@@
 The genes represent the top 500 differentially expressed genes between Lp-NP and luminal cells.
-\textbf{(c)} Volcano plot illustrates differential expression between Lp-NP and Lp-PI, coloured dots represent significant genes with known function in lactation and immunity, dashed lines highlight the P value threshold of 0.01 and a log fold change of 1.
+\textbf{(b)} Volcano plot illustrates differential expression between Lp-NP and Lp-PI, coloured dots represent significant genes with known function in lactation and immunity, dashed lines highlight the P value threshold of 0.01 and a log fold change of 1.
 P values are adjusted for multiple testing using Benjamini-Hochberg.
@@
 Expression values correspond to normalized UMI counts.
+\textbf{(e)} Cells from PI mapped onto the diffusion map from the NP and G time-point. Cells from NP and G are colored in grey.
 %-------------------------------------------------------------------------------------------------------------------
@@
 \textbf{(a)} Frequency of each cluster in all 8 samples, the threshold of 7\% is indicated by the red dashed line.
-\textbf{(b)} Correlation of Bsl-G2 with Avd-G (left panels), Bsl-G (center panel) and the mean of Avd-G and Bsl-G (right panel) for both samples G1 and G2.
+\textbf{(b)} Correlation of Bsl-G2 (on the y-axes) with Avd-G (left panels), Bsl-G (center panel) and the mean of Avd-G and Bsl-G (right panel) for both samples G1 and G2.
 \textbf{(c-d)} Number of genes detected \textbf{(c)} and total number of molecules \textbf{(d)}.
@@
 This identified a total of 13 clusters.
-Yet, the majority of clusters still contained substructure that was evident from inspecting the t-SNE (Supplementary Figure 5a).
-In order to resolve theses structures further we then applied agglomerative hierarchical clustering on each of the 13 clusters (Supplementary Figure 5b).
+Yet, the majority of clusters still contained substructure as evident from the t-SNE (Supplementary Figure 5a).
+In order to resolve the cellular heterogeneity further we then applied agglomerative hierarchical clustering on each of the 13 clusters (Supplementary Figure 5b).
 For this, we first computed the pairwise distances between cells based on the Spearman's rank correlation of log-transformed HVG counts.
@@
 With this we identified a total number of 21 clusters.
+As the last step we performed a \textit{post-hoc} test as previously reported\autocite{Macosko2015} by merging clusters with less than 10 differentially expressed genes (at a P value threshold of 0.01 and minimum log fold change of 1), resulting in one merged cluster (Myo).
 Next, we flagged clusters that expressed clear markers of non-epithelial cells as contaminating cells and removed them from the downstream analysis.
@@
 We retained cluster 10.2 as it expressed markers of a previously described celltype in the gland (CITATION), yet it was also positive for 2 out of 4 (\textit{Des} and \textit{Cspg4}) pericyte markers (CITATION) (Supplementary Figure 5e).
-The high number of cells that was loaded on to the 10x per sample increased the chance of doublet formation in the droplet encapsulation (CITATION).
-Especially doublets of two cell types can confound downstream analysis.
-We thus identified clusters that appeared at low frequencies in the samples were it was detected (less than the maximum doublet rate of 7\%, Supplementary Figure 9b) and was highly similar to at least two other clusters present in the same sample ($p > 0.9$ of mean log2(counts+1)).
-The only cluster that fulfilled this criterion was cluster Bsl-G2, which was strongly correlated with Bsl-G and Avd-G (Supplementary Figure 9b)
-In addition, the cluster was almost perfectly correlated with the average expression values of Bsl-G and Avd-G, supporting the hypothesis that it is indeed a mixture of the two clusters (Supplementary Figure 9b).
-This was further supported by the fact that Bsl-G and Avd-G are the most prevalent clusters in these samples (Table 1).
+The high number of cells that was loaded on the 10x increased the chance of doublet formation in the droplet encapsulation (CITATION).
+A major concern a doublets of two cell types as they can substantially confound downstream analysis.
+We thus identified clusters that appeared at low frequencies (less than the maximum doublet rate of 7\%, Supplementary Figure 9b) and that were highly similar to at least two other clusters present in the same sample ($p > 0.9$ of mean log2(counts+1)).
+The only cluster that fulfilled this criterion was cluster Bsl-G2, whose gene expression was strongly correlated with Bsl-G and Avd-G (Supplementary Figure 9b)
+In addition, the gene expression profile of this cluster is virtually identical to the average expression values of Bsl-G and Avd-G, supporting the hypothesis that it is indeed a mixture of the two clusters (Supplementary Figure 9b).
+As Bsl-G and Avd-G are the most prevalent clusters in these samples doublets would be expected to be enriched for a combination of the two (Table 1).
 Finally, Bsl-G2 also belonged to the clusters with a high number of genes detected and a high number of unique molecules, which is also indicative for doublet clusters (Supplementary Figure 9c).
 We hence excluded Bsl-G2 from any downstream analysis.
-As the last step we grouped clusters of cells together in super structures based on the expression of marker genes for known celltypes.
 
@@
 
-A gene set enrichment analysis based on gene-ontology (GO) terms was conducted to characterize genes that were up-regulated in Lp-PI compared to Lp.
-Genes with positive log fold changes were compared to all genes that were tested for differential expression using topGO with default settings\autocite{Alexa2016}.
+A gene set enrichment analysis based on gene-ontology (GO) terms was conducted to characterize various genesets in the analysis.
+The genes of interest were compared to all genes that were tested for differential expression using topGO with default settings\autocite{Alexa2016}.
 

### 2017-08-12 src/Figure1.R

@@ -70,7 +70,7 @@ dendr <- dis %>% hclust(.,method="ward.D2") %>% as.dendrogram %>%
     set("leaves_pch",19) %>%
     set("leaves_cex",2) %>%
     set("branches_lwd",3)
-par(cex=.7)
+par(cex=.6)
 plot(dendr,horiz=TRUE,yaxt="n")
 p.dendr <- grab_grob()
 p.dendr <- grid.arrange(p.dendr)
@@ -83,7 +83,6 @@ general <- c("Acta2","Mylk","Krt5","Krt14","Cnn1","Trp63","Epcam","Krt18","Krt8"
 c1 <- c("Cited1","Prlr","Esr1","Areg")
 c2 <- c("Rspo1","Atp6v1c2","Fabp3","Thrsp","Wap","Glycam1","Olah")
 c3 <- c("Foxa1","Ly6a","Aldh1a3","Kit","Cd14")
-# c4 <- c("Ltf","Hp","Pdk4","C4b","Chil1","Vegfa","Slpi")
 c5 <- c("Lypd3")
 c6 <- c("1500015O10Rik","Col7a1","Moxd1","Mia","Emid1","Pdpn","Col9a2","Fbln2","Igfbp3","Fst","Il17b")
 c7 <- c("Oxtr","Krt15","Igfbp6","Igfbp2","Tns1")
@@ -144,7 +143,7 @@ p <-  pheatmap(mheat,
          show_colnames=FALSE,
          annotation_legend=FALSE,
 	 annotation_col=annoCol,
-	 gaps_col=c(100,200,263,363,463,563,663,763,863,963,1052,1152,1252,1352),
+	 gaps_col=c(263,463,663,863,1052,1252,1352),
          gaps_row=9,
 	 annotation_colors=annoColors,
 	 fontsize=8)

### 2017-08-12 src/Figure3.R

@@ -234,7 +234,7 @@ htmps <- plot_grid(res[[1]][[1]][[4]],
 		   res[[2]][[1]][[4]],
 		   NULL,
 		   nrow=2,
-		   scale=0.8,
+		   scale=0.9,
 		   labels=c("e","","f",""),
 		   rel_widths=c(1,0.75,1,0.75))
 fullP <- plot_grid(expPlot,htmps,ncol=2)

### 2017-08-12 src/Figure4.R

@@ -144,7 +144,7 @@ pD <- right_join(pD,dms,by="barcode")
 levels(pD$SubCluster) <- c(levels(pD$SubCluster),"NP and G")
 pD$SubCluster[pD$Condition %in% c("NP","G")] <- "NP and G"
 pD$SubCluster <- factor(as.character(pD$SubCluster), levels=unique(as.character(pD$SubCluster)))
-cols <- unique(as.character(pD$Colors))[c(5,7,3,1,6,8,9,4,2)]
+cols <- unique(as.character(pD$Colors))[c(5,7,3,1,6,8,4,9,2)]
 cols[9] <- "#D3D3D3"
 
 p3 <- ggplot(filter(pD, Condition %in% c("L","PI")), aes(DC1,DC2,color=SubCluster)) +

### 2017-08-12 src/S6.R

@@ -19,7 +19,7 @@ fD <- dataList[[3]]
 
 # ---- Robustness -----
 condComb <- c("NP","G")
-set <- c("Bsl-G1","Bsl","Myo","Prc")
+set <- c("Bsl-G","Bsl","Myo","Prc")
 keepCells <- pD$keep & !(pD$SuperCluster %in% set) & pD$Condition %in% condComb
 
 # Cell filtering
@@ -90,7 +90,7 @@ for (feats in features) {
 }
 
 # Set color scheme
-cols <- levels(factor(pD.vp$Colors))
+cols <- levels(pD.vp$Colors)[levels(pD.vp$SubCluster) %in% unique(pD.vp$SubCluster)]
 
 # Plot
 out$SubSample <- factor(out$SubSample, levels=c("100%","50%","25%"))

### 2017-08-12 src/S8.R

@@ -131,10 +131,10 @@ pc <- prcomp(m.sub)
 pD.sub$PC1 <- pc$x[,1]
 pD.sub$PC2 <- pc$x[,2]
 
-cols <- levels(pD$Colors)
+cols <- levels(pD.sub$Colors)[levels(pD.sub$SubCluster) %in% unique(pD.sub$SubCluster)]
 pcplot <- ggplot(pD.sub, aes(PC1,PC2,color=SubCluster)) +
-    geom_point() 
-    #     scale_color_manual(values=cols)
+    geom_point() +
+    scale_color_manual(values=cols)
 
 loadngs <- pc$rotation[,1]
 loddf <- data.frame("id"=names(loadngs),

### 2017-08-12 src/S9.R

@@ -89,7 +89,7 @@ for (smpl in c("G1","G2")) {
 		     "Singleton"=rowMeans(tmp[,c("Bsl-G","Avd-G")]),
 		     "Cluster"="Mean(Avd-G,Bsl-G)")
     x4 <- rbind(x2,x3)
-    x4$Sample <- smpl
+    x4$Sample <- paste0("Sample-",smpl)
     xout <- rbind(xout,x4)
 }
 
@@ -100,7 +100,7 @@ p2 <- ggplot(xout, aes(Singleton, DoubletCluster)) +
     xlab("Log-Expression") +
     ylab("Log-Expression") +
     theme_bw() +
-    ylab("Total number of molecules detected")
+    ylab("Log-Expression of Bsl-G2")
 
 pD <- group_by(pD, SampleID) %>%
     dplyr::mutate(GenesDetected=GenesDetected/median(GenesDetected)) %>%
@@ -135,7 +135,8 @@ fp4$SubCluster <- factor(fp4$SubCluster, levels=ordr)
 p4 <- ggplot(fp4, aes(y=UmiSums, x=SubCluster)) +
     geom_boxplot() +
     scale_y_log10() +
-    theme_bw()
+    theme_bw() +
+    ylab("Total number of molecules detected")
 
 library(cowplot)
 

### 2017-08-21 paper/main.tex
added a few refs

@@
 In addition, all cells with 5\% or more of UMIs mapping to mitochondrial genes were defined as non-viable or apoptotic and removed from the analysis (Supplementary Fig 1d).
-We finally also ensured that none of the reads in our dataset derived from index swapping (CITATION).
+We finally also ensured that none of the reads in our dataset derived from index swapping\autocite{Sinha2017}.
 For this we excluded cells from the analysis with barcodes that appeared in more than one sample (non-unique barcodes).
@@
 The SNN-graph was constructed using the \texttt{buildSNNGraph} function from scran with the number of nearest-neighbors set to 20.
-Community-based clustering was then performed on the SNN-graph by multi-level modularity optimization using the \texttt{cluster\_louvain} function in igraph (CITATION).
+Community-based clustering was then performed on the SNN-graph by multi-level modularity optimization using the \texttt{cluster\_louvain} function in igraph\autocite{Csardi2006}.
 This identified a total of 13 clusters.
@@
 The dissimilarity matrix was then used to perform hierarchical clustering with the 'hclust' function in R using average linkage.
-Clusters were defined using the \texttt{cutree} function and the optimal $k$ was determined by maximizing the gap statistic using the \texttt{clusGap} function from the cluster package (CITATION).
+Clusters were defined using the \texttt{cutree} function and the optimal $k$ was determined by maximizing the gap statistic using the \texttt{clusGap} function from the cluster package\autocite{Maechler2017}.
 With this we identified a total number of 21 clusters.
@@
 Next, we flagged clusters that expressed clear markers of non-epithelial cells as contaminating cells and removed them from the downstream analysis.
-Cluster 6.1 and 6.2 were tagged as immune cells based on the expression of \textit{Cd74}, \textit{Cd72} and \textit{Cd54}\autocite{Scheele2017} and cluster 10.3 as endothelial cells based on the expression of \textit{Eng}, \textit{S1pr1} and \textit{Emcn} (CITATION, Supplementary Figure 5c, Supplementary Figure 5d).
-We retained cluster 10.2 as it expressed markers of a previously described celltype in the gland (CITATION), yet it was also positive for 2 out of 4 (\textit{Des} and \textit{Cspg4}) pericyte markers (CITATION) (Supplementary Figure 5e).
-The high number of cells that was loaded on the 10x increased the chance of doublet formation in the droplet encapsulation (CITATION).
+Cluster 6.1 and 6.2 were tagged as immune cells based on the expression of \textit{Cd74}, \textit{Cd72} and \textit{Cd54}\autocite{Scheele2017} and cluster 10.3 as endothelial cells based on the expression of \textit{Eng}, \textit{S1pr1} and \textit{Emcn} (Supplementary Figure 5c, Supplementary Figure 5d).
+We retained cluster 10.2 as it expressed markers of a previously described celltype in the gland\autocite{Wang2015}, yet it was also positive for 2 out of 4 (\textit{Des} and \textit{Cspg4}) pericyte markers\autocite{Armulik2011} (Supplementary Figure 5e).
+The high number of cells that was loaded on the 10x increased the chance of doublet formation in the droplet encapsulation\autocite{Zheng2017}.
 A major concern a doublets of two cell types as they can substantially confound downstream analysis.

### 2017-08-22 paper/main.tex
added comments from Dave

@@
 Therefore, there is a need for an unbiased and comprehensive description of mammary epithelial cells within the gland at different developmental stages.
-To this end we herein use single cell RNA sequencing (scRNAseq) to determine the gene expression profile of individual mammary epithelial cells across four adult developmental stages; nulliparous, mid gestation, lactation and post weaning (full natural involution).
+To this end we herein used single cell RNA sequencing (scRNAseq) to determine the gene expression profile of individual mammary epithelial cells across four adult developmental stages; nulliparous, mid gestation, lactation and post weaning (full natural involution).
 Our data from 25,010 individual cells identifies 8 distinct mammary epithelial cell populations and allows their hierarchical structure across development to be charted.
@@
 For example, our analysis revealed a cluster of luminal progenitor cells in post involution glands, which is distinct from progenitors found in nulliparous glands.
-The data also showed that only few clusters could be fully characterized by a single marker gene.
+The data also showed that only a few clusters could be fully characterized by a single marker gene.
 We argue instead that the epithelial cells -- especially in the luminal compartment -- should rather be conceptualized as being part of a continuous spectrum of differentiation.
@@
 \section{Introduction}
-The purpose of the mammary gland is to provide nourishment and passive immunity to the young until they are capable of feeding themselves.
+The purpose of the mammary gland is to provide nourishment and passive immunity for the young until they are capable of feeding themselves.
 From a developmental biology perspective, the mammary gland is a unique organ as it undergoes most of its development during puberty and adulthood\autocite{Watson2008,Hennighausen2005,Hennighausen1998,Inman2015}
@@
 However, in response to cessation of suckling the gland undergoes involution, which is characterised by extensive cell death and tissue remodelling \autocite{Inman2015,Watson2006} .
-Towards the end of involution the gland reaches a morphology resembling that prior to pregnancy and subsequent pregnancies will trigger the same chain of events.
+Towards the end of involution the gland reaches a morphology resembling that of glands prior to pregnancy and subsequent pregnancies will trigger the same chain of events.
 
@@
 Pioneering fat-pad transplantation studies nearly 70 years ago were the first to demonstrate the regenerative and differentiation capacity of small numbers of cells\autocite{Faulkin1960,Daniel197,Smalley1998}.
-More recently the use of cell surface markers coupled with flow cytometry has been used to enrich for various progenitor and stem cell compartments\autocite{Smalley1998,Stingl2006,Shackleton2006,Asselin2007} and shown that the imbalance of such cell populations results in cellular transformation and subsequently breast cancer\autocite{Lim2009,Molyneux2010}.
-Other studies, inspired by breast cancer transcriptomic profiling, have identified transcriptional regulators of mammary epithelial cell types such as \textit{GATA3} in the luminal cells\autocite{Asselin2007,Kouros2006}.
+More recently the use of cell surface markers coupled with flow cytometry has been used to enrich for various progenitor and stem cell compartments\autocite{Smalley1998,Stingl2006,Shackleton2006,Asselin2007} and showed the imbalance of such cell populations results in cellular transformation and subsequently breast cancer\autocite{Lim2009,Molyneux2010}.
+Other studies, inspired by breast cancer transcriptomic profiling, have identified transcriptional regulators of mammary epithelial cell types such as \textit{GATA3} in luminal cells\autocite{Asselin2007,Kouros2006}.
 More recently, elegant lineage-tracing studies used key markers to address the contribution of each lineage to adult mammary epithelial cell homeostasis\autocite{Inman2015}.
@@
 
-To this end we herein use single cell RNA sequencing (scRNAseq) to map the cellular dynamics of mammary epithelial cells across four adult developmental stages; nulliparous, mid gestation, lactation and post weaning (full natural involution).
+Here we used single cell RNA sequencing (scRNAseq) to map the cellular dynamics of mammary epithelial cells across four adult developmental stages; nulliparous, mid gestation, lactation and post weaning (full natural involution).
 Our data from 25,010 individual cells, identifies 15 distinct cell populations within the gland and allows their hierarchical structure across developmental time points to be charted.
@@
 In a second step we used hierarchical clustering on each of the identified groups to further resolve the cellular heterogeneity (\autoref{F1}c, Supplementary Figure 5b).
-After removal of enodthelial and immune cell clusters this resulted in a total of 15 clusters of mammary epithelial cells.
+After removal of endothelial and immune cell clusters this resulted in a total of 15 clusters of mammary epithelial cells.
 Differential gene expression analysis allowed us to further characterise the clusters and infer putative identities (\autoref{F1}d, \autoref{T1} and Supplementary Table 1).
@@
 We would like to thank the staff at Sanger Institute, Research Service Facility (RSF) for their assistance.
-We would like to thank Dr. Aaron T. Lun (CRUK CI) for helpful discussions and comments on manuscript.
+We would like to thank Dr. Aaron T. Lun (CRUK CI) for helpful discussions and comments on the manuscript.
 K.B. is funded by a Cambridge Cancer Centre studentship.

### 2017-08-22 src/ShinyApp.R

@@ -1,193 +1,0 @@
-library(topGO)
-library(org.Mm.eg.db)
-library(shiny)
-library(dplyr)
-library(gtools)
-library(knitr)
-library(ggplot2)
-library(DT)
-source(file.path("functions.R"))
-
-dataList <- readRDS(file.path("../data/Robjects/secondRun_2500/ExpressionList_Clustered.rds"))
-deList <- readRDS(file.path("../data/Robjects/secondRun_2500/DEList.rds"))
-#Commence
-m <- dataList[[1]]
-pD <- dataList[[2]]
-fD <- dataList[[3]]
-
-#Pre-Filtering before DE-Analysis
-# Cells
-keepCells <- pD$PassAll & ! (pD$isImmuneCell | pD$isOutlier)
-m <- m[,keepCells]
-pD <- pD[keepCells,]
-
-# Genes
-keep <- rowSums(m)>10
-m <- m[keep,]
-fD <- fD[keep,]
-
-#Normalization
-m <- t(t(m)/pD$sf)
-
-#Add cell surface annotation
-library(scLVM)
-surface <- getEnsembl("GO:0009986")
-cellCycle <- getEnsembl("GO:0007049")
-tfs <- filter(fD, TranscriptionFactor) %>% .$id
-
-deAdd <- select(fD, id,symbol) 
-colnames(deAdd) <- c("Gene","GeneSymbol")
-deList <- lapply(deList, function(x) {
-		      nw <- full_join(x,deAdd)
-		      nw$Surface <- nw$Gene %in% surface
-		      nw$CC <- nw$Gene %in% cellCycle
-		      nw$TF <- nw$Gene %in% tfs
-		      nw <- nw[,-2]
-		      cols <- c("GeneSymbol",grep("top",colnames(nw),value=TRUE, #reorder cols
-						  ignore.case=TRUE))
-		      nw <- nw[,c(cols,setdiff(colnames(nw), cols))]})
-names(deList) <- c(1,2,3,4,5,6,7,9,10)
-
-ui <- shinyUI(fluidPage(
-
-  titlePanel("Tabsets"),
-
-  sidebarLayout(
-    
-    sidebarPanel(
-    inputPanel(
-	       selectInput("colorBy", label = "Color By",
-			   choices= c("cluster","Condition"), selected="cluster"),
-	       selectInput("cNumber", label = "Cluster",
-			   choices= as.character(c(1,2,3,4,5,6,7,9,10)), selected="1"),
-	       selectInput("ontology", label = "Ontology",
[…]

### 2017-08-22 paper/main.tex
incluced Saras comments

@@
 Characterising the hierarchy of the various mammary epithelial cells and how they are regulated in response to gestation, lactation and involution is important for understanding how breast cancer develops.
-Recent studies have used various markers to enrich, isolate and characterise the different epithelial cell compartments within the adult mammary gland.
+Recent studies have used numerous markers to enrich, isolate and characterise the different epithelial cell compartments within the adult mammary gland.
 However, in all of these studies only a handful of markers were used to isolate, define and trace cell populations.
@@
 The purpose of the mammary gland is to provide nourishment and passive immunity for the young until they are capable of feeding themselves.
-From a developmental biology perspective, the mammary gland is a unique organ as it undergoes most of its development during puberty and adulthood\autocite{Watson2008,Hennighausen2005,Hennighausen1998,Inman2015}
-In the pre-pubertal mouse the mammary gland consists of a rudimentary epithelial ductal structure embedded within a mammary fatpad, which is connected to the nipple\autocite{Mikkola2006,Hens2005}
-At the onset of puberty and in response to hormonal changes, the rudimentary ductal structure will proliferate and migrate to fill the entire mammary fatpad leaving a developed network of ductal structures that later serve as channels for milk transport during lactation.
+From a developmental biology perspective, the mammary gland is a unique organ as it undergoes most of its development during puberty and adulthood\autocite{Watson2008,Hennighausen2005,Hennighausen1998,Inman2015}.
+In the pre-pubertal mouse the mammary gland consists of a rudimentary epithelial ductal structure embedded within a mammary fat pad, which is connected to the nipple\autocite{Mikkola2006,Hens2005}.
+At the onset of puberty and in response to hormonal changes, the rudimentary ductal structure will proliferate and migrate to fill the entire mammary fat pad leaving a developed network of ductal structures that later serve as channels for milk transport during lactation.
 At the onset of pregnancy a highly proliferative stage is initiated characterised by further ductal side-branching and widespread lobuloalveolar development\autocite{Watson2008}.
@@
 This morphology is largely maintained throughout lactation.
-However, in response to cessation of suckling the gland undergoes involution, which is characterised by extensive cell death and tissue remodelling \autocite{Inman2015,Watson2006} .
+However, in response to cessation of suckling the gland undergoes involution, which is characterised by extensive cell death and tissue remodelling \autocite{Inman2015,Watson2006}.
 Towards the end of involution the gland reaches a morphology resembling that of glands prior to pregnancy and subsequent pregnancies will trigger the same chain of events.
@@
 Recent efforts have focused on the identification and characterisation of the various mammary epithelial cell lineages within the gland that contribute to this developmental homeostasis.
-Pioneering fat-pad transplantation studies nearly 70 years ago were the first to demonstrate the regenerative and differentiation capacity of small numbers of cells\autocite{Faulkin1960,Daniel197,Smalley1998}.
+Pioneering fat pad transplantation studies nearly 70 years ago were the first to demonstrate the regenerative and differentiation capacity of small numbers of cells\autocite{Faulkin1960,Daniel197,Smalley1998}.
 More recently the use of cell surface markers coupled with flow cytometry has been used to enrich for various progenitor and stem cell compartments\autocite{Smalley1998,Stingl2006,Shackleton2006,Asselin2007} and showed the imbalance of such cell populations results in cellular transformation and subsequently breast cancer\autocite{Lim2009,Molyneux2010}.
@@
 
-First, we isolated mammary epithelial cells from four developmental time points; nulliparous (NP), day 14.5 gestation (G), day 6 lactation (L) and two weeks post natural involution (PI) (Supplementary Figure 2).
-For each time point we sorted mammary epithelial cells based on the Epcam cell surface marker from two independent mice (\autoref{F1}a).
+First, we isolated mammary epithelial cells from four developmental time points; nulliparous (NP), day 14.5 gestation (G), day 6 lactation (L) and 11 days post natural involution (PI) (Supplementary Figure 2).
+For each time point we sorted mammary epithelial cells based on the EpCAM cell surface marker from two independent mice (\autoref{F1}a).
 All samples were then prepared for single-cell RNA-sequencing using the 10x Chromium platform\autocite{Zheng2017}.
@@
 \textbf{Supplementary Figure 1. Estrus cycle}
-Vaginal smears of the mice from the NP and PI timepoints. The animals were classified as (CHECK). The scale bar represents 200$\mu m$.
+Vaginal smears of the mice from the NP and PI timepoints. The animals were classified as being in estrus (NP1, NP2, PI1) or diestrus (PI1). The scale bar represents 200$\mu m$.
 
 \textbf{Supplementary Figure 2. Gating strategy to isolate MECs}
-\textbf{(a)} Gating strategy used to select live, lineage\textsuperscript{-} singlets.
-\textbf{(b)} Sorting of epithelial cells based on Epcam. Representative plots for each of the four timepoints are shown.
+\textbf{(a)} Gating strategy used to select live, lineage\textsuperscript{negative} singlets.
+\textbf{(b)} Representative plots for each of the four timepoints are shown. FSC-W: forward scatter width, FSC-A: forward scatter area, BSC-A: backward scatter area.
 
@@
 
-Lymph node divested mammary glands (excluding the cervical pair and the proximal region of one of the fourth glands) were dissected from the mice and mechanically dissociated.
-The finely minced tissue was transferred to a digestion mix consisting of DMEM/F12 (Gibco) + 10 mM HEPES (Gibco) + 1 mg ml\textsuperscript{-1} collagenase (Roche) + 100 U ml\textsuperscript{-1} hyaluronidase (Sigma) + 50 mycg ml\textsuperscript{-1} gentamicin (Gibco) for 2.5 hours at \SI{37}{\celsius} and vortexed every 30 minutes.
-After the lysis of red blood cells in NH\textsubscript{4}Cl, cells were briefly digested with warm 0.05\% Trypsin-EDTA (Gibco), 5 mg ml\textsuperscript{-1} dispase (Sigma) and 1 mg ml\textsuperscript{-1} DNase (Sigma), and filtered through a 40 mycm cell strainer (BD Biosciences).
+Lymph node divested mammary glands (excluding the cervical pair) were dissected from the mice and mechanically dissociated.
+The finely minced tissue was transferred to a digestion mix consisting of DMEM/F12 (Gibco) + 10 mM HEPES (Gibco) + \SI{2}{\milli\gram\per\milli\litre} collagenase (Roche) + 200 U ml\textsuperscript{-1} hyaluronidase (Sigma) + \SI{50}{\micro\gram\per\milli\litre} gentamicin (Gibco) for 3 hours at \SI{37}{\celsius} and vortexed every 30 minutes.
+After the lysis of red blood cells in NH\textsubscript{4}Cl, cells were briefly digested with warm 0.05\% Trypsin-EDTA (Gibco), 5 mg ml\textsuperscript{-1} dispase (Sigma) and 1 mg ml\textsuperscript{-1} DNase (Sigma), and filtered through a \SI{40}{\micro\meter} cell strainer (BD Biosciences).
 
@@
 All antibody incubations were performed for 10 min on ice in HF media.
-Mammary cells were stained with the following primary antibodies: 1 mycg ml\textsuperscript{-1} CD31-biotin (eBioscience); 1 mycg ml\textsuperscript{-1} CD45-biotin (eBioscience); 1 mycg ml\textsuperscript{-1} Ter119-biotin (eBioscience) and 0.5 mycg ml\textsuperscript{-1} EpCAM--PE (Biolegend).
-Cells were then stained with 0.4 mycg ml\textsuperscript{-1} streptavidin-PE-CF594 (BD-Biosciences).
-Propidium Iodide (PI, 1 mycg ml\textsuperscript{-1}; Sigma) was used to detect dead cells.
-Cells were filtered through a 30 mycm cell strainer (Partec) before sorting.
-Sorting of cells was done using a SH800 sorter (SONY).
+Mammary cells were stained with the following primary antibodies: \SI{1}{\micro\gram\per\milli\litre} CD31-biotin (eBioscience, clone 390); \SI{1}{\micro\gram\per\milli\litre} CD45-biotin (eBioscience, clone 30F11); \SI{1}{\micro\gram\per\milli\litre} Ter119-biotin (eBioscience, clone Ter119) and \SI{0.5}{\micro\gram\per\milli\litre} EpCAM-PE (Biolegend, clone G8.8).
+Cells were then stained with \SI{0.4}{\micro\gram\per\milli\litre} streptavidin-PE-CF594 (BD-Biosciences).
+Propidium Iodide (PI, \SI{1}{\micro\gram\per\milli\litre}; Sigma) was used to detect dead cells.
+Cells were filtered through a \SI{40}{\micro\meter} cell strainer (Partec) before sorting.
+Sorting of cells was done using a SH800Z sorter (SONY).
 Single-stained control cells were used to perform compensation manually and unstained cells were used to set gates.
 Chip alignment and sorting calibration was performed with automatic setup beads (SONY) immediately prior to sorting.
-Doublets, dead cells and contaminating haematopoietic, endothelial and stromal cells were gated out and EpCAM positive cells were sorted in FACS tubes (BD Biosciences) containing 1 mL of HF.
-After sorting, cells were transferred to a 1.5 ml tube through a 30 mycm cell strainer, spun down and resuspended in 15-50 mycl of HF according to the number of cells sorted per sample.
-Representative samples were manually counted using an improved Neubauer chamber to estimate sample loss from centrifugation and filtering.
+Doublets, dead cells and contaminating haematopoietic, endothelial and stromal cells were gated out and EpCAM positive cells were sorted in LoBind\textsuperscript{\textregistered} microcentrifuges tubes (Eppendorf) with \SI{100}{\micro\litre} HF.
+After sorting cells were spun down and resuspended in \SI{50}{\micro\litre} of HF. 
+Samples were manually counted using an improved Neubauer chamber and the cell concentration was normalized by addition of HF.
 Equal numbers of cells per sample were processed for scRNA library preparation.
-Samples were processed for scRNA library preparation within 10 hours from tissue isolation.
+Samples were processed for scRNA library preparation within 9 hours from tissue isolation.
 

### 2017-08-24 paper/main.tex
included Johns comments

@@
 Recent studies have used numerous markers to enrich, isolate and characterise the different epithelial cell compartments within the adult mammary gland.
-However, in all of these studies only a handful of markers were used to isolate, define and trace cell populations.
+However, in all of these studies only a handful of markers were used to define and trace cell populations.
 Therefore, there is a need for an unbiased and comprehensive description of mammary epithelial cells within the gland at different developmental stages.
-To this end we herein used single cell RNA sequencing (scRNAseq) to determine the gene expression profile of individual mammary epithelial cells across four adult developmental stages; nulliparous, mid gestation, lactation and post weaning (full natural involution).
+To this end we used single cell RNA sequencing (scRNAseq) to determine the gene expression profile of individual mammary epithelial cells across four adult developmental stages; nulliparous, mid gestation, lactation and post weaning (full natural involution).
 Our data from 25,010 individual cells identifies 8 distinct mammary epithelial cell populations and allows their hierarchical structure across development to be charted.
@@
 For example, our analysis revealed a cluster of luminal progenitor cells in post involution glands, which is distinct from progenitors found in nulliparous glands.
-The data also showed that only a few clusters could be fully characterized by a single marker gene.
+The data also showed that few clusters could be fully characterized by a single marker gene.
 We argue instead that the epithelial cells -- especially in the luminal compartment -- should rather be conceptualized as being part of a continuous spectrum of differentiation.
@@
 In the pre-pubertal mouse the mammary gland consists of a rudimentary epithelial ductal structure embedded within a mammary fat pad, which is connected to the nipple\autocite{Mikkola2006,Hens2005}.
-At the onset of puberty and in response to hormonal changes, the rudimentary ductal structure will proliferate and migrate to fill the entire mammary fat pad leaving a developed network of ductal structures that later serve as channels for milk transport during lactation.
-At the onset of pregnancy a highly proliferative stage is initiated characterised by further ductal side-branching and widespread lobuloalveolar development\autocite{Watson2008}.
+At the onset of puberty and in response to hormonal changes, the rudimentary ductal structure will proliferate and migrate to fill the entire mammary fat pad, leaving a developed network of ductal structures that later serve as channels for milk transport during lactation.
+At the onset of pregnancy a highly proliferative stage is initiated, characterised by further ductal side-branching and widespread lobuloalveolar development\autocite{Watson2008}.
 Differentiation of the epithelial cells within alveoli prepares the gland for milk production and secretion.
@@
 This resulted in a coarse clustering into 13 groups (Supplementary Figure 5a).
-In a second step we used hierarchical clustering on each of the identified groups to further resolve the cellular heterogeneity (\autoref{F1}c, Supplementary Figure 5b).
+In a second step we applied hierarchical clustering to each of the identified groups to further resolve the cellular heterogeneity (\autoref{F1}c, Supplementary Figure 5b).
 After removal of endothelial and immune cell clusters this resulted in a total of 15 clusters of mammary epithelial cells.
@@
 Interestingly, \textit{Hmga1} is a known chromatin remodeller with reprogramming activity that has been implicated in basal-like breast cancer\autocite{Shah2012,Shah2013}.
-In addition, we identified 1005 genes with branch-specific expression patterns (\autoref{F3}d, \autoref{F3}e Supplementary Table 3).
-We further clustered the gene expression trend on each of the two branches and identified sets of genes that change over pseudotime (Supplementary Figure 7).
+In addition, we identified 1005 genes with branch-specific expression patterns (\autoref{F3}d, \autoref{F3}e, Supplementary Table 3).
+We further clustered the gene expression trend on each of the two branches and identified sets of genes that change over pseudotime (\fixme{Supplementary Figure 7}).
 On the alveolar branch we found two clusters of genes that increase in expression during differentiation, one that is switched on early and another that is activated later in the final stages of differentiation.
@@
 Genes with increasing expression are involved in hormone metabolic processes and the regulation of morphogenesis of an epithelium (e.g. \textit{Vdr}, \textit{Esr1}, \textit{Pgr} and \textit{Msx2}).
-Amongst the branch-specific transcription factors we found for example \textit{Runx1, Tox2} and \textit{Bhlhe41} (also known as \textit{Sharp1}) to be transcribed during differentiation towards the hormone-sensing lineage (\autoref{F3}d) and \textit{Elf5, Foxs1} and \textit{Ehf} in the secretory lineage (\autoref{F3}e).
+Amongst the branch-specific transcription factors we found for example \textit{Runx1, Tox2} and \textit{Bhlhe41} (also known as \textit{Sharp1}) to be transcribed during differentiation towards the hormone-sensing lineage (\autoref{F3}c) and \textit{Elf5, Foxs1} and \textit{Ehf} in the secretory lineage (\autoref{F3}d).
 \textit{Runx1} is a known repressor of \textit{Elf5} and its deletion has been shown to be deleterious for ductal morphogenesis\autocite{VanBragt2014}.
-The expression of the known progenitor master regulator \textit{Elf5} is maintained and further increased during secretory differentiation suggesting that its transcriptional level is fine-tuned in luminal progenitors (\autoref{F3}e).
+The expression of the known progenitor master regulator \textit{Elf5} is maintained and further increased during secretory differentiation suggesting that its transcriptional level is fine-tuned in luminal progenitors (\autoref{F3}d).
 
@@
 Some of the genes that are involved in the immune response are known regulators of the involution process such as \textit{Ctsc}, \textit{Tgfb3} or \textit{Mfge8}.
-The up-regulation of these genesets was also present in the Hsp and in some cases even the differentiated Hsd cluster, yet the effect remained strongest in the progenitor cells (\autoref{F4}c).
+The up-regulation of these genesets was also present in the Hsp and in some cases even the differentiated Hsd cluster, yet the effect remained strongest in the progenitor cells (\autoref{F4}c-d).
 Finally, we compared the differentiation processes of the luminal compartment of the parous gland to the nulliparous gland by mapping the luminal cells from PI to the trajectory of NP and G cells.
-Interestingly, the Hsp and Hsd cluster generally maintain their position of their nulliparous counterpart in the differentiation hierarchy whereas the Lp-PI cluster is stretched out from the origin down towards the alveolar branch, suggesting that these are biased towards the alveolar fate (\autoref{F4}d).
+Interestingly, the Hsp and Hsd cluster generally maintain the position of their nulliparous counterpart in the differentiation hierarchy whereas the Lp-PI cluster is stretched out from the origin down towards the alveolar branch, suggesting that these are biased towards the alveolar fate (\autoref{F4}e).
 Together, the data suggest that luminal progenitor cells maintain memory of having undergone gestation and involution.
@@
 Furthermore, we characterised gene expression patterns along the differentiation hierarchy.
-Here we identified genes that are lineage-specific, thus enabling to disentangle the transcriptional events that regulate differentiation of the luminal compartment.
+Here we identified genes that are lineage-specific, thus enabling us to disentangle the transcriptional events that regulate differentiation of the luminal compartment.
 By analysing the mammary gland at various stages of development we were also able to describe the effects of parity at cellular resolution.
@@
 Cells are coloured by the four developmental time points Pink=NP, Dark Green=G, Light Green=L, Purple=PI.
-\textbf{(c)} Dendrogram of clusters based on the mean expression values of the 15 clusters. The tree was computed based on Spearman's rank correlation with ward linkage.
+\textbf{(c)} Dendrogram of clusters based on the mean expression values of the 15 clusters. The tree was computed based on Spearman's rank correlation with Ward linkage.
 \textbf{(d)} Same as \textbf{(b)} but colouring cells by clusters.
@@
 \textbf{(b)} Differentiation trajectory of the luminal compartment based on the first two diffusion components.
-\textbf{(c)} Same plot as in \textbf{(b)}, colored by the normalized and scaled expression values of various genes.
+\textbf{(c)} Same plot as in \textbf{(b)}, coloured by the normalized and scaled expression values of various genes.
 %-------------------------------------------------------------------------------------------------------------------
@@
 In total the Cell Ranger software identified 25,806 barcodes that contained enough unique molecules to be considered as cells (4,376 in NP, 6,021 in G, 9,603 in L and 5,806 in PI).
-Libraries prepared from all time points showed high quality that was reproducible between the two biological replicates (Supplementary 1a).
-We used the following metrics to flag poor quality cells: number of genes detected, total number of unique molecules (UMIs) and percentage of molecules mapped to mitochondrial genes.
+Libraries prepared from all time points showed high quality that was reproducible between the two biological replicates (Supplementary 2a).
+We used the following metrics to flag poor quality cells: number of genes detected, total number of unique molecular identifiers (UMIs) and percentage of molecules mapped to mitochondrial genes.
 Poor quality cells were then identified by setting a threshold on the number of genes and number of UMIs that was defined as three median absolute deviations (MAD) below the median for each sample, while requiring a minimum value of 1,000 total molecules and 500 genes detected.
 This resulted in the following thresholds for total number of genes detected: 1,042 for NP, 836 for G, 500 for L and 759 for PI; and the following for total number of molecules : 2,012 in NP, 1,479 in G, 1,000 in L and 1,379 in G.
-In addition, all cells with 5\% or more of UMIs mapping to mitochondrial genes were defined as non-viable or apoptotic and removed from the analysis (Supplementary Fig 1d).
-We finally also ensured that none of the reads in our dataset derived from index swapping\autocite{Sinha2017}.
-For this we excluded cells from the analysis with barcodes that appeared in more than one sample (non-unique barcodes).
-The reasoning being that any index swapped read between samples will also share the same cellular barcode.
+In addition, all cells with 5\% or more of UMIs mapping to mitochondrial genes were defined as non-viable or apoptotic and removed from the analysis (Supplementary Figure 2d).
+We finally also ensured that none of the reads in our dataset derived from index swapping\autocite{Sinha2017, Griffiths2017}.
+For this we excluded cells with barcodes that appeared in more than one sample (non-unique barcodes).
+The reasoning behind this being that any index swapped read between samples will also share the same cellular barcode.
 Exclusion of these cells hence offers protection against index swapping.
@@
 In a first step we clustered cells using a shared-nearest-neighbor graph (SNN-graph) based approach.
-This has the advantage of not requiring the computation of a distance matrix which is computationally inhibitive for a large number of cells.
+This has the advantage of not requiring the computation of a distance matrix which is computationally prohibitive for large numbers of cells.
 The SNN-graph was constructed using the \texttt{buildSNNGraph} function from scran with the number of nearest-neighbors set to 20.
@@
 This identified a total of 13 clusters.
-Yet, the majority of clusters still contained substructure as evident from the t-SNE (Supplementary Figure 5a).
-In order to resolve the cellular heterogeneity further we then applied agglomerative hierarchical clustering on each of the 13 clusters (Supplementary Figure 5b).
-For this, we first computed the pairwise distances between cells based on the Spearman's rank correlation of log-transformed HVG counts.
-The dissimilarity matrix was then used to perform hierarchical clustering with the 'hclust' function in R using average linkage.
+Despite this, the majority of clusters still contained substructure as evident from the t-SNE (Supplementary Figure 5a).
+To resolve the cellular heterogeneity further we then applied agglomerative hierarchical clustering on each of the 13 clusters (Supplementary Figure 5b).
+For this, we first computed the pairwise distances between cells based on Spearman's rank correlation of log-transformed HVG counts.
+The dissimilarity matrix was then used to perform hierarchical clustering with the \texttt{hclust} function in R using average linkage.
 Clusters were defined using the \texttt{cutree} function and the optimal $k$ was determined by maximizing the gap statistic using the \texttt{clusGap} function from the cluster package\autocite{Maechler2017}.
@@
 As the last step we performed a \textit{post-hoc} test as previously reported\autocite{Macosko2015} by merging clusters with less than 10 differentially expressed genes (at a P value threshold of 0.01 and minimum log fold change of 1), resulting in one merged cluster (Myo).
-Next, we flagged clusters that expressed clear markers of non-epithelial cells as contaminating cells and removed them from the downstream analysis.
-Cluster 6.1 and 6.2 were tagged as immune cells based on the expression of \textit{Cd74}, \textit{Cd72} and \textit{Cd54}\autocite{Scheele2017} and cluster 10.3 as endothelial cells based on the expression of \textit{Eng}, \textit{S1pr1} and \textit{Emcn} (Supplementary Figure 5c, Supplementary Figure 5d).
-We retained cluster 10.2 as it expressed markers of a previously described celltype in the gland\autocite{Wang2015}, yet it was also positive for 2 out of 4 (\textit{Des} and \textit{Cspg4}) pericyte markers\autocite{Armulik2011} (Supplementary Figure 5e).
-The high number of cells that was loaded on the 10x increased the chance of doublet formation in the droplet encapsulation\autocite{Zheng2017}.
-A major concern a doublets of two cell types as they can substantially confound downstream analysis.
-We thus identified clusters that appeared at low frequencies (less than the maximum doublet rate of 7\%, Supplementary Figure 9b) and that were highly similar to at least two other clusters present in the same sample ($p > 0.9$ of mean log2(counts+1)).
+Next, we flagged clusters that expressed clear markers of non-epithelial cells as ``contaminating cells'' and removed them from the downstream analysis.
+Clusters 6.1 and 6.2 were tagged as immune cells based on the expression of \textit{Cd74}, \textit{Cd72} and \textit{Cd54}\autocite{Scheele2017} and cluster 10.3 as endothelial cells based on the expression of \textit{Eng}, \textit{S1pr1} and \textit{Emcn} (Supplementary Figure 5c, Supplementary Figure 5d).
+We retained cluster 10.2 as it expressed markers of a previously described celltype in the gland\autocite{Wang2015}, despite being positive for 2 out of 4 (\textit{Des} and \textit{Cspg4}) pericyte markers\autocite{Armulik2011} (Supplementary Figure 5e).
+The high number of cells loaded on the 10x increased the chance of doublet formation in the droplet encapsulation\autocite{Zheng2017}, \fixme{which can result in artificial clusters}.
+We thus identified clusters that appeared at low frequencies (less than the maximum doublet rate of 7\%, Supplementary Figure 9b) and that were highly similar to at least two other clusters present in the same sample ($\rho > 0.9$ of mean log2(counts+1)).
 The only cluster that fulfilled this criterion was cluster Bsl-G2, whose gene expression was strongly correlated with Bsl-G and Avd-G (Supplementary Figure 9b)
 In addition, the gene expression profile of this cluster is virtually identical to the average expression values of Bsl-G and Avd-G, supporting the hypothesis that it is indeed a mixture of the two clusters (Supplementary Figure 9b).
-As Bsl-G and Avd-G are the most prevalent clusters in these samples doublets would be expected to be enriched for a combination of the two (Table 1).
+As Bsl-G and Avd-G are the most prevalent clusters in these samples (Table 1), doublets would be expected to be enriched for a combination of the two.
 Finally, Bsl-G2 also belonged to the clusters with a high number of genes detected and a high number of unique molecules, which is also indicative for doublet clusters (Supplementary Figure 9c).
@@
 Differential gene expression analysis was performed using ``edgeR''\autocite{Robinson2010}.
-Genes with a mean expression level below 0.1 counts were removed from the analysis.
+\fixme{Genes with a mean expression level below 0.1 counts were removed from the analysis.}
 A negative binomial generalized log-linear model was fitted to the remaining genes with the cluster assignments as covariate(s).
@@
 First, we selected all cells from the NP and G time point (\autoref{F2}a) and detected the HVGs as described above.
-The log-transformed (log2(count+1)) gene counts were then used to compute the first twenty diffusion components using the \texttt{DiffusionMap} function (default parameters as in 'destiny'\autocite{Angerer2016}.
+The log-transformed (log2(count+1)) gene counts were then used to compute the first twenty diffusion components using the \texttt{DiffusionMap} function (default parameters as in 'destiny'\autocite{Angerer2016}).
 After inspecting the fraction of explained variation for the first twenty components, we decided to retain the first 3.
 In \autoref{F2}b we then focused on the luminal compartment and recomputed the diffusion map based only on the luminal cells, using the aforementioned procedure.
-Here we retained only the first two components (Supplementary Fig 3a).
+\fixme{Here we retained only the first two components (Supplementary Fig 3a).
 Notably, the structure inferred by the diffusion map algorithm was robust to the choice of features and down-sampling of cells (Supplementary Fig 3d).
 The structure of a common origin and the two branches could also be inferred using Monocle with standard settings\autocite{Trapnell2014} (Supplementary Fig 3c).
+}
 For inferring the branches and pseudotime ordering, we defined the following three tips, the cell with the smallest value for the second eigenvector (which was set as root) and the cells with the largest and smallest values for the first eigenvector (compare \autoref{F2}b).
@@
 
-All computational analyses were performed in R (Version 3.3.3) using standard functions unless otherwise indicated.
+\fixme{All computational analyses were performed in R (Version 3.3.3) using standard functions unless otherwise indicated.
 Code is available online at \href{https://github.com/MarioniLab/MammaryGland}{\textit{https://github.com/MarioniLab/MammaryGland}}.
 Data can be accessed at this link (\url{https://karstenbach.shinyapps.io/webapp/}).
+}
 
@@
 
-\textbf{Acknowledgements }
+\textbf{Acknowledgements}
 

### 2017-08-25 paper/main.tex
changed supp order

@@
 
-First, we isolated mammary epithelial cells from four developmental time points; nulliparous (NP), day 14.5 gestation (G), day 6 lactation (L) and 11 days post natural involution (PI) (Supplementary Figure 2).
+First, we isolated mammary epithelial cells from four developmental time points; nulliparous (NP), day 14.5 gestation (G), day 6 lactation (L) and 11 days post natural involution (PI) (Supplementary Figure 1).
 For each time point we sorted mammary epithelial cells based on the EpCAM cell surface marker from two independent mice (\autoref{F1}a).
 All samples were then prepared for single-cell RNA-sequencing using the 10x Chromium platform\autocite{Zheng2017}.
-Following quality control (materials and methods), this yielded an average of 6,175 unique transcripts and 2,118 genes detected from 25,010 cells (4,223 in NP, 5,826 in G, 9,319 in L and 5,642 in P) (Supplementary Figure 3).
-Visual inspection of the data using t-distributed stochastic neighbor embedding (t-SNE) suggested that although there is grouping of cells by time-point, there are also other factors that underlie structure within the dataset (\autoref{F1}b and Supplementary Figure 4).
+Following quality control (materials and methods), this yielded an average of 6,175 unique transcripts and 2,118 genes detected from 25,010 cells (4,223 in NP, 5,826 in G, 9,319 in L and 5,642 in P) (Supplementary Figure 2).
+Visual inspection of the data using t-distributed stochastic neighbor embedding (t-SNE) suggested that although there is grouping of cells by time-point, there are also other factors that underlie structure within the dataset (\autoref{F1}b and Supplementary Figure 3).
 First, we dissected the global structure by unsupervised clustering using a shared-nearest-neighbor clustering approach.
-This resulted in a coarse clustering into 13 groups (Supplementary Figure 5a).
-In a second step we applied hierarchical clustering to each of the identified groups to further resolve the cellular heterogeneity (\autoref{F1}c, Supplementary Figure 5b).
+This resulted in a coarse clustering into 13 groups (Supplementary Figure 4a).
+In a second step we applied hierarchical clustering to each of the identified groups to further resolve the cellular heterogeneity (\autoref{F1}c, Supplementary Figure 4b).
 After removal of endothelial and immune cell clusters this resulted in a total of 15 clusters of mammary epithelial cells.
@@
 Notably, some clusters mainly represent the same putative cell type at different developmental time points (indicated by the NP, G, L or PI suffix, in the following the suffix might be omitted to refer to a group of clusters).
-Based on the expression of \textit{Krt18,} \textit{Krt8, Krt5, Krt14} and \textit{Acta2} we noted that 11 clusters show a luminal profile (Lp, Hsp, Hsd, Avp \& Avd clusters) and 4 a basal profile (Bsl, Myo \& Prc clusters) (\autoref{F1}D, Supplementary Fig 2b and 2c).
+Based on the expression of \textit{Krt18,} \textit{Krt8, Krt5, Krt14} and \textit{Acta2} we noted that 11 clusters show a luminal profile (Lp, Hsp, Hsd, Avp \& Avd clusters) and 4 a basal profile (Bsl, Myo \& Prc clusters) (\autoref{F1}D, Supplementary Fig 3b and 3c).
 The Avd and Avp clusters express high levels of beta-casein, \textit{Csn2}, and are exclusively composed of cells from the gestation and lactation time-points, suggesting that these cells are secretory alveolar cells (\autoref{F1}d).
@@
 In contrast, the luminal compartment showed a distinct structure, with gradual transitions between different clusters and cells originating from a common origin (\autoref{F2}b).
-We confirmed the robustness of this bifurcation by verifying that it was present when different methods of feature selection, algorithms for trajectory inference and down-sampling were employed (Supplementary Figure 6).
+We confirmed the robustness of this bifurcation by verifying that it was present when different methods of feature selection, algorithms for trajectory inference and down-sampling were employed (Supplementary Figure 5).
 
@@
 In addition, we identified 1005 genes with branch-specific expression patterns (\autoref{F3}d, \autoref{F3}e, Supplementary Table 3).
-We further clustered the gene expression trend on each of the two branches and identified sets of genes that change over pseudotime (\fixme{Supplementary Figure 7}).
+We further clustered the gene expression trend on each of the two branches and identified sets of genes that change over pseudotime (\fixme{Supplementary Figure 6}).
 On the alveolar branch we found two clusters of genes that increase in expression during differentiation, one that is switched on early and another that is activated later in the final stages of differentiation.
@@
 However, we found that in particular the luminal compartment differs from its nulliparous counterpart (\autoref{F1}b and \autoref{F1}d).
-Interestingly, the effect of parity is most pronounced for the less differentiated and more progenitor-like cell types (\autoref{F1}c, Supplementary Figure 8).
+Interestingly, the effect of parity is most pronounced for the less differentiated and more progenitor-like cell types (\autoref{F1}c, Supplementary Figure 7).
 To ensure that Lp-PI still represents the progenitor population in the post-parous gland, we identified genes that distinguish Lp-NP from the rest of the luminal compartment to see if they are also characteristic for the proposed post-parity progenitor population Lp-PI.
 Indeed, we find genes that are differentially expressed between Lp-NP and the rest of the luminal compartment to show the same trend between Lp-PI and the luminal compartment (\autoref{F4}a).
-In a similar manner we can distinguish Lp-PI from the rest of the PI gland in a principal component analysis (PCA) using the identified progenitor genes (Supplementary Fig 8b and c).
+In a similar manner we can distinguish Lp-PI from the rest of the PI gland in a principal component analysis (PCA) using the identified progenitor genes (Supplementary Fig 7b and c).
 From this we conclude, that Lp-PI indeed represents the post-parity luminal progenitor population.
-Interestingly, genes that were up-regulated in Lp-PI compared to Lp-NP were significantly enriched for pathways that are involved in the immune response and lactation (\autoref{F4}b, Supplementary Fig 8d).
+Interestingly, genes that were up-regulated in Lp-PI compared to Lp-NP were significantly enriched for pathways that are involved in the immune response and lactation (\autoref{F4}b, Supplementary Fig 7d).
 These included genes that play roles in various processes during lactation, e.g.
@@
 
-\textbf{Supplementary Figure 1. Estrus cycle}
-Vaginal smears of the mice from the NP and PI timepoints. The animals were classified as being in estrus (NP1, NP2, PI1) or diestrus (PI1). The scale bar represents 200$\mu m$.
-
-\textbf{Supplementary Figure 2. Gating strategy to isolate MECs}
+\textbf{Supplementary Figure 1. Gating strategy to isolate MECs}
 \textbf{(a)} Gating strategy used to select live, lineage\textsuperscript{negative} singlets.
@@
 
-\textbf{Supplementary Figure 3. Quality control of sequencing data}
+\textbf{Supplementary Figure 2. Quality control of sequencing data}
 \textbf{(a)} Table summarizing quality control criteria per sample.
@@
 
-\textbf{Supplementary Figure 4. Biological replicates account for little structure in the data.}
+\textbf{Supplementary Figure 3. Biological replicates account for little structure in the data.}
 \textbf{(a)} t-SNE plot coloured by the eight different samples. All biological replicates are well mixed and none of the identified clusters is composed of cells from only one replicate.
@@
 
-\textbf{Supplementary Figure 5. Clustering strategy}
+\textbf{Supplementary Figure 4. Clustering strategy}
 \textbf{(a)} First round of clustering using an SNN-Graph and modularity maximization for cluster identification.
@@
 
-\textbf{Supplementary Figure 6. Robustness of bifurcation event in diffusion map}
+\textbf{Supplementary Figure 5. Robustness of bifurcation event in diffusion map}
 \textbf{(a)} The differentiation trajectory as determined by Monocle (see materials and methods) coloured by cluster assignment.
@@
 
-\textbf{Supplementary Figure 7. Gene set enrichment analysis of pseudo-time dependent genes}
+\textbf{Supplementary Figure 6. Gene set enrichment analysis of pseudo-time dependent genes}
 \textbf{(a)} Significantly enriched GO-Terms for genes in the two clusters of the hormone-sensing branch.
@@
 
-\textbf{Supplementary Figure 8. Lp-PI is a post-parity luminal progenitor cluster}
+\textbf{Supplementary Figure 7. Lp-PI is a post-parity luminal progenitor cluster}
 \textbf{(a)} Number of differentially expressed genes between the NP and PI clusters of Hsd, Hsp and Lp.
@@
 Dashed line indicates P value threshold at 0.001.
+
+\textbf{Supplementary Figure 8. Estrus cycle}
+Vaginal smears of the mice from the NP and PI timepoints. The animals were classified as being in estrus (NP1, NP2, PI1) or diestrus (PI1). The scale bar represents 200$\mu m$.
 
@@
 Two individual mice per developmental time point were used in the study.
+The estrus cycle stage of animals from the NP and PI timepoint were determined by vaginal smears (Supplementary Figure 8).
 
@@
 This identified a total of 13 clusters.
-Despite this, the majority of clusters still contained substructure as evident from the t-SNE (Supplementary Figure 5a).
-To resolve the cellular heterogeneity further we then applied agglomerative hierarchical clustering on each of the 13 clusters (Supplementary Figure 5b).
+Despite this, the majority of clusters still contained substructure as evident from the t-SNE (Supplementary Figure 4a).
+To resolve the cellular heterogeneity further we then applied agglomerative hierarchical clustering on each of the 13 clusters (Supplementary Figure 4b).
 For this, we first computed the pairwise distances between cells based on Spearman's rank correlation of log-transformed HVG counts.
@@
 Next, we flagged clusters that expressed clear markers of non-epithelial cells as ``contaminating cells'' and removed them from the downstream analysis.
-Clusters 6.1 and 6.2 were tagged as immune cells based on the expression of \textit{Cd74}, \textit{Cd72} and \textit{Cd54}\autocite{Scheele2017} and cluster 10.3 as endothelial cells based on the expression of \textit{Eng}, \textit{S1pr1} and \textit{Emcn} (Supplementary Figure 5c, Supplementary Figure 5d).
-We retained cluster 10.2 as it expressed markers of a previously described celltype in the gland\autocite{Wang2015}, despite being positive for 2 out of 4 (\textit{Des} and \textit{Cspg4}) pericyte markers\autocite{Armulik2011} (Supplementary Figure 5e).
+Clusters 6.1 and 6.2 were tagged as immune cells based on the expression of \textit{Cd74}, \textit{Cd72} and \textit{Cd54}\autocite{Scheele2017} and cluster 10.3 as endothelial cells based on the expression of \textit{Eng}, \textit{S1pr1} and \textit{Emcn} (Supplementary Figure 4c, Supplementary Figure 4d).
+We retained cluster 10.2 as it expressed markers of a previously described celltype in the gland\autocite{Wang2015}, despite being positive for 2 out of 4 (\textit{Des} and \textit{Cspg4}) pericyte markers\autocite{Armulik2011} (Supplementary Figure 4e).
 The high number of cells loaded on the 10x increased the chance of doublet formation in the droplet encapsulation\autocite{Zheng2017}, \fixme{which can result in artificial clusters}.

### 2017-08-25 src/S2.R

@@ -1,0 +1,49 @@
+# Figure S1
+library(scran)
+library(plyr)
+library(dplyr)
+library(knitr)
+library(ggplot2)
+library(gridExtra)
+library(Rtsne)
+library(cowplot)
+source("functions.R")
+
+# Load Data
+dataList <- readRDS("../data/Robjects/secondRun_2500/ExpressionList_QC.rds")
+sumDat <- read.csv("../data/CellRangerData/secondRun_2500/Summary.csv")
+m <- dataList[["counts"]]
+pD <- dataList[["phenoData"]]
+fD <- dataList[["featureData"]]
+
+# Summary table 
+sumry <- group_by(pD, SampleID) %>%
+    summarize("Number of cells"=n(),
+	      "Total molecules"=median(UmiSums),
+	      "Genes Detected"=median(GenesDetected))
+sumry <- left_join(sumry, sumDat[,c("SampleID","NumberOfReads","Saturation")])
+
+p1 <- tableGrob(sumry,rows=NULL,)
+
+# Load plots from QCAnalysis
+# Illustrate thresholds
+gdHist <- ggplot(pD, aes(x=GenesDetected,y=..density..)) +
+    geom_histogram(fill="white",color="black",bins=100) +
+    scale_x_log10() +
+    xlab("Total number of genes detected") +
+    facet_wrap(~Condition) 
+
+libSizeHist <- ggplot(pD, aes(x=UmiSums,y=..density..)) +
+    geom_histogram(fill="white",color="black",bins=100) +
+    scale_x_log10() +
+    facet_wrap(~Condition) +
+    xlab("Total number of unique molecules") 
+
+cellViability <- ggplot(pD, aes(x=prcntMito, y=GenesDetected, color=Condition, shape=Replicate))+
+    geom_point() +
+    xlab("Percentage of Mitochondrial RNA molecules") +
+    ylab("Total number of genes detected")
+
+cairo_pdf("../paper/figures/S2.pdf",height=12.41,width=17.54)
+plot_grid(p1,gdHist,libSizeHist,cellViability, labels="auto")
+dev.off()

### 2017-08-25 src/S3.R

@@ -1,49 +1,70 @@
-# Figure S1
-library(scran)
+# S4 and S7
+
 library(plyr)
 library(dplyr)
-library(knitr)
 library(ggplot2)
-library(gridExtra)
-library(Rtsne)
 library(cowplot)
+library(viridis)
+library(RColorBrewer)
+library(reshape2)
+library(pheatmap)
 source("functions.R")
 
 # Load Data
-dataList <- readRDS("../data/Robjects/secondRun_2500/ExpressionList_QC.rds")
-sumDat <- read.csv("../data/CellRangerData/secondRun_2500/Summary.csv")
-m <- dataList[["counts"]]
-pD <- dataList[["phenoData"]]
-fD <- dataList[["featureData"]]
+rnd_seed <- 300
+dataList <- readRDS("../data/Robjects/secondRun_2500/ExpressionList_QC_norm_clustered_clean.rds")
+m <- dataList[[1]]
+pD <- dataList[[2]]
+fD <- dataList[[3]]
 
-# Summary table 
-sumry <- group_by(pD, SampleID) %>%
-    summarize("Number of cells"=n(),
-	      "Total molecules"=median(UmiSums),
-	      "Genes Detected"=median(GenesDetected))
-sumry <- left_join(sumry, sumDat[,c("SampleID","NumberOfReads","Saturation")])
 
-p1 <- tableGrob(sumry,rows=NULL,)
+# ---- S4 ----
 
-# Load plots from QCAnalysis
-# Illustrate thresholds
-gdHist <- ggplot(pD, aes(x=GenesDetected,y=..density..)) +
-    geom_histogram(fill="white",color="black",bins=100) +
-    scale_x_log10() +
-    xlab("Total number of genes detected") +
-    facet_wrap(~Condition) 
+# Rename Condition for plot
+pD$Condition <- mapvalues(pD$Condition, from=c("NP","G","L","PI"),
+			  to=c("Nulliparous", "14.5d Gestation",
+			       "6d Lactation", "11d Post Natural Involution"))
 
-libSizeHist <- ggplot(pD, aes(x=UmiSums,y=..density..)) +
-    geom_histogram(fill="white",color="black",bins=100) +
-    scale_x_log10() +
-    facet_wrap(~Condition) +
-    xlab("Total number of unique molecules") 
+# Remove previously identified outlier and immune cells
+m <- m[,pD$keep]
[…]

### 2017-08-25 src/S4.R

@@ -1,4 +1,4 @@
-# S4 and S7
+# S5
 
 library(plyr)
 library(dplyr)
@@ -6,65 +6,60 @@ library(ggplot2)
 library(cowplot)
 library(viridis)
 library(RColorBrewer)
-library(reshape2)
 library(pheatmap)
+library(gridExtra)
+library(gridGraphics)
 source("functions.R")
 
 # Load Data
 rnd_seed <- 300
-dataList <- readRDS("../data/Robjects/secondRun_2500/ExpressionList_QC_norm_clustered_clean.rds")
+dataList <- readRDS("../data/Robjects/secondRun_2500/ExpressionList_QC_norm_clustered.rds")
 m <- dataList[[1]]
 pD <- dataList[[2]]
 fD <- dataList[[3]]
 
+# Remove outlier and immune cells
+m <- m[,pD$PassAll]
+pD <- pD[pD$PassAll,]
 
-# ---- S4 ----
+m <- t(t(m)/pD$sf)
 
-# Rename Condition for plot
-pD$Condition <- mapvalues(pD$Condition, from=c("NP","G","L","PI"),
-			  to=c("Nulliparous", "14.5d Gestation",
-			       "6d Lactation", "11d Post Natural Involution"))
 
-# Remove previously identified outlier and immune cells
-m <- m[,pD$keep]
-pD <- pD[pD$keep,]
+# First step
+p1 <- ggplot(pD,aes(tSNE1,tSNE2,color=Cluster)) +
+    geom_point()
 
-# put in rnd order for plotting
-set.seed(rnd_seed)
-fp1 <- pD[sample(c(1:nrow(pD)),nrow(pD)),]
-#t-SNE colored by SampleID
-p1 <- ggplot(fp1, aes(x=tSNE1, y=tSNE2, color=SampleID)) +
-    geom_point(size=1) +
-    scale_color_brewer(palette="Paired")+
-    #     ggtitle("Cluster") +
-    theme_void(base_size=12) +
-    guides(colour = guide_legend(override.aes = list(size=3))) +
-    theme(legend.position="bottom",legend.direction="horizontal",
-	  legend.title=element_blank()) 
+# Second round
+fp2 <- pD
+fp2$SubCluster <- as.character(fp2$SubCluster)
[…]

### 2017-08-25 src/S5.R

@@ -1,65 +1,122 @@
-# S5
-
+# Figure S6
 library(plyr)
+library(destiny)
 library(dplyr)
 library(ggplot2)
 library(cowplot)
-library(viridis)
 library(RColorBrewer)
-library(pheatmap)
-library(gridExtra)
 library(gridGraphics)
+library(gridExtra)
+library(scran)
 source("functions.R")
 
 # Load Data
 rnd_seed <- 300
-dataList <- readRDS("../data/Robjects/secondRun_2500/ExpressionList_QC_norm_clustered.rds")
+dataList <- readRDS("../data/Robjects/secondRun_2500/ExpressionList_QC_norm_clustered_clean.rds")
 m <- dataList[[1]]
 pD <- dataList[[2]]
 fD <- dataList[[3]]
 
-# Remove outlier and immune cells
-m <- m[,pD$PassAll]
-pD <- pD[pD$PassAll,]
+# ---- Robustness -----
+condComb <- c("NP","G")
+set <- c("Bsl-G","Bsl","Myo","Prc")
+keepCells <- pD$keep & !(pD$SuperCluster %in% set) & pD$Condition %in% condComb
 
-m <- t(t(m)/pD$sf)
+# Cell filtering
+m.vp <- m[,keepCells]
+pD.vp <- pD[keepCells,]
 
+# Gene filtering
+keep <- rowMeans(m.vp)>0.01
+m.vp <- m.vp[keep,]
+fD.vp <- fD[keep,]
 
-# First step
-p1 <- ggplot(pD,aes(tSNE1,tSNE2,color=Cluster)) +
-    geom_point()
+# Normalize
+m.norm <- t(t(m.vp)/pD.vp$sf)
 
-# Second round
-fp2 <- pD
-fp2$SubCluster <- as.character(fp2$SubCluster)
-fp2$SubCluster <- substr(fp2$SubCluster,nchar(fp2$SubCluster),nchar(fp2$SubCluster))
-p2 <- ggplot(fp2,aes(tSNE1,tSNE2,color=SubCluster)) +
-    geom_point(size=.8) +
-    facet_wrap(~Cluster)
+out <- NULL
+features <- c("HVG","selected","PCA","all") 
+subsampls <- c(1,0.5,0.25)
 
-subP1 <- plot_grid(p1,p2, labels=c("a","b"))
+for (subsampl in subsampls) {
+for (feats in features) {
+    pD.cur <- pD.vp
+    
+    if (feats=="HVG") {
[…]

### 2017-08-25 src/S6.R

@@ -1,122 +1,72 @@
-# Figure S6
-library(plyr)
-library(destiny)
-library(dplyr)
+# GO-Analysis
 library(ggplot2)
 library(cowplot)
-library(RColorBrewer)
-library(gridGraphics)
-library(gridExtra)
-library(scran)
-source("functions.R")
+library("topGO")
+library(org.Mm.eg.db)
 
-# Load Data
-rnd_seed <- 300
-dataList <- readRDS("../data/Robjects/secondRun_2500/ExpressionList_QC_norm_clustered_clean.rds")
-m <- dataList[[1]]
-pD <- dataList[[2]]
-fD <- dataList[[3]]
+# ---- DEAnalysis4vs5 ----
+output <- data.frame()
+input <- read.csv("../data/Robjects/secondRun_2500/BranchDECluster.R",
+		  stringsAsFactor=FALSE)
+input$BranchCluster <- paste(input$Branch,input$Cluster,sep="-")
 
-# ---- Robustness -----
-condComb <- c("NP","G")
-set <- c("Bsl-G","Bsl","Myo","Prc")
-keepCells <- pD$keep & !(pD$SuperCluster %in% set) & pD$Condition %in% condComb
+for (clust in unique(input$BranchCluster)) { 
 
-# Cell filtering
-m.vp <- m[,keepCells]
-pD.vp <- pD[keepCells,]
+    # DeGenes
+    deGenes <- input[input$BranchCluster==clust,"Gene"]
 
-# Gene filtering
-keep <- rowMeans(m.vp)>0.01
-m.vp <- m.vp[keep,]
-fD.vp <- fD[keep,]
+    # Gene universe
+    if (substr(clust,1,3)=="Hrm") i <- 1 else i <- 2
+    m.smooth <- readRDS("../data/Robjects/secondRun_2500/BranchDEList.rds")[[i]][["mSmooth"]]
+    univrs <- rownames(m.smooth[rowMeans(m.smooth)>0.1,])
 
-# Normalize
-m.norm <- t(t(m.vp)/pD.vp$sf)
+    # ---- Data ----
+    alG <- factor(as.numeric(univrs %in% deGenes))
+    names(alG) <- univrs
 
-out <- NULL
-features <- c("HVG","selected","PCA","all") 
-subsampls <- c(1,0.5,0.25)
+    # ---- GOanalysis ----
 
-for (subsampl in subsampls) {
-for (feats in features) {
-    pD.cur <- pD.vp
-    
[…]

### 2017-08-25 src/S7.R

@@ -3,24 +3,44 @@ library(ggplot2)
 library(cowplot)
 library("topGO")
 library(org.Mm.eg.db)
+library(dplyr)
+library(edgeR)
+library(reshape2)
+source("functions.R")
+
+# ---- ImpactStrongestOnProgenitor ----
+
+out <- data.frame()
+for (clust in c("Hsd","Hsp","Lp")) {
+    filen <- sprintf("../data/Robjects/secondRun_2500/%s_NPvsPI.csv",clust)
+    tmp <- read.csv(file=filen,stringsAsFactors=FALSE)
+    tmp$Cluster <- clust
+    out <- rbind(out,tmp)
+}
+out <- group_by(out, Cluster) %>%
+    summarize(Up=sum(FDR < 0.01 & logFC > 0),
+	      Down=sum(FDR < 0.01 & logFC < 0)) %>%
+    melt()
+
+out$Cluster <- factor(out$Cluster,levels=c("Lp","Hsp","Hsd"))
+p1 <- ggplot(out, aes(x=Cluster,y=value,fill=variable)) +
+    geom_bar(stat="identity",position="dodge") +
+    coord_flip() +
+    ylab("Number of Genes DE") +
+    labs(fill="")
+
 
 # ---- DEAnalysis4vs5 ----
 output <- data.frame()
-input <- read.csv("../data/Robjects/secondRun_2500/BranchDECluster.R",
-		  stringsAsFactor=FALSE)
-input$BranchCluster <- paste(input$Branch,input$Cluster,sep="-")
+for (clust in c("Hsd","Hsp","Lp")) { # Lp has to be last for rest of script
+    filen <- sprintf("../data/Robjects/secondRun_2500/%s_NPvsPI.csv",clust)
+    topTab <- read.csv(file=filen,stringsAsFactors=FALSE)
 
-for (clust in unique(input$BranchCluster)) { 
-
-    # DeGenes
-    deGenes <- input[input$BranchCluster==clust,"Gene"]
-
-    # Gene universe
-    if (substr(clust,1,3)=="Hrm") i <- 1 else i <- 2
-    m.smooth <- readRDS("../data/Robjects/secondRun_2500/BranchDEList.rds")[[i]][["mSmooth"]]
-    univrs <- rownames(m.smooth[rowMeans(m.smooth)>0.1,])
[…]

### 2017-08-25 src/S8.R

@@ -1,158 +1,0 @@
-# GO-Analysis
-library(ggplot2)
-library(cowplot)
-library("topGO")
-library(org.Mm.eg.db)
-library(dplyr)
-library(edgeR)
-library(reshape2)
-source("functions.R")
-
-# ---- ImpactStrongestOnProgenitor ----
-
-out <- data.frame()
-for (clust in c("Hsd","Hsp","Lp")) {
-    filen <- sprintf("../data/Robjects/secondRun_2500/%s_NPvsPI.csv",clust)
-    tmp <- read.csv(file=filen,stringsAsFactors=FALSE)
-    tmp$Cluster <- clust
-    out <- rbind(out,tmp)
-}
-out <- group_by(out, Cluster) %>%
-    summarize(Up=sum(FDR < 0.01 & logFC > 0),
-	      Down=sum(FDR < 0.01 & logFC < 0)) %>%
-    melt()
-
-out$Cluster <- factor(out$Cluster,levels=c("Lp","Hsp","Hsd"))
-p1 <- ggplot(out, aes(x=Cluster,y=value,fill=variable)) +
-    geom_bar(stat="identity",position="dodge") +
-    coord_flip() +
-    ylab("Number of Genes DE") +
-    labs(fill="")
-
-
-# ---- DEAnalysis4vs5 ----
-output <- data.frame()
-for (clust in c("Hsd","Hsp","Lp")) { # Lp has to be last for rest of script
-    filen <- sprintf("../data/Robjects/secondRun_2500/%s_NPvsPI.csv",clust)
-    topTab <- read.csv(file=filen,stringsAsFactors=FALSE)
-
-    deGenes <- topTab[topTab$FDR < 0.01 & topTab$logFC > 0,"symbol"] 
-    allGenes <- topTab$symbol
-
-    # ---- Data ----
-    univrs <- allGenes
-    alG <- factor(as.numeric(univrs %in% deGenes))
-    names(alG) <- univrs
-
-    # ---- GOanalysis ----
-
-    # prepare Data for topGO
-    GO.data <- new("topGOdata", description="Lib GO",ontology="BP", allGenes=alG, 
-		   annot=annFUN.org, mapping="org.Mm.eg.db",
-		   nodeSize=20, ID="symbol")
-    result.classic <- runTest(GO.data, statistic="fisher")
[…]

### 2017-08-26 paper/main.tex
split F1 and edited text accordingly

@@
 In a second step we applied hierarchical clustering to each of the identified groups to further resolve the cellular heterogeneity (\autoref{F1}c, Supplementary Figure 4b).
-After removal of endothelial and immune cell clusters this resulted in a total of 15 clusters of mammary epithelial cells.
-Differential gene expression analysis allowed us to further characterise the clusters and infer putative identities (\autoref{F1}d, \autoref{T1} and Supplementary Table 1).
-Notably, some clusters mainly represent the same putative cell type at different developmental time points (indicated by the NP, G, L or PI suffix, in the following the suffix might be omitted to refer to a group of clusters).
-Based on the expression of \textit{Krt18,} \textit{Krt8, Krt5, Krt14} and \textit{Acta2} we noted that 11 clusters show a luminal profile (Lp, Hsp, Hsd, Avp \& Avd clusters) and 4 a basal profile (Bsl, Myo \& Prc clusters) (\autoref{F1}D, Supplementary Fig 3b and 3c).
-The Avd and Avp clusters express high levels of beta-casein, \textit{Csn2}, and are exclusively composed of cells from the gestation and lactation time-points, suggesting that these cells are secretory alveolar cells (\autoref{F1}d).
-Hsd and Hsp on the other hand represent the clusters with the highest expression of hormone receptors, suggesting hormone sensing function.
-Lp expresses high levels of the luminal progenitor marker \textit{Aldh1a3} suggesting progenitor function, while Hsp and Avp also express progenitor markers but to a lesser extent.
-Myo shows characteristics of myoepithelial cells such as high levels of \textit{Acta2}, \textit{Oxtr} and \textit{Krt15} (\autoref{F1}d).
+After removal of endothelial and immune cell clusters this resulted in a total of 15 clusters of mammary epithelial cells (C1-C15).
+Based on the expression of \textit{Krt18,} \textit{Krt8, Krt5, Krt14} and \textit{Acta2} we noted that 11 clusters show a luminal profile (C1-C11) and 4 a basal profile (C12-C15).\\
+To further charactarise the clusters we identified differentially expressed genes and inferred putative identities based on known marker (\autoref{F2}a and b).
+In the luminal compartment we found two larger subgroups.
+The first one expressed hormone receptors (\textit{Pgr}, \textit{Esr1}, \textit{Prlr} and \textit{Vdr)} (\autoref{F2}a).
+We noted that one group of hormone-sensing clusters showed a more differentiated state (Hsd), whereas others also expressed progenitor markers (e.g.
+\textit{Aldh1a3}), suggesting progenitor function (Hsp) (\autoref{F2}a).
+The second subgroup within the luminal compartment was devoid of hormone-receptors.
+Here, the Avd and Avp clusters expressed genes associated with the production of milk (e.g.
+\textit{Wap}, \textit{Csn2}), and were exclusively composed of cells from gestation and lactation, suggesting that these are secretory alveolar cells (\autoref{F2}a).
+Similar to the hormone-sensing cells we found alveolar cells that appeared to be in a more differentiated (Avd) or progenitor-like state (Avp).
+In addition to the alveolar cells, the subgroup that is devoid of hormone-receptors also contained the cells that showed the highest expression of progenitor genes and were devoid of lactation associated genes, suggesting that these are luminal progenitor cells (Lp) (\autoref{F2}a).
+In the basal compartment we found a group of cells (Myo) that showed characteristics of myoepithelial cells such as high levels of \textit{Acta2}, \textit{Oxtr} and \textit{Krt15} (\autoref{F1}d).
+Another group of cluster appeared to be less differentiated and expressed genes associated with basal progenitor cells (Bsl) (\textbf{Examples and Citation}).
+In addition the basal compartment also contained a cluster of cells that expressed high levels of \textit{Procr}, \textit{Gng11} and \textit{Zeb2} suggesting that these represent the previously identified Procr\textsuperscript{+}  basal cells, which are enriched for mammary repopulating units (\autoref{F2}a).
+Notably, the above described group of clusters were in some instances further separated by the different developmental time points (indicated by the NP, G, L or PI suffix), suggesting that the developmental stage of the animal has specific effects on the transcriptional landscape of certain mammary epithelial cell types (\autoref{F2}b).
 
@@
 Briefly, the method embeds the data in a low-dimensional space, where distances between cells represent the progression through a gradual but stochastic process such as differentiation.
-In diffusion maps constructed from all epithelial cells, we observed a clear segregation between the luminal and basal clusters, with virtually no transition states between the two (\autoref{F2}a).
+In diffusion maps constructed from all epithelial cells, we observed a clear segregation between the luminal and basal clusters, with virtually no transition states between the two (\autoref{F3}a).
 This supports the hypothesis that during normal tissue homoeostasis the two lineages are largely self-maintained, which is in agreement with the majority of the lineage tracing studies\autocite{VanKeymeulen2011,VanAmerongen2012,Davis2016}.
-In contrast, the luminal compartment showed a distinct structure, with gradual transitions between different clusters and cells originating from a common origin (\autoref{F2}b).
+In contrast, the luminal compartment showed a distinct structure, with gradual transitions between different clusters and cells originating from a common origin (\autoref{F3}b).
 We confirmed the robustness of this bifurcation by verifying that it was present when different methods of feature selection, algorithms for trajectory inference and down-sampling were employed (Supplementary Figure 5).
 
-The expression of the progenitor marker \textit{Aldh1a3} gradually decreased as cells progressed away from their common origin (\autoref{F2}c), which was largely composed of Lp.
-We further noted that the left arm of the differentiation trajectory terminates at Avd and shows increasing expression of \textit{Csn2} and \textit{Glycam1} (\autoref{F2}c), consistent with a secretory phenotype.
-Between Lp and Avd we found cells from Avp (\autoref{F2}b).
-On the right arm of the differentiation trajectory, cells in Lp transitioned to Hsp and Hsd, during which the expression of \textit{Cited1} and \textit{Pgr} increase suggesting that this branch represents differentiation towards hormone-sensing luminal cells (\autoref{F2}c).
+The expression of the progenitor marker \textit{Aldh1a3} gradually decreased as cells progressed away from their common origin (\autoref{F3}c), which was largely composed of Lp.
+We further noted that the left arm of the differentiation trajectory terminates at Avd and shows increasing expression of \textit{Csn2} and \textit{Glycam1} (\autoref{F3}c), consistent with a secretory phenotype.
+Between Lp and Avd we found cells from Avp (\autoref{F3}b).
+On the right arm of the differentiation trajectory, cells in Lp transitioned to Hsp and Hsd, during which the expression of \textit{Cited1} and \textit{Pgr} increase suggesting that this branch represents differentiation towards hormone-sensing luminal cells (\autoref{F3}c).
 
-Being confident that the diffusion map recapitulates the luminal differentiation process, we then computationally inferred the two branches and ordered the cells according to their progression through ``pseudotime''\autocite{Haghverdi2016} (see materials and methods for further explanation) (\autoref{F3}a).
+Being confident that the diffusion map recapitulates the luminal differentiation process, we then computationally inferred the two branches and ordered the cells according to their progression through ``pseudotime''\autocite{Haghverdi2016} (see materials and methods for further explanation) (\autoref{F4}a).
 This allows us to identify genes whose expression changes during the process of differentiation.
 We found 456 genes that showed a pseudotime-dependent expression with the same directionality along both differentiation trajectories (Supplementary Table 2).
-These included genes associated with known progenitor characteristics such as \textit{Aldh1a3} and \textit{Tspan8} as well as transcription factors that have not previously been associated with luminal differentiation such as \textit{Creb5}, \textit{Hmga1} and \textit{Fosl1} (\autoref{F3}b).
+These included genes associated with known progenitor characteristics such as \textit{Aldh1a3} and \textit{Tspan8} as well as transcription factors that have not previously been associated with luminal differentiation such as \textit{Creb5}, \textit{Hmga1} and \textit{Fosl1} (\autoref{F4}b).
 Interestingly, \textit{Hmga1} is a known chromatin remodeller with reprogramming activity that has been implicated in basal-like breast cancer\autocite{Shah2012,Shah2013}.
-In addition, we identified 1005 genes with branch-specific expression patterns (\autoref{F3}d, \autoref{F3}e, Supplementary Table 3).
+In addition, we identified 1005 genes with branch-specific expression patterns (\autoref{F4}d, \autoref{F4}e, Supplementary Table 3).
 We further clustered the gene expression trend on each of the two branches and identified sets of genes that change over pseudotime (\fixme{Supplementary Figure 6}).
@@
 Genes with increasing expression are involved in hormone metabolic processes and the regulation of morphogenesis of an epithelium (e.g. \textit{Vdr}, \textit{Esr1}, \textit{Pgr} and \textit{Msx2}).
-Amongst the branch-specific transcription factors we found for example \textit{Runx1, Tox2} and \textit{Bhlhe41} (also known as \textit{Sharp1}) to be transcribed during differentiation towards the hormone-sensing lineage (\autoref{F3}c) and \textit{Elf5, Foxs1} and \textit{Ehf} in the secretory lineage (\autoref{F3}d).
+Amongst the branch-specific transcription factors we found for example \textit{Runx1, Tox2} and \textit{Bhlhe41} (also known as \textit{Sharp1}) to be transcribed during differentiation towards the hormone-sensing lineage (\autoref{F4}c) and \textit{Elf5, Foxs1} and \textit{Ehf} in the secretory lineage (\autoref{F4}d).
 \textit{Runx1} is a known repressor of \textit{Elf5} and its deletion has been shown to be deleterious for ductal morphogenesis\autocite{VanBragt2014}.
-The expression of the known progenitor master regulator \textit{Elf5} is maintained and further increased during secretory differentiation suggesting that its transcriptional level is fine-tuned in luminal progenitors (\autoref{F3}d).
+The expression of the known progenitor master regulator \textit{Elf5} is maintained and further increased during secretory differentiation suggesting that its transcriptional level is fine-tuned in luminal progenitors (\autoref{F4}d).
 
@@
 To ensure that Lp-PI still represents the progenitor population in the post-parous gland, we identified genes that distinguish Lp-NP from the rest of the luminal compartment to see if they are also characteristic for the proposed post-parity progenitor population Lp-PI.
-Indeed, we find genes that are differentially expressed between Lp-NP and the rest of the luminal compartment to show the same trend between Lp-PI and the luminal compartment (\autoref{F4}a).
+Indeed, we find genes that are differentially expressed between Lp-NP and the rest of the luminal compartment to show the same trend between Lp-PI and the luminal compartment (\autoref{F5}a).
 In a similar manner we can distinguish Lp-PI from the rest of the PI gland in a principal component analysis (PCA) using the identified progenitor genes (Supplementary Fig 7b and c).
 From this we conclude, that Lp-PI indeed represents the post-parity luminal progenitor population.
-Interestingly, genes that were up-regulated in Lp-PI compared to Lp-NP were significantly enriched for pathways that are involved in the immune response and lactation (\autoref{F4}b, Supplementary Fig 7d).
+Interestingly, genes that were up-regulated in Lp-PI compared to Lp-NP were significantly enriched for pathways that are involved in the immune response and lactation (\autoref{F5}b, Supplementary Fig 7d).
 These included genes that play roles in various processes during lactation, e.g.
@@
 Some of the genes that are involved in the immune response are known regulators of the involution process such as \textit{Ctsc}, \textit{Tgfb3} or \textit{Mfge8}.
-The up-regulation of these genesets was also present in the Hsp and in some cases even the differentiated Hsd cluster, yet the effect remained strongest in the progenitor cells (\autoref{F4}c-d).
+The up-regulation of these genesets was also present in the Hsp and in some cases even the differentiated Hsd cluster, yet the effect remained strongest in the progenitor cells (\autoref{F5}c-d).
 Finally, we compared the differentiation processes of the luminal compartment of the parous gland to the nulliparous gland by mapping the luminal cells from PI to the trajectory of NP and G cells.
-Interestingly, the Hsp and Hsd cluster generally maintain the position of their nulliparous counterpart in the differentiation hierarchy whereas the Lp-PI cluster is stretched out from the origin down towards the alveolar branch, suggesting that these are biased towards the alveolar fate (\autoref{F4}e).
+Interestingly, the Hsp and Hsd cluster generally maintain the position of their nulliparous counterpart in the differentiation hierarchy whereas the Lp-PI cluster is stretched out from the origin down towards the alveolar branch, suggesting that these are biased towards the alveolar fate (\autoref{F5}e).
 Together, the data suggest that luminal progenitor cells maintain memory of having undergone gestation and involution.
@@
 We note, however, that only some of the clusters can be fully characterized by a single marker gene.
-Instead we argue that the epithelial cells - especially in the luminal compartment - should rather be conceptualized as being part of a continuous spectrum of differentiation as visualised in \autoref{F2}.
+Instead we argue that the epithelial cells - especially in the luminal compartment - should rather be conceptualized as being part of a continuous spectrum of differentiation as visualised in \autoref{F3}.
 This view highlights the plasticity of the tissue and might help to explain some of the conflicting results from lineage tracing studies\autocite{Inman2015}.
@@
 Cells are coloured by the four developmental time points Pink=NP, Dark Green=G, Light Green=L, Purple=PI.
-\textbf{(c)} Dendrogram of clusters based on the mean expression values of the 15 clusters. The tree was computed based on Spearman's rank correlation with Ward linkage.
-\textbf{(d)} Same as \textbf{(b)} but colouring cells by clusters.
-\textbf{(e)} Heatmap highlighting some key marker genes to distinguish the main classes of clusters.
+\textbf{(c)} Same as \textbf{(c)} but colouring cells by clusters.
+The smaller t-SNEs are coloured by the normalized log-transformed expression of the basal marker \textit{Krt5} and the luminal marker \textit{Krt18}.
+%-------------------------------------------------------------------------------------------------------------------
+
+%---------------------------------------------Figure 2--------------------------------------------------------------
+\textbf{Figure 2. Single cell RNA sequencing identifies 15 clusters of mammary epithelial cells }
+\fakefigure{F2}
+\textbf{(a)} Dendrogram of clusters based on the mean expression values of the 15 clusters. The tree was computed based on Spearman's rank correlation with Ward linkage.
+\textbf{(b)} Heatmap highlighting some key marker genes that were used to infer putative identities.
 Colour scale represents log-transformed and normalized UMI counts scaled to a maximum of 1 per row.
@@
 
-%---------------------------------------------Figure 2--------------------------------------------------------------
-\textbf{Figure 2. Computational reconstruction of differentiation processes in the mammary gland}
-\fakefigure{F2}
+%---------------------------------------------Figure 3--------------------------------------------------------------
+\textbf{Figure 3. Computational reconstruction of differentiation processes in the mammary gland}
+\fakefigure{F3}
 \textbf{(a)} Diffusion map of epithelial cells from the NP and G time points, showing the first three diffusion components.
@@
 
-%---------------------------------------------Figure 3--------------------------------------------------------------
-\textbf{Figure 3. Pseudotime ordering identifies genes associated with luminal differentiation}
-\fakefigure{F3}
+%---------------------------------------------Figure 4--------------------------------------------------------------
+\textbf{Figure 4. Pseudotime ordering identifies genes associated with luminal differentiation}
+\fakefigure{F4}
 \textbf{(a)} Definition of the hormone-sensing and secretory differentiation branch.
@@
 
-%---------------------------------------------Figure 4--------------------------------------------------------------
-\textbf{Figure 4. The effect of parity on the transcriptomic landscape of the luminal progenitor compartment}
-\fakefigure{F4}
+%---------------------------------------------Figure 5--------------------------------------------------------------
+\textbf{Figure 5. The effect of parity on the transcriptomic landscape of the luminal progenitor compartment}
+\fakefigure{F5}
 \textbf{(a)} Comparison of fold changes from Lp-NP versus luminal compartment and fold changes from Lp-PI versus the luminal cells.
@@
 For inferring the differentiation trajectory we used diffusion maps.
-First, we selected all cells from the NP and G time point (\autoref{F2}a) and detected the HVGs as described above.
+First, we selected all cells from the NP and G time point (\autoref{F3}a) and detected the HVGs as described above.
 The log-transformed (log2(count+1)) gene counts were then used to compute the first twenty diffusion components using the \texttt{DiffusionMap} function (default parameters as in 'destiny'\autocite{Angerer2016}).
 After inspecting the fraction of explained variation for the first twenty components, we decided to retain the first 3.
-In \autoref{F2}b we then focused on the luminal compartment and recomputed the diffusion map based only on the luminal cells, using the aforementioned procedure.
+In \autoref{F3}b we then focused on the luminal compartment and recomputed the diffusion map based only on the luminal cells, using the aforementioned procedure.
 \fixme{Here we retained only the first two components (Supplementary Fig 3a).
@@
 }
-For inferring the branches and pseudotime ordering, we defined the following three tips, the cell with the smallest value for the second eigenvector (which was set as root) and the cells with the largest and smallest values for the first eigenvector (compare \autoref{F2}b).
+For inferring the branches and pseudotime ordering, we defined the following three tips, the cell with the smallest value for the second eigenvector (which was set as root) and the cells with the largest and smallest values for the first eigenvector (compare \autoref{F3}b).
 

### 2017-08-26 src/Figure1.R

@@ -36,7 +36,7 @@ p0 <- ggplot(pD, aes(x=tSNE1, y=tSNE2, color=Condition)) +
 	  legend.title=element_blank()) 
 
 # t-SNE colored by cluster
-p1 <- ggplot(pD, aes(x=tSNE1, y=tSNE2, color=SubCluster)) +
+p1 <- ggplot(pD, aes(x=tSNE1, y=tSNE2, color=SubClusterNumbers)) +
     geom_point(size=1.5) +
     scale_color_manual(values=levels(pD$Colors))+
     #     ggtitle("Cluster") +

### 2017-08-26 src/Figure2.R

@@ -33,11 +33,11 @@ pD$Condition <- mapvalues(pD$Condition, from=c("NP","G","L","PI"),
 
 
 # ---- Dendrogram ----
-pD$SubCluster <- factor(pD$SubCluster) #drop unused levels
+pD$SubClusterNumbers <- factor(pD$SubClusterNumbers) #drop unused levels
 out <- data.frame(numeric(nrow(m)))
-colnames(out) <- levels(pD$SubCluster)[1]
-for (clust in levels(pD$SubCluster)) {
-    expr <- rowMeans(log2(m[,pD$SubCluster==clust]+1))
+colnames(out) <- levels(pD$SubClusterNumbers)[1]
+for (clust in levels(pD$SubClusterNumbers)) {
+    expr <- rowMeans(log2(m[,pD$SubClusterNumbers==clust]+1))
     colname <- clust
     out[,colname] <- expr
 }
@@ -75,7 +75,7 @@ genes <- c(c1,c3,c5,c2,c6,c7,c9)
 set.seed(rnd_seed)
 subsP <- filter(pD, !SubCluster %in% c("Hsd-G","Avd-L")) %>%
     group_by(SubCluster) %>%
-    do(sample_n(.,200))
+    do(sample_n(.,100))
 
 # Combine with remaining clusters and relevel factor according to order in plot
 ord <- filter(pD, SubCluster %in% c("Hsd-G","Avd-L")) %>%

### 2017-09-01 paper/main.tex
minor updates

@@
 To this end we used single cell RNA sequencing (scRNAseq) to determine the gene expression profile of individual mammary epithelial cells across four adult developmental stages; nulliparous, mid gestation, lactation and post weaning (full natural involution).
-Our data from 25,010 individual cells identifies 8 distinct mammary epithelial cell populations and allows their hierarchical structure across development to be charted.
+Our data from 25,010 individual cells identifies 15 distinct mammary epithelial cell populations and allows their hierarchical structure across development to be charted.
 Interestingly, the effect of gestation and lactation appeared to be more pronounced for some cell types.
@@
 In the basal compartment we found a group of cells (Myo) that showed characteristics of myoepithelial cells such as high levels of \textit{Acta2}, \textit{Oxtr} and \textit{Krt15} (\autoref{F1}d).
-Another group of cluster appeared to be less differentiated and expressed genes associated with basal progenitor cells (Bsl) (\textbf{Examples and Citation}).
+Another group of cluster appeared to be less differentiated and expressed genes associated with basal progenitor cells (Bsl) (e.g. \textit{Pdpn}, \autoref{F1}d).
 In addition the basal compartment also contained a cluster of cells that expressed high levels of \textit{Procr}, \textit{Gng11} and \textit{Zeb2} suggesting that these represent the previously identified Procr\textsuperscript{+}  basal cells, which are enriched for mammary repopulating units (\autoref{F2}a).
@@
 Next, we flagged clusters that expressed clear markers of non-epithelial cells as ``contaminating cells'' and removed them from the downstream analysis.
-Clusters 6.1 and 6.2 were tagged as immune cells based on the expression of \textit{Cd74}, \textit{Cd72} and \textit{Cd54}\autocite{Scheele2017} and cluster 10.3 as endothelial cells based on the expression of \textit{Eng}, \textit{S1pr1} and \textit{Emcn} (Supplementary Figure 4c, Supplementary Figure 4d).
-We retained cluster 10.2 as it expressed markers of a previously described celltype in the gland\autocite{Wang2015}, despite being positive for 2 out of 4 (\textit{Des} and \textit{Cspg4}) pericyte markers\autocite{Armulik2011} (Supplementary Figure 4e).
+Clusters C16 and C17 were tagged as immune cells based on the expression of \textit{Cd74}, \textit{Cd72} and \textit{Cd54}\autocite{Scheele2017}, C18 as fibroblasts based on the expression of collagens and fibronectin and C19 as endothelial cells based on the expression of \textit{Eng}, \textit{S1pr1} and \textit{Emcn} (Supplementary Figure 4c-e).
+We retained cluster 10.2 as it expressed markers of a previously described celltype in the gland\autocite{Wang2015}, despite being positive for 2 out of 4 (\textit{Des} and \textit{Cspg4}) pericyte markers\autocite{Armulik2011} (Supplementary Figure 4f).
 The high number of cells loaded on the 10x increased the chance of doublet formation in the droplet encapsulation\autocite{Zheng2017}, \fixme{which can result in artificial clusters}.
 We thus identified clusters that appeared at low frequencies (less than the maximum doublet rate of 7\%, Supplementary Figure 9b) and that were highly similar to at least two other clusters present in the same sample ($\rho > 0.9$ of mean log2(counts+1)).
-The only cluster that fulfilled this criterion was cluster Bsl-G2, whose gene expression was strongly correlated with Bsl-G and Avd-G (Supplementary Figure 9b)
-In addition, the gene expression profile of this cluster is virtually identical to the average expression values of Bsl-G and Avd-G, supporting the hypothesis that it is indeed a mixture of the two clusters (Supplementary Figure 9b).
-As Bsl-G and Avd-G are the most prevalent clusters in these samples (Table 1), doublets would be expected to be enriched for a combination of the two.
-Finally, Bsl-G2 also belonged to the clusters with a high number of genes detected and a high number of unique molecules, which is also indicative for doublet clusters (Supplementary Figure 9c).
-We hence excluded Bsl-G2 from any downstream analysis.
+The only cluster that fulfilled this criterion was cluster C20, whose gene expression was strongly correlated with C13 and C8 (Supplementary Figure 9b)
+In addition, the gene expression profile of this cluster is virtually identical to the average expression values of C13 and C8, supporting the hypothesis that it is indeed a mixture of the two clusters (Supplementary Figure 9b).
+As C13 and C8 are the most prevalent clusters in these samples (Table 1), doublets would be expected to be enriched for a combination of the two.
+Finally, C20 also belonged to the clusters with a high number of genes detected and a high number of unique molecules, which is also indicative for doublet clusters (Supplementary Figure 9c).
+We hence excluded C20 from any downstream analysis.
 

### 2017-09-01 paper/main.tex
changed to WTK labeling

@@
 To further charactarise the clusters we identified differentially expressed genes and inferred putative identities based on known marker (\autoref{F2}a and b).
+As shown in Table 1 multiple clusters of cells have been assigned to the same putative cell type.
+We found that in these instances the cells showed a high similarity and expressed the same marker genes but are from different developmental stages, suggesting that the stage of the gland has specific effects on the transcriptional landscape of certain cell types.\\
+
 In the luminal compartment we found two larger subgroups.
-The first one expressed hormone receptors (\textit{Pgr}, \textit{Esr1}, \textit{Prlr} and \textit{Vdr)} (\autoref{F2}a).
-We noted that one group of hormone-sensing clusters showed a more differentiated state (Hsd), whereas others also expressed progenitor markers (e.g.
-\textit{Aldh1a3}), suggesting progenitor function (Hsp) (\autoref{F2}a).
-The second subgroup within the luminal compartment was devoid of hormone-receptors.
-Here, the Avd and Avp clusters expressed genes associated with the production of milk (e.g.
-\textit{Wap}, \textit{Csn2}), and were exclusively composed of cells from gestation and lactation, suggesting that these are secretory alveolar cells (\autoref{F2}a).
-Similar to the hormone-sensing cells we found alveolar cells that appeared to be in a more differentiated (Avd) or progenitor-like state (Avp).
-In addition to the alveolar cells, the subgroup that is devoid of hormone-receptors also contained the cells that showed the highest expression of progenitor genes and were devoid of lactation associated genes, suggesting that these are luminal progenitor cells (Lp) (\autoref{F2}a).
-In the basal compartment we found a group of cells (Myo) that showed characteristics of myoepithelial cells such as high levels of \textit{Acta2}, \textit{Oxtr} and \textit{Krt15} (\autoref{F1}d).
-Another group of cluster appeared to be less differentiated and expressed genes associated with basal progenitor cells (Bsl) (e.g. \textit{Pdpn}, \autoref{F1}d).
-In addition the basal compartment also contained a cluster of cells that expressed high levels of \textit{Procr}, \textit{Gng11} and \textit{Zeb2} suggesting that these represent the previously identified Procr\textsuperscript{+}  basal cells, which are enriched for mammary repopulating units (\autoref{F2}a).
-Notably, the above described group of clusters were in some instances further separated by the different developmental time points (indicated by the NP, G, L or PI suffix), suggesting that the developmental stage of the animal has specific effects on the transcriptional landscape of certain mammary epithelial cell types (\autoref{F2}b).
+The first one expressed hormone receptors (\textit{Pgr}, \textit{Esr1}, \textit{Prlr}) (\autoref{F2}a) suggesting that these cells are putative hormone sensing (Hs) cells.
+Of the putative Hs cells we noted that C1/C2 expressed progenitor markers (e.g.
+\textit{Aldh1a3}), suggesting progenitor function (Hsp) (\autoref{F2}a) whereas C3/4/5 showed a more differentiated state (Hsd) (\autoref{F2}b and \autoref{T1}).
+The second subgroup within the luminal compartment (C6-C11) was devoid of hormone-receptors.
+C8-C11 expressed expressed genes associated with the production of milk (e.g.
+\textit{Wap}, \textit{Csn2}), and were exclusively composed of cells from gestation and lactation, suggesting that these are secretory alveolar cells (Av) (\autoref{F2}a).
+Of the putative alveolar cells we noted that C10/C11 expressed progenitor markers suggesting a progenitor like function (Avp) wherease C8/C9 appeared to be in a more differentiated state (Avd) (\autoref{F2}a and \autoref{T1}).
+We also found C6/7 which were devoid of hormone-receptors and milk genes but expressed high levels of progenitor markers, suggesting that these are luminal progenitor cells (Lp) (\autoref{F2}a).\\
+
+In the basal compartment we found that cells from C14 showed characteristics of myoepithelial cells such as high levels of \textit{Acta2}, \textit{Oxtr} and \textit{Krt15} (\autoref{F2}a).
+Whereas C12/13 cells appeared to be less differntiated and expressed genes associated with basal cells (Bsl) such as \textit{Pdpn} (REF).
+In addition the basal compartment also contained a cluster of cells, C15, that expressed high levels of \textit{Procr}, \textit{Gng11} and \textit{Zeb2} suggesting that these represent the previously identified Procr\textsuperscript{+}  basal cells, which have been shown to be enriched for mammary repopulating units (\autoref{F2}a).
 
@@
 
-The expression of the progenitor marker \textit{Aldh1a3} gradually decreased as cells progressed away from their common origin (\autoref{F3}c), which was largely composed of Lp.
-We further noted that the left arm of the differentiation trajectory terminates at Avd and shows increasing expression of \textit{Csn2} and \textit{Glycam1} (\autoref{F3}c), consistent with a secretory phenotype.
-Between Lp and Avd we found cells from Avp (\autoref{F3}b).
-On the right arm of the differentiation trajectory, cells in Lp transitioned to Hsp and Hsd, during which the expression of \textit{Cited1} and \textit{Pgr} increase suggesting that this branch represents differentiation towards hormone-sensing luminal cells (\autoref{F3}c).
+The expression of the progenitor marker \textit{Aldh1a3} gradually decreased as cells progressed away from their common origin (\autoref{F3}c), which was largely composed of cells from C6 (Lp).  
+We further noted that the left arm of the differentiation trajectory terminates at C8 (Avd) and shows increasing expression of \textit{Csn2} and \textit{Glycam1} (\autoref{F3}c), consistent with a secretory phenotype.
+Between C6 and C8 we found cells from C10 (\autoref{F3}b), suggesting the presence of an alveolar restricted progenitor (Avp).
+On the right arm of the differentiation trajectory, cells in C6 transitioned to C2 and C4/5(Hsp and Hsd), during which the expression of \textit{Cited1} and \textit{Pgr} increase suggesting that this branch represents differentiation towards hormone-sensing luminal cells (\autoref{F3}c).
 
@@
 In addition, we identified 1005 genes with branch-specific expression patterns (\autoref{F4}d, \autoref{F4}e, Supplementary Table 3).
-We further clustered the gene expression trend on each of the two branches and identified sets of genes that change over pseudotime (\fixme{Supplementary Figure 6}).
+We further clustered the gene expression trend on each of the two branches and identified sets of genes that change over pseudotime (Supplementary Figure 6).
 On the alveolar branch we found two clusters of genes that increase in expression during differentiation, one that is switched on early and another that is activated later in the final stages of differentiation.
@@
 However, we found that in particular the luminal compartment differs from its nulliparous counterpart (\autoref{F1}b and \autoref{F1}d).
-Interestingly, the effect of parity is most pronounced for the less differentiated and more progenitor-like cell types (\autoref{F1}c, Supplementary Figure 7).
-To ensure that Lp-PI still represents the progenitor population in the post-parous gland, we identified genes that distinguish Lp-NP from the rest of the luminal compartment to see if they are also characteristic for the proposed post-parity progenitor population Lp-PI.
-Indeed, we find genes that are differentially expressed between Lp-NP and the rest of the luminal compartment to show the same trend between Lp-PI and the luminal compartment (\autoref{F5}a).
-In a similar manner we can distinguish Lp-PI from the rest of the PI gland in a principal component analysis (PCA) using the identified progenitor genes (Supplementary Fig 7b and c).
-From this we conclude, that Lp-PI indeed represents the post-parity luminal progenitor population.
-Interestingly, genes that were up-regulated in Lp-PI compared to Lp-NP were significantly enriched for pathways that are involved in the immune response and lactation (\autoref{F5}b, Supplementary Fig 7d).
+Interestingly, the effect of parity is most pronounced for the less differentiated and more progenitor-like cell types C6/7 (\autoref{F1}c, Supplementary Figure 7).
+To ensure that C7 still represents the progenitor population in the post-parous gland, we identified genes that distinguish C6 from the rest of the luminal compartment to see if they are also characteristic for the proposed post-parity progenitor population C7.
+Indeed, we find genes that are differentially expressed between C6 and the rest of the luminal compartment to show the same trend between C7 and the luminal compartment (\autoref{F5}a).
+In a similar manner we can distinguish C7 from the rest of the PI gland in a principal component analysis (PCA) using the identified progenitor genes (Supplementary Fig 7b and c).
+From this we conclude, that C7 indeed represents the post-parity luminal progenitor population.
+Interestingly, genes that were up-regulated in C7 compared to C6 were significantly enriched for pathways that are involved in the immune response and lactation (\autoref{F5}b, Supplementary Fig 7d).
 These included genes that play roles in various processes during lactation, e.g.
@@
 Some of the genes that are involved in the immune response are known regulators of the involution process such as \textit{Ctsc}, \textit{Tgfb3} or \textit{Mfge8}.
-The up-regulation of these genesets was also present in the Hsp and in some cases even the differentiated Hsd cluster, yet the effect remained strongest in the progenitor cells (\autoref{F5}c-d).
+The up-regulation of these genesets was also present in C1 (Hsp) and in some cases even the differentiated C3 (Hsd) cluster, yet the effect remained strongest in the progenitor cells (\autoref{F5}c-d).
 Finally, we compared the differentiation processes of the luminal compartment of the parous gland to the nulliparous gland by mapping the luminal cells from PI to the trajectory of NP and G cells.
-Interestingly, the Hsp and Hsd cluster generally maintain the position of their nulliparous counterpart in the differentiation hierarchy whereas the Lp-PI cluster is stretched out from the origin down towards the alveolar branch, suggesting that these are biased towards the alveolar fate (\autoref{F5}e).
+Interestingly, C1 and C3 cluster generally maintain the position of their nulliparous counterpart in the differentiation hierarchy whereas the C7 cluster is stretched out from the origin down towards the alveolar branch, suggesting that these are biased towards the alveolar fate (\autoref{F5}e).
 Together, the data suggest that luminal progenitor cells maintain memory of having undergone gestation and involution.
@@
 In this study we could not provide any evidence for contribution of a putative mutlipotent stem cell to the day-to-day homoeostasis of the gland.
-We do note, however, that the Prc cluster expresses low levels of both luminal and basal markers, high levels of the stem cell marker \textit{Procr}\autocite{Wang2015} and high levels of the luminal progenitor marker \textit{Notch3}\autocite{Lafkas2013} (\autoref{F1}d).
+We do note, however, that cells form C15 (Prc) cluster expresses low levels of both luminal and basal markers, high levels of the stem cell marker \textit{Procr}\autocite{Wang2015} and high levels of the luminal progenitor marker \textit{Notch3}\autocite{Lafkas2013} (\autoref{F1}d).
 Yet, our data does not support a central role of this cluster in the day-to-day homeostasis of the gland. 
-In addition, cells from the Prc cluster also express some but not all markers of pericytes and we thus cannot exclude that these are non-epithelial cells. 
+In addition, cells from the C15 also express some but not all markers of pericytes and we thus cannot exclude that these are non-epithelial cells. 
 Based on the gene expression data presented here, the luminal compartment appears to have one common progenitor population (Lp).
@@
 The majority of the changes were related to pathways involved in immunity and lactation, suggesting that, in particular, the luminal progenitors maintain a memory of gestation and involution.
-It is reasonable to assume that Lp-PI overlaps with the previously described parity-induced mammary epithelial cells (PI-MECs)\autocite{Wagner2002}.
+It is reasonable to assume that C7 overlaps with the previously described parity-induced mammary epithelial cells (PI-MECs)\autocite{Wagner2002}.
 

### 2017-09-04 paper/main.tex
version before proofread

@@
 \large
-\textbf{Differentiation dynamics of mammary epithelial cells revealed by
-single-cell RNA-sequencing}
+\textbf{Differentiation dynamics of mammary epithelial cells revealed by single-cell RNA-sequencing}
 \normalsize
@@
 We argue instead that the epithelial cells -- especially in the luminal compartment -- should rather be conceptualized as being part of a continuous spectrum of differentiation.
-This view highlights the plasticity of the tissue and might help to explain some of the conflicting results from lineage tracing studies.}
+This view highlights the plasticity of the tissue and might help to explain the homeostasis of adult mammary epithelial cells.}
 \newpage
@@
 \section{Introduction}
+
 The purpose of the mammary gland is to provide nourishment and passive immunity for the young until they are capable of feeding themselves.
@@
 Here we used single cell RNA sequencing (scRNAseq) to map the cellular dynamics of mammary epithelial cells across four adult developmental stages; nulliparous, mid gestation, lactation and post weaning (full natural involution).
-Our data from 25,010 individual cells, identifies 15 distinct cell populations within the gland and allows their hierarchical structure across developmental time points to be charted.
+Our data from 25,010 individual cells, identifies 15 distinct cell populations within the gland and allows their hierarchical structure across developmental time-points to be charted.
 
@@
 
-First, we isolated mammary epithelial cells from four developmental time points; nulliparous (NP), day 14.5 gestation (G), day 6 lactation (L) and 11 days post natural involution (PI) (Supplementary Figure 1).
-For each time point we sorted mammary epithelial cells based on the EpCAM cell surface marker from two independent mice (\autoref{F1}a).
+First, we isolated mammary epithelial cells from four developmental time-points; nulliparous (NP), day 14.5 gestation (G), day 6 lactation (L) and 11 days post natural involution (PI) (Supplementary Figure 1).
+For each time-point we sorted mammary epithelial cells based on the EpCAM cell surface marker from two independent mice (\autoref{F1}a).
 All samples were then prepared for single-cell RNA-sequencing using the 10x Chromium platform\autocite{Zheng2017}.
-Following quality control (materials and methods), this yielded an average of 6,175 unique transcripts and 2,118 genes detected from 25,010 cells (4,223 in NP, 5,826 in G, 9,319 in L and 5,642 in P) (Supplementary Figure 2).
-Visual inspection of the data using t-distributed stochastic neighbor embedding (t-SNE) suggested that although there is grouping of cells by time-point, there are also other factors that underlie structure within the dataset (\autoref{F1}b and Supplementary Figure 3).
+Following quality control (materials and methods), this yielded an average of 6,175 unique transcripts and 2,118 genes detected from 25,010 cells (4,223 in NP, 5,826 in G, 9,319 in L and 5,642 in P) with high reproducibility (Supplementary Figure 2,3).
+Visual inspection of the data using t-distributed stochastic neighbor embedding (t-SNE) suggested that although there is grouping of cells by time-point, there are also other factors that underlie structure within the dataset (\autoref{F1}b).
 First, we dissected the global structure by unsupervised clustering using a shared-nearest-neighbor clustering approach.
@@
 After removal of endothelial and immune cell clusters this resulted in a total of 15 clusters of mammary epithelial cells (C1-C15).
-Based on the expression of \textit{Krt18,} \textit{Krt8, Krt5, Krt14} and \textit{Acta2} we noted that 11 clusters show a luminal profile (C1-C11) and 4 a basal profile (C12-C15).\\
+Based on the expression of \textit{Krt18,} \textit{Krt8, Krt5, Krt14} and \textit{Acta2} we noted that 11 clusters show a luminal profile (C1-C11) and 4 a basal profile (C12-C15) \fixme{F1d?}.\\
 To further charactarise the clusters we identified differentially expressed genes and inferred putative identities based on known marker (\autoref{F2}a and b).
@@
 
-In the luminal compartment we found two larger subgroups.
-The first one expressed hormone receptors (\textit{Pgr}, \textit{Esr1}, \textit{Prlr}) (\autoref{F2}a) suggesting that these cells are putative hormone sensing (Hs) cells.
+In the luminal compartment we found two larger subgroups (\autoref{F2}a).
+The first one expressed high levels of hormone receptors (\textit{Pgr}, \textit{Esr1}, \textit{Prlr}) (\autoref{F2}b-c) suggesting that these cells are putative hormone sensing (Hs) cells.
 Of the putative Hs cells we noted that C1/C2 expressed progenitor markers (e.g.
-\textit{Aldh1a3}), suggesting progenitor function (Hsp) (\autoref{F2}a) whereas C3/4/5 showed a more differentiated state (Hsd) (\autoref{F2}b and \autoref{T1}).
-The second subgroup within the luminal compartment (C6-C11) was devoid of hormone-receptors.
-C8-C11 expressed expressed genes associated with the production of milk (e.g.
-\textit{Wap}, \textit{Csn2}), and were exclusively composed of cells from gestation and lactation, suggesting that these are secretory alveolar cells (Av) (\autoref{F2}a).
-Of the putative alveolar cells we noted that C10/C11 expressed progenitor markers suggesting a progenitor like function (Avp) wherease C8/C9 appeared to be in a more differentiated state (Avd) (\autoref{F2}a and \autoref{T1}).
-We also found C6/7 which were devoid of hormone-receptors and milk genes but expressed high levels of progenitor markers, suggesting that these are luminal progenitor cells (Lp) (\autoref{F2}a).\\
+\textit{Aldh1a3}, \textit{Cd14}, \textit{Kit})\autocite{Shehata2012}, suggesting progenitor function (Hsp) (\autoref{F2}b-c) whereas C3/4/5 showed a more differentiated state (Hsd) (\autoref{F2}b-c and \autoref{T1}).
+The second subgroup within the luminal compartment (C6-C11) expressed only low levels of hormone receptors.
+C8-C11 expressed expressed milk proteins (e.g.
+\textit{Wap}, \textit{Csn2}), and were exclusively composed of cells from gestation and lactation, suggesting that these are secretory alveolar cells (Av) (\autoref{F2}b-c).
+Of the putative alveolar cells we noted that C10/C11 expressed progenitor markers suggesting a progenitor like function (Avp) wherease C8/C9 appeared to be in a more differentiated state (Avd) (\autoref{F2}b-c and \autoref{T1}).
+We also found C6/7 which were devoid of hormone-receptors and milk proteins but expressed high levels of progenitor markers, suggesting that these are luminal progenitor cells (Lp) (\autoref{F2}b-c).\\
 
-In the basal compartment we found that cells from C14 showed characteristics of myoepithelial cells such as high levels of \textit{Acta2}, \textit{Oxtr} and \textit{Krt15} (\autoref{F2}a).
-Whereas C12/13 cells appeared to be less differntiated and expressed genes associated with basal cells (Bsl) such as \textit{Pdpn} (REF).
-In addition the basal compartment also contained a cluster of cells, C15, that expressed high levels of \textit{Procr}, \textit{Gng11} and \textit{Zeb2} suggesting that these represent the previously identified Procr\textsuperscript{+}  basal cells, which have been shown to be enriched for mammary repopulating units (\autoref{F2}a).
+In the basal compartment we found that cells from C14 showed characteristics of myoepithelial cells such as high levels of \textit{Acta2}, \textit{Oxtr} and \textit{Krt15} (\autoref{F2}b-c).
+Whereas C12/13 cells appeared to be less differntiated and expressed genes associated with basal cells (Bsl) such as \textit{Pdpn} (\textbf{NEED REFERENCES}) (\autoref{F2}b).
+In addition the basal compartment also contained a cluster of cells, C15, that expressed high levels of \textit{Procr}, \textit{Gng11} and \textit{Zeb2} suggesting that these represent the previously identified Procr\textsuperscript{+}  basal cells, which have been shown to be enriched for mammary repopulating units\autocite{Wang2015} (\autoref{F2}b,c).
 
@@
 Between C6 and C8 we found cells from C10 (\autoref{F3}b), suggesting the presence of an alveolar restricted progenitor (Avp).
-On the right arm of the differentiation trajectory, cells in C6 transitioned to C2 and C4/5(Hsp and Hsd), during which the expression of \textit{Cited1} and \textit{Pgr} increase suggesting that this branch represents differentiation towards hormone-sensing luminal cells (\autoref{F3}c).
+On the right arm of the differentiation trajectory, cells from C6 transitioned to C2 and C4/5 (Hsp and Hsd), during which the expression of \textit{Cited1} and \textit{Pgr} increased suggesting that this branch represents differentiation towards hormone-sensing luminal cells (\autoref{F3}c).
 
 Being confident that the diffusion map recapitulates the luminal differentiation process, we then computationally inferred the two branches and ordered the cells according to their progression through ``pseudotime''\autocite{Haghverdi2016} (see materials and methods for further explanation) (\autoref{F4}a).
-This allows us to identify genes whose expression changes during the process of differentiation.
+This allows us to identify genes whose expression changes during the process of differentiation.\\
+
 We found 456 genes that showed a pseudotime-dependent expression with the same directionality along both differentiation trajectories (Supplementary Table 2).
@@
 Interestingly, \textit{Hmga1} is a known chromatin remodeller with reprogramming activity that has been implicated in basal-like breast cancer\autocite{Shah2012,Shah2013}.
-In addition, we identified 1005 genes with branch-specific expression patterns (\autoref{F4}d, \autoref{F4}e, Supplementary Table 3).
-We further clustered the gene expression trend on each of the two branches and identified sets of genes that change over pseudotime (Supplementary Figure 6).
+In addition, we identified 1005 genes with branch-specific expression patterns (\autoref{F4}e-f, Supplementary Table 3).
+We further clustered the gene expression trend on each of the two branches and identified sets of genes that change over pseudotime (\autoref{F4}e-f,Supplementary Figure 6).
 On the alveolar branch we found two clusters of genes that increase in expression during differentiation, one that is switched on early and another that is activated later in the final stages of differentiation.
-The early cluster is enriched for genes involved in fatty-acid oxidation and lipid biosynthetic processes (e.g. \textit{Lipa}, \textit{Dbi} and \textit{Acox1}) .
+The early cluster is enriched for genes involved in fatty-acid oxidation and lipid biosynthetic processes (e.g. \textit{Lipa}, \textit{Dbi} and \textit{Lpl}) .
 The second, smaller group of genes that is switched on during the late stages of differentiation was enriched for genes involved in fatty-acid transport and lipid homeostasis (e.g.
 \textit{Olah}, \textit{Cd36}).
+Of note, milk proteins vary in their time of activation, so are caseins expressed early on, whereas \textit{Wap}, \textit{Glycam1} and \textit{Lalba} are expressed in the final stages (\textbf{Supp?}).
 Genes involved in cell division and translation show a transient phase of up-regulation during the alveolar differentiation process (e.g.
@@
 At the end of lactation the gland then reverts back to a stage that resembles the pre-pregnancy state.
-However, we found that in particular the luminal compartment differs from its nulliparous counterpart (\autoref{F1}b and \autoref{F1}d).
+However, we found that in particular the luminal compartment differs from its nulliparous counterpart (\autoref{F1}b and \autoref{F1}c).
 Interestingly, the effect of parity is most pronounced for the less differentiated and more progenitor-like cell types C6/7 (\autoref{F1}c, Supplementary Figure 7).
@@
 These included genes that play roles in various processes during lactation, e.g.
-milk-proteins (\textit{Csn2}, \textit{Lalba}), lipases (\textit{Lipa}), proteins involved in milk secretion (\textit{Xdh}, \textit{Cd36}) and transcriptional regulators of lactation (\textit{Cidea}).
+milk-proteins (\textit{Csn2}, \textit{Lalba}), lipases (\textit{Lipa}), proteins involved in milk secretion (\textit{Xdh}, \textit{Cd36}) and transcriptional regulators of lactation (\textit{Cidea})\autocite{Wang2012}.
 Of note, the genes of the casein locus (\textit{Csn2}, \textit{Csn1s1}, \textit{Csn1s2a}, \textit{Csn3}) have previously been reported to be up-regulated in the parous gland, most likely due to changes in chromatin accessibility\autocite{Dos2015,Rijnkels2013}.
 However, it has not been shown before that this effect is confined to the progenitor population of the luminal compartment.
-Some of the genes that are involved in the immune response are known regulators of the involution process such as \textit{Ctsc}, \textit{Tgfb3} or \textit{Mfge8}.
+Some of the genes that are involved in the immune response are known regulators of the involution process such as \textit{Ctsc}, \textit{Tgfb3} or \textit{Mfge8}\autocite{Clarkson2003}.
 The up-regulation of these genesets was also present in C1 (Hsp) and in some cases even the differentiated C3 (Hsd) cluster, yet the effect remained strongest in the progenitor cells (\autoref{F5}c-d).
@@
 We have reported here the use of single cell RNA sequencing to comprehensively map the transcriptomes of thousands of mammary epithelial cells across four developmental time-points.
-Our analysis identified 15 clusters of epithelial cells, some of which are only present during specific developmental stage (e.g. Avd and Lp-PI).
+Our analysis identified 15 clusters of epithelial cells, some of which are only present during specific developmental stage.
 This study provides a rich dataset that can be mined online (see link in materials and methods) to identify marker genes and lineage specific promoters that can be used to trace populations of cells \textit{in vivo}.
@@
 Instead we argue that the epithelial cells - especially in the luminal compartment - should rather be conceptualized as being part of a continuous spectrum of differentiation as visualised in \autoref{F3}.
-This view highlights the plasticity of the tissue and might help to explain some of the conflicting results from lineage tracing studies\autocite{Inman2015}.
+This view highlights the plasticity of the tissue and might help to explain the tissue homeostasis of the adult mammary gland\autocite{Inman2015}.
 In this study we could not provide any evidence for contribution of a putative mutlipotent stem cell to the day-to-day homoeostasis of the gland.
-We do note, however, that cells form C15 (Prc) cluster expresses low levels of both luminal and basal markers, high levels of the stem cell marker \textit{Procr}\autocite{Wang2015} and high levels of the luminal progenitor marker \textit{Notch3}\autocite{Lafkas2013} (\autoref{F1}d).
+We do note, however, that cells form C15 (Prc) expresses low levels of high levels of the stem cell marker \textit{Procr}\autocite{Wang2015} and high levels of the luminal progenitor marker \textit{Notch3}\autocite{Lafkas2013} (\autoref{F1}d).
 Yet, our data does not support a central role of this cluster in the day-to-day homeostasis of the gland. 
-In addition, cells from the C15 also express some but not all markers of pericytes and we thus cannot exclude that these are non-epithelial cells. 
+In addition, cells from the C15 also express some but not all markers of pericytes and it thus cannot be excluded that these are non-epithelial cells. 
 Based on the gene expression data presented here, the luminal compartment appears to have one common progenitor population (Lp).
-Lp gives rise to intermediate states of either hormone-sensing (Hsp) or secretory (Avp) cells that express differentiation markers as well as progenitor markers and appear to represent meta-stable states on the differentiation path from luminal progenitors towards fully differentiated cells (Hsd).
+Lp gives rise to intermediate states of either hormone-sensing or secretory cells that express differentiation markers as well as progenitor markers and appear to represent meta-stable states on the differentiation path from luminal progenitors towards fully differentiated cells.
 Furthermore, we characterised gene expression patterns along the differentiation hierarchy.
@@
 %---------------------------------------------Figure 1--------------------------------------------------------------
-\textbf{Figure 1. Single cell RNA sequencing identifies 15 clusters of mammary epithelial cells }
+\textbf{Figure 1. Single cell RNA sequencing identifies 15 clusters of mammary epithelial cells}
 \fakefigure{F1}
@@
 \textbf{(b)} t-SNE plot of 25,010 cells visualizes general structure in the data.
-Cells are coloured by the four developmental time points Pink=NP, Dark Green=G, Light Green=L, Purple=PI.
-\textbf{(c)} Same as \textbf{(c)} but colouring cells by clusters.
-The smaller t-SNEs are coloured by the normalized log-transformed expression of the basal marker \textit{Krt5} and the luminal marker \textit{Krt18}.
+Cells are coloured by the four developmental time-points Pink=NP, Dark Green=G, Light Green=L, Purple=PI.
+\textbf{(c)} Same as \textbf{(b)} but colouring cells by clusters.
+The smaller t-SNEs are coloured by the normalised log-transformed expression of the basal marker \textit{Krt5} and the luminal marker \textit{Krt18}.
 %-------------------------------------------------------------------------------------------------------------------
@@
 %---------------------------------------------Figure 2--------------------------------------------------------------
-\textbf{Figure 2. Single cell RNA sequencing identifies 15 clusters of mammary epithelial cells }
+\textbf{Figure 2. Single cell RNA sequencing identifies 15 clusters of mammary epithelial cells}
 \fakefigure{F2}
 \textbf{(a)} Dendrogram of clusters based on the mean expression values of the 15 clusters. The tree was computed based on Spearman's rank correlation with Ward linkage.
-\textbf{(b)} Heatmap highlighting some key marker genes that were used to infer putative identities.
-Colour scale represents log-transformed and normalized UMI counts scaled to a maximum of 1 per row.
+\textbf{(b)} t-SNE with overlaid expression of cluster specific genes.
+\textbf{(b)} Heatmap highlighting key marker genes that were used to infer putative identities.
+Colour scale represents log-transformed and normalised UMI counts scaled to a maximum of 1 per row.
 The upper panel shows genes that were used to distinguish between luminal and basal cells.
@@
 \fakefigure{F3}
-\textbf{(a)} Diffusion map of epithelial cells from the NP and G time points, showing the first three diffusion components.
+\textbf{(a)} Diffusion map of epithelial cells from the NP and G time-points, showing the first three diffusion components.
 \textbf{(b)} Differentiation trajectory of the luminal compartment based on the first two diffusion components.
-\textbf{(c)} Same plot as in \textbf{(b)}, coloured by the normalized and scaled expression values of various genes.
+\textbf{(c)} Same plot as in \textbf{(b)}, coloured by the normalised and scaled expression values of various genes.
 %-------------------------------------------------------------------------------------------------------------------
@@
 \fakefigure{F5}
-\textbf{(a)} Comparison of fold changes from Lp-NP versus luminal compartment and fold changes from Lp-PI versus the luminal cells.
-The genes represent the top 500 differentially expressed genes between Lp-NP and luminal cells.
-\textbf{(b)} Volcano plot illustrates differential expression between Lp-NP and Lp-PI, coloured dots represent significant genes with known function in lactation and immunity, dashed lines highlight the P value threshold of 0.01 and a log fold change of 1.
+\textbf{(a)} Comparison of fold changes from C6 versus luminal compartment and fold changes from C7 versus the luminal cells.
+The genes represent the top 500 differentially expressed genes between C6 and luminal cells.
+\textbf{(b)} Volcano plot illustrates differential expression between C6 and C7, coloured dots represent significant genes with known function in lactation and immunity, dashed lines highlight the P value threshold of 0.01 and a log fold change of 1.
 P values are adjusted for multiple testing using Benjamini-Hochberg.
 \textbf{(c-d)} Visualization of expression difference for genes related to lactation \textbf{(c)} or immunity \textbf{(d)} for the six luminal clusters in NP and PI.
-Expression values correspond to normalized UMI counts.
+Expression values correspond to normalised UMI counts.
 \textbf{(e)} Cells from PI mapped onto the diffusion map from the NP and G time-point. Cells from NP and G are colored in grey.
@@
 Overview of the different clusters including number of cells captured for each time-point and key genes that were used to infer their identities.
-\textbf{Missing!}
 
@@
 
-\textbf{Supplementary Figure 7. Lp-PI is a post-parity luminal progenitor cluster}
+\textbf{Supplementary Figure 7. C7 is a post-parity luminal progenitor cluster}
 \textbf{(a)} Number of differentially expressed genes between the NP and PI clusters of Hsd, Hsp and Lp.
@@
 \textbf{(a)} Frequency of each cluster in all 8 samples, the threshold of 7\% is indicated by the red dashed line.
-\textbf{(b)} Correlation of Bsl-G2 (on the y-axes) with Avd-G (left panels), Bsl-G (center panel) and the mean of Avd-G and Bsl-G (right panel) for both samples G1 and G2.
+\textbf{(b)} Correlation of C20 (on the y-axes) with C8 (left panels), C13 (center panel) and the mean of C8 and C13 (right panel) for both samples G1 and G2.
 \textbf{(c-d)} Number of genes detected \textbf{(c)} and total number of molecules \textbf{(d)}.
-The values for each cell were normalized to the median value of the sample in which the cell was captured.
+The values for each cell were normalised to the median value of the sample in which the cell was captured.
 
@@
 
-\textbf{Table S4.} Genes differentially expressed between Lp-PI and Lp-NP
+\textbf{Table S4.} Genes differentially expressed between C6 and C7
 
@@
 C57BL/6N mice were housed in individually ventilated cages under a 12:12~h light-dark cycle, with water and food available \textit{ad libitum}.
-The experiment was set up to allow for all of the developmental time points to be collected and tissues to be processed at the same time.
+The experiment was set up to allow for all of the developmental time-points to be collected and tissues to be processed at the same time.
 Mice were euthanized by terminal anaesthesia.
@@
 Tissue from NP females was harvested at 8 weeks of age.
-Two individual mice per developmental time point were used in the study.
+Two individual mice per developmental time-point were used in the study.
 The estrus cycle stage of animals from the NP and PI timepoint were determined by vaginal smears (Supplementary Figure 8).
@@
 After sorting cells were spun down and resuspended in \SI{50}{\micro\litre} of HF. 
-Samples were manually counted using an improved Neubauer chamber and the cell concentration was normalized by addition of HF.
+Samples were manually counted using an improved Neubauer chamber and the cell concentration was normalised by addition of HF.
 Equal numbers of cells per sample were processed for scRNA library preparation.
@@
 In total the Cell Ranger software identified 25,806 barcodes that contained enough unique molecules to be considered as cells (4,376 in NP, 6,021 in G, 9,603 in L and 5,806 in PI).
-Libraries prepared from all time points showed high quality that was reproducible between the two biological replicates (Supplementary 2a).
+Libraries prepared from all time-points showed high quality that was reproducible between the two biological replicates (Supplementary 2a).
 We used the following metrics to flag poor quality cells: number of genes detected, total number of unique molecular identifiers (UMIs) and percentage of molecules mapped to mitochondrial genes.
@@
 For inferring the differentiation trajectory we used diffusion maps.
-First, we selected all cells from the NP and G time point (\autoref{F3}a) and detected the HVGs as described above.
+First, we selected all cells from the NP and G time-point (\autoref{F3}a) and detected the HVGs as described above.
 The log-transformed (log2(count+1)) gene counts were then used to compute the first twenty diffusion components using the \texttt{DiffusionMap} function (default parameters as in 'destiny'\autocite{Angerer2016}).

### 2017-09-04 src/Figure1.R

@@ -58,7 +58,7 @@ forPlot <- left_join(add,pD[,c("barcode","tSNE1","tSNE2")])
 forPlot <- melt(forPlot,id=c("barcode","tSNE1","tSNE2")) %>%
     dplyr::rename(Expression=value)
 plots <- list()
-pal <- colorRampPalette(brewer.pal(n=7,name="YlOrRd"))(200)
+pal <- colorRampPalette(rev(brewer.pal(n=7,name="RdYlBu")))(200)
 
 for(i in seq_along(genes)) {
     gene <- genes[i]

### 2017-09-04 src/Figure3.R

@@ -95,7 +95,7 @@ fPlot <- data.frame(exps,
 fPlot <- full_join(fPlot, pD[,c("barcode","DC1","DC2")], by="barcode")
 
 # color
-pal <- colorRampPalette(brewer.pal(n=7,name="YlOrRd"))(200)
+pal <- colorRampPalette(rev(brewer.pal(n=7,name="RdYlBu")))(200)
 
 #  forloop for all genes
 pltlist <- list()

### 2017-09-04 paper/main.tex
added my own suggestions

@@
 
-First, we isolated mammary epithelial cells from four developmental time-points; nulliparous (NP), day 14.5 gestation (G), day 6 lactation (L) and 11 days post natural involution (PI) (Supplementary Figure 1).
-For each time-point we sorted mammary epithelial cells based on the EpCAM cell surface marker from two independent mice (\autoref{F1}a).
+We isolated mammary epithelial cells from four developmental time-points; nulliparous (NP), day 14.5 gestation (G), day 6 lactation (L) and 11 days post natural involution (PI).
+For each time-point we sorted mammary epithelial cells based on the EpCAM cell surface marker from two independent mice (Supplementary Figure 1, \autoref{F1}a).
 All samples were then prepared for single-cell RNA-sequencing using the 10x Chromium platform\autocite{Zheng2017}.
-Following quality control (materials and methods), this yielded an average of 6,175 unique transcripts and 2,118 genes detected from 25,010 cells (4,223 in NP, 5,826 in G, 9,319 in L and 5,642 in P) with high reproducibility (Supplementary Figure 2,3).
+Following quality control (materials and methods), this yielded an average of 6,175 unique transcripts and 2,118 genes detected from 25,010 cells (4,223 in NP, 5,826 in G, 9,319 in L and 5,642 in P) with high reproducibility between the biological replicates (Supplementary Figure 2,3).
 Visual inspection of the data using t-distributed stochastic neighbor embedding (t-SNE) suggested that although there is grouping of cells by time-point, there are also other factors that underlie structure within the dataset (\autoref{F1}b).
@@
 In a second step we applied hierarchical clustering to each of the identified groups to further resolve the cellular heterogeneity (\autoref{F1}c, Supplementary Figure 4b).
-After removal of endothelial and immune cell clusters this resulted in a total of 15 clusters of mammary epithelial cells (C1-C15).
-Based on the expression of \textit{Krt18,} \textit{Krt8, Krt5, Krt14} and \textit{Acta2} we noted that 11 clusters show a luminal profile (C1-C11) and 4 a basal profile (C12-C15) \fixme{F1d?}.\\
-To further charactarise the clusters we identified differentially expressed genes and inferred putative identities based on known marker (\autoref{F2}a and b).
+After removal of endothelial and immune cell clusters (Supplementary Figure 4c-e) this resulted in a total of 15 mammary epithelial cell clusters (C1-C15).
+Based on the expression of \textit{Krt18,} \textit{Krt8, Krt5, Krt14} and \textit{Acta2} we noted that 11 clusters show a luminal profile (C1-C11) and 4 a basal profile (C12-C15) (\autoref{F1}d).\\
+To further characterise the clusters we identified differentially expressed genes and inferred putative identities based on known marker (\autoref{F2}a and b).
 As shown in Table 1 multiple clusters of cells have been assigned to the same putative cell type.
-We found that in these instances the cells showed a high similarity and expressed the same marker genes but are from different developmental stages, suggesting that the stage of the gland has specific effects on the transcriptional landscape of certain cell types.\\
+We found that in these instances the cells showed a high similarity and expressed the same marker genes but derived from different developmental stages, suggesting that the stage of the gland has specific effects on the transcriptional landscape of certain cell types.\\
 
 In the luminal compartment we found two larger subgroups (\autoref{F2}a).
-The first one expressed high levels of hormone receptors (\textit{Pgr}, \textit{Esr1}, \textit{Prlr}) (\autoref{F2}b-c) suggesting that these cells are putative hormone sensing (Hs) cells.
+C1-C5 showed characteristics of hormone-sensing cells (Hs) such as high expression levels of hormone receptors (\textit{Pgr}, \textit{Esr1}, \textit{Prlr}) (\autoref{F2}b-c).
 Of the putative Hs cells we noted that C1/C2 expressed progenitor markers (e.g.
@@
 C8-C11 expressed expressed milk proteins (e.g.
-\textit{Wap}, \textit{Csn2}), and were exclusively composed of cells from gestation and lactation, suggesting that these are secretory alveolar cells (Av) (\autoref{F2}b-c).
-Of the putative alveolar cells we noted that C10/C11 expressed progenitor markers suggesting a progenitor like function (Avp) wherease C8/C9 appeared to be in a more differentiated state (Avd) (\autoref{F2}b-c and \autoref{T1}).
-We also found C6/7 which were devoid of hormone-receptors and milk proteins but expressed high levels of progenitor markers, suggesting that these are luminal progenitor cells (Lp) (\autoref{F2}b-c).\\
+\textit{Wap}, \textit{Csn2}), and were exclusively composed of cells from gestation and lactation, supporting the hypothesis that these are secretory alveolar cells (Av) (\autoref{F2}b-c).
+Of the putative alveolar cells we noted that C10/C11 expressed genes associated with a progenitor function (Avp) whereas C8/C9 appeared to be in a more differentiated state (Avd) (\autoref{F2}b-c and \autoref{T1}).
+We also found C6/7 which were devoid of hormone-receptors and milk proteins but expressed high levels of progenitor markers, which is consistent with a luminal progenitor (Lp) phenotype (\autoref{F2}b-c).\\
 
 In the basal compartment we found that cells from C14 showed characteristics of myoepithelial cells such as high levels of \textit{Acta2}, \textit{Oxtr} and \textit{Krt15} (\autoref{F2}b-c).
-Whereas C12/13 cells appeared to be less differntiated and expressed genes associated with basal cells (Bsl) such as \textit{Pdpn} (\textbf{NEED REFERENCES}) (\autoref{F2}b).
-In addition the basal compartment also contained a cluster of cells, C15, that expressed high levels of \textit{Procr}, \textit{Gng11} and \textit{Zeb2} suggesting that these represent the previously identified Procr\textsuperscript{+}  basal cells, which have been shown to be enriched for mammary repopulating units\autocite{Wang2015} (\autoref{F2}b,c).
+In addition, it also contained a cluster of cells, C15, that expressed high levels of \textit{Procr}, \textit{Gng11} and \textit{Zeb2} suggesting that these represent the previously identified Procr\textsuperscript{+}  basal cells, which have been shown to be enriched for mammary repopulating units\autocite{Wang2015} (\autoref{F2}b,c).
 
@@
 We further noted that the left arm of the differentiation trajectory terminates at C8 (Avd) and shows increasing expression of \textit{Csn2} and \textit{Glycam1} (\autoref{F3}c), consistent with a secretory phenotype.
-Between C6 and C8 we found cells from C10 (\autoref{F3}b), suggesting the presence of an alveolar restricted progenitor (Avp).
+In between C6 and C8  cells mainly derived from C10 (Avp, \autoref{F3}b) .
 On the right arm of the differentiation trajectory, cells from C6 transitioned to C2 and C4/5 (Hsp and Hsd), during which the expression of \textit{Cited1} and \textit{Pgr} increased suggesting that this branch represents differentiation towards hormone-sensing luminal cells (\autoref{F3}c).
@@
 
-We found 456 genes that showed a pseudotime-dependent expression with the same directionality along both differentiation trajectories (Supplementary Table 2).
+We found 456 genes that showed a pseudotime-dependent expression with the same directionality along both differentiation trajectories (see materials and methods for definition, Supplementary Table 2).
 These included genes associated with known progenitor characteristics such as \textit{Aldh1a3} and \textit{Tspan8} as well as transcription factors that have not previously been associated with luminal differentiation such as \textit{Creb5}, \textit{Hmga1} and \textit{Fosl1} (\autoref{F4}b).
-Interestingly, \textit{Hmga1} is a known chromatin remodeller with reprogramming activity that has been implicated in basal-like breast cancer\autocite{Shah2012,Shah2013}.
-In addition, we identified 1005 genes with branch-specific expression patterns (\autoref{F4}e-f, Supplementary Table 3).
+\textit{Hmga1} is a known chromatin remodeller with reprogramming activity that has been implicated in basal-like breast cancer\autocite{Shah2012,Shah2013}.
+In addition, we identified 1005 genes with branch-specific expression patterns (\autoref{F4}e and f, Supplementary Table 3).
 We further clustered the gene expression trend on each of the two branches and identified sets of genes that change over pseudotime (\autoref{F4}e-f,Supplementary Figure 6).
 On the alveolar branch we found two clusters of genes that increase in expression during differentiation, one that is switched on early and another that is activated later in the final stages of differentiation.
-The early cluster is enriched for genes involved in fatty-acid oxidation and lipid biosynthetic processes (e.g. \textit{Lipa}, \textit{Dbi} and \textit{Lpl}) .
+The early cluster is enriched for genes involved in fatty-acid oxidation and lipid biosynthetic processes (e.g. \textit{Lipa}, \textit{Dbi} and \textit{Lpl}, \autoref{F4}f) .
 The second, smaller group of genes that is switched on during the late stages of differentiation was enriched for genes involved in fatty-acid transport and lipid homeostasis (e.g.
-\textit{Olah}, \textit{Cd36}).
-Of note, milk proteins vary in their time of activation, so are caseins expressed early on, whereas \textit{Wap}, \textit{Glycam1} and \textit{Lalba} are expressed in the final stages (\textbf{Supp?}).
+\textit{Olah}, \textit{Cd36}, \autoref{F4}f).
+Of note, milk proteins vary in their time of activation, so were caseins expressed early on, whereas \textit{Wap}, \textit{Glycam1} and \textit{Lalba} were expressed in the final stages.
 Genes involved in cell division and translation show a transient phase of up-regulation during the alveolar differentiation process (e.g.
@@
 However, we found that in particular the luminal compartment differs from its nulliparous counterpart (\autoref{F1}b and \autoref{F1}c).
-Interestingly, the effect of parity is most pronounced for the less differentiated and more progenitor-like cell types C6/7 (\autoref{F1}c, Supplementary Figure 7).
+Here, the effect of parity is most pronounced for the luminal progenitor cell types (\autoref{F1}c, Supplementary Figure 7).
+C6 and C7 both showed characteristics of luminal progenitor cells, but we identified C6 practically only in the nulliparous gland and C7 in the gland from parous animals.
 To ensure that C7 still represents the progenitor population in the post-parous gland, we identified genes that distinguish C6 from the rest of the luminal compartment to see if they are also characteristic for the proposed post-parity progenitor population C7.
@@
 Interestingly, C1 and C3 cluster generally maintain the position of their nulliparous counterpart in the differentiation hierarchy whereas the C7 cluster is stretched out from the origin down towards the alveolar branch, suggesting that these are biased towards the alveolar fate (\autoref{F5}e).
+This is in agreement with the overexpression of genes related to the production and secretion of milk.
 Together, the data suggest that luminal progenitor cells maintain memory of having undergone gestation and involution.
@@
 This view highlights the plasticity of the tissue and might help to explain the tissue homeostasis of the adult mammary gland\autocite{Inman2015}.
-In this study we could not provide any evidence for contribution of a putative mutlipotent stem cell to the day-to-day homoeostasis of the gland.
-We do note, however, that cells form C15 (Prc) expresses low levels of high levels of the stem cell marker \textit{Procr}\autocite{Wang2015} and high levels of the luminal progenitor marker \textit{Notch3}\autocite{Lafkas2013} (\autoref{F1}d).
+In this study we could not provide any evidence for the contribution of a putative mutlipotent stem cell to the day-to-day homoeostasis of the gland.
+We do note, however, that cells form C15 (Prc) expresses low levels of high levels of the stem cell marker \textit{Procr}\autocite{Wang2015} and high levels of the luminal progenitor marker \textit{Notch3}\autocite{Lafkas2013} (\autoref{F2}c).
 Yet, our data does not support a central role of this cluster in the day-to-day homeostasis of the gland. 
-In addition, cells from the C15 also express some but not all markers of pericytes and it thus cannot be excluded that these are non-epithelial cells. 
+In addition, cells from the C15 also express some but not all markers of pericytes and it thus cannot be excluded that these are non-epithelial cells (Supplementary Figure 4f). 
 Based on the gene expression data presented here, the luminal compartment appears to have one common progenitor population (Lp).
@@
 The majority of the changes were related to pathways involved in immunity and lactation, suggesting that, in particular, the luminal progenitors maintain a memory of gestation and involution.
-It is reasonable to assume that C7 overlaps with the previously described parity-induced mammary epithelial cells (PI-MECs)\autocite{Wagner2002}.
+We assume that C7 overlaps at least partially with the previously described parity-induced mammary epithelial cells (PI-MECs)\autocite{Wagner2002}.
 

### 2017-09-04 paper/main.tex
included Walids suggestions

@@
 After removal of endothelial and immune cell clusters (Supplementary Figure 4c-e) this resulted in a total of 15 mammary epithelial cell clusters (C1-C15).
-Based on the expression of \textit{Krt18,} \textit{Krt8, Krt5, Krt14} and \textit{Acta2} we noted that 11 clusters show a luminal profile (C1-C11) and 4 a basal profile (C12-C15) (\autoref{F1}d).\\
+Based on the expression of \textit{Krt18,} \textit{Krt8, Krt5, Krt14} and \textit{Acta2} we noted that 11 clusters showed a luminal profile (C1-C11) and 4 showed a basal profile (C12-C15) (\autoref{F1}d).\\
 To further characterise the clusters we identified differentially expressed genes and inferred putative identities based on known marker (\autoref{F2}a and b).
@@
 The second subgroup within the luminal compartment (C6-C11) expressed only low levels of hormone receptors.
-C8-C11 expressed expressed milk proteins (e.g.
+C6/7 expressed high levels of progenitor markers, which is consistent with a luminal progenitor (Lp) phenotype (\autoref{F2}b-c).\\
+In contrast, C8-C11 expressed expressed milk proteins (e.g.
 \textit{Wap}, \textit{Csn2}), and were exclusively composed of cells from gestation and lactation, supporting the hypothesis that these are secretory alveolar cells (Av) (\autoref{F2}b-c).
 Of the putative alveolar cells we noted that C10/C11 expressed genes associated with a progenitor function (Avp) whereas C8/C9 appeared to be in a more differentiated state (Avd) (\autoref{F2}b-c and \autoref{T1}).
-We also found C6/7 which were devoid of hormone-receptors and milk proteins but expressed high levels of progenitor markers, which is consistent with a luminal progenitor (Lp) phenotype (\autoref{F2}b-c).\\
 
@@
 Finally, we compared the differentiation processes of the luminal compartment of the parous gland to the nulliparous gland by mapping the luminal cells from PI to the trajectory of NP and G cells.
-Interestingly, C1 and C3 cluster generally maintain the position of their nulliparous counterpart in the differentiation hierarchy whereas the C7 cluster is stretched out from the origin down towards the alveolar branch, suggesting that these are biased towards the alveolar fate (\autoref{F5}e).
+Interestingly, C1 and C3 generally maintain the position of their nulliparous counterpart in the differentiation hierarchy whereas the C7 cluster is stretched out from the origin down towards the alveolar branch, suggesting that these are biased towards the alveolar fate (\autoref{F5}e).
 This is in agreement with the overexpression of genes related to the production and secretion of milk.
@@
 We note, however, that only some of the clusters can be fully characterized by a single marker gene.
-Instead we argue that the epithelial cells - especially in the luminal compartment - should rather be conceptualized as being part of a continuous spectrum of differentiation as visualised in \autoref{F3}.
+Instead we argue that the epithelial cells - especially in the luminal compartment - should rather be conceptualized as being part of a continuous spectrum of differentiation as visualised in \autoref{F3} and \autoref{F4}.
 This view highlights the plasticity of the tissue and might help to explain the tissue homeostasis of the adult mammary gland\autocite{Inman2015}.
 In this study we could not provide any evidence for the contribution of a putative mutlipotent stem cell to the day-to-day homoeostasis of the gland.
-We do note, however, that cells form C15 (Prc) expresses low levels of high levels of the stem cell marker \textit{Procr}\autocite{Wang2015} and high levels of the luminal progenitor marker \textit{Notch3}\autocite{Lafkas2013} (\autoref{F2}c).
+We do note, however, that cells form C15 (Prc) expresses high levels of the stem cell marker \textit{Procr}\autocite{Wang2015} and high levels of the luminal progenitor marker \textit{Notch3}\autocite{Lafkas2013} (\autoref{F2}c).
 Yet, our data does not support a central role of this cluster in the day-to-day homeostasis of the gland. 
@@
 \textbf{(c)} Same as \textbf{(b)} but colouring cells by clusters.
-The smaller t-SNEs are coloured by the normalised log-transformed expression of the basal marker \textit{Krt5} and the luminal marker \textit{Krt18}.
+\textbf{(d)} t-SNEs coloured by the normalised log-transformed expression of the basal marker \textit{Krt5} and the luminal marker \textit{Krt18}.
 %-------------------------------------------------------------------------------------------------------------------
@@
 \textbf{(a)} Dendrogram of clusters based on the mean expression values of the 15 clusters. The tree was computed based on Spearman's rank correlation with Ward linkage.
-\textbf{(b)} t-SNE with overlaid expression of cluster specific genes.
-\textbf{(b)} Heatmap highlighting key marker genes that were used to infer putative identities.
+\textbf{(b)} t-SNEs with overlaid expression of cluster specific genes.
+\textbf{(c)} Heatmap highlighting key marker genes that were used to infer putative identities.
 Colour scale represents log-transformed and normalised UMI counts scaled to a maximum of 1 per row.

### 2017-09-05 src/S7.R
bugfix!

@@ -96,7 +96,7 @@ m <- m[,pD$keep]
 pD <- pD[pD$keep,]
 
 # DE Test on NP cells progenitor versus rest
-pD.sub <- filter(pD,Condition %in% "NP", !(SubCluster %in% c("Bsl","Bsl-G1","Myo","Prc")))
+pD.sub <- filter(pD,Condition %in% "NP", !(SubCluster %in% c("Bsl","Bsl-G","Myo","Prc")))
 m.sub <- m[,as.character(pD.sub$barcode)]
 keep <- rowMeans(m.sub) > 0.01
 m.sub <- m.sub[keep,]
@@ -120,7 +120,7 @@ resTab <- topTags(res,n=Inf,sort.by="PValue")
 tabNulPar <- resTab$table[1:500,]
 
 # ---- PCA ----
-pD.sub <- filter(pD,Condition %in% "PI", !(SubCluster %in% c("Bsl","Bsl-G1","Myo","Prc")))
+pD.sub <- filter(pD,Condition %in% "PI", !(SubCluster %in% c("Bsl","Bsl-G","Myo","Prc")))
 m.sub <- m[,as.character(pD.sub$barcode)]
 
 genes <- tabNulPar$id %in% rownames(m.sub)
@@ -132,7 +132,7 @@ pD.sub$PC1 <- pc$x[,1]
 pD.sub$PC2 <- pc$x[,2]
 
 cols <- levels(pD.sub$Colors)[levels(pD.sub$SubCluster) %in% unique(pD.sub$SubCluster)]
-pcplot <- ggplot(pD.sub, aes(PC1,PC2,color=SubCluster)) +
+pcplot <- ggplot(pD.sub, aes(PC1,PC2,color=SubClusterNumbers)) +
     geom_point() +
     scale_color_manual(values=cols)
 

### 2017-09-05 paper/main.tex
near final version

@@
 Based on the expression of \textit{Krt18,} \textit{Krt8, Krt5, Krt14} and \textit{Acta2} we noted that 11 clusters showed a luminal profile (C1-C11) and 4 showed a basal profile (C12-C15) (\autoref{F1}d).\\
+
 To further characterise the clusters we identified differentially expressed genes and inferred putative identities based on known marker (\autoref{F2}a and b).
@@
 The second subgroup within the luminal compartment (C6-C11) expressed only low levels of hormone receptors.
-C6/7 expressed high levels of progenitor markers, which is consistent with a luminal progenitor (Lp) phenotype (\autoref{F2}b-c).\\
+C6/7 expressed high levels of progenitor markers, which is consistent with a luminal progenitor (Lp) phenotype (\autoref{F2}b-c).
 In contrast, C8-C11 expressed expressed milk proteins (e.g.
-\textit{Wap}, \textit{Csn2}), and were exclusively composed of cells from gestation and lactation, supporting the hypothesis that these are secretory alveolar cells (Av) (\autoref{F2}b-c).
-Of the putative alveolar cells we noted that C10/C11 expressed genes associated with a progenitor function (Avp) whereas C8/C9 appeared to be in a more differentiated state (Avd) (\autoref{F2}b-c and \autoref{T1}).
+\textit{Wap}, \textit{Csn2}), and were exclusively composed of cells from gestation and lactation, supporting the hypothesis that these are secretory alveolar cells (Av) (\autoref{F2}b-c and \autoref{T1}).
+Of the putative alveolar cells we noted that C10/C11 expressed genes associated with a progenitor function (Avp) whereas C8/C9 appeared to be in a more differentiated state (Avd) (\autoref{F2}b-c).
 
@@
 \textit{Hmga1} is a known chromatin remodeller with reprogramming activity that has been implicated in basal-like breast cancer\autocite{Shah2012,Shah2013}.
-In addition, we identified 1005 genes with branch-specific expression patterns (\autoref{F4}e and f, Supplementary Table 3).
-We further clustered the gene expression trend on each of the two branches and identified sets of genes that change over pseudotime (\autoref{F4}e-f,Supplementary Figure 6).
+In addition, we identified 1005 genes with branch-specific expression patterns.
+We then clustered the gene expression trend on each of the two branches and identified sets of genes that change over pseudotime (\autoref{F4}e-f, Supplementary Table 3, Supplementary Figure 6).
 On the alveolar branch we found two clusters of genes that increase in expression during differentiation, one that is switched on early and another that is activated later in the final stages of differentiation.
@@
 %---------------------------------------------Figure 2--------------------------------------------------------------
-\textbf{Figure 2. Single cell RNA sequencing identifies 15 clusters of mammary epithelial cells}
+\textbf{Figure 2. Putative identities of mammary epithelial cell clusters}
 \fakefigure{F2}
-\textbf{(a)} Dendrogram of clusters based on the mean expression values of the 15 clusters. The tree was computed based on Spearman's rank correlation with Ward linkage.
+\textbf{(a)} Dendrogram of clusters based on the log transformed mean expression values of the 15 clusters. The tree was computed based on Spearman's rank correlation with Ward linkage.
 \textbf{(b)} t-SNEs with overlaid expression of cluster specific genes.
 \textbf{(c)} Heatmap highlighting key marker genes that were used to infer putative identities.
-Colour scale represents log-transformed and normalised UMI counts scaled to a maximum of 1 per row.
+Colour scale represents log-transformed and normalised counts scaled to a maximum of 1 per row.
 The upper panel shows genes that were used to distinguish between luminal and basal cells.
@@
 The genes represent the top 500 differentially expressed genes between C6 and luminal cells.
-\textbf{(b)} Volcano plot illustrates differential expression between C6 and C7, coloured dots represent significant genes with known function in lactation and immunity, dashed lines highlight the P value threshold of 0.01 and a log fold change of 1.
+\textbf{(b)} Volcano plot illustrating differential expression between C6 and C7, coloured dots represent significant genes with known function in lactation and immunity, dashed lines highlight the P value threshold of 0.01 and a log fold change of 1.
 P values are adjusted for multiple testing using Benjamini-Hochberg.
@@
 Expression values correspond to normalised UMI counts.
-\textbf{(e)} Cells from PI mapped onto the diffusion map from the NP and G time-point. Cells from NP and G are colored in grey.
+\textbf{(e)} Cells from PI projected on the diffusion map from the NP and G time-point. Cells from NP and G are colored in grey.
 %-------------------------------------------------------------------------------------------------------------------
@@
 \textbf{(a)} Gating strategy used to select live, lineage\textsuperscript{negative} singlets.
-\textbf{(b)} Representative plots for each of the four timepoints are shown. FSC-W: forward scatter width, FSC-A: forward scatter area, BSC-A: backward scatter area.
+\textbf{(b)} Representative plots for each of the four time-points are shown. FSC-W: forward scatter width, FSC-A: forward scatter area, BSC-A: backward scatter area.
 
@@
 Number of unique molecules, genes detected and number of reads represent the median value for each sample.
-\textbf{(b-c)} Histograms for the four conditions showing the distributions of the number of genes detected \textbf{(b)} or total number of molecules \textbf{(c)} metric.
+\textbf{(b-c)} Histograms for the four conditions showing the distributions of the number of genes detected \textbf{(b)} or total number of molecules \textbf{(c)}.
 \textbf{(d)} Scatterplot of number of genes detected versus percentage of mitochondrial RNA molecules.
@@
 \textbf{(a)} t-SNE plot coloured by the eight different samples. All biological replicates are well mixed and none of the identified clusters is composed of cells from only one replicate.
-\textbf{(b-c)} t-SNE plots highlighting the expression of luminal \textbf{(b)} and basal \textbf{(c)} genes in all 25,010 cells.
+\textbf{(b)} Correlation of log transformed mean expression values of biological replicates for all four time-points. The identity line is represented in blue.
 
 \textbf{Supplementary Figure 4. Clustering strategy}
-\textbf{(a)} First round of clustering using an SNN-Graph and modularity maximization for cluster identification.
+\textbf{(a)} First round of clustering using an SNN-Graph and modularity maximization for cluster identification. Cluster numbers are unrelated to the naming scheme used in the manuscript.
 \textbf{(b)} Second round of clustering using hierarchical clustering based on Spearman's rank correlation with average linkage.
 The gap statistic was used as optimization criterion to choose $k$.
-\textbf{(c-e)} Marker genes that were used to remove immune \textbf{(c)} and endothelial \textbf{(d)} cells from the analysis.
-Cluster 10.2 (Prc) showed some degree of expression for some pericyte markers \textbf{(e)}.
+\textbf{(c-f)} Marker genes that were used to remove immune cells \textbf{(c)}, fibroblasts \textbf{(d)} and endothelial \textbf{(e)} cells from the analysis.
+Cluster C15 (Prc) showed some degree of expression for some pericyte markers \textbf{(f)}.
 
@@
 \textbf{Supplementary Figure 6. Gene set enrichment analysis of pseudo-time dependent genes}
-\textbf{(a)} Significantly enriched GO-Terms for genes in the two clusters of the hormone-sensing branch.
-\textbf{(b)} Significantly enriched GO-Terms for genes in the four clusters of the secretory branch.
+\textbf{(a)} Significantly enriched GO-Terms for genes in the four clusters of the secretory branch.
+\textbf{(b)} Significantly enriched GO-Terms for genes in the two clusters of the hormone-sensing branch.
 
@@
 \textbf{Supplementary Figure 8. Estrus cycle}
-Vaginal smears of the mice from the NP and PI timepoints. The animals were classified as being in estrus (NP1, NP2, PI1) or diestrus (PI1). The scale bar represents 200$\mu m$.
+Vaginal smears of the mice from the NP and PI time-points. The animals were classified as being in estrus (NP1, NP2, PI1) or diestrus (PI1). The scale bar represents 200$\mu m$.
 
@@
 In total the Cell Ranger software identified 25,806 barcodes that contained enough unique molecules to be considered as cells (4,376 in NP, 6,021 in G, 9,603 in L and 5,806 in PI).
-Libraries prepared from all time-points showed high quality that was reproducible between the two biological replicates (Supplementary 2a).
+Libraries prepared from all time-points showed high quality that was reproducible between the two biological replicates (Supplementary Figure 2,3).
 We used the following metrics to flag poor quality cells: number of genes detected, total number of unique molecular identifiers (UMIs) and percentage of molecules mapped to mitochondrial genes.
@@
 \textbf{Clustering}
+
 Due to the large number of cells we used a two-step approach to identify clusters of cells.
@@
 This has the advantage of not requiring the computation of a distance matrix which is computationally prohibitive for large numbers of cells.
-The SNN-graph was constructed using the \texttt{buildSNNGraph} function from scran with the number of nearest-neighbors set to 20.
+The SNN-graph was constructed using the \texttt{buildSNNGraph} function from scran with the number of nearest-neighbors set to 20\autocite{Lun2016a}.
 Community-based clustering was then performed on the SNN-graph by multi-level modularity optimization using the \texttt{cluster\_louvain} function in igraph\autocite{Csardi2006}.
@@
 With this we identified a total number of 21 clusters.
-As the last step we performed a \textit{post-hoc} test as previously reported\autocite{Macosko2015} by merging clusters with less than 10 differentially expressed genes (at a P value threshold of 0.01 and minimum log fold change of 1), resulting in one merged cluster (Myo).
+As the last step we performed a \textit{post-hoc} test as previously reported\autocite{Macosko2015} by merging clusters with less than 10 differentially expressed genes (at a P value threshold of 0.01 and minimum log fold change of 1).
+This resulted in combining cluster 7 and 8 (Supplementary Figure 4a) in C14.
 Next, we flagged clusters that expressed clear markers of non-epithelial cells as ``contaminating cells'' and removed them from the downstream analysis.
 Clusters C16 and C17 were tagged as immune cells based on the expression of \textit{Cd74}, \textit{Cd72} and \textit{Cd54}\autocite{Scheele2017}, C18 as fibroblasts based on the expression of collagens and fibronectin and C19 as endothelial cells based on the expression of \textit{Eng}, \textit{S1pr1} and \textit{Emcn} (Supplementary Figure 4c-e).
-We retained cluster 10.2 as it expressed markers of a previously described celltype in the gland\autocite{Wang2015}, despite being positive for 2 out of 4 (\textit{Des} and \textit{Cspg4}) pericyte markers\autocite{Armulik2011} (Supplementary Figure 4f).
-The high number of cells loaded on the 10x increased the chance of doublet formation in the droplet encapsulation\autocite{Zheng2017}, \fixme{which can result in artificial clusters}.
-We thus identified clusters that appeared at low frequencies (less than the maximum doublet rate of 7\%, Supplementary Figure 9b) and that were highly similar to at least two other clusters present in the same sample ($\rho > 0.9$ of mean log2(counts+1)).
+We retained C15 as it expressed markers of a previously described celltype in the gland\autocite{Wang2015}, despite being positive for 2 out of 4 (\textit{Des} and \textit{Cspg4}) pericyte markers\autocite{Armulik2011} (Supplementary Figure 4f).
+The high number of cells loaded on the 10x increased the chance of doublet formation in the droplet encapsulation\autocite{Zheng2017}. 
+We thus identified clusters that appeared at low frequencies (less than the maximum doublet rate of 7\%, Supplementary Figure 9a) and that were highly similar to at least two other clusters present in the same sample ($\rho > 0.9$ of mean log2(counts+1)).
 The only cluster that fulfilled this criterion was cluster C20, whose gene expression was strongly correlated with C13 and C8 (Supplementary Figure 9b)
@@
 As C13 and C8 are the most prevalent clusters in these samples (Table 1), doublets would be expected to be enriched for a combination of the two.
-Finally, C20 also belonged to the clusters with a high number of genes detected and a high number of unique molecules, which is also indicative for doublet clusters (Supplementary Figure 9c).
+Finally, C20 also belonged to the clusters with a high number of genes detected and a high number of unique molecules, which is also indicative for doublet clusters (Supplementary Figure 9c and d).
 We hence excluded C20 from any downstream analysis.
@@
 Differential gene expression analysis was performed using ``edgeR''\autocite{Robinson2010}.
-\fixme{Genes with a mean expression level below 0.1 counts were removed from the analysis.}
+For pairwise comparisons between clusters genes with a mean expression level below 0.1 were removed from the analysis.
 A negative binomial generalized log-linear model was fitted to the remaining genes with the cluster assignments as covariate(s).
 The 'glmTreat' function was used to identify genes that have a significantly higher log fold change than 1 at an FDR of 0.01.
+The marker genes visualized in \autoref{F2} were identified using the 'findMarkers' function in scran with default settings on genes that had a median expression of 1 in at least one cluster\autocite{Lun2016a}.
 
@@
 The log-transformed (log2(count+1)) gene counts were then used to compute the first twenty diffusion components using the \texttt{DiffusionMap} function (default parameters as in 'destiny'\autocite{Angerer2016}).
-After inspecting the fraction of explained variation for the first twenty components, we decided to retain the first 3.
 In \autoref{F3}b we then focused on the luminal compartment and recomputed the diffusion map based only on the luminal cells, using the aforementioned procedure.
-\fixme{Here we retained only the first two components (Supplementary Fig 3a).
-Notably, the structure inferred by the diffusion map algorithm was robust to the choice of features and down-sampling of cells (Supplementary Fig 3d).
-The structure of a common origin and the two branches could also be inferred using Monocle with standard settings\autocite{Trapnell2014} (Supplementary Fig 3c).
-}
-For inferring the branches and pseudotime ordering, we defined the following three tips, the cell with the smallest value for the second eigenvector (which was set as root) and the cells with the largest and smallest values for the first eigenvector (compare \autoref{F3}b).
+Notably, the structure inferred by the diffusion map algorithm was robust to the choice of features and down-sampling of cells (Supplementary Fig 5b).
+The structure of a common origin and the two branches could also be inferred using Monocle with standard settings\autocite{Trapnell2014} (Supplementary Fig 5a).
+For inferring the branches and pseudotime ordering, we defined the following three tips, the cell with the largest value for the second eigenvector (which was set as root) and the cells with the largest and smallest values for the first eigenvector (compare \autoref{F4}a).
 
@@
 
-\fixme{All computational analyses were performed in R (Version 3.3.3) using standard functions unless otherwise indicated.
-Code is available online at \href{https://github.com/MarioniLab/MammaryGland}{\textit{https://github.com/MarioniLab/MammaryGland}}.
-Data can be accessed at this link (\url{https://karstenbach.shinyapps.io/webapp/}).
-}
+All computational analyses were performed in R (Version 3.4.1) using standard functions unless otherwise indicated.
+Code is available online at \url{https://github.com/MarioniLab/MammaryGland}.
+Data can be explored at \url{https://karstenbach.shinyapps.io/webapp/}.
 
@@
 
-K.B. performed the experiments and all the computational analysis.
+K.B. performed the experiments and all the computational analyses.
 S.P. and K.B. setup and collected the mammary epithelial cells.

### 2017-09-05 src/Figure3.R

@@ -84,7 +84,7 @@ m <- m[,pD$barcode]
 m.norm <- t(t(m)/pD$sf)
 
 # Genes to plot for trends
-genes <- c("Aldh1a3","Csn2","Pgr","Glycam1","Cited1")
+genes <- c("Aldh1a3","Csn2","Pgr","Glycam1","Esr1")
 rownames(m.norm) <- fD$symbol
 exps <- log2(m.norm[genes,]+1)
 exps <- t(exps/apply(exps,1,max))

### 2017-09-10 paper/main.tex
fixed mistakes

@@
 This is in agreement with the overexpression of genes related to the production and secretion of milk.
-Together, the data suggest that luminal progenitor cells maintain memory of having undergone gestation and involution.
+Together, the data suggest that luminal progenitor cells maintain memory of having undergone gestation and lactation.
 This memory could potentially prime progenitor cells towards the alveolar fate to facilitate alveologenesis in subsequent pregnancies.
@@
 This is especially interesting in the light of the protective effect of pregnancies against breast cancer and the role of luminal progenitors as cell of origin.
-The majority of the changes were related to pathways involved in immunity and lactation, suggesting that, in particular, the luminal progenitors maintain a memory of gestation and involution.
+The majority of the changes were related to pathways involved in immunity and lactation, suggesting that, in particular, the luminal progenitors maintain a memory of gestation and lactation.
 We assume that C7 overlaps at least partially with the previously described parity-induced mammary epithelial cells (PI-MECs)\autocite{Wagner2002}.
@@
 \textbf{Supplementary Figure 8. Estrus cycle}
-Vaginal smears of the mice from the NP and PI time-points. The animals were classified as being in estrus (NP1, NP2, PI1) or diestrus (PI1). The scale bar represents 200$\mu m$.
+Vaginal smears of the mice from the NP and PI time-points. The animals were classified as being in estrus (NP1, NP2, PI2) or diestrus (PI1). The scale bar represents 200$\mu m$.
 
@@
 We used the following metrics to flag poor quality cells: number of genes detected, total number of unique molecular identifiers (UMIs) and percentage of molecules mapped to mitochondrial genes.
-Poor quality cells were then identified by setting a threshold on the number of genes and number of UMIs that was defined as three median absolute deviations (MAD) below the median for each sample, while requiring a minimum value of 1,000 total molecules and 500 genes detected.
+Poor quality cells were then identified by setting a threshold on the number of genes and number of UMIs that was defined as three median absolute deviations (MAD) below the median for each time-point, while requiring a minimum value of 1,000 total molecules and 500 genes detected.
 This resulted in the following thresholds for total number of genes detected: 1,042 for NP, 836 for G, 500 for L and 759 for PI; and the following for total number of molecules : 2,012 in NP, 1,479 in G, 1,000 in L and 1,379 in G.

### 2017-09-10 paper/main.tex
Walids suggestions

@@
 After removal of endothelial and immune cell clusters (Supplementary Figure 4c-e) this resulted in a total of 15 mammary epithelial cell clusters (C1-C15).
-Based on the expression of \textit{Krt18,} \textit{Krt8, Krt5, Krt14} and \textit{Acta2} we noted that 11 clusters showed a luminal profile (C1-C11) and 4 showed a basal profile (C12-C15) (\autoref{F1}d).\\
+Based on the expression of \textit{Krt18,} \textit{Krt8, Krt5, Krt14} and \textit{Acta2} we noted that 11 clusters showed a luminal profile (C1-C11) and 4 showed a basal profile (C12-C15) (\autoref{F1}d, Supplementary Table 1).\\
 
-To further characterise the clusters we identified differentially expressed genes and inferred putative identities based on known marker (\autoref{F2}a and b).
+To further characterise the clusters, we identified differentially expressed genes and inferred putative identities based on known marker (\autoref{F2}a and b).
 As shown in Table 1 multiple clusters of cells have been assigned to the same putative cell type.
@@
 We focused on cells from the NP and G time-points to investigate mammary epithelial differentiation states of the gland.
-These differentiation states and the transitions between them can be computationally reconstructed using diffusion maps.
+These differentiation states and the transitions between them can be computationally reconstructed using diffusion maps\autocite{Haghverdi2015}.
 Briefly, the method embeds the data in a low-dimensional space, where distances between cells represent the progression through a gradual but stochastic process such as differentiation.
@@
 In between C6 and C8  cells mainly derived from C10 (Avp, \autoref{F3}b) .
-On the right arm of the differentiation trajectory, cells from C6 transitioned to C2 and C4/5 (Hsp and Hsd), during which the expression of \textit{Cited1} and \textit{Pgr} increased suggesting that this branch represents differentiation towards hormone-sensing luminal cells (\autoref{F3}c).
+On the right arm of the differentiation trajectory, cells from C6 transitioned to C2 and C4/5 (Hsp and Hsd), during which the expression of \textit{Esr1} and \textit{Pgr} increased suggesting that this branch represents differentiation towards hormone-sensing luminal cells (\autoref{F3}c).
 
@@
 At the end of lactation the gland then reverts back to a stage that resembles the pre-pregnancy state.
-However, we found that in particular the luminal compartment differs from its nulliparous counterpart (\autoref{F1}b and \autoref{F1}c).
-Here, the effect of parity is most pronounced for the luminal progenitor cell types (\autoref{F1}c, Supplementary Figure 7).
-C6 and C7 both showed characteristics of luminal progenitor cells, but we identified C6 practically only in the nulliparous gland and C7 in the gland from parous animals.
+However, we found that in particular the post-parous luminal compartment differs from its nulliparous counterpart (\autoref{F1}b and \autoref{F1}c).
+Here, the effect of parity is most pronounced for the luminal progenitor cell types (\autoref{F1}c), which show larger numbers of differentially expressed genes as compared to the differentiated cell types (Supplementary Figure 7a).
+C6 and C7 both showed characteristics of luminal progenitor cells, but we identified C6 predominantly in the nulliparous gland and C7 predominantly in the post-parous gland.
 To ensure that C7 still represents the progenitor population in the post-parous gland, we identified genes that distinguish C6 from the rest of the luminal compartment to see if they are also characteristic for the proposed post-parity progenitor population C7.

### 2017-09-10 paper/main.tex
added Johns suggestions

@@
 Pioneering fat pad transplantation studies nearly 70 years ago were the first to demonstrate the regenerative and differentiation capacity of small numbers of cells\autocite{Faulkin1960,Daniel197,Smalley1998}.
-More recently the use of cell surface markers coupled with flow cytometry has been used to enrich for various progenitor and stem cell compartments\autocite{Smalley1998,Stingl2006,Shackleton2006,Asselin2007} and showed the imbalance of such cell populations results in cellular transformation and subsequently breast cancer\autocite{Lim2009,Molyneux2010}.
+More recently the use of cell surface markers coupled with flow cytometry has been used to enrich for various progenitor and stem cell compartments\autocite{Smalley1998,Stingl2006,Shackleton2006,Asselin2007} and showed that imbalance of such cell populations results in cellular transformation and subsequently breast cancer\autocite{Lim2009,Molyneux2010}.
 Other studies, inspired by breast cancer transcriptomic profiling, have identified transcriptional regulators of mammary epithelial cell types such as \textit{GATA3} in luminal cells\autocite{Asselin2007,Kouros2006}.
@@
 In diffusion maps constructed from all epithelial cells, we observed a clear segregation between the luminal and basal clusters, with virtually no transition states between the two (\autoref{F3}a).
-This supports the hypothesis that during normal tissue homoeostasis the two lineages are largely self-maintained, which is in agreement with the majority of the lineage tracing studies\autocite{VanKeymeulen2011,VanAmerongen2012,Davis2016}.
+This supports the hypothesis that during normal tissue homoeostasis the two lineages are largely self-maintained, which is in agreement with the majority of lineage tracing studies\autocite{VanKeymeulen2011,VanAmerongen2012,Davis2016}.
 In contrast, the luminal compartment showed a distinct structure, with gradual transitions between different clusters and cells originating from a common origin (\autoref{F3}b).
@@
 \textit{Olah}, \textit{Cd36}, \autoref{F4}f).
-Of note, milk proteins vary in their time of activation, so were caseins expressed early on, whereas \textit{Wap}, \textit{Glycam1} and \textit{Lalba} were expressed in the final stages.
+Of note, milk proteins vary in their time of activation, with caseins expressed early on, whereas \textit{Wap}, \textit{Glycam1} and \textit{Lalba} were expressed in the final stages.
 Genes involved in cell division and translation show a transient phase of up-regulation during the alveolar differentiation process (e.g.
@@
 Genes with increasing expression are involved in hormone metabolic processes and the regulation of morphogenesis of an epithelium (e.g. \textit{Vdr}, \textit{Esr1}, \textit{Pgr} and \textit{Msx2}).
-Amongst the branch-specific transcription factors we found for example \textit{Runx1, Tox2} and \textit{Bhlhe41} (also known as \textit{Sharp1}) to be transcribed during differentiation towards the hormone-sensing lineage (\autoref{F4}c) and \textit{Elf5, Foxs1} and \textit{Ehf} in the secretory lineage (\autoref{F4}d).
+Amongst the branch-specific transcription factors we found, for example, \textit{Runx1, Tox2} and \textit{Bhlhe41} (also known as \textit{Sharp1}) to be transcribed during differentiation towards the hormone-sensing lineage (\autoref{F4}c) and \textit{Elf5, Foxs1} and \textit{Ehf} in the secretory lineage (\autoref{F4}d).
 \textit{Runx1} is a known repressor of \textit{Elf5} and its deletion has been shown to be deleterious for ductal morphogenesis\autocite{VanBragt2014}.
@@
 In this study we could not provide any evidence for the contribution of a putative mutlipotent stem cell to the day-to-day homoeostasis of the gland.
-We do note, however, that cells form C15 (Prc) expresses high levels of the stem cell marker \textit{Procr}\autocite{Wang2015} and high levels of the luminal progenitor marker \textit{Notch3}\autocite{Lafkas2013} (\autoref{F2}c).
+We do note, however, that cells from C15 (Prc) expresses high levels of the stem cell marker \textit{Procr}\autocite{Wang2015} and high levels of the luminal progenitor marker \textit{Notch3}\autocite{Lafkas2013} (\autoref{F2}c).
 Yet, our data does not support a central role of this cluster in the day-to-day homeostasis of the gland. 
-In addition, cells from the C15 also express some but not all markers of pericytes and it thus cannot be excluded that these are non-epithelial cells (Supplementary Figure 4f). 
+In addition, cells from C15 also express some but not all markers of pericytes and it thus cannot be excluded that these are non-epithelial cells (Supplementary Figure 4f). 
 Based on the gene expression data presented here, the luminal compartment appears to have one common progenitor population (Lp).
@@
 Colour scale represents log-transformed and normalised counts scaled to a maximum of 1 per row.
-The upper panel shows genes that were used to distinguish between luminal and basal cells.
 Upper bars represent the cluster assignment and stages for the individual cells.
@@
 \textbf{(a)} The differentiation trajectory as determined by Monocle (see materials and methods) coloured by cluster assignment.
-\textbf{(b)} The diffusion map is robust towards down-sampling of cells (100, 50 or 25\% of all cells were used in the left, middle or right panel, respectively) as well as the method of feature selection (all= all genes with mean expression level above 0.1, HVG= highly variable genes, PCA= first 50 components of PCA, selected= a manually selected choice of genes that are known to be involved in luminal cell differentiation).
+\textbf{(b)} The diffusion map is robust to down-sampling of cells (100, 50 or 25\% of all cells were used in the left, middle or right panel, respectively) as well as the method of feature selection (all= all genes with mean expression level above 0.1, HVG= highly variable genes, PCA= first 50 components of PCA, selected= a manually selected choice of genes that are known to be involved in luminal cell differentiation).
 The gene list included: \textit{Csn2, Gata3, Prlr, Elf5, Esr1, Pgr, Aldh1a3, Wap, Tspan8, Krt18, Krt8, Areg, Fgfr1, Fgfr2, Notch1, Notch3, Foxc1} and \textit{Zeb2}.
@@
 All lanes per sample were processed using the ``cellranger count'' function.
-The output from different lanes were then aggregated using ``cellranger aggr'' with --normalize set to ``none''.
+The output from different lanes was then aggregated using ``cellranger aggr'' with --normalize set to ``none''.
 
@@
 Clusters C16 and C17 were tagged as immune cells based on the expression of \textit{Cd74}, \textit{Cd72} and \textit{Cd54}\autocite{Scheele2017}, C18 as fibroblasts based on the expression of collagens and fibronectin and C19 as endothelial cells based on the expression of \textit{Eng}, \textit{S1pr1} and \textit{Emcn} (Supplementary Figure 4c-e).
-We retained C15 as it expressed markers of a previously described celltype in the gland\autocite{Wang2015}, despite being positive for 2 out of 4 (\textit{Des} and \textit{Cspg4}) pericyte markers\autocite{Armulik2011} (Supplementary Figure 4f).
-The high number of cells loaded on the 10x increased the chance of doublet formation in the droplet encapsulation\autocite{Zheng2017}. 
+We retained C15 as it expressed markers of a previously described cell type in the gland\autocite{Wang2015}, despite being positive for 2 out of 4 (\textit{Des} and \textit{Cspg4}) pericyte markers\autocite{Armulik2011} (Supplementary Figure 4f).
+The high number of cells loaded on the 10X increased the chance of doublet formation in the droplet encapsulation\autocite{Zheng2017}. 
 We thus identified clusters that appeared at low frequencies (less than the maximum doublet rate of 7\%, Supplementary Figure 9a) and that were highly similar to at least two other clusters present in the same sample ($\rho > 0.9$ of mean log2(counts+1)).
-The only cluster that fulfilled this criterion was cluster C20, whose gene expression was strongly correlated with C13 and C8 (Supplementary Figure 9b)
+The only cluster that fulfilled this criterion was cluster C20, whose gene expression was strongly correlated with C13 and C8 (Supplementary Figure 9b).
 In addition, the gene expression profile of this cluster is virtually identical to the average expression values of C13 and C8, supporting the hypothesis that it is indeed a mixture of the two clusters (Supplementary Figure 9b).

### 2017-09-13 paper/main.tex
minor change

@@
 As shown in Table 1 multiple clusters of cells have been assigned to the same putative cell type.
-We found that in these instances the cells showed a high similarity and expressed the same marker genes but derived from different developmental stages, suggesting that the stage of the gland has specific effects on the transcriptional landscape of certain cell types.\\
+We found that in these instances the cells showed a high similarity and expressed the same marker genes but derived from different developmental stages (see NP, G, L or PI suffix in \autoref{F2}a), suggesting that the stage of the gland has specific effects on the transcriptional landscape of certain cell types.\\
 

### 2017-09-14 paper/main.tex
WTK final comments

@@
 For example, our analysis revealed a cluster of luminal progenitor cells in post involution glands, which is distinct from progenitors found in nulliparous glands.
-The data also showed that few clusters could be fully characterized by a single marker gene.
+The data also showed that few clusters could be fully characterised by a single marker gene.
 We argue instead that the epithelial cells -- especially in the luminal compartment -- should rather be conceptualized as being part of a continuous spectrum of differentiation.
@@
 Visual inspection of the data using t-distributed stochastic neighbor embedding (t-SNE) suggested that although there is grouping of cells by time-point, there are also other factors that underlie structure within the dataset (\autoref{F1}b).
-First, we dissected the global structure by unsupervised clustering using a shared-nearest-neighbor clustering approach.
+First, we dissected the global structure by unsupervised clustering using a shared-nearest-neighbor clustering approach (see materials and methods).
 This resulted in a coarse clustering into 13 groups (Supplementary Figure 4a).
@@
 Of the putative Hs cells we noted that C1/C2 expressed progenitor markers (e.g.
-\textit{Aldh1a3}, \textit{Cd14}, \textit{Kit})\autocite{Shehata2012}, suggesting progenitor function (Hsp) (\autoref{F2}b-c) whereas C3/4/5 showed a more differentiated state (Hsd) (\autoref{F2}b-c and \autoref{T1}).
+\textit{Aldh1a3}, \textit{Cd14}, \textit{Kit})\autocite{Shehata2012}, suggesting restricted progenitor function (Hsp) (\autoref{F2}b-c) whereas C3/4/5 showed a more differentiated state (Hsd) (\autoref{F2}b-c and \autoref{T1}).
 The second subgroup within the luminal compartment (C6-C11) expressed only low levels of hormone receptors.
 C6/7 expressed high levels of progenitor markers, which is consistent with a luminal progenitor (Lp) phenotype (\autoref{F2}b-c).
-In contrast, C8-C11 expressed expressed milk proteins (e.g.
+In contrast, C8-C11 expressed proteins (e.g.
 \textit{Wap}, \textit{Csn2}), and were exclusively composed of cells from gestation and lactation, supporting the hypothesis that these are secretory alveolar cells (Av) (\autoref{F2}b-c and \autoref{T1}).
-Of the putative alveolar cells we noted that C10/C11 expressed genes associated with a progenitor function (Avp) whereas C8/C9 appeared to be in a more differentiated state (Avd) (\autoref{F2}b-c).
+Of the putative alveolar cells we noted that C10/C11 expressed genes associated with a restricted alveolar progenitor function (Avp) whereas C8/C9 appeared to be in a more differentiated alveolar state (Avd) (\autoref{F2}b-c).
 
 In the basal compartment we found that cells from C14 showed characteristics of myoepithelial cells such as high levels of \textit{Acta2}, \textit{Oxtr} and \textit{Krt15} (\autoref{F2}b-c).
-In addition, it also contained a cluster of cells, C15, that expressed high levels of \textit{Procr}, \textit{Gng11} and \textit{Zeb2} suggesting that these represent the previously identified Procr\textsuperscript{+}  basal cells, which have been shown to be enriched for mammary repopulating units\autocite{Wang2015} (\autoref{F2}b,c).
+In addition, the basal compartment also contained a cluster of cells, C15, that expressed high levels of \textit{Procr}, \textit{Gng11} and \textit{Zeb2} suggesting that these represent the previously identified Procr\textsuperscript{+}  basal cells, which have been shown to be enriched for mammary repopulating units\autocite{Wang2015} (\autoref{F2}b,c).
 
@@
 
-The expression of the progenitor marker \textit{Aldh1a3} gradually decreased as cells progressed away from their common origin (\autoref{F3}c), which was largely composed of cells from C6 (Lp).  
+We found the expression of the progenitor marker \textit{Aldh1a3} gradually decreased as cells progressed away from their common origin (\autoref{F3}c), which was largely composed of cells from C6 (Lp).
 We further noted that the left arm of the differentiation trajectory terminates at C8 (Avd) and shows increasing expression of \textit{Csn2} and \textit{Glycam1} (\autoref{F3}c), consistent with a secretory phenotype.
-In between C6 and C8  cells mainly derived from C10 (Avp, \autoref{F3}b) .
+In between C6 and C8, we found cells that were mainly derived from C10 (Avp, \autoref{F3}b).
 On the right arm of the differentiation trajectory, cells from C6 transitioned to C2 and C4/5 (Hsp and Hsd), during which the expression of \textit{Esr1} and \textit{Pgr} increased suggesting that this branch represents differentiation towards hormone-sensing luminal cells (\autoref{F3}c).
@@
 We found 456 genes that showed a pseudotime-dependent expression with the same directionality along both differentiation trajectories (see materials and methods for definition, Supplementary Table 2).
-These included genes associated with known progenitor characteristics such as \textit{Aldh1a3} and \textit{Tspan8} as well as transcription factors that have not previously been associated with luminal differentiation such as \textit{Creb5}, \textit{Hmga1} and \textit{Fosl1} (\autoref{F4}b).
+These included genes associated with known progenitor characteristics such as \textit{Aldh1a3} and \textit{Tspan8}\autocite{Fu2017} as well as transcription factors that have not previously been associated with luminal differentiation such as \textit{Creb5}, \textit{Hmga1} and \textit{Fosl1} (\autoref{F4}b).
 \textit{Hmga1} is a known chromatin remodeller with reprogramming activity that has been implicated in basal-like breast cancer\autocite{Shah2012,Shah2013}.
@@
 On the alveolar branch we found two clusters of genes that increase in expression during differentiation, one that is switched on early and another that is activated later in the final stages of differentiation.
-The early cluster is enriched for genes involved in fatty-acid oxidation and lipid biosynthetic processes (e.g. \textit{Lipa}, \textit{Dbi} and \textit{Lpl}, \autoref{F4}f) .
+The early cluster is enriched for genes involved in fatty-acid oxidation and lipid biosynthetic processes (e.g. \textit{Lipa}, \textit{Dbi} and \textit{Lpl}, \autoref{F4}f).
 The second, smaller group of genes that is switched on during the late stages of differentiation was enriched for genes involved in fatty-acid transport and lipid homeostasis (e.g.
 \textit{Olah}, \textit{Cd36}, \autoref{F4}f).
-Of note, milk proteins vary in their time of activation, with caseins expressed early on, whereas \textit{Wap}, \textit{Glycam1} and \textit{Lalba} were expressed in the final stages.
+Of note, milk genes vary in their time of activation, with caseins switched on earler during differentiation, whereas \textit{Wap}, \textit{Glycam1} and \textit{Lalba} were only expressed in the final stages.
 Genes involved in cell division and translation show a transient phase of up-regulation during the alveolar differentiation process (e.g.
@@
 \textit{Runx1} is a known repressor of \textit{Elf5} and its deletion has been shown to be deleterious for ductal morphogenesis\autocite{VanBragt2014}.
-The expression of the known progenitor master regulator \textit{Elf5} is maintained and further increased during secretory differentiation suggesting that its transcriptional level is fine-tuned in luminal progenitors (\autoref{F4}d).
+The expression of the known progenitor master regulator \textit{Elf5}\autocite{Oakes2008,Rios2014} is maintained and further increased during secretory differentiation suggesting that its transcriptional level is fine-tuned in luminal progenitors (\autoref{F4}d).
 
@@
 This study provides a rich dataset that can be mined online (see link in materials and methods) to identify marker genes and lineage specific promoters that can be used to trace populations of cells \textit{in vivo}.
-We note, however, that only some of the clusters can be fully characterized by a single marker gene.
+We note, however, that only some of the clusters can be fully characterised by a single marker gene.
 Instead we argue that the epithelial cells - especially in the luminal compartment - should rather be conceptualized as being part of a continuous spectrum of differentiation as visualised in \autoref{F3} and \autoref{F4}.
 This view highlights the plasticity of the tissue and might help to explain the tissue homeostasis of the adult mammary gland\autocite{Inman2015}.
-In this study we could not provide any evidence for the contribution of a putative mutlipotent stem cell to the day-to-day homoeostasis of the gland.
+In this study we could not detect any cells in transition between the basal and luminal cells.
+Therefore, we have no evidence to support the contribution of a putative multipotent stem cell to the day-to-day homoeostasis of the gland.
 We do note, however, that cells from C15 (Prc) expresses high levels of the stem cell marker \textit{Procr}\autocite{Wang2015} and high levels of the luminal progenitor marker \textit{Notch3}\autocite{Lafkas2013} (\autoref{F2}c).
@@
 Expression values correspond to normalised UMI counts.
-\textbf{(e)} Cells from PI projected on the diffusion map from the NP and G time-point. Cells from NP and G are colored in grey.
+\textbf{(e)} Cells from PI projected on the diffusion map from the NP and G time-point. Cells from NP and G are coloured in grey.
 %-------------------------------------------------------------------------------------------------------------------
@@
 \textbf{Supplementary Figure 2. Quality control of sequencing data}
-\textbf{(a)} Table summarizing quality control criteria per sample.
+\textbf{(a)} Table summarising quality control criteria per sample.
 Number of unique molecules, genes detected and number of reads represent the median value for each sample.
@@
 \textbf{Supplementary Figure 7. C7 is a post-parity luminal progenitor cluster}
-\textbf{(a)} Number of differentially expressed genes between the NP and PI clusters of Hsd, Hsp and Lp.
+\textbf{(a)} Number of differentially expressed genes between the NP and PI clusters of Hsd (C1 vs C2), Hsp (C3 vs C4) and Lp (C7 vs. C6).
 \textbf{(b)} Principal component analysis (PCA) on all luminal cells from the PI time-point.
@@
 \textbf{(c)} Genes that are higher expressed in Lp compared to the rest of the NP gland also have negative PC1 loadings.
-\textbf{(d)} Top GO-terms (biological processes) that are significantly enriched in upregulated genes from \textbf{(a)}.
+\textbf{(d)} Top GO-terms (biological processes) that are significantly enriched in upregulated genes from the same differential expression in analysis visualized in \textbf{(a)}.
 Dashed line indicates P value threshold at 0.001.
@@
 expression with different trends in the two branches
-
-\textbf{Table S4.} Genes differentially expressed between C6 and C7
 
@@
 Poor quality cells were then identified by setting a threshold on the number of genes and number of UMIs that was defined as three median absolute deviations (MAD) below the median for each time-point, while requiring a minimum value of 1,000 total molecules and 500 genes detected.
-This resulted in the following thresholds for total number of genes detected: 1,042 for NP, 836 for G, 500 for L and 759 for PI; and the following for total number of molecules : 2,012 in NP, 1,479 in G, 1,000 in L and 1,379 in G.
+This resulted in the following thresholds for total number of genes detected: 1,042 for NP, 836 for G, 500 for L and 759 for PI; and the following for total number of molecules: 2,012 in NP, 1,479 in G, 1,000 in L and 1,379 in G.
 In addition, all cells with 5\% or more of UMIs mapping to mitochondrial genes were defined as non-viable or apoptotic and removed from the analysis (Supplementary Figure 2d).
@@
 
-A gene set enrichment analysis based on gene-ontology (GO) terms was conducted to characterize various genesets in the analysis.
+A gene set enrichment analysis based on gene-ontology (GO) terms was conducted to characterise various genesets in the analysis.
 The genes of interest were compared to all genes that were tested for differential expression using topGO with default settings\autocite{Alexa2016}.
@@
 Code is available online at \url{https://github.com/MarioniLab/MammaryGland}.
-Data can be explored at \url{https://karstenbach.shinyapps.io/webapp/}.
+Data can be explored at \url{http://marionilab.cruk.cam.ac.uk/mammaryGland}.
 

### 2017-09-14 paper/main.tex
fixed 25k bug

@@
 To this end we used single cell RNA sequencing (scRNAseq) to determine the gene expression profile of individual mammary epithelial cells across four adult developmental stages; nulliparous, mid gestation, lactation and post weaning (full natural involution).
-Our data from 25,010 individual cells identifies 15 distinct mammary epithelial cell populations and allows their hierarchical structure across development to be charted.
+Our data from 23,184 individual cells identifies 15 distinct mammary epithelial cell populations and allows their hierarchical structure across development to be charted.
 Interestingly, the effect of gestation and lactation appeared to be more pronounced for some cell types.
@@
 Here we used single cell RNA sequencing (scRNAseq) to map the cellular dynamics of mammary epithelial cells across four adult developmental stages; nulliparous, mid gestation, lactation and post weaning (full natural involution).
-Our data from 25,010 individual cells, identifies 15 distinct cell populations within the gland and allows their hierarchical structure across developmental time-points to be charted.
+Our data from 23,184 individual cells, identifies 15 distinct cell populations within the gland and allows their hierarchical structure across developmental time-points to be charted.
 
@@
 In a second step we applied hierarchical clustering to each of the identified groups to further resolve the cellular heterogeneity (\autoref{F1}c, Supplementary Figure 4b).
-After removal of endothelial and immune cell clusters (Supplementary Figure 4c-e) this resulted in a total of 15 mammary epithelial cell clusters (C1-C15).
+After removal of endothelial and immune cell clusters (Supplementary Figure 4c-e) this resulted in a total of 15 mammary epithelial cell clusters (C1-C15) of 23,184 cells.
 Based on the expression of \textit{Krt18,} \textit{Krt8, Krt5, Krt14} and \textit{Acta2} we noted that 11 clusters showed a luminal profile (C1-C11) and 4 showed a basal profile (C12-C15) (\autoref{F1}d, Supplementary Table 1).\\
@@
 \textit{Olah}, \textit{Cd36}, \autoref{F4}f).
-Of note, milk genes vary in their time of activation, with caseins switched on earler during differentiation, whereas \textit{Wap}, \textit{Glycam1} and \textit{Lalba} were only expressed in the final stages.
+Of note, milk genes vary in their time of activation, with caseins switched on earlier during differentiation, whereas \textit{Wap}, \textit{Glycam1} and \textit{Lalba} were only expressed in the final stages.
 Genes involved in cell division and translation show a transient phase of up-regulation during the alveolar differentiation process (e.g.
@@
 \textbf{(a)} Schematic diagram highlighting the experimental setup for isolating and sequencing the RNA of single cells using the 10X chromium system.
-\textbf{(b)} t-SNE plot of 25,010 cells visualizes general structure in the data.
+\textbf{(b)} t-SNE plot of 23,184 cells visualizes general structure in the data.
 Cells are coloured by the four developmental time-points Pink=NP, Dark Green=G, Light Green=L, Purple=PI.
